### 2025-01-14 endofquarterpaper.qmd
Initial commit

---
title: "Geometric Signs on Paleolithic Mobile Objects as Cultural Group Markers"
author: Sophie Cooper
format: html
echo: false
execute: 
  warning: false
bibliography: references.bib
zotero: "Signbase-study"
link-citations: true
---

# Introduction

The Upper Paleolithic is characterized by a rise in early Homo sapiens, and a decline for other hominids. During this time, we see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe. One of the most distinctive qualities of the Upper Paleolithic is the large rise in artwork available from that era, both parietal and otherwise. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012].

We still lack a complete knowledge of the cultural diversity and movement patterns of the early modern humans from this era, despite a great deal of research into the subject. In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe during the Aurignacian technocomplex. The basis of this analysis is the idea that variations on signs can serve as an indicator of group membership and cultural similarity. This approach is supported by both ethnographic evidence throughout history, as well as similar archaeological studies done on ornamental and parietal art [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d]. In this study, we use our analyses to attempt to answer the following questions: Do identifiable and relevant patterns of cultural groups exist in the data, and if so, what can they tell us about the patterns of diversity and movement among Upper Paleolithic hominids?

<!--SC: I'm still having a difficult time figuring out just how to end this paragraph - you mentioned that I should have a research question ending in a question mark, but I'm struggling to figure out how to phrase it in a way that is appropriately formal  -->

# Background

Our understanding of the patterns of cultural and genetic diversity among the hominids of the Upper Paleolithic remains incomplete. Previous works have performed various cluster analysis methods upon variations in artefacts to represent and understand patterns of groups and cultural diversity. The studies have primarily relied upon personal ornamentation and parietal art, with few taking mobile objects into account.

Multivariate analyses of Gravettian beads from burial and occupation sites in Europe suggests an east to west cline of nine cultural groups [@baker2024evidence]. Seriation analysis, PCoA, and network analysis are used to identify the groups. The geographically discrete clusters identified encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and Western Mediterranean regions.

Similar analyses performed upon bead work patterns from the Aurignacian technocomplex as an indicator for linguistic diversity identify fourteen geographically cohesive clusters [@Vanhaeren_d]. Specifically, the groups are identified using seriation, correspondence and GIS analyses. The clusters identified are geographically discrete and sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.

Following the established precedent of using ornamentation and signs to identify cultural groups, we look to see if designs on mobile objects can be applicable here, and what it can add to the research being done. The additional dimension of mobility in this study allows us to potentially identify more geographically dispersed cultural groups, which might help visualize the migration patterns of early modern humans. As said previously, previous studies have primarily excluded mobile objects, and tend to find more geographically compact cultural groups as a result.

Our use of geometric signs in this study is supported by @Sauvet_2018, in which used quadrilateral geometric parietal signs are used to identify Upper Paleolithic cultural groups in the Cantabrian region of Spain. The authors identified and classified the signs based on variations in form, then performed a factorial correspondence analysis and ascending hierarchical clustering upon the data in order to identify two clusters, which they argued represents distinct and separate groups of hominids. In their study, @Sauvet_2018 made a case for the relevance of signs as identity markers, pointing to both the relevant qualities of signs, which are similar to those explained above, and the case studies throughout history where they served a similar purpose.

@Whallon_2006 proposes a framework for understanding hunter-gatherer mobility, arguing that mobility occurs among hunter-gatherers for multiple reasons. The first two reasons, referred to by Whallon as logistical and residential mobility, occur when groups move in order to procure a new supply of resources once the more immediate ones have been depleted. Another type of mobility is network mobility, in which groups, or subsets of groups, move in order to make contact with other groups. This occurs in order to form a "safety net", in times of scarce resources, so that groups are assured of a positive reception if they are forced to move. Lastly, and similarly, is informational mobility, in which groups, or individuals, move with the primary goal of gathering information, typically about the environmental conditions and resources in other areas. Again, this operates as a "safety net" of sorts, in that these groups then posses knowledge on where to move in times of resource scarcity.

In this paper, we test the hypothesis that cultural groups are discernible in a big data set of geometric signs on mobile objects. Firstly, we use mantel tests to evaluate the null hypothesis of isolation-by-distance. We then identify clusters using seriation, network, and Principal Coordinate analyses. Finally, we test the strength of the clusters using modularity, perMANOVA, and PERMDISP2 tests.

# Methods and Materials

```{r}
## clean the data
library(tidyverse)
library(magrittr)
library(rcarbon)
signbase_full <- read_csv("data/signBase_Version1.0.csv")
nsites <- signbase_full %>% 
  pull(site_name ) %>% 
  unique() %>% 
  length()
nartifacts <- nrow(signbase_full)
ncountries <- signbase_full %>% 
  pull(country) %>% 
  unique() %>% 
  length()
nsigntypes <- ncol(signbase_full)
signbase_full_clean <- signbase_full %>% 
  filter(site_name != "Willendorf",
         site_name != "Riparo di Fontana Nuova",
         site_name != "Muralovka",
         site_name != "Shanidar Cave",
         site_name != "Hayonim Cave",
         site_name != "El Salitre",
         site_name != "Grotte De La Princesse Pauline",
         site_name != "Šandalja II") %>% 
  dplyr::select(-other, -rectangle)
nsites_clean <- signbase_full_clean%>% 
  pull(site_name ) %>% 
  unique() %>% 
  length()
nartifacts_clean <- nrow(signbase_full_clean)
ncountries_clean <- signbase_full_clean %>% 
  pull(country) %>% 
  unique() %>% 
  length()
nsigntypes_clean <- nsigntypes - 1
signbase_years <- signbase_full_clean %>% 
  drop_na(date_bp_max_min) %>% 
  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+\\/\\-", "±")) %>% 
  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+", "±")) %>% 
  separate(date_bp_max_min,
           sep = " - ",
           c("date_bp_max", "date_bp_min"),
           remove = FALSE) %>% 
  mutate(date_bp_max = ifelse(str_detect(date_bp_max, "\\/"),
                              str_extract(date_bp_max, ".*?(?=\\/)"),
                              date_bp_max)) %>% 
  mutate(date_bp_min = ifelse(str_detect(date_bp_min, "\\/"),
                              str_extract(date_bp_min, ".*?(?=\\/)"),
                              date_bp_min)) %>% 
  separate(date_bp_max,
           sep = "±",
           c("date_bp_max_age",
             "date_bp_max_error")) %>% 
  separate(date_bp_min,
           sep = "±",
           c("date_bp_min_age",
             "date_bp_min_error")) %>% 
  drop_na(date_bp_max_age,
          date_bp_max_error) %>% 
  mutate(date_bp_max_age = parse_number(date_bp_max_age),
         date_bp_max_error = parse_number(date_bp_max_error))
signbase_years_cal <- 
  rcarbon::calibrate(signbase_years$date_bp_max_age,
            signbase_years$date_bp_max_error,
            verbose = FALSE) %>% 
  summary() %>% 
  data_frame()
pretty_print_dates <- 
  function(x) formatC(x, 
                      decimal.mark=".", 
                      big.mark=",", 
                      digits = 0, 
                      format = "f")
date_min <- signbase_years_cal %>% 
  pull(MedianBP) %>% 
  min()
date_max <- signbase_years_cal %>% 
  pull(MedianBP) %>% 
  max()
date_range <- date_max - date_min
date_min_pretty <- pretty_print_dates(date_min)
date_max_pretty <- pretty_print_dates(date_max)
date_range_pretty <- pretty_print_dates(date_range)
```

```{r}
groups <- list("1" = c("Abri Pataud", "El Rascaño", 
                         "Fumane", "La Viña",
                         "Lhotka", "Pod Hradem",
                         "Riparo Bombrini", "Sirgenstein Cave"),
                "2" = c("Kvasice", "Labeko Koba", 
                        "Les Rois", "Slatinice",
                        "Göpfelsteinhöhle"),
                "3" = c("Aurignac", "d'Engihoul",
                         "Istállóskő Cave", "Hohlenstein-Stadel",
                         "Gargas", "Grotte de la Verpillière I", 
                         "Mladeč", "Brassempouy",
                         "Gatzarria"),
                "4" = c("Wildscheuer", "Breitenbach",
                        "Peskő Cave", "Vindija Cave",
                        "El Castillo"),
                "5" = c("Trou Magrite",
                        "Grotte de Goyet", "Bockstein-Törle"),
                "6" = c("Trou al'Wesse", "Maisières-Canal",
                        "Les Cottés", "Le Terme Pialat"),
                "7" = c("Vogelherd","Hohle Fels",
                        "Geissenklösterle"),
               "8" = c("Tuto de Camalhot","Abri Lartet/Gorge d'Enfer",
                        "Castanet", "Le Trou Du Renard", 
                        "Grottes de Fonds-de-Forêt" ),
               "9" = c("Nová Dědina", "Menton/Grottes du Grimaldi",
                        "Grotte du Renne", "Solutré", 
                        "Žlutava", "Spy"),
                "10" = c("Cellier", "Blanchard", 
                         "La Ferrassie", "Hornos de la Pena",
                         "Otaslavice"),
                "11" = c("La Souquette", "Shelter Birów IV",
                         "Abri de Laussel", "Abri du Poisson"))
group_df <- 
  enframe(groups, 
          name = "group", 
          value = "site_name") %>%
  unnest(cols = site_name)

lat_long_df <- signbase_full_clean %>% 
  dplyr::select(site_name, longitude, latitude) %>% 
  distinct(site_name, .keep_all = TRUE)

signbase_unique_groups <- signbase_full_clean %>% 
  mutate(longitude = as.character(longitude),
         latitude = as.character(latitude)) %>% 
  group_by(site_name) %>%
  summarize(across(where(is.numeric), sum)) %>% 
  left_join(group_df) %>% 
  left_join(lat_long_df)

artifact_data <- signbase_unique_groups %>% 
  column_to_rownames("site_name") %>% 
  dplyr::select(line:star)
```

```{r}
# mantel abundance
library(vegan)
library(ade4)

set.seed(200)
euc_data <- artifact_data %>% 
  rownames_to_column("site_name") %>% 
  group_by(site_name) %>% 
  mutate(sign_count = sum(line:star)) %>% 
  column_to_rownames("site_name") %>% 
  dplyr::select(sign_count)

euc_dist <- dist(x = euc_data, method = "euclidean")
jac <- vegan::vegdist(artifact_data, method = "jaccard", binary = TRUE)
euc_mantel <- vegan::mantel(jac, euc_dist, permutations = 1000)
p_value_abundance_correlation <- round(euc_mantel$signif, 3)
r_value_abundance_correlation <- round(euc_mantel$statistic, 3)
```

## Data

```{r fig.height= 10, fig.width=10}
#| label: fig-site-map
#| fig-cap: "Map of Palaeolithic sites containing signs, sign abundance is represented by point size, data from Dutkiewicz et al. (2020)"
#| 
library(rnaturalearth)
library(rnaturalearthdata)
library(ggrepel)
library(sf)

library(ggpubr)

abundance_data <- signbase_full_clean %>% 
  pivot_longer(cols= line:star,
               values_to = "sign_total") %>% 
  filter(sign_total != 0) %>% 
  mutate(longitude = as.character(longitude),
         latitude = as.character(latitude)) %>% 
  group_by(site_name) %>%
  summarize(across(where(is.numeric), sum))

signbase_sf <- 
  st_as_sf(signbase_unique_groups, 
           coords = c("longitude", "latitude"),
           remove = FALSE,
           crs = 4326)
world <- ne_countries(scale = "medium", returnclass = "sf")
Europe <- world[which(world$continent == "Europe"),]

artifact_map <- ggplot(Europe) +
  geom_sf() +
  geom_sf(data = signbase_sf) +
  coord_sf(xlim = c(-10,30), 
           ylim = c(35,53), 
           expand = FALSE) +
  geom_text_repel(data = signbase_sf,
                  aes(x = longitude ,
                      y = latitude,
                      label = site_name),
                  max.overlaps = 30,
                  size = 2) +
  theme_minimal() +
  labs(size = "Sign Count")

date_graph <- ggplot(signbase_years_cal) +
  aes(x = MedianBP) +
  geom_bar(width = 150) +
  labs(x = "Median calibrated BP")
signbase_3000 <- signbase_years_cal
signbase_3000$site_name <- signbase_years$site_name
signbase_3000 <- signbase_3000 %>% 
  filter(MedianBP < 30000)

abundance_graph <- ggplot(abundance_data) +
  aes(x = reorder(site_name, sign_total),
      y = sign_total) +
  geom_col() +
  rotate_x_text(angle = 90, hjust = NULL, vjust = NULL) +
  labs(x = "Site name",
       y = "Sign count")

library(cowplot)
plot_grid(date_graph, abundance_graph, artifact_map,
          labels = "AUTO")

```

Our data for this paper comes from SignBase, an open-access database collection of geometric signs on mobile objects from Paleolithic-era Europe, which was launched in 2020, with the explicit goal being to enable quantitative analyses of abstract geographical representation [@dutkiewicz2020]. The creators of the database identified around 30 distinct geometric sign types, and recorded their occurrences on mobile artefacts from the Paleolithic. Identification and categorization of the sign types had already been performed by the creators of the SignBase database, the methods for which are described in @dutkiewicz2020, and we adhered to those categories and placements.

The data set consists of `r nartifacts` artifacts found at `r nsites` sites in `r ncountries` countries. The database records `r nsigntypes` geometric sign type found on these artifacts. We excluded signs if they fell under the "other" category, as the wide range of possibilities that could qualify as "other" means that it would be difficult to establish a definitive connection. We also excluded all data from Willendorf (only one sign present), and Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave (due to their extreme geographic distance from the other sites). After those sites were excluded, all signs under the "rectangle" category were also removed as a side effect of that exclusion. Lastly, data from Grotte De La Princesse Pauline and Šandalja II was excluded, due to being dated much earlier than the rest of the artifacts.

These exclusions resulted in `r nartifacts_clean` artifacts found at `r nsites_clean` sites in `r ncountries_clean` countries, encapsulating `r nsigntypes_clean` sign types, as seen in @fig-site-map. All our tests were run with presence/absence data, in order to keep the data from being affected by outliers like Vogelherd. We ran a Mantel test between an abundance distance matrix and and a Euclidean distance matrix in order to make sure that the exclusion of abundance would not significantly impact our results, and found that it was not statistically significant [@Lycett_2019] (P = `r p_value_abundance_correlation`, Mantel R = `r r_value_abundance_correlation`)

The artifacts in question are determined to have been made in the Paleolithic era, specifically in the Aurignacien technocomplex. The calibrated radiocarbon dates of the artifacts range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years. As seen in @fig-site-map, the dates have a few outliers, but tend to cluster between 35000 - 42500 BP.

## Testing for Cultural Groupings

Seriation analysis is the arrangement of data into a linear order to reveal patterns, and is a common technique in analyzing sets of archaeological data [@Hahsler_Hornik_Buchta]. Seriation in this paper was performed using code designed to mimic the Browner-Kile seriation algorithm [@Brower_Kile_1988], in which the data is presented in a matrix diagram, and a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal. This is done so by calculating the mean row and column positions of presences, and arranging the rows and columns by respectively.

Network analysis represents the relationship between nodes [@Mills_2017]. In this network analysis, each node is a site, and each edge is a measure of sign similarity, with the direct value represented by edge weight. The results were plotted with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths.

Principal coordinate decomposition analysis (PCoA) analyzes the eigenvalue decomposition of the Jaccard matrix and arranges the sites by dissimilarity [@Legendre_Legendre_2012]. This provides a visual of sign similarity as visualized by spacing and arrangement. The PCoA in this paper is calibrated to be plotted on a scatterplot, as opposed to a biplot.

## Testing Strength of Cultural Groupings

The modularity test quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006, @Newman_Girvan_2004]. The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings. The PerMANOVA test quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017], with the results expressed by an R-statistic, and a P-score.

PERMDISP is a multivariate extension of Levene's test [@Anderson_2006, @Anderson_Ellingsen_McArdle_2006], for the analysis of multivariate homogeneity of group dispersion. In this study, we perform the PERMDISP test upon the Jaccard matrix as a measure of compositional dissimilarity. We did so using the "betadisper" function from the RStudio package 'vegan', which mimics the PERMDISP2 procedure [@Vegan2001]. In this code, the non-euclidean distance matrix is reduced to principal coordinates to embed them within a Euclidean space. The results are used to produce a visual of the level of overlap between the groups in terms of sign content.

## Testing Isolation-by-Distance Hypothesis

The isolation-by-distance framework proposes that cultural difference can be primarily explained by geographic distance. Assuming that this framework is true, we would hypothetically see patterns of cultural features of Upper Paleolithic hominids that are dispersed according to geography. Conversely, if geographic distance cannot account for the distribution of the sign types, we would have to assume that other factors are at play, namely, movement of groups or the creation of social networks.

Mantel tests are used here to determine whether there is a correlation between sign distribution and geography. The Mantel test is run between a geographic distance matrix and a Jaccard dissimilarity matrix. The test seeks to determine the level of correspondence between the matrices through permuational evaluation of the null distribution [@Mantel_Smouse], producing a p-score as a result.

# Results

## Testing for Cultural Groupings

```{r}
#| label: fig-seriation
#| fig-cap: "Seriation analysis of the data"
## seriation
library(ggpubr)
artifact_matrix <- as.matrix(artifact_data %>% 
 mutate(across(everything(), ~replace(., . > 1, 1))))
rownames(artifact_matrix) <- rownames(artifact_data)
mplot <- function(x,...){
  d <- data.frame(x = rep(1:ncol(x), each = nrow(x)),
                  y = rep(nrow(x):1, ncol(x)),
                  z = as.integer(x))
  ggplot(d, aes(x, y, fill = factor(z))) +
    geom_tile(col = "grey50", linewidth = 0.5) +
    scale_fill_manual(values = c("0" = "white", "1" = "black")) +
    coord_equal(expand = FALSE) +
    scale_x_continuous(breaks = 1:ncol(x), 
                       labels = colnames(x),
                       position = "top") +
    scale_y_continuous(breaks = 1:nrow(x), 
                       labels = rev(rownames(x))) +
    theme(axis.text = element_text(size = rel(0.5)),
          axis.ticks = element_blank(),
          legend.position = "none",
          plot.title = element_text(hjust = 0.25)) +
    labs(x = "Sign Type",
         y = "Site Name",
         fill = "Count") +
    rotate_x_text(angle = 90, hjust = NULL, vjust = NULL)}
increase_focus <- function(x){
  mcp <- apply(
    x, 
    MARGIN = 1, 
    FUN = \(z){ mean(which(z == 1)) }, 
    simplify = FALSE
  ) |> unlist()
  x <- x[order(mcp), ]
  mrp <- apply(
    x, 
    MARGIN = 2, 
    FUN = \(z){ mean(which(z == 1)) }, 
    simplify = FALSE) |> unlist()
  x[, order(mrp)]}
concentrate <- function(x){
  old <- x
  not_identical <- TRUE
  while(not_identical){
    new <- increase_focus(old)
    not_identical <- !identical(old, new)
    old <- new}
  new}
artifact_matrix |> concentrate() |> mplot()

```

```{r}
#| label: fig-neighbor-net
#| fig-cap: "Neighbor-Net analysis of the data"
library(phangorn)
library(ggraph)
library(statnet)
jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
dm <- as.matrix(jac)
disim <- 1 - dm
disim[disim<0.5] <- 0
disim_net <- network(disim,
                    directed=F,
                    ignore.eval=F,
                    names.eval='weight')
disim_net %v% 'vertex.names' <- 
  row.names(artifact_data)

set.seed(500)

ggraph(disim_net, layout="fr") +
  geom_edge_link(aes(width = weight, 
                     alpha=weight), 
                 edge_colour = "black", 
                 show.legend=T) +
  scale_edge_width(range=c(0.5,1.5)) +
  scale_edge_colour_gradient(low = "#CCCCCC",
                             high = "#000000") +
  geom_node_point(alpha=1, size=5) +
  geom_node_text(aes(label=row.names(artifact_data)), 
                 size=6, 
                 family= "Arial", 
                 repel=T) +
  theme(text= element_text(size = 15), 
        panel.background = element_rect(fill = 'white'))
```

```{r}
#| label: fig-pcoa
#| fig-cap: "PCoA analysis of the data"
library(ecodist)
library(ape)
pcoa <- pcoa(jac)
pcoa_df <- data.frame(pcoa1 = pcoa$vectors[,1], 
                      pcoa2 = pcoa$vectors[,2])
pcoa_df <- pcoa_df %>% 
  mutate(site_name = rownames(artifact_data))
library(ggrepel)
ggplot(pcoa_df) +
  aes(x=pcoa1, 
      y=pcoa2) +
  geom_point() +
  labs(x = "PC1",
       y = "PC2",) +
  theme(title = element_text(size = 10)) +
  geom_text_repel(aes(label = site_name), 
                  size = 2,
                  max.overlaps = 20)
```

Combining the results of the above tests allowed us to find 11 groups of sites, as shown in @fig-group-table. The geographic distributions of the groups are shown in @fig-group-map, and the sign makeup of each group is shown in @fig-group-sign.

## Groups

```{r}
#| label: fig-group-table
#| fig-cap: "Table of the sites by group"
library(knitr)
group_1 <- c("Abri Pataud, El Rascaño, Fumane, La Viña, Lhotka, Pod Hradem, Riparo Bombrini, Sirgenstein Cave, Šandalja II")
group_2 <- c("Kvasice, Labeko Koba, Les Rois, Slatinice, Göpfelsteinhöhle")
group_3 <- c("Aurignac, d'Engihoul, Istállóskő Cave, Hohlenstein-Stadel, Gargas, Grotte de la Verpillière I, Mladeč, Brassempouy, Gatzarria")
group_4 <- c("Wildscheuer, Breitenbach, Peskő Cave, Vindija Cave, El Castillo")
group_5 <- c("Tuto de Camalhot, Abri Lartet/Gorge d'Enfer, Castanet, Le Trou Du Renard, Grottes de Fonds-de-Forêt")
group_6 <- c("Nová Dědina, Menton/Grottes du Grimaldi, Grotte du Renne, Solutré, Žlutava, Spy")
group_7 <- c("Trou Magrite, Grotte De La Princesse Pauline, Grotte de Goyet, Bockstein-Törle")
group_8 <- c("Trou al'Wesse, Maisières-Canal, Les Cottés, Le Terme Pialat")
group_9 <- c("Vogelherd, Hohle Fels, Geissenklösterle")
group_10 <- c("Cellier, Blanchard, La Ferrassie, Hornos de la Pena, Otaslavice")
group_11 <- c("La Souquette, Shelter Birów IV, Abri de Laussel, Abri du Poisson")
group_table <- data_frame("Group" = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11"), "Site Names" = c(group_1, group_2, group_3, group_4, group_5, group_6, group_7, group_8, group_9, group_10, group_11))
kable(group_table)

```

```{r fig.height= 5, fig.width=10}
#| label: fig-group-map
#| fig-cap: "Map of sites faceted and colored by group number"

signbase_sf_group <- 
  st_as_sf(signbase_unique_groups, 
           coords = c("longitude", "latitude"),
           remove = FALSE,
           crs = 4326)
ggplot(Europe) +
  geom_sf() +
  geom_sf(data = signbase_sf_group,
          aes(colour = as.factor(group))) +
  coord_sf(xlim = c(-10,30), 
           ylim = c(35,53), 
           expand = FALSE) +
  geom_text_repel(data = signbase_sf_group,
                  aes(x = longitude ,
                      y = latitude,
                      label = site_name),
                  max.overlaps = 100,
                  size = 2) +
  theme_minimal() +
  facet_wrap(~group) +
  labs(color = "Group Number")

```

```{r fig.height= 10, fig.width=10}
#| label: fig-group-sign
#| fig-cap: "Bar plot of signs by group"
graph_data <- artifact_data %>% 
  mutate(across(everything(), ~replace(., . > 1, 1))) %>% 
  concentrate() %>% 
  rownames_to_column("site_name") %>% 
  left_join(group_df)
sign_type_data <- graph_data %>% 
  pivot_longer(cols= notch:hashtag,
               names_to = "sign_type",
               values_to = "count") %>% 
  dplyr::filter(count != 0) %>% 
  mutate(group = parse_number(group))
x_order <- graph_data %>% 
  dplyr::select(-site_name,
         -group) %>% 
  colnames() 
ggplot(sign_type_data) +
  aes(x = sign_type) +
  geom_bar() +
  rotate_x_text(angle = 90, hjust = NULL, vjust = NULL) +
  labs(x = x_order) +
  scale_x_discrete(limits = c(x_order)) +
  scale_y_continuous(labels = NULL) +
  facet_wrap(~group, nrow = 11)

```

## Testing Strength of Cultural Groupings

```{r}
##Modularity test
library(igraph)
set.seed(500)
modul_matrix <-
  graph_from_adjacency_matrix(1 - dm, mode = "undirected")
mem <- as_membership(parse_number(signbase_unique_groups$group))
as_group <- modularity(modul_matrix, membership = mem)
```

```{r}
##permanova test
perman <- adonis2(jac~as.factor(signbase_unique_groups$group), 
                  method = "jaccard",
                  sqrt.dist = TRUE)
perman <- perman %>% 
  data_frame() %>% 
  slice_head()
perman_r <- perman$R2
perman_p <- perman$`Pr(>F)`

```

The modularity test, when run on the above groupings, resulted in a score of `r as_group`. The perMANOVA test produced an R-squared value of `r perman_r` and a P value of `r perman_p`, which tells us that the groupings are statisically significant, and that 52% of the variability in the sign distribution can be attributed to group membership. The multivariate homogeneity test of group dispersions produced @fig-perm-disp, showing that the overlap between groups is fairly minimal.

```{r fig.height= 8, fig.width=5}
#| label: fig-perm-disp
#| fig-cap: "Multivariate homogeneity of group dispersions"

set.seed(500)
dispersion <- betadisper(jac, group = signbase_unique_groups$group)
plot(dispersion, hull = FALSE, ellipse=TRUE)

```

## Testing Isolation-by-Distance Hypothesis

```{r}
## Mantel test for geography vs signs
set.seed(200)
site_distances <- dist(cbind(signbase_unique_groups$longitude, signbase_unique_groups$latitude))
mantel_distance <- vegan::mantel(jac, site_distances, permutations = 1000)
p_value_distance_correlation <- round(mantel_distance$signif, 3)
r_value_distance_correlation <- round(mantel_distance$statistic, 3)
```

```{r}
#| label: fig-distance-mantel
#| fig-cap: "Mantel correlogram showing the results of a mantel test run between a geographic distance matrix and a dissimilarity matrix for site sign types"
distance_corr <- mantel.correlog(jac, site_distances, n.class=50, break.pts=NULL, cutoff=FALSE, r.type="pearson", nperm=999, mult="holm", progressive=TRUE)
plot(distance_corr)
```

In terms of geographic correlation, there is no significant statistical correlation between the geographic distance matrix and the sign type dissimilarity matrix (P = `r p_value_distance_correlation`, Mantel R = `r r_value_distance_correlation`). Again, the results are displayed in a mantel correlogram (@fig-distance-mantel), which demonstrates that similarity and geographic distance do not experience a clearly correlated relationship.

# Discussion

In this paper, we set out to ask if identifiable and relevant patterns existed in the data, and if so, what could they tell us about the patterns of diversity and movement among Upper Paleolithic hominids?

Through our various analyses, we saw that the isolation-by-distance framework could not account for the distribution of sign types. Specifically, our Mantel test (@fig-distance-mantel) found no statistically significant relationship between geographic distance and sign variation. Therefore, we must look to other factors to explain the distribution. We also saw distinct groupings in the data, as demonstrated in @fig-seriation, @fig-neighbor-net, and @fig-pcoa. We divided the sites into approximately ten groups based on sign types with a modularity score of `r as_group`. @fig-group-map shows groups with sites both clustered geographically speaking, and more distant from each other, potentially showing us cultural groups and patterns of movement.

Mobility for various reasons is commonly found in hunter-gatherer groups. Therefore, one explanation for the sign type distribution could be that the groupings of the data represent cultural groups, as indicated by sign type patterns, and the geographic distribution between sites represent group mobility. The mobility may have occurred for all the above reasons. One point of note is that, according to @Whallon_2006, decorative objects may be exchanged as a method of network formation, which may account for the presence of some signs at some sites.

Further research will need to be done on this topic in order to understand these patterns. In particular, the ecological and biogeographical conditions at the sites will need to be analyzed for potential connections, as mobility is heavily environmentally-motivated. Additionally, further analysis of the sign type categories as created by @dutkiewicz2020, as it is possible, when it comes to the simpler signs, for identical creation to occur independently.

However, this preliminary research has been able to produce some conclusions to be reasonably confident in. Firstly, given that we know definitively that distribution and variation of sign types is not the result of geographic distance, we can conclude that the data is likely relevant to understanding patterns of cultural diversity. Secondly, we have found clusters of sites that correspond in sign type, possibly indicating mobility of specific cultural groups. These conclusions contribute to, and broaden our current understanding of cultural and biological diversity among Upper Paleolithic hominids in Europe.

# Conclusion

### 2025-01-14 Extra/download-images-from-database.R

@@ -1,0 +1,52 @@
+
+library(tidyverse)
+library(rvest)
+library(httr)
+
+signbase_full <- read_csv("signBase_Version1.0.csv")
+
+# find objects with hatching or grid
+sb_grid_hatch <- 
+signbase_full %>% 
+  filter(grid == 1 | hatching == 1)
+
+# urls to each object data webpage
+urls_object_webpage <- 
+  paste0("https://www.signbase.org/artifacts/",
+         sb_grid_hatch$access_id)
+
+# get images of objects with hatching or grid
+img_urls <- 
+  map(
+  urls_object_webpage,
+  ~.x %>%
+  read_html() %>% 
+  html_nodes("img.img-fluid.img-thumbnail") %>%  
+  html_attr("src")
+  )
+
+# add to data frame so we have urls together with other data
+# about the objects
+sb_grid_hatch$img_urls  <- unlist(img_urls)
+
+# drop objects with no image
+sb_grid_hatch_img <- 
+sb_grid_hatch %>% 
+  filter(img_urls != "../../static/images/signbase_logo.png" )
+
+for(i in 1:length(sb_grid_hatch_img$img_urls)){
+  
+  # work on this image
+  img_url <- sb_grid_hatch_img$img_urls[i]
+  
+  # download the image with an informative 
+  # filename
+if (!is.na(img_url)) {
+  img <- GET(img_url)
+  writeBin(content(img, "raw"), 
+           paste0("images/signbase-", 
+                  sb_grid_hatch_img$object_id[i], 
+           ".jpg"))  
+           }
+}
+  

### 2025-01-14 Extra/signbase-network-options.R

@@ -1,0 +1,115 @@
+#| label: fig-neighbor-net
+#| fig-cap: "Neighbor-Net analysis of the data"
+##network analysis
+library(vegan)
+library(phangorn)
+
+## neighborNet
+jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
+
+dm <- as.matrix(jac)
+nnet <- neighborNet(dm)
+
+par("mar" = rep(1, 4))
+
+# Reorder the tips according to the seriation grouping
+nnet$tip.label <- nnet$tip.label[match(group_df$site_name, nnet$tip.label)]  
+
+# Plot the network without tip labels initially
+plot(nnet, 
+     show.tip.label = TRUE,
+     cex = 0.5)
+
+
+## try new neighbor net
+library(ggraph)
+library(tidygraph)
+library(igraph)
+library(statnet)
+library(intergraph)
+library(FastKNN)
+library(ggpubr)
+library(reshape2)
+library(patchwork)
+library(ggmap)
+library(seriation)
+library(purrr)
+library(tnet)
+library(argosfilter)
+
+disim <- 1 - dm
+disim[disim<0.5] <- 0
+disim_net <- network(disim,
+                     directed=F,
+                     ignore.eval=F,
+                     names.eval='weight')
+
+disim_net %v% 'vertex.names' <- 
+  row.names(artifact_data)
+
+set.seed(500)
+
+ggraph(disim_net, layout="fr") +
+  geom_edge_link(aes(width = weight, 
+                     alpha=weight), 
+                 edge_colour = "black", 
+                 show.legend=T) +
+  scale_edge_width(range=c(0.5,1.5)) +
+  scale_edge_colour_gradient(low = "#CCCCCC",
+                             high = "#000000") +
+  geom_node_point(alpha=1, size=5) +
+  geom_node_text(aes(label=row.names(artifact_data)), 
+                 size=6, 
+                 family= "Arial", 
+                 repel=T) +
+  theme(text= element_text(size = 20), 
+        panel.background = element_rect(fill = 'white'))
+
+# BM try yet another
+
+# Ensure all edge lengths are positive
+nnet$edge.length <- pmax(nnet$edge.length, 1e-6)
+
+# Create the graph object
+g <- graph_from_edgelist(nnet$edge, directed = FALSE)
+E(g)$weight <- nnet$edge.length
+
+# Create a layout for the nodes
+layout <- layout_with_fr(g)  # Fruchterman-Reingold layout
+
[…]

### 2025-01-14 Extra/signbase_groups.R

@@ -1,0 +1,47 @@
+jac_df <- as_tibble(dm)
+rownames(jac_df) <- rownames(artifact_data)
+jac_df <- jac_df %>% 
+  rownames_to_column("site_name") %>% 
+  pivot_longer(cols = `Abri Lartet/Gorge d'Enfer`: Žlutava) %>% 
+  filter(value < 0.6)
+
+
+pia <- jac_df %>% 
+  filter(site_name =="Nová Dědina")
+
+deng <- 
+
+groups <- list("1" = c("Abri Pataud", "El Rascaño", 
+                         "Fumane", "La Viña",
+                         "Lhotka", "Pod Hradem",
+                         "Riparo Bombrini", "Sirgenstein Cave",
+                         "Šandalja II"),
+                "2" = c("Kvasice", "Labeko Koba", 
+                        "Les Rois", "Slatinice",
+                        "Göpfelsteinhöhle"),
+                "3" = c("Aurignac", "d'Engihoul",
+                         "Istállóskő Cave", "Hohlenstein-Stadel",
+                         "Gargas", "Grotte de la Verpillière I", 
+                         "Mladeč", "Brassempouy",
+                         "Gatzarria"),
+                "4" = c("Wildscheuer", "Breitenbach",
+                        "Peskő Cave", "Vindija Cave",
+                        "El Castillo"),
+                "5" = c("Tuto de Camalhot","Abri Lartet/Gorge d'Enfer",
+                        "Castanet", "Le Trou Du Renard", 
+                        "Grottes de Fonds-de-Forêt"),
+                "6" = c("Nová Dědina", "Menton/Grottes du Grimaldi",
+                        "Grotte du Renne", "Solutré", 
+                        "Žlutava", "Spy"),
+                "7" = c("Trou Magrite", "Grotte De La Princesse Pauline",
+                        "Grotte de Goyet", "Bockstein-Törle"),
+                "8" = c("Trou al'Wesse", "Maisières-Canal",
+                        "Les Cottés", "Le Terme Pialat"),
+                "9" = c("Vogelherd","Hohle Fels",
+                        "Geissenklösterle"),
+                "10" = c("Cellier", "Blanchard", 
+                         "La Ferrassie", "Hornos de la Pena",
+                         "Otaslavice"),
+                "11" = c("La Souquette", "Shelter Birów IV",
+                         "Abri de Laussel", "Abri du Poisson"))
+  

### 2025-01-14 eco-regions-signbase.R

@@ -1,0 +1,165 @@
+library(tidyverse)
+library(magrittr)
+library(rnaturalearth)
+library(rnaturalearthdata)
+library(ggrepel)
+library(sf)
+library(terra)
+signbase_full <- read_csv("data/signBase_Version1.0.csv")
+signbase_full_clean <- signbase_full %>% 
+  filter(site_name != "Willendorf",
+         site_name != "Riparo di Fontana Nuova",
+         site_name != "Muralovka",
+         site_name != "Shanidar Cave",
+         site_name != "Hayonim Cave",
+         site_name != "El Salitre") %>% 
+  select(-other, -rectangle)
+
+lat_long_df <- signbase_full_clean %>% 
+  select(site_name, longitude, latitude) %>% 
+  distinct(site_name, .keep_all = TRUE)
+
+signbase_unique_groups <- signbase_full_clean %>% 
+  mutate(longitude = as.character(longitude),
+         latitude = as.character(latitude)) %>% 
+  group_by(site_name) %>%
+  summarize(across(where(is.numeric), sum)) %>% 
+  left_join(lat_long_df) %>% 
+  column_to_rownames("site_name") 
+
+artifact_div <- signbase_unique_groups %>% 
+  select(line:star) %>% 
+  vegan::diversity(index = "shannon") %>% 
+  tibble() 
+
+colnames(artifact_div) <- "diversity"
+
+abundance_data <- signbase_full_clean %>% 
+  pivot_longer(cols= line:star,
+               values_to = "sign_total",
+              names_to = "sign_type") %>% 
+  filter(sign_total != 0) %>% 
+  mutate(longitude = as.character(longitude),
+         latitude = as.character(latitude)) %>% 
+  group_by(site_name) %>%
+  summarize(across(where(is.numeric), sum)) %>% 
+  left_join(lat_long_df) %>% 
+  select(site_name, sign_total,  longitude, latitude)
+
+signbase_sf <- 
+  st_as_sf(abundance_data, 
+           coords = c("longitude", "latitude"),
+           remove = FALSE,
+           crs = 4326)
+
+signbase_sf <- signbase_sf %>% 
+  add_column(artifact_div)
+
+##Environmental Zones in Europe, 
+##as categorized by the Environmental Stratification of Europe (EnS) dataset (European Environmental Agency)
[…]

### 2025-01-14 species_dist.R

@@ -1,0 +1,139 @@
+library(geodata)
+library(predicts)
+library(dismo)
+list_of_rasters <- c("data/cclgmbi_2-5m/cclgmbi1.tif", "data/cclgmbi_2-5m/cclgmbi10.tif", 
+                     "data/cclgmbi_2-5m/cclgmbi11.tif", "data/cclgmbi_2-5m/cclgmbi12.tif",
+                     "data/cclgmbi_2-5m/cclgmbi13.tif", "data/cclgmbi_2-5m/cclgmbi14.tif",
+                     "data/cclgmbi_2-5m/cclgmbi15.tif", "data/cclgmbi_2-5m/cclgmbi16.tif",
+                     "data/cclgmbi_2-5m/cclgmbi17.tif", "data/cclgmbi_2-5m/cclgmbi18.tif",
+                     "data/cclgmbi_2-5m/cclgmbi19.tif", "data/cclgmbi_2-5m/cclgmbi2.tif",
+                     "data/cclgmbi_2-5m/cclgmbi3.tif", "data/cclgmbi_2-5m/cclgmbi4.tif",
+                     "data/cclgmbi_2-5m/cclgmbi5.tif", "data/cclgmbi_2-5m/cclgmbi6.tif",
+                     "data/cclgmbi_2-5m/cclgmbi7.tif", "data/cclgmbi_2-5m/cclgmbi8.tif",
+                     "data/cclgmbi_2-5m/cclgmbi9.tif")
+clim_data <- raster::stack(list_of_rasters)
+clim_data <- rast(clim_data)
+
+max_lat <- ceiling(max(obs_data$latitude))
+min_lat <- floor(min(obs_data$latitude))
+max_lon <- ceiling(max(obs_data$longitude))
+min_lon <- floor(min(obs_data$longitude))
+geographic_extent <- ext(x = c(min_lon, max_lon, min_lat, max_lat))
+
+
+sample_extent <- geographic_extent * 1.25
+
+clim_data <- terra::crop(x = clim_data, y = sample_extent)
+
+obs_data <- st_as_sf(signbase_unique_groups, 
+           coords = c("longitude", "latitude"),
+           remove = FALSE,
+           crs = 4326)
+
+obs_data$climate <- extract(clim_data, obs_data)
+
+obs_data1 <- obs_data %>% 
+  filter(group == 1) %>% 
+  dplyr::select(latitude, longitude)
+
[…]

### 2025-01-14 Extra/eco-regions-signbase.R
Initial commit

@@ -1,0 +1,87 @@
+##Environmental Zones in Europe, 
+##as categorized by the Environmental Stratification of Europe (EnS) dataset (European Environmental Agency)
+##link to download : https://www.eea.europa.eu/en/datahub/datahubitem-view/c8c4144a-8c0e-4686-9422-80dbf86bc0cb?activeAccordion=1082728
+library(raster)
+library(terra)
+eco_raster <- terra::rast("data/eea_r_3035_1_km_env-zones_p_2018_v01_r00.tif")
+
+signbase_sf$environmental_zone <- extract(eco_raster, signbase_sf)
+
+max_lat <- ceiling(max(signbase_sf$latitude))
+min_lat <- floor(min(signbase_sf$latitude))
+max_lon <- ceiling(max(signbase_sf$longitude))
+min_lon <- floor(min(signbase_sf$longitude))
+# Store boundaries in a single extent object
+geographic_extent <- ext(x = c(min_lon, max_lon, min_lat, max_lat))
+
+eco_raster_crop = crop(x = eco_raster, y = geographic_extent)
+
+
+signbase_sf <- signbase_sf %>% 
+  mutate(environmental_zone = environmental_zone$eea_r_3035_1_km_env)
+
+plot(eco_raster)
+points(x = signbase_sf$longitude, 
+       y = signbase_sf$latitude, add = TRUE)
+
+##looking at distribution of signs within environmental zones
+signbase_unique_groups$environmental_zone<- signbase_sf$environmental_zone
+
+abundance_data_environment <- signbase_unique_groups %>% 
+  pivot_longer(cols= line:star,
+               values_to = "sign_total",
+               names_to = "sign_type") %>% 
+  filter(sign_total != 0) 
+
+library(ggpubr)
+ggplot(abundance_data_environment) +
+  aes(x = sign_type) +
+  geom_bar() +
+  facet_wrap(~environmental_zone) +
+  rotate_x_text(angle = 90, hjust = NULL, vjust = NULL)
+
+## Biogeographical regions of Europe - (https://pmc.ncbi.nlm.nih.gov/articles/PMC7340631/#sec24)
+#3 as determined by the European Environment Agency
+
+bio_geo_raster <- rast("data/bdj-08-e53720-s002.tif")
+
+signbase_sf$biogeographical_region <- extract(bio_geo_raster, signbase_sf)
+
+signbase_sf <- signbase_sf %>% 
[…]

### 2025-01-14 Extra/signbase-diversity.R

@@ -1,0 +1,57 @@
+library(vegan)
+
+artifact_div <- vegan::diversity(artifact_data, index = "shannon")
+
+abundance_data$diversity <- artifact_div
+
+##plot diversity against abundance
+
+ggplot(abundance_data) +
+  aes(x = sign_total,
+      y = diversity) +
+  geom_point() +
+  geom_smooth(se = FALSE)
+
+## plot diversity by site
+library(ggpubr)
+ggplot(abundance_data) +
+  aes(x = site_name, y = sign_total, fill = diversity) +
+  geom_col() +
+  rotate_x_text(angle = 90, hjust = NULL, vjust = NULL)
+
+## map diversity by site
+signbase_sf <- 
+  st_as_sf(abundance_data, 
+           coords = c("longitude", "latitude"),
+           remove = FALSE,
+           crs = 4326)
+
+
+world <- ne_countries(scale = "medium", returnclass = "sf")
+Europe <- world[which(world$continent == "Europe"),]
+
+ggplot(Europe) +
+  geom_sf() +
+  geom_sf(data = signbase_sf,
+          aes(size = diversity)) +
+  scale_size_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1, 1.25, 1.5, 1.75, 2)) +
+  coord_sf(xlim = c(-10,30), 
+           ylim = c(35,53), 
+           expand = FALSE) +
+  geom_text_repel(data = signbase_sf,
+                  aes(x = longitude ,
+                      y = latitude,
+                      label = site_name),
+                  max.overlaps = 100,
+                  size = 2) +
+  theme_minimal() +
+  facet_wrap(~group) +
+  labs(color = "Diversity",
+       size = "Sign Count")
+
+
+## Calculate the diversity of groups
+div_groups <- abundance_data %>% pull(group)
+artifact_div_groups <- diversity(artifact_data, index = "shannon", groups = div_groups)
+group_diversity <- data_frame(artifact_div_groups) %>% 
+  rownames_to_column("group")

### 2025-01-14 Extra/signbase_map.R

@@ -1,0 +1,125 @@
+glibrary(magrittr)
+library(tidyverse)
+
+
+abundance_data <- signbase_full_clean %>% 
+  pivot_longer(cols= line:star,
+               values_to = "sign_total") %>% 
+  filter(sign_total != 0) %>% 
+  mutate(longitude = as.character(longitude),
+         latitude = as.character(latitude)) %>% 
+  group_by(site_name) %>%
+  summarize(across(where(is.numeric), sum))
+
+
+abundance_data <- abundance_data %>% 
+  left_join(lat_long_df) %>% 
+  left_join(group_df) %>% 
+  select(site_name, sign_total, longitude, latitude, group)
+
+
+library(rnaturalearth)
+library(rnaturalearthdata)
+library(ggrepel)
+library(sf)
+
+signbase_sf <- 
+  st_as_sf(abundance_data, 
+           coords = c("longitude", "latitude"),
+           remove = FALSE,
+           crs = 4326)
+
+
+world <- ne_countries(scale = "medium", returnclass = "sf")
+Europe <- world[which(world$continent == "Europe"),]
+
+ggplot(Europe) +
+  geom_sf() +
+  geom_sf(data = signbase_sf,
+          aes(colour = group,
+              size = sign_total)) +
+  geom_text_repel(data = signbase_sf,
+                  aes(x = longitude ,
+                      y = latitude,
+                      label = site_name),
+                  max.overlaps = 100,
+                  size = 2) +
+  coord_sf(xlim = c(-10,30), 
+           ylim = c(35,53), 
+           expand = FALSE) +
+  theme_minimal() +
+  facet_wrap(~group)
+
+
+## GIS with raster data - data from https://ec.europa.eu/eurostat/web/gisco/geodata/digital-elevation-model/eu-dem (DD dataset)
+library(terra)
+library(tidyterra)
+
+raster_data <- rast("large-files/EU_DEM_mosaic_5deg/eudem_dem_4258_europe.tif")
+
+signbase_sf$elevation <- extract(raster_data, signbase_sf)
+
+raster_data <- crop(raster_data, signbase_sf)
+
+aspect_data <- terrain(raster_data, v = "aspect", unit = "degrees", neighbors=8)
+slope_data <- terrain(raster_data, v = "slope", unit = "degrees", neighbors=8)
+
[…]

### 2025-01-15 Extra/eco-regions-signbase.R
update

@@ -1,87 +1,65 @@
 ##Environmental Zones in Europe, 
 ##as categorized by the Environmental Stratification of Europe (EnS) dataset (European Environmental Agency)
 ##link to download : https://www.eea.europa.eu/en/datahub/datahubitem-view/c8c4144a-8c0e-4686-9422-80dbf86bc0cb?activeAccordion=1082728
-library(raster)
+
 library(terra)
-eco_raster <- terra::rast("data/eea_r_3035_1_km_env-zones_p_2018_v01_r00.tif")
+library(tidyterra)
+env_raster <- rast("large-files//eea_r_3035_1_km_env-zones_p_2018_v01_r00.tif")
 
-signbase_sf$environmental_zone <- extract(eco_raster, signbase_sf)
+env_raster_reproj <- terra::project(env_raster, "epsg:4326")
 
-max_lat <- ceiling(max(signbase_sf$latitude))
-min_lat <- floor(min(signbase_sf$latitude))
-max_lon <- ceiling(max(signbase_sf$longitude))
-min_lon <- floor(min(signbase_sf$longitude))
-# Store boundaries in a single extent object
-geographic_extent <- ext(x = c(min_lon, max_lon, min_lat, max_lat))
-
-eco_raster_crop = crop(x = eco_raster, y = geographic_extent)
-
+signbase_sf$environmental_zone <- terra::extract(env_raster, signbase_sf)
 
 signbase_sf <- signbase_sf %>% 
   mutate(environmental_zone = environmental_zone$eea_r_3035_1_km_env)
 
-plot(eco_raster)
-points(x = signbase_sf$longitude, 
-       y = signbase_sf$latitude, add = TRUE)
 
-##looking at distribution of signs within environmental zones
-signbase_unique_groups$environmental_zone<- signbase_sf$environmental_zone
+##Plotting out distribution of signs over environmental zone raster
+ggplot(Europe) + 
+  geom_spatraster(data = env_raster_reproj) +
+  scale_fill_grass_c(palette = "viridis") +
+  geom_sf(data = signbase_sf) +
+  coord_sf(xlim = c(-10,30), 
+           ylim = c(35,53), 
+           expand = FALSE) +
+  facet_wrap(~group)
 
-abundance_data_environment <- signbase_unique_groups %>% 
-  pivot_longer(cols= line:star,
-               values_to = "sign_total",
[…]

### 2025-01-15 Extra/signbase_map.R

@@ -1,125 +1,0 @@
-glibrary(magrittr)
-library(tidyverse)
-
-
-abundance_data <- signbase_full_clean %>% 
-  pivot_longer(cols= line:star,
-               values_to = "sign_total") %>% 
-  filter(sign_total != 0) %>% 
-  mutate(longitude = as.character(longitude),
-         latitude = as.character(latitude)) %>% 
-  group_by(site_name) %>%
-  summarize(across(where(is.numeric), sum))
-
-
-abundance_data <- abundance_data %>% 
-  left_join(lat_long_df) %>% 
-  left_join(group_df) %>% 
-  select(site_name, sign_total, longitude, latitude, group)
-
-
-library(rnaturalearth)
-library(rnaturalearthdata)
-library(ggrepel)
-library(sf)
-
-signbase_sf <- 
-  st_as_sf(abundance_data, 
-           coords = c("longitude", "latitude"),
-           remove = FALSE,
-           crs = 4326)
-
-
-world <- ne_countries(scale = "medium", returnclass = "sf")
-Europe <- world[which(world$continent == "Europe"),]
-
-ggplot(Europe) +
-  geom_sf() +
-  geom_sf(data = signbase_sf,
-          aes(colour = group,
-              size = sign_total)) +
-  geom_text_repel(data = signbase_sf,
-                  aes(x = longitude ,
-                      y = latitude,
-                      label = site_name),
-                  max.overlaps = 100,
-                  size = 2) +
-  coord_sf(xlim = c(-10,30), 
-           ylim = c(35,53), 
-           expand = FALSE) +
-  theme_minimal() +
-  facet_wrap(~group)
-
-
-## GIS with raster data - data from https://ec.europa.eu/eurostat/web/gisco/geodata/digital-elevation-model/eu-dem (DD dataset)
-library(terra)
-library(tidyterra)
-
-raster_data <- rast("large-files/EU_DEM_mosaic_5deg/eudem_dem_4258_europe.tif")
-
-signbase_sf$elevation <- extract(raster_data, signbase_sf)
-
-raster_data <- crop(raster_data, signbase_sf)
-
-aspect_data <- terrain(raster_data, v = "aspect", unit = "degrees", neighbors=8)
-slope_data <- terrain(raster_data, v = "slope", unit = "degrees", neighbors=8)
-
[…]

### 2025-01-15 endofquarterpaper.qmd

@@
 This approach is supported by both ethnographic evidence throughout history, as well as similar archaeological studies done on ornamental and parietal art [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d].
-In this study, we use our analyses to attempt to answer the following questions: Do identifiable and relevant patterns of cultural groups exist in the data, and if so, what can they tell us about the patterns of diversity and movement among Upper Paleolithic hominids?
+In this study, we address the following questions: Are cultural groups evident in the distribution of geometric signs, and if so, what can they tell us about the cultural diversity and movement of Upper Paleolithic humans?
 
-<!--SC: I'm still having a difficult time figuring out just how to end this paragraph - you mentioned that I should have a research question ending in a question mark, but I'm struggling to figure out how to phrase it in a way that is appropriately formal  -->
+<!--BM: our position is that cultural groups exist, our contribution is to see if/how they are evident in the sign data  -->
 

### 2025-01-18 Extra/eco-regions-signbase.R
edits

@@ -8,11 +8,10 @@ env_raster <- rast("large-files//eea_r_3035_1_km_env-zones_p_2018_v01_r00.tif")
 
 env_raster_reproj <- terra::project(env_raster, "epsg:4326")
 
-signbase_sf$environmental_zone <- terra::extract(env_raster, signbase_sf)
+signbase_sf$environmental_zone <- terra::extract(env_raster_reproj, signbase_sf)
 
 signbase_sf <- signbase_sf %>% 
   mutate(environmental_zone = environmental_zone$eea_r_3035_1_km_env)
-
 
 ##Plotting out distribution of signs over environmental zone raster
 ggplot(Europe) + 
@@ -26,8 +25,8 @@ ggplot(Europe) +
 
 ## plot out env zone by group 
 ggplot(signbase_sf) +
-  aes(x = group) +
-  geom_bar(fill = environmental_zone)
+  aes(x = group, fill= as.factor(environmental_zone)) +
+  geom_bar()
 
 ## Biogeographical regions of Europe - (https://pmc.ncbi.nlm.nih.gov/articles/PMC7340631/#sec24)
 #3 as determined by the European Environment Agency
@@ -49,17 +48,18 @@ ggplot(Europe) +
   facet_wrap(~group)
 
 ## elevation data
+library(geodata)
+elev_rast <- elevation_global(res = 0.5, path = "/Users/sophiecooper/Independent Research Study/SignBase/large-files")
 
-elev_rast <- rast("large-files/EU_DEM_mosaic_5deg/eudem_dem_4258_europe.tif")
-elev_raster_reproj <- terra::project(elev_rast, "epsg:4326")
 
-signbase_sf$elevation <- extract(elev_raster_reproj, signbase_sf)
+signbase_sf$elevation <- extract(elev_rast, signbase_sf)
 
-elev_crop <- crop(elev_raster_reproj, signbase_sf)
 
-aspect_data <- terrain(elev_crop, v = "aspect", unit = "degrees", neighbors=8)
-slope_data <- terrain(elev_crop, v = "slope", unit = "degrees", neighbors=8)
+aspect_data <- terrain(elev_rast, v = "aspect", unit = "degrees", neighbors=8)
+slope_data <- terrain(elev_rast, v = "slope", unit = "degrees", neighbors=8)
 
 signbase_sf$aspect <- extract(aspect_data, signbase_sf)
 signbase_sf$slope <- extract(slope_data, signbase_sf)
+
+
 

### 2025-01-18 endofquarterpaper.qmd
edits

@@
 
-The Upper Paleolithic is characterized by a rise in early Homo sapiens, and a decline for other hominids.
-During this time, we see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe.
-One of the most distinctive qualities of the Upper Paleolithic is the large rise in artwork available from that era, both parietal and otherwise.
-Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012].
+One of the most distinctive qualities of the Upper Paleolithic is the large rise in available artwork, both parietal and mobile.
+Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the grouping patterns of early modern humans.
+During this period, we also see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe.
+Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca.
+43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].
 
@@
 In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe during the Aurignacian technocomplex.
-The basis of this analysis is the idea that variations on signs can serve as an indicator of group membership and cultural similarity.
+The basis of this analysis is the idea that variations in signs can serve as an indicator of group membership and cultural similarity.
 This approach is supported by both ethnographic evidence throughout history, as well as similar archaeological studies done on ornamental and parietal art [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d].
@@
 
-<!--BM: our position is that cultural groups exist, our contribution is to see if/how they are evident in the sign data  -->
-
 # Background
 
-Our understanding of the patterns of cultural and genetic diversity among the hominids of the Upper Paleolithic remains incomplete.
+Our understanding of the patterns of cultural and genetic diversity among the early modern humans of the Upper Paleolithic remains incomplete.
 Previous works have performed various cluster analysis methods upon variations in artefacts to represent and understand patterns of groups and cultural diversity.
@@
 Our use of geometric signs in this study is supported by @Sauvet_2018, in which used quadrilateral geometric parietal signs are used to identify Upper Paleolithic cultural groups in the Cantabrian region of Spain.
-The authors identified and classified the signs based on variations in form, then performed a factorial correspondence analysis and ascending hierarchical clustering upon the data in order to identify two clusters, which they argued represents distinct and separate groups of hominids.
+The authors identified and classified the signs based on variations in form, then performed a factorial correspondence analysis and ascending hierarchical clustering upon the data in order to identify two clusters, which they argued represents distinct and separate groups of humans.
 In their study, @Sauvet_2018 made a case for the relevance of signs as identity markers, pointing to both the relevant qualities of signs, which are similar to those explained above, and the case studies throughout history where they served a similar purpose.
@@
 abundance_graph <- ggplot(abundance_data) +
-  aes(x = reorder(site_name, sign_total),
-      y = sign_total) +
+  aes(x = sign_total, y = reorder(site_name, sign_total)) +
   geom_col() +
@@
   labs(x = "Site name",
-       y = "Sign count")
+       y = "Sign count") +
+  theme(axis.text.y = element_text(size = 4.25))
 
@@
 plot_grid(date_graph, abundance_graph, artifact_map,
-          labels = "AUTO")
+          labels = "AUTO", align = "v",
+          nrow = 3, scale = c(0.75, 0.75, 1.25))
 
@@
 The isolation-by-distance framework proposes that cultural difference can be primarily explained by geographic distance.
-Assuming that this framework is true, we would hypothetically see patterns of cultural features of Upper Paleolithic hominids that are dispersed according to geography.
+Assuming that this framework is true, we would hypothetically see patterns of cultural features of Upper Paleolithic early modern humans that are dispersed according to geography.
 Conversely, if geographic distance cannot account for the distribution of the sign types, we would have to assume that other factors are at play, namely, movement of groups or the creation of social networks.
@@
 
-In this paper, we set out to ask if identifiable and relevant patterns existed in the data, and if so, what could they tell us about the patterns of diversity and movement among Upper Paleolithic hominids?
+In this paper, we set out to ask if identifiable and relevant patterns existed in the data, and if so, what could they tell us about the patterns of diversity and movement among Upper Paleolithic humans?
 
@@
 Secondly, we have found clusters of sites that correspond in sign type, possibly indicating mobility of specific cultural groups.
-These conclusions contribute to, and broaden our current understanding of cultural and biological diversity among Upper Paleolithic hominids in Europe.
+These conclusions contribute to, and broaden our current understanding of cultural and biological diversity among Upper Paleolithic Homo sapiens in Europe.
 

### 2025-01-19 Extra/eco-regions-signbase.R
clim

@@ -62,4 +62,23 @@ signbase_sf$aspect <- extract(aspect_data, signbase_sf)
 signbase_sf$slope <- extract(slope_data, signbase_sf)
 
 
+##climate data
+
+clim_files <- c("large-files/cclgmbi_2-5m/cclgmbi1.tif", "large-files/cclgmbi_2-5m/cclgmbi10.tif",
+                "large-files/cclgmbi_2-5m/cclgmbi11.tif", "large-files/cclgmbi_2-5m/cclgmbi12.tif",
+                "large-files/cclgmbi_2-5m/cclgmbi13.tif", "large-files/cclgmbi_2-5m/cclgmbi14.tif",
+                "large-files/cclgmbi_2-5m/cclgmbi15.tif", "large-files/cclgmbi_2-5m/cclgmbi16.tif",
+                "large-files/cclgmbi_2-5m/cclgmbi17.tif", "large-files/cclgmbi_2-5m/cclgmbi18.tif",
+                "large-files/cclgmbi_2-5m/cclgmbi19.tif", "large-files/cclgmbi_2-5m/cclgmbi2.tif",
+                "large-files/cclgmbi_2-5m/cclgmbi3.tif", "large-files/cclgmbi_2-5m/cclgmbi4.tif",
+                "large-files/cclgmbi_2-5m/cclgmbi5.tif", "large-files/cclgmbi_2-5m/cclgmbi6.tif",
+                "large-files/cclgmbi_2-5m/cclgmbi7.tif", "large-files/cclgmbi_2-5m/cclgmbi8.tif",
+                "large-files/cclgmbi_2-5m/cclgmbi9.tif")
+
+clim_stack <- raster::stack(clim_files)
+
+clim_raster <- rast(clim_stack)
+
+
+
 

### 2025-01-19 eco-regions-signbase.R

@@ -1,165 +1,0 @@
-library(tidyverse)
-library(magrittr)
-library(rnaturalearth)
-library(rnaturalearthdata)
-library(ggrepel)
-library(sf)
-library(terra)
-signbase_full <- read_csv("data/signBase_Version1.0.csv")
-signbase_full_clean <- signbase_full %>% 
-  filter(site_name != "Willendorf",
-         site_name != "Riparo di Fontana Nuova",
-         site_name != "Muralovka",
-         site_name != "Shanidar Cave",
-         site_name != "Hayonim Cave",
-         site_name != "El Salitre") %>% 
-  select(-other, -rectangle)
-
-lat_long_df <- signbase_full_clean %>% 
-  select(site_name, longitude, latitude) %>% 
-  distinct(site_name, .keep_all = TRUE)
-
-signbase_unique_groups <- signbase_full_clean %>% 
-  mutate(longitude = as.character(longitude),
-         latitude = as.character(latitude)) %>% 
-  group_by(site_name) %>%
-  summarize(across(where(is.numeric), sum)) %>% 
-  left_join(lat_long_df) %>% 
-  column_to_rownames("site_name") 
-
-artifact_div <- signbase_unique_groups %>% 
-  select(line:star) %>% 
-  vegan::diversity(index = "shannon") %>% 
-  tibble() 
-
-colnames(artifact_div) <- "diversity"
-
-abundance_data <- signbase_full_clean %>% 
-  pivot_longer(cols= line:star,
-               values_to = "sign_total",
-              names_to = "sign_type") %>% 
-  filter(sign_total != 0) %>% 
-  mutate(longitude = as.character(longitude),
-         latitude = as.character(latitude)) %>% 
-  group_by(site_name) %>%
-  summarize(across(where(is.numeric), sum)) %>% 
-  left_join(lat_long_df) %>% 
-  select(site_name, sign_total,  longitude, latitude)
-
-signbase_sf <- 
-  st_as_sf(abundance_data, 
-           coords = c("longitude", "latitude"),
-           remove = FALSE,
-           crs = 4326)
-
-signbase_sf <- signbase_sf %>% 
-  add_column(artifact_div)
-
-##Environmental Zones in Europe, 
-##as categorized by the Environmental Stratification of Europe (EnS) dataset (European Environmental Agency)
[…]

### 2025-01-19 Extra/eco-regions-signbase.R
climate

@@ -62,6 +62,8 @@ signbase_sf$aspect <- extract(aspect_data, signbase_sf)
 signbase_sf$slope <- extract(slope_data, signbase_sf)
 
 
+
+
 ##climate data
 
 clim_files <- c("large-files/cclgmbi_2-5m/cclgmbi1.tif", "large-files/cclgmbi_2-5m/cclgmbi10.tif",
@@ -75,10 +77,23 @@ clim_files <- c("large-files/cclgmbi_2-5m/cclgmbi1.tif", "large-files/cclgmbi_2-
                 "large-files/cclgmbi_2-5m/cclgmbi7.tif", "large-files/cclgmbi_2-5m/cclgmbi8.tif",
                 "large-files/cclgmbi_2-5m/cclgmbi9.tif")
 
-clim_stack <- raster::stack(clim_files)
+mean_temp_rast <- rast("large-files/cclgmbi_2-5m/cclgmbi1.tif")
 
-clim_raster <- rast(clim_stack)
+clim_raster<- rast(clim_files)
 
 
+signbase_sf$climate <- extract(clim_raster, signbase_sf)
 
+mean_temp_rast <- rast("large-files/cclgmbi_2-5m/cclgmbi1.tif")
+
+signbase_sf$mean_temperature <- extract(mean_temp_rast, signbase_sf)
+
+ggplot(Europe) + 
+  geom_spatraster(data = mean_temp_rast) +
+  scale_fill_gradient(low = "blue", high = "red") +
+  geom_sf(data = signbase_sf) +
+  coord_sf(xlim = c(-10,30), 
+           ylim = c(35,53), 
+           expand = FALSE) +
+  facet_wrap(~group)
 

### 2025-01-22 Extra/eco-regions-signbase.R
eco edits

@@ -13,20 +13,24 @@ signbase_sf$environmental_zone <- terra::extract(env_raster_reproj, signbase_sf)
 signbase_sf <- signbase_sf %>% 
   mutate(environmental_zone = environmental_zone$eea_r_3035_1_km_env)
 
+
 ##Plotting out distribution of signs over environmental zone raster
 ggplot(Europe) + 
   geom_spatraster(data = env_raster_reproj) +
-  scale_fill_grass_c(palette = "viridis") +
+  scale_fill_grass_d(palette = "viridis") +
   geom_sf(data = signbase_sf) +
   coord_sf(xlim = c(-10,30), 
            ylim = c(35,53), 
            expand = FALSE) +
   facet_wrap(~group)
 
+unique(signbase_sf$environmental_zone)
 ## plot out env zone by group 
 ggplot(signbase_sf) +
   aes(x = group, fill= as.factor(environmental_zone)) +
   geom_bar()
+
+#3 plot 
 
 ## Biogeographical regions of Europe - (https://pmc.ncbi.nlm.nih.gov/articles/PMC7340631/#sec24)
 #3 as determined by the European Environment Agency
@@ -61,21 +65,22 @@ slope_data <- terrain(elev_rast, v = "slope", unit = "degrees", neighbors=8)
 signbase_sf$aspect <- extract(aspect_data, signbase_sf)
 signbase_sf$slope <- extract(slope_data, signbase_sf)
 
+ggplot(Europe) + 
+  geom_spatraster(data = elev_rast) +
+  geom_sf(data = signbase_sf, color = "red") +
+  scale_fill_grass_c(palette = "viridis") +
+  coord_sf(xlim = c(-10,30), 
+           ylim = c(35,53), 
+           expand = FALSE) +
+  facet_wrap(~group)
 
 
 
 ##climate data
 
-clim_files <- c("large-files/cclgmbi_2-5m/cclgmbi1.tif", "large-files/cclgmbi_2-5m/cclgmbi10.tif",
-                "large-files/cclgmbi_2-5m/cclgmbi11.tif", "large-files/cclgmbi_2-5m/cclgmbi12.tif",
-                "large-files/cclgmbi_2-5m/cclgmbi13.tif", "large-files/cclgmbi_2-5m/cclgmbi14.tif",
[…]

### 2025-01-22 endofquarterpaper.qmd
eco edits

@@
   labs(x = "Median calibrated BP")
-signbase_3000 <- signbase_years_cal
-signbase_3000$site_name <- signbase_years$site_name
-signbase_3000 <- signbase_3000 %>% 
-  filter(MedianBP < 30000)
 

### 2025-01-23 Extra/eco-regions-signbase.R
env update

@@ -13,38 +13,64 @@ signbase_sf$environmental_zone <- terra::extract(env_raster_reproj, signbase_sf)
 signbase_sf <- signbase_sf %>% 
   mutate(environmental_zone = environmental_zone$eea_r_3035_1_km_env)
 
-
 ##Plotting out distribution of signs over environmental zone raster
+
+env_zone_data <- signbase_sf %>% 
+  mutate(environmental_zone_clean = as.factor(environmental_zone))
+  
+
 ggplot(Europe) + 
   geom_spatraster(data = env_raster_reproj) +
-  scale_fill_grass_d(palette = "viridis") +
-  geom_sf(data = signbase_sf) +
+  geom_sf(data = env_zone_data, mapping = aes(fill = as.factor(environmental_zone_clean))) +
   coord_sf(xlim = c(-10,30), 
            ylim = c(35,53), 
            expand = FALSE) +
   facet_wrap(~group)
 
-unique(signbase_sf$environmental_zone)
 ## plot out env zone by group 
 ggplot(signbase_sf) +
   aes(x = group, fill= as.factor(environmental_zone)) +
   geom_bar()
 
-#3 plot 
+## diversity vs env zone
+
+signbase_div <- vegan::diversity(artifact_data, index = "shannon")
+
+signbase_sf$diversity <- signbase_div
+
+ggplot(signbase_sf) +
+  aes(x = environmental_zone, y = diversity, color = group) +
+  geom_point()
+
+ggplot(Europe) + 
+  geom_spatraster(data = env_raster_reproj) +
+  geom_sf(data = signbase_sf, 
+          mapping = aes(size = diversity, )) +
+  scale_size_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1, 1.25, 1.5, 1.75, 2, 2.25)) +
+  coord_sf(xlim = c(-10,30), 
+           ylim = c(35,53), 
+           expand = FALSE) +
+  facet_wrap(~group)
+
+
 
 ## Biogeographical regions of Europe - (https://pmc.ncbi.nlm.nih.gov/articles/PMC7340631/#sec24)
 #3 as determined by the European Environment Agency
 
 bio_geo_raster <- rast("data/bdj-08-e53720-s002.tif")
 
+res(bio_geo_raster)
+
 signbase_sf$biogeographical_region <- terra::extract(bio_geo_raster, signbase_sf)
+
+
 
[…]

### 2025-01-24 endofquarterpaper.qmd
editing

@@
 
-One of the most distinctive qualities of the Upper Paleolithic is the large rise in available artwork, both parietal and mobile.
-Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the grouping patterns of early modern humans.
-During this period, we also see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe.
+~One of the most~ A distinctive quality~ies~ of the Upper Paleolithic is the large rise in ~available~ artwork, both parietal and mobile.
+Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the grouping patterns of early modern humans <!--BM: scholarly citations needed here-->.
+During this period, we also see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe <!--BM: scholarly citations needed here-->.
 Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca.
@@
 
-We still lack a complete knowledge of the cultural diversity and movement patterns of the early modern humans from this era, despite a great deal of research into the subject.
-In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe during the Aurignacian technocomplex.
+We still lack a ~complete knowledge~ <!--BM: I previously mentioned, this is not a compelling motivation, nothing can ever have a 'complete knowledge', so this is not a unique condition for our topic.
+Look at how other papers identify the gap in knowledge and follow their example and replace this sentence --> of the cultural diversity and movement patterns of the early modern humans from this era, despite a great deal of research into the subject.
+In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe during the Aurignacian ~technocomplex~ period.
 The basis of this analysis is the idea that variations in signs can serve as an indicator of group membership and cultural similarity.
-This approach is supported by both ethnographic evidence throughout history, as well as similar archaeological studies done on ornamental and parietal art [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d].
-In this study, we address the following questions: Are cultural groups evident in the distribution of geometric signs, and if so, what can they tell us about the cultural diversity and movement of Upper Paleolithic humans?
+This approach is supported by both ethnographic evidence ~throughout history~ <!--BM: put scholarly citations on ethnography here-->, as well as similar archaeological studies ~done~ of~n~ ornamental and parietal art <!--BM: put scholarly citations on archaeology here--> [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d].
+In this study, we address the following questions: Are cultural groups evident in the distribution of portable geometric signs, and if so, what can they tell us about the cultural diversity and movement of Upper Paleolithic humans?
 
@@
 
-Our understanding of the patterns of cultural and genetic diversity among the early modern humans of the Upper Paleolithic remains incomplete.
-Previous works have performed various cluster analysis methods upon variations in artefacts to represent and understand patterns of groups and cultural diversity.
-The studies have primarily relied upon personal ornamentation and parietal art, with few taking mobile objects into account.
+Our understanding of the patterns of cultural and genetic diversity among the early modern humans of the Upper Paleolithic ~remains incomplete~ <!--BM: as noted above and previously, not a strong opening, replace -->.
+Previous works have performed various <!--BM: avoid this weasel word 'various' it gives the reader no information --> cluster analysis methods upon variations in artefacts to represent and understand patterns of groups and cultural diversity.
+The studies <!--BM: not informative, this first paragraph is better replaced by a single sentence that very specifically introduces the previous work that the next paragraphs with describe in more detail: Gravettian beads, Aurignacian beads, quadrilateral geometric parietal signs  --> have primarily relied upon personal ornamentation and parietal art, with few taking mobile objects into account.
 
 Multivariate analyses of Gravettian beads from burial and occupation sites in Europe suggests an east to west cline of nine cultural groups [@baker2024evidence].
-Seriation analysis, PCoA, and network analysis are used to identify the groups.
+Seriation analysis, PCoA, and network analysis ~are~ were <!--BM: everything in this background section needs to be in past tense, they have already done their work and it is complete --> used to identify the groups.
 The geographically discrete clusters identified encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and Western Mediterranean regions.
+<!--BM: this is a 3 sentence paragraph, that's a sign that something is wrong.
+Either add more detail, or combine with another para.
+I think adding more detail is ok here -->
 
-Similar analyses performed upon bead work patterns from the Aurignacian technocomplex as an indicator for linguistic diversity identify fourteen geographically cohesive clusters [@Vanhaeren_d].
-Specifically, the groups are identified using seriation, correspondence and GIS analyses.
+Similar analyses performed upon bead work patterns from the Aurignacian ~technocomplex~ period as an indicator for linguistic diversity identify fourteen geographically cohesive clusters [@Vanhaeren_d].
+Specifically, the groups ~are~ were identified using seriation, correspondence and GIS analyses.
 The clusters identified are geographically discrete and sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.

### 2025-01-24 endofquarterpaper.qmd
edits

@@
 During this period, we also see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe <!--BM: scholarly citations needed here-->.
-Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca.
+Considered to be the starting point of the Upper Paleolithic, the Aurignacian ~technocomplex~ period (ca.
 43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].
@@
 Either add more detail, or combine with another para.
-I think adding more detail is ok here -->
+I think adding more detail is ok here.
+What was their data, what were their methods, what were their results, in detail.
+ratio of words to information, mention 'isolation-by-distance' -->
 
@@
 The additional dimension of mobility in this study allows us to potentially identify more geographically dispersed cultural groups, which might help visualize the migration patterns of early modern humans.
-As said previously, previous studies have primarily excluded mobile objects, and tend to find more geographically compact cultural groups as a result.
+~As said previously~ <!--BM: no, we write so logically  that we don't need this -->, previous studies have primarily excluded mobile objects, and tend to find more geographically compact cultural groups as a result.
 
@@
 
-@Whallon_2006 proposes a framework for understanding hunter-gatherer mobility, arguing that mobility occurs among hunter-gatherers for multiple reasons.
+~@Whallon_2006 proposes a framework for understanding hunter-gatherer mobility, arguing that mobility occurs among hunter-gatherers for multiple reasons.
 The first two reasons, referred to by Whallon as logistical and residential mobility, occur when groups move in order to procure a new supply of resources once the more immediate ones have been depleted.
@@
 Lastly, and similarly, is informational mobility, in which groups, or individuals, move with the primary goal of gathering information, typically about the environmental conditions and resources in other areas.
-Again, this operates as a "safety net" of sorts, in that these groups then posses knowledge on where to move in times of resource scarcity.
+Again, this operates as a "safety net" of sorts, in that these groups then posses knowledge on where to move in times of resource scarcity.~ <!--BM: better in the discuss as context and interpretation -->
 

### 2025-01-25 Extra/eco-regions-signbase.R
env zones

@@ -13,11 +13,6 @@ signbase_sf$environmental_zone <- terra::extract(env_raster_reproj, signbase_sf)
 signbase_sf <- signbase_sf %>% 
   mutate(environmental_zone = environmental_zone$eea_r_3035_1_km_env)
 
-##Plotting out distribution of signs over environmental zone raster
-
-env_zone_data <- signbase_sf %>% 
-  mutate(environmental_zone_clean = as.factor(environmental_zone))
-  
 
 ggplot(Europe) + 
   geom_spatraster(data = env_raster_reproj) +
@@ -45,8 +40,8 @@ ggplot(signbase_sf) +
 ggplot(Europe) + 
   geom_spatraster(data = env_raster_reproj) +
   geom_sf(data = signbase_sf, 
-          mapping = aes(size = diversity, )) +
-  scale_size_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1, 1.25, 1.5, 1.75, 2, 2.25)) +
+          mapping = aes(size = diversity), 
+          scale_size_manual(breaks = c(0, 0.25, 0.5, 0.75, 1, 1.25, 1.5, 1.75, 2))) +
   coord_sf(xlim = c(-10,30), 
            ylim = c(35,53), 
            expand = FALSE) +
@@ -54,23 +49,28 @@ ggplot(Europe) +
 
 
 
+
 ## Biogeographical regions of Europe - (https://pmc.ncbi.nlm.nih.gov/articles/PMC7340631/#sec24)
 #3 as determined by the European Environment Agency
 
 bio_geo_raster <- rast("data/bdj-08-e53720-s002.tif")
+bio_geo_raster_df <- 
+  as.data.frame(bio_geo_raster, xy = TRUE) %>% 
+  mutate(`bdj-08-e53720-s002` = as.factor(`bdj-08-e53720-s002`))
 
 res(bio_geo_raster)
 
 signbase_sf$biogeographical_region <- terra::extract(bio_geo_raster, signbase_sf)
 
-
-
 signbase_sf <- signbase_sf %>% 
   mutate(biogeographical_region = biogeographical_region$bdj)
 
 ggplot(Europe) + 
-  geom_spatraster(data = bio_geo_raster) +
-  scale_fill_grass_d(palette = "viridis") +
+  geom_raster(data = bio_geo_raster_df,
+                  aes(x = x,
+                      y = y,
+                      fill = `bdj-08-e53720-s002`)) +
[…]

### 2025-01-26 Extra/eco-regions-signbase.R
eco changes

@@ -1,83 +1,7 @@
-##Environmental Zones in Europe, 
-##as categorized by the Environmental Stratification of Europe (EnS) dataset (European Environmental Agency)
-##link to download : https://www.eea.europa.eu/en/datahub/datahubitem-view/c8c4144a-8c0e-4686-9422-80dbf86bc0cb?activeAccordion=1082728
-
+## elevation data
 library(terra)
 library(tidyterra)
-env_raster <- rast("large-files//eea_r_3035_1_km_env-zones_p_2018_v01_r00.tif")
-
-env_raster_reproj <- terra::project(env_raster, "epsg:4326")
-
-signbase_sf$environmental_zone <- terra::extract(env_raster_reproj, signbase_sf)
-
-signbase_sf <- signbase_sf %>% 
-  mutate(environmental_zone = environmental_zone$eea_r_3035_1_km_env)
-
-
-ggplot(Europe) + 
-  geom_spatraster(data = env_raster_reproj) +
-  geom_sf(data = env_zone_data, mapping = aes(fill = as.factor(environmental_zone_clean))) +
-  coord_sf(xlim = c(-10,30), 
-           ylim = c(35,53), 
-           expand = FALSE) +
-  facet_wrap(~group)
-
-## plot out env zone by group 
-ggplot(signbase_sf) +
-  aes(x = group, fill= as.factor(environmental_zone)) +
-  geom_bar()
-
-## diversity vs env zone
-
-signbase_div <- vegan::diversity(artifact_data, index = "shannon")
-
-signbase_sf$diversity <- signbase_div
-
-ggplot(signbase_sf) +
-  aes(x = environmental_zone, y = diversity, color = group) +
-  geom_point()
-
-ggplot(Europe) + 
-  geom_spatraster(data = env_raster_reproj) +
-  geom_sf(data = signbase_sf, 
-          mapping = aes(size = diversity), 
-          scale_size_manual(breaks = c(0, 0.25, 0.5, 0.75, 1, 1.25, 1.5, 1.75, 2))) +
-  coord_sf(xlim = c(-10,30), 
-           ylim = c(35,53), 
-           expand = FALSE) +
-  facet_wrap(~group)
-
-
-
-
-## Biogeographical regions of Europe - (https://pmc.ncbi.nlm.nih.gov/articles/PMC7340631/#sec24)
-#3 as determined by the European Environment Agency
-
[…]

### 2025-01-28 endofquarterpaper.qmd
minor edits

--- 2025-01-22 endofquarterpaper.qmd
@@
 
-One of the most distinctive qualities of the Upper Paleolithic is the large rise in available artwork, both parietal and mobile.
-Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the grouping patterns of early modern humans.
-During this period, we also see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe.
-Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca.
+A distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile.
+Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the grouping patterns of early modern humans [@McDonald_Harper_2016; @Pereira_Manen_2023; @Fuentes_Lucas_Robert_2019].
+During this period, we also see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe [@Boyd_Silk_2020].
+Considered to be the starting point of the Upper Paleolithic, the Aurignacian period (ca.
 43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].
 
-We still lack a complete knowledge of the cultural diversity and movement patterns of the early modern humans from this era, despite a great deal of research into the subject.
-In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe during the Aurignacian technocomplex.
+In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe during the Aurignacian period.
 The basis of this analysis is the idea that variations in signs can serve as an indicator of group membership and cultural similarity.
-This approach is supported by both ethnographic evidence throughout history, as well as similar archaeological studies done on ornamental and parietal art [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d].
-In this study, we address the following questions: Are cultural groups evident in the distribution of geometric signs, and if so, what can they tell us about the cultural diversity and movement of Upper Paleolithic humans?
+This approach is supported by ethnographic and historical evidence [@texier2010howiesons; @Hodder1979; @Hodder1977; @Vanhaeren_2005] as well as similar archaeological studies of ornamental and parietal art [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d; @McDonald_Harper_2016].
+In this study, we address the following questions: Are cultural groups evident in the distribution of portable geometric signs, and if so, what can they tell us about the cultural diversity and movement of Upper Paleolithic humans?
 
@@
 
-Our understanding of the patterns of cultural and genetic diversity among the early modern humans of the Upper Paleolithic remains incomplete.
-Previous works have performed various cluster analysis methods upon variations in artefacts to represent and understand patterns of groups and cultural diversity.
-The studies have primarily relied upon personal ornamentation and parietal art, with few taking mobile objects into account.
+The purpose of this study is to contribute to our understanding of the patterns of cultural and genetic diversity among the early modern humans of the Upper Paleolithic.
+Previous works have performed cluster analysis methods upon variations in artefacts to represent and understand patterns of groups and cultural diversity, and have primarily relied upon personal ornamentation and parietal art, with few taking mobile objects into account.
+For example, multivariate analyses of Gravettian beads suggests an east to west cline of nine cultural groups [@baker2024evidence].
+The data set consisted of 134 discrete beads types from both burial and occupation sites, as distinguished by shape and material.
+Seriation analysis, PCoA, and network analysis were used to identify the groups.
+The geographically discrete clusters identified encompassed the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and Western Mediterranean regions.
+The isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance, is tested using a Mantel test, determining that geographic distance cannot solely account for the bead distribution.
 
-Multivariate analyses of Gravettian beads from burial and occupation sites in Europe suggests an east to west cline of nine cultural groups [@baker2024evidence].
-Seriation analysis, PCoA, and network analysis are used to identify the groups.
-The geographically discrete clusters identified encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and Western Mediterranean regions.
-
-Similar analyses performed upon bead work patterns from the Aurignacian technocomplex as an indicator for linguistic diversity identify fourteen geographically cohesive clusters [@Vanhaeren_d].
-Specifically, the groups are identified using seriation, correspondence and GIS analyses.
+Similar analyses (seriation, correspondence and GIS) were also performed upon personal ornament types from the Aurignacian period as a proxy for linguistic diversity.
+The data set consisted of 157 distinct ornament types, and the analyses identified fourteen geographically cohesive clusters sharing similar ornament type associations [@Vanhaeren_d].
 The clusters identified are geographically discrete and sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.
+Availability of raw materials was considered ~in regards to~ the potential explanation that observed patterns are determined by ~such~.
+~It was concluded that, due to observed patterns in the data that countered that explanation~, [BM: They found that] the differences in personal ornamentation could not be explained by material availability.
 
@@
 The additional dimension of mobility in this study allows us to potentially identify more geographically dispersed cultural groups, which might help visualize the migration patterns of early modern humans.
-As said previously, previous studies have primarily excluded mobile objects, and tend to find more geographically compact cultural groups as a result.
+Previous studies have primarily excluded mobile objects, and tend to find more geographically compact cultural groups as a result.
 
@@
 
-@Whallon_2006 proposes a framework for understanding hunter-gatherer mobility, arguing that mobility occurs among hunter-gatherers for multiple reasons.
-The first two reasons, referred to by Whallon as logistical and residential mobility, occur when groups move in order to procure a new supply of resources once the more immediate ones have been depleted.
-Another type of mobility is network mobility, in which groups, or subsets of groups, move in order to make contact with other groups.
-This occurs in order to form a "safety net", in times of scarce resources, so that groups are assured of a positive reception if they are forced to move.
-Lastly, and similarly, is informational mobility, in which groups, or individuals, move with the primary goal of gathering information, typically about the environmental conditions and resources in other areas.
-Again, this operates as a "safety net" of sorts, in that these groups then posses knowledge on where to move in times of resource scarcity.
-
 In this paper, we test the hypothesis that cultural groups are discernible in a big data set of geometric signs on mobile objects.
-Firstly, we use mantel tests to evaluate the null hypothesis of isolation-by-distance.
-We then identify clusters using seriation, network, and Principal Coordinate analyses.
-Finally, we test the strength of the clusters using modularity, perMANOVA, and PERMDISP2 tests.
+Firstly, we identify clusters using seriation analysis, network analysis, and PCoA.
+We then test the strength of the clusters using modularity, perMANOVA, and PERMDISP2 tests.
+Lastly, we use mantel tests to evaluate the null hypothesis of isolation-by-distance.
 
@@
 
-Mobility for various reasons is commonly found in hunter-gatherer groups.
-Therefore, one explanation for the sign type distribution could be that the groupings of the data represent cultural groups, as indicated by sign type patterns, and the geographic distribution between sites represent group mobility.
-The mobility may have occurred for all the above reasons.
+One explanation for the sign type distribution could be that the groupings of the data represent cultural groups, as indicated by sign type patterns, and the geographic distribution between sites represent group mobility.
+@Whallon_2006 proposes a framework for understanding hunter-gatherer mobility, arguing that mobility occurs among hunter-gatherers for multiple reasons.
+The first two reasons, referred to by Whallon as logistical and residential mobility, occur when groups move in order to procure a new supply of resources once the more immediate ones have been depleted.
+Another type of mobility is network mobility, in which groups, or subsets of groups, move in order to make contact with other groups.
+This occurs in order to form a "safety net", in times of scarce resources, so that groups are assured of a positive reception if they are forced to move.
+Lastly, and similarly, is informational mobility, in which groups, or individuals, move with the primary goal of gathering information, typically about the environmental conditions and resources in other areas.
+Similarly, this operates as a "safety net" of sorts, in that these groups then posses knowledge on where to move in times of resource scarcity.
 One point of note is that, according to @Whallon_2006, decorative objects may be exchanged as a method of network formation, which may account for the presence of some signs at some sites.

### 2025-01-29 Extra/eco-regions-signbase.R
pca

@@ -6,7 +6,6 @@ library(geodata)
 elev_rast <- elevation_global(res = 0.5, path = "/Users/sophiecooper/Independent Research Study/SignBase/large-files") %>% 
   crop(rast_ext)
 
-
 signbase_sf$elevation <- extract(elev_rast, signbase_sf)
 
 signbase_sf <- signbase_sf %>% 
@@ -24,12 +23,10 @@ signbase_sf <- signbase_sf %>%
 signbase_sf <- signbase_sf %>% 
   mutate(slope = slope$slope)
 
-
 ##climate data
 
 ## creation of Koggen Geiger climate zones
 library(climetrics)
-)
 
 month_avg_min_tmp_files <- list.files(path = "large-files/cclgmtn_2-5m/", full.names = TRUE)
 
@@ -48,7 +45,10 @@ month_precip <- rast(month_precip_files) %>%
 
 mean_month_temp_rast <- ((month_avg_min_tmp + month_avg_max_tmp)/2)
 
-koppen_geiger_zones <- kgc(p = month_precip, tmin = month_avg_min_tmp, tmax = month_avg_max_tmp, tmean = mean_month_temp_rast)
+koppen_geiger_zones <- kgc(p = month_precip, 
+                           tmin = month_avg_min_tmp, 
+                           tmax = month_avg_max_tmp, 
+                           tmean = mean_month_temp_rast)
 
 koppen_geiger_zones_df <- as.data.frame(koppen_geiger_zones, xy = TRUE)
 
@@ -67,8 +67,7 @@ ggplot(Europe) +
   facet_wrap(~group)
 
 
-##pca on climate data
-
+## pca on climate data
 
 clim_pca_df <- as.data.frame(month_avg_min_tmp, xy= TRUE) %>% 
   st_as_sf(coords = c("x", "y"),
@@ -80,7 +79,6 @@ clim_pca_df$elevation <- extract(elev_rast, clim_pca_df)
 clim_pca_df$max_monthly_temp <- extract(month_avg_max_tmp, clim_pca_df)
 
 clim_pca_df$slope <- extract(slope_data, clim_pca_df)
-
 
 clim_pca_df$precipitation <- extract(month_precip, clim_pca_df)
 
@@ -113,20 +111,15 @@ clim_pca_df$x <- clim_xy$x
 
 clim_pca_df$y <- clim_xy$y
 
-
 clim_pca_sf <- clim_pca_df %>% 
[…]

### 2025-01-29 Extra/eco-regions-signbase.R
climate pca

@@ -69,31 +69,31 @@ ggplot(Europe) +
 
 ## pca on climate data
 
-clim_pca_df <- as.data.frame(month_avg_min_tmp, xy= TRUE) %>% 
+clim_df <- as.data.frame(month_avg_min_tmp, xy= TRUE) %>% 
   st_as_sf(coords = c("x", "y"),
            remove = FALSE,
            crs = 4326)
 
-clim_pca_df$elevation <- extract(elev_rast, clim_pca_df)
+clim_df$elevation <- extract(elev_rast, clim_df)
 
-clim_pca_df$max_monthly_temp <- extract(month_avg_max_tmp, clim_pca_df)
+clim_df$max_monthly_temp <- extract(month_avg_max_tmp, clim_df)
 
-clim_pca_df$slope <- extract(slope_data, clim_pca_df)
+clim_df$slope <- extract(slope_data, clim_df)
 
-clim_pca_df$precipitation <- extract(month_precip, clim_pca_df)
+clim_df$precipitation <- extract(month_precip, clim_df)
 
-clim_pca_df <- clim_pca_df %>% 
+clim_df <- clim_df %>% 
   st_drop_geometry() %>% 
   drop_na() %>% 
   unnest(c(max_monthly_temp, precipitation, elevation, slope), names_repair = "universal") 
 
-clim_xy <- clim_pca_df %>% 
+clim_xy <- clim_df %>% 
   select(x, y)
 
-clim_pca_df <- clim_pca_df %>% 
+clim_df <- clim_df%>% 
   select(-x)
 
-colnames(clim_pca_df) <-  c("northing", "min_temp_jan", "min_temp_oct", "min_temp_nov", 
+colnames(clim_df) <-  c("northing", "min_temp_jan", "min_temp_oct", "min_temp_nov", 
                                "min_temp_dec", "min_temp_feb", "min_temp_mar", "min_temp_apr",
                                "min_temp_may", "min_temp_june", "min_temp_jul", "min_temp_aug",
                                "min_temp_sep", "elevID", "elevation", "maxID", "max_temp_jan", "max_temp_oct", 
@@ -104,22 +104,31 @@ colnames(clim_pca_df) <-  c("northing", "min_temp_jan", "min_temp_oct", "min_tem
                                "precip_apr", "precip_may", "precip_june", "precip_jul", "precip_aug",
[…]

### 2025-02-03 Extra/eco-regions-signbase.R
date

@@ -88,10 +88,10 @@ clim_df <- clim_df %>%
   unnest(c(max_monthly_temp, precipitation, elevation, slope), names_repair = "universal") 
 
 clim_xy <- clim_df %>% 
-  select(x, y)
+  dplyr::select(x, y)
 
 clim_df <- clim_df%>% 
-  select(-x)
+  dplyr::select(-x)
 
 colnames(clim_df) <-  c("northing", "min_temp_jan", "min_temp_oct", "min_temp_nov", 
                                "min_temp_dec", "min_temp_feb", "min_temp_mar", "min_temp_apr",
@@ -105,7 +105,7 @@ colnames(clim_df) <-  c("northing", "min_temp_jan", "min_temp_oct", "min_temp_no
                                "precip_sep")
 
 clim_df <- clim_df %>% 
-  select(-maxID, -precipID, -elevID, -slopeID)
+  dplyr::select(-maxID, -precipID, -elevID, -slopeID)
 
 
 clim_df$x <- clim_xy$x
@@ -124,11 +124,34 @@ clim_pca_rast <- clim_pca$rasters
 clim_pca_df<- as.data.frame(clim_pca_rast, xy = TRUE) %>% 
   mutate(PC1 = as.factor(PC1))
 
+
+clim_pca_comp <- clim_pca$pca.object
+
+x <- clim_pca_comp$x %>% 
+  as.data.frame()
+
+x$PC1
+
+
+
 ggplot(Europe) +
   geom_spatraster(data = clim_pca_rast)+
-  scale_fill_grass_c() +
+  scale_fill_grass_b(breaks = c(-17, -16, -15, -14, -13, -12, -11, -10, -9, -8, -7, -6, -5, -4, -3, -2, -1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13)) +
   geom_sf(data = signbase_sf) +
   coord_sf(xlim = c(-10,30),
            ylim = c(35,53), 
-           expand = FALSE)
+           expand = FALSE) +
+  facet_wrap(~group) +
+  theme()
+
+
+signbase_sf$PCA <- extract(clim_pca_rast, signbase_sf) 
+
+signbase_sf <- signbase_sf %>% 
+  mutate(PCA = PCA$PC1)
+
+ggplot(signbase_sf) +
+  aes(x = group, y = PCA) +
+  geom_boxplot()
+
 

### 2025-02-04 Extra/image-analysis.R
explore images with ML

@@ -1,0 +1,69 @@
+# Do object recognition on the  images
+
+library(keras) # install_keras()
+
+# Check where Python's looked for
+reticulate::py_config() 
+# https://stackoverflow.com/a/46820701/1036500
+model <- application_xception(weights = "imagenet")
+# pip install pip-system-certs
+# for certificate error: https://stackoverflow.com/a/58525755/1036500
+# I got an error the first time, then downloaded the model file an 
+# put this in ~/.keras/models : "inception_v3_weights_tf_dim_ordering_tf_kernels.h5"
+# get a list of the cropped images
+imgs_cropped <- list.files("Extra/images", full.names = TRUE)
+# loop over all  images and get a list of the ten highest 
+# probability objects recognised in each image
+output_list <- vector("list", length = length(imgs_cropped))
+for(i in seq_along(imgs_cropped)){
+  f <- paste0(getwd(), "/", imgs_cropped[i])
+  # load the image
+  img_path <- f
+  img <- image_load(img_path, target_size = c(299,299))
+  x <- image_to_array(img)
+  # ensure we have a 4d tensor with single element in the batch dimension,
+  # the preprocess the input for prediction using resnet50
+  x <- array_reshape(x, c(1, dim(x)))
+  x <- xception_preprocess_input(x)
+  # make predictions then decode and print them
+  preds <- model %>% predict(x)
+  output_list[[i]] <- imagenet_decode_predictions(preds, top = 20)[[1]]
+}
+# tidy output into data frame
+library(tidyverse)
+wide_out <-
+  bind_rows(output_list,
+            .id = "image")  %>%
+  select(-class_name) %>%
+  pivot_wider(image,
+              names_from = "class_description",
+              values_from = "score",
+              values_fill = 0)
+
+
+# Now we have attributes for each image, in the form of the objects recognised
+# in them, we can see how images with similar objects cluster together. We use
+# t-SNE for reducing the dimensionality, and then plot the images to see how 
+# groupings appear. 
+
+
+library(Rtsne)
+set.seed(42)
+tsne_out <-
+  Rtsne(wide_out,
+        pca=TRUE,
+        perplexity=30,
+        theta=0.0) # Run TSNE
+tsne_out_plot <-
+  tibble(x = tsne_out$Y[,1],
+         y = tsne_out$Y[,2],
+         img = paste0(getwd(), "/",  imgs_cropped)
+  )
+
[…]

### 2025-02-04 Extra/download-images-from-database.R
explore images with ML

@@ -3,12 +3,12 @@ library(tidyverse)
 library(rvest)
 library(httr)
 
-signbase_full <- read_csv("signBase_Version1.0.csv")
+signbase_full <- read_csv("data/signBase_Version1.0.csv")
 
 # find objects with hatching or grid
 sb_grid_hatch <- 
 signbase_full %>% 
-  filter(grid == 1 | hatching == 1)
+  filter(notch == 1)
 
 # urls to each object data webpage
 urls_object_webpage <- 
@@ -44,7 +44,7 @@ for(i in 1:length(sb_grid_hatch_img$img_urls)){
 if (!is.na(img_url)) {
   img <- GET(img_url)
   writeBin(content(img, "raw"), 
-           paste0("images/signbase-", 
+           paste0("Extra/images/signbase-", 
                   sb_grid_hatch_img$object_id[i], 
            ".jpg"))  
            }

### 2025-02-04 endofquarterpaper.qmd

@@
 library(rcarbon)
+library(readxl)
 signbase_full <- read_csv("data/signBase_Version1.0.csv")

### 2025-02-04 endofquarterpaper.qmd
seriation now plots with colour for country of site

@@
 rownames(artifact_matrix) <- rownames(artifact_data)
-mplot <- function(x,...){
-  d <- data.frame(x = rep(1:ncol(x), each = nrow(x)),
+
+# Load required libraries
+library(ggplot2)
+
+# Define the mplot function
+mplot <- function(x, ..., fill_colour = "black", title = NULL, row_labels = NULL, col_labels = NULL, country_colors = NULL) {
+  # Create the data frame
+  d <- data.frame(
+    x = rep(1:ncol(x), each = nrow(x)),
     y = rep(nrow(x):1, ncol(x)),
-                  z = as.integer(x))
-  ggplot(d, aes(x, y, fill = factor(z))) +
-    geom_tile(col = "grey50", linewidth = 0.5) +
-    scale_fill_manual(values = c("0" = "white", "1" = "black")) +
+    z = as.integer(x)
+  )
+
+  # Add country colors to the data frame if provided
+  if (!is.null(country_colors)) {
+    d$country_color <- rep(country_colors, ncol(x))
+  }
+
+  # Use custom row/column labels if provided, otherwise use defaults
+  if (is.null(col_labels)) col_labels <- colnames(x)
+  if (is.null(row_labels)) row_labels <- rev(rownames(x))
+
+  # Plot
+  p <- ggplot(d, aes(x, y)) +
+    geom_tile(col = "grey50", linewidth = 0.5, aes(fill = ifelse(z == 1, country_color, "white"))) +
+    scale_fill_identity() +  # Use the fill values directly
     coord_equal(expand = FALSE) +
-    scale_x_continuous(breaks = 1:ncol(x), 
-                       labels = colnames(x),
-                       position = "top") +
-    scale_y_continuous(breaks = 1:nrow(x), 
-                       labels = rev(rownames(x))) +
-    theme(axis.text = element_text(size = rel(0.5)),
+    scale_x_continuous(
+      breaks = 1:ncol(x),
+      labels = col_labels,
+      position = "top"
+    ) +
+    scale_y_continuous(
+      breaks = 1:nrow(x),
+      labels = row_labels
+    ) +
+    theme(
+      axis.text = element_text(size = rel(0.5)),
       axis.ticks = element_blank(),
       legend.position = "none",
-          plot.title = element_text(hjust = 0.25)) +
-    labs(x = "Sign Type",
-         y = "Site Name",
-         fill = "Count") +
-    rotate_x_text(angle = 90, hjust = NULL, vjust = NULL)}
-increase_focus <- function(x){
+      plot.title = element_text(hjust = 0.25)
+    ) +
+    labs(x = "Sign Type", y = "Site Name", fill = "Count", title = title) +
+    rotate_x_text(angle = 90, hjust = NULL, vjust = NULL)
+
+  p
+}
+
+# Define the increase_focus function
+increase_focus <- function(x, country = NULL) {
+  # Reorder rows by the mean position of 1s in each row
   mcp <- apply(
@@
     MARGIN = 1, 
-    FUN = \(z){ mean(which(z == 1)) }, 
+    FUN = \(z) { mean(which(z == 1)) }, 
     simplify = FALSE
   ) |> unlist()
-  x <- x[order(mcp), ]
+  row_order <- order(mcp, na.last = TRUE)
+  x <- x[row_order, ]
+
+  # Reorder the country vector if provided
+  if (!is.null(country)) {
+    country <- country[row_order]
+  }
+
+  # Reorder columns by the mean position of 1s in each column
   mrp <- apply(
@@
     MARGIN = 2, 
-    FUN = \(z){ mean(which(z == 1)) }, 
-    simplify = FALSE) |> unlist()
-  x[, order(mrp)]}
-concentrate <- function(x){
+    FUN = \(z) { mean(which(z == 1)) }, 
+    simplify = FALSE
+  ) |> unlist()
+  x <- x[, order(mrp, na.last = TRUE)]
+
+  # Return the reordered matrix and country vector
+  list(matrix = x, country = country)
+}
+
+# Define the concentrate function
+concentrate <- function(x, country = NULL, max_iter = 100) {
   old <- x
   not_identical <- TRUE
-  while(not_identical){
-    new <- increase_focus(old)
+  iter <- 0
+
+  while (not_identical && iter < max_iter) {
+    result <- increase_focus(old, country)
+    new <- result$matrix
+    country <- result$country
     not_identical <- !identical(old, new)
-    old <- new}
-  new}
-artifact_matrix |> concentrate() |> mplot()
+    old <- new
+    iter <- iter + 1
+  }
 
+  if (iter == max_iter) {
+    warning("Reached maximum iterations without convergence.")
+  }
+
+  list(matrix = new, country = country)
+}
+
+# get country names, for example
+x_df <- 
+  artifact_data %>% 
+  rownames_to_column("site_name") %>% 
+  left_join(signbase_full_clean %>% 
+              distinct(site_name, .keep_all = TRUE) %>% 
+              select(site_name,
+                     country),
+            by = "site_name")
+
+countries <- x_df$country
+
+# Generate a named vector of colors, one for each unique country
+unique_countries <- unique(countries)
+color_palette <-  RColorBrewer::brewer.pal(length(unique_countries), "Set3")   # Generate distinct colors
+row_colors <- setNames(color_palette, unique_countries) 
+countries_colors <- row_colors[countries]
+
+# Concentrate the matrix and reorder the country vector
+result <- concentrate(artifact_matrix, country = countries)
+concentrated_matrix <- result$matrix
+reordered_countries <- result$country
+
+# Map countries to colors
+country_colors <- row_colors[reordered_countries]
+
+# Plot the concentrated matrix with country colors
+mplot(
+  concentrated_matrix,
+  country_colors = country_colors
+)
 ```
@@
   mutate(across(everything(), ~replace(., . > 1, 1))) %>% 
-  concentrate() %>% 
+ # concentrate() %>% 
   rownames_to_column("site_name") %>% 
   left_join(group_df)
+
 sign_type_data <- graph_data %>% 
@@
   mutate(group = parse_number(group))
+
 x_order <- graph_data %>% 
@@
   colnames()
+
 ggplot(sign_type_data) +

### 2025-02-06 Extra/date-change.R
experimentation

@@ -15,7 +15,7 @@ signbase_full_clean <- signbase_full %>%
          site_name != "Šandalja II",
          site_name != "Les Cottés",
          site_name != "Trou al'Wesse") %>% 
-  dplyr::select(-other, -rectangle, -circumline, -pinleft, -pinright, -star, -circumspiral)
+  dplyr::select(-other, -rectangle, -circumline, -pinleft, -pinright, -star, -circumspiral, -circumnotch, -maccaroni)
 
 signbase_years <- signbase_full_clean %>% 
   drop_na(date_bp_max_min) %>% 
@@ -151,128 +151,5 @@ ggplot(Europe) +
                   size = 2) +
   theme_minimal()
 
-# network analysis
-
-library(phangorn)
-library(ggraph)
-library(statnet)
-library(vegan)
-jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
-dm <- as.matrix(jac)
-disim <- 1 - dm
-disim[disim<0.35] <- 0
-disim_net <- network(disim,
-                     directed=F,
-                     ignore.eval=F,
-                     names.eval='weight')
-disim_net %v% 'vertex.names' <- 
-  row.names(artifact_data)
-
-set.seed(500)
-
-ggraph(disim_net, layout="fr") +
-  geom_edge_link(aes(width = weight, 
-                     alpha=weight), 
-                 edge_colour = "black", 
-                 show.legend=T) +
-  scale_edge_width(range=c(0.5,1.5)) +
-  scale_edge_colour_gradient(low = "#CCCCCC",
-                             high = "#000000") +
-  geom_node_point(alpha=1, size=5) +
-  geom_node_text(aes(label=row.names(artifact_data)), 
-                 size=6, 
-                 family= "Arial", 
-                 repel=T) +
-  theme(text= element_text(size = 15), 
-        panel.background = element_rect(fill = 'white'))
-
-
-#pcoa
-library(ecodist)
-library(ape)
-pcoa <- pcoa(jac)
-pcoa_df <- data.frame(pcoa1 = pcoa$vectors[,1], 
-                      pcoa2 = pcoa$vectors[,2])
-pcoa_df <- pcoa_df %>% 
-  mutate(site_name = rownames(artifact_data))
-library(ggrepel)
-ggplot(pcoa_df) +
-  aes(x=pcoa1, 
-      y=pcoa2) +
-  geom_point() +
-  labs(x = "PC1",
-       y = "PC2",) +
[…]

### 2025-02-06 indiv_artifact-groups.R

@@ -1,0 +1,82 @@
+library(tidyverse)
+library(magrittr)
+library(rcarbon)
+signbase_full <- read_csv("data/signBase_Version1.0.csv")
+
+signbase_full_clean <- signbase_full %>% 
+  filter(site_name != "Willendorf",
+         site_name != "Riparo di Fontana Nuova",
+         site_name != "Muralovka",
+         site_name != "Shanidar Cave",
+         site_name != "Hayonim Cave",
+         site_name != "El Salitre",
+         site_name != "Grotte De La Princesse Pauline",
+         site_name != "Šandalja II") %>% 
+  dplyr::select(-other, -rectangle, - circumnotch, - zigzag, -hashtag, -maccaroni, -circumspiral, -anthropomorph, -concenline, -zigzagrow, -pinleft, -pinright)
+
+
+signbase_years <- signbase_full_clean %>% 
+  drop_na(date_bp_max_min) %>% 
+  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+\\/\\-", "±")) %>% 
+  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+", "±")) %>% 
+  separate(date_bp_max_min,
+           sep = " - ",
+           c("date_bp_max", "date_bp_min"),
+           remove = FALSE) %>% 
+  mutate(date_bp_max = ifelse(str_detect(date_bp_max, "\\/"),
+                              str_extract(date_bp_max, ".*?(?=\\/)"),
+                              date_bp_max)) %>% 
+  mutate(date_bp_min = ifelse(str_detect(date_bp_min, "\\/"),
+                              str_extract(date_bp_min, ".*?(?=\\/)"),
+                              date_bp_min)) %>% 
+  separate(date_bp_max,
+           sep = "±",
+           c("date_bp_max_age",
+             "date_bp_max_error")) %>% 
+  separate(date_bp_min,
+           sep = "±",
+           c("date_bp_min_age",
+             "date_bp_min_error")) %>% 
+  drop_na(date_bp_max_age,
+          date_bp_max_error) %>% 
+  mutate(date_bp_max_age = parse_number(date_bp_max_age),
+         date_bp_max_error = parse_number(date_bp_max_error))
+signbase_years_cal <- 
+  rcarbon::calibrate(signbase_years$date_bp_max_age,
+                     signbase_years$date_bp_max_error,
[…]

### 2025-02-06 pipeline-change.R

@@ -1,0 +1,171 @@
+## most abundant signs
+library(tidyverse)
+library(magrittr)
+library(rcarbon)
+library(readxl)
+signbase_full <- read_csv("data/signBase_Version1.0.csv")
+
+signbase_full_clean <- signbase_full %>% 
+  filter(site_name != "Willendorf",
+         site_name != "Riparo di Fontana Nuova",
+         site_name != "Muralovka",
+         site_name != "Shanidar Cave",
+         site_name != "Hayonim Cave",
+         site_name != "El Salitre",
+         site_name != "Grotte De La Princesse Pauline",
+         site_name != "Šandalja II",
+         site_name!= "Le Terme Pialat",
+         site_name != "Les Cottés",
+         site_name != "Otaslavice",
+         site_name != "Trou al'Wesse",
+         site_name != "Hornos de la Pena",
+         site_name != "Göpfelsteinhöhle") %>% 
+  dplyr::select(-other, -rectangle)
+
+
+sign_long <- signbase_full_clean %>% 
+  pivot_longer(cols = line:star) %>% 
+  dplyr::select(name, value) %>% 
+  filter(value!= 0) %>% 
+  group_by(name) %>% 
+  summarise(total = sum(value)) %>% 
+  dplyr::arrange(desc(total)) %>% 
+  dplyr::slice(1:10)
+  
+abundant_sign_names <- c(sign_long$name)
+
+signbase_full_clean <- signbase_full_clean %>% 
+  dplyr::select(object_id, site_name, longitude, latitude,
+                all_of(abundant_sign_names), date_bp_max_min, country)
+
+lat_long_df <- signbase_full_clean %>% 
+  dplyr::select(site_name, longitude, latitude) %>% 
+  distinct(site_name, .keep_all = TRUE)
+
+signbase_unique_groups <- signbase_full_clean %>% 
+  mutate(longitude = as.character(longitude),
+         latitude = as.character(latitude)) %>% 
+  group_by(site_name) %>%
+  summarize(across(where(is.numeric), sum)) %>% 
+  left_join(lat_long_df)
+
+artifact_data <- signbase_unique_groups %>% 
+  column_to_rownames("site_name") %>% 
+  dplyr::select(notch:obnotch)
+
+
+groups <- list("1" = c("Abri Pataud", "El Rascaño", 
+                       "Fumane", "La Viña",
+                       "Lhotka", "Pod Hradem",
+                       "Riparo Bombrini", "Sirgenstein Cave",
[…]

### 2025-02-08 endofquarterpaper.qmd
experimentation
new-groups

@@
                        "Riparo Bombrini", "Sirgenstein Cave"),
-                "2" = c("Kvasice", "Labeko Koba", 
-                        "Les Rois", "Slatinice",
-                        "Göpfelsteinhöhle"),
-                "3" = c("Aurignac", "d'Engihoul",
-                         "Istállóskő Cave", "Hohlenstein-Stadel",
-                         "Gargas", "Grotte de la Verpillière I", 
-                         "Mladeč", "Brassempouy",
-                         "Gatzarria"),
-                "4" = c("Wildscheuer", "Breitenbach",
-                        "Peskő Cave", "Vindija Cave",
-                        "El Castillo"),
-                "5" = c("Trou Magrite",
-                        "Grotte de Goyet", "Bockstein-Törle"),
-                "6" = c("Trou al'Wesse", "Maisières-Canal",
-                        "Les Cottés", "Le Terme Pialat"),
-                "7" = c("Vogelherd","Hohle Fels",
-                        "Geissenklösterle"),
-               "8" = c("Tuto de Camalhot","Abri Lartet/Gorge d'Enfer",
-                        "Castanet", "Le Trou Du Renard", 
-                        "Grottes de Fonds-de-Forêt" ),
-               "9" = c("Nová Dědina", "Menton/Grottes du Grimaldi",
-                        "Grotte du Renne", "Solutré", 
-                        "Žlutava", "Spy"),
-                "10" = c("Cellier", "Blanchard", 
-                         "La Ferrassie", "Hornos de la Pena",
-                         "Otaslavice"),
-                "11" = c("La Souquette", "Shelter Birów IV",
-                         "Abri de Laussel", "Abri du Poisson"))
+               "2" = c("Aurignac", "Gargas",
+                       "Grotte de la Verpillière I","Hohlenstein-Stadel",
+                       "Istállóskő Cave", "Mladeč",
+                       "d'Engihoul"),
+               "3" = c("Breitenbach", "Peskő Cave",
+                       "Vindija Cave", "Wildscheuer"),
+               "4" = c("Kvasice", "Labeko Koba", 
+                       "Les Rois"),
+               "5" = c("Le Trou Du Renard", "Abri Lartet/Gorge d'Enfer",
+                       "Tuto de Camalhot", "Castanet", 
+                       "Gatzarria", "Abri du Poisson"),
+               "6" = c("Bockstein-Törle", "Grotte de Goyet",
+                        "Trou Magrite", "Brassempouy"),
+               "7" = c( "Göpfelsteinhöhle",  "Slatinice", 
+                        "Grotte du Renne"),
+               "8" = c( "El Castillo", "Trou al'Wesse", 
+                        "Maisières-Canal"),
+               "9" = c("Vogelherd","Hohle Fels",
+                       "Geissenklösterle", "Le Terme Pialat", 
+                       "Les Cottés", "Spy",  "Grottes de Fonds-de-Forêt"),
+               "10" = c("Nová Dědina", "Menton/Grottes du Grimaldi",
+                        "Solutré", "Žlutava", 
+                        "Hornos de la Pena", "Otaslavice"),
+               "11" = c("Cellier", "Blanchard", 
+                        "La Ferrassie"),
+               "12" = c("La Souquette", "Shelter Birów IV",
+                        "Abri de Laussel"))
 group_df <- 
@@
 
-# Load required libraries
-library(ggplot2)
-
 # Define the mplot function
@@
   rownames_to_column("site_name") %>% 
-  left_join(signbase_full_clean %>% 
-              distinct(site_name, .keep_all = TRUE) %>% 
+  left_join(signbase_unique_groups %>% 
               select(site_name,
-                     country),
+                     group),
             by = "site_name")
 
-countries <- x_df$country
+countries <- x_df$group
 
@@
                  family= "Arial", 
-                 repel=T) +
+                 repel=T, max.overlaps = 100) +
   theme(text= element_text(size = 15), 

### 2025-02-11 endofquarterpaper.qmd
time period

@@
 The modularity test, when run on the above groupings, resulted in a score of `r as_group`.
-The perMANOVA test produced an R-squared value of `r perman_r` and a P value of `r perman_p`, which tells us that the groupings are statisically significant, and that 52% of the variability in the sign distribution can be attributed to group membership.
+The perMANOVA test produced an R-squared value of `r perman_r` and a P value of `r perman_p`, which tells us that the groupings are statistically significant, and that 52% of the variability in the sign distribution can be attributed to group membership.
 The multivariate homogeneity test of group dispersions produced @fig-perm-disp, showing that the overlap between groups is fairly minimal.

### 2025-02-11 Extra/time-periods.R

@@ -1,0 +1,291 @@
+library(tidyverse)
+library(magrittr)
+library(rcarbon)
+library(readxl)
+signbase_full <- read_csv("data/signBase_Version1.0.csv")
+
+signbase_full_clean <- signbase_full %>% 
+  filter(site_name != "Willendorf",
+         site_name != "Riparo di Fontana Nuova",
+         site_name != "Muralovka",
+         site_name != "Shanidar Cave",
+         site_name != "Hayonim Cave",
+         site_name != "El Salitre",
+         site_name != "Grotte De La Princesse Pauline",
+         site_name != "Šandalja II") %>% 
+  dplyr::select(-other, -rectangle)
+
+signbase_years <- signbase_full_clean %>% 
+  drop_na(date_bp_max_min) %>% 
+  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+\\/\\-", "±")) %>% 
+  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+", "±")) %>% 
+  separate(date_bp_max_min,
+           sep = " - ",
+           c("date_bp_max", "date_bp_min"),
+           remove = FALSE) %>% 
+  mutate(date_bp_max = ifelse(str_detect(date_bp_max, "\\/"),
+                              str_extract(date_bp_max, ".*?(?=\\/)"),
+                              date_bp_max)) %>% 
+  mutate(date_bp_min = ifelse(str_detect(date_bp_min, "\\/"),
+                              str_extract(date_bp_min, ".*?(?=\\/)"),
+                              date_bp_min)) %>% 
+  separate(date_bp_max,
+           sep = "±",
+           c("date_bp_max_age",
+             "date_bp_max_error")) %>% 
+  separate(date_bp_min,
+           sep = "±",
+           c("date_bp_min_age",
+             "date_bp_min_error")) %>% 
+  drop_na(date_bp_max_age,
+          date_bp_max_error) %>% 
+  mutate(date_bp_max_age = parse_number(date_bp_max_age),
+         date_bp_max_error = parse_number(date_bp_max_error))
+signbase_years_cal <- 
+  rcarbon::calibrate(signbase_years$date_bp_max_age,
+                     signbase_years$date_bp_max_error,
+                     verbose = FALSE) %>% 
+  summary() %>% 
+  data_frame()
+
+signbase_years$MedianBP <- signbase_years_cal$MedianBP
+
+
+signbase_full_clean <- signbase_full_clean %>% 
[…]

### 2025-02-11 pipeline-change.R

@@ -1,8 +1,9 @@
-## most abundant signs
+## Most distinct signs
+
 library(tidyverse)
 library(magrittr)
 library(rcarbon)
-library(readxl)
+
 signbase_full <- read_csv("data/signBase_Version1.0.csv")
 
 signbase_full_clean <- signbase_full %>% 
@@ -14,29 +15,63 @@ signbase_full_clean <- signbase_full %>%
          site_name != "El Salitre",
          site_name != "Grotte De La Princesse Pauline",
          site_name != "Šandalja II",
-         site_name!= "Le Terme Pialat",
-         site_name != "Les Cottés",
-         site_name != "Otaslavice",
-         site_name != "Trou al'Wesse",
-         site_name != "Hornos de la Pena",
-         site_name != "Göpfelsteinhöhle") %>% 
-  dplyr::select(-other, -rectangle)
+         site_name != "Žlutava",
+         site_name != "Trou Magrite",
+         site_name != "Le Terme Pialat",
+         site_name != "Le Trou Du Renard",
+         site_name !=  "Otaslavice",
+         site_name != "Brassempouy") %>% 
+  dplyr::select(site_name, object_id, country, longitude, latitude, date_bp_max_min, 
+                cross, v, star, zigzag, zigzagrow, grid, hashtag, hatching, 
+                anthropomorph,zoomorph, vulva, rhombus, rectangle)
 
 
-sign_long <- signbase_full_clean %>% 
-  pivot_longer(cols = line:star) %>% 
-  dplyr::select(name, value) %>% 
-  filter(value!= 0) %>% 
-  group_by(name) %>% 
-  summarise(total = sum(value)) %>% 
-  dplyr::arrange(desc(total)) %>% 
-  dplyr::slice(1:10)
-  
-abundant_sign_names <- c(sign_long$name)
+signbase_years <- signbase_full_clean %>% 
+  drop_na(date_bp_max_min) %>% 
+  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+\\/\\-", "±")) %>% 
+  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+", "±")) %>% 
+  separate(date_bp_max_min,
+           sep = " - ",
+           c("date_bp_max", "date_bp_min"),
+           remove = FALSE) %>% 
+  mutate(date_bp_max = ifelse(str_detect(date_bp_max, "\\/"),
+                              str_extract(date_bp_max, ".*?(?=\\/)"),
[…]

### 2025-02-11 indiv_artifact-groups.R
time period

@@ -12,8 +12,13 @@ signbase_full_clean <- signbase_full %>%
          site_name != "El Salitre",
          site_name != "Grotte De La Princesse Pauline",
          site_name != "Šandalja II") %>% 
-  dplyr::select(-other, -rectangle, - circumnotch, - zigzag, -hashtag, -maccaroni, -circumspiral, -anthropomorph, -concenline, -zigzagrow, -pinleft, -pinright)
-
+  dplyr::select(-other, -rectangle, -circumline, -pinleft, -pinright, -star, -circumspiral, -circumnotch, -maccaroni, -obnotch, -hashtag, -anthropomorph) %>%
+  filter(object_id != "laf0022",
+       object_id != "gkl0025",
+       object_id != "vhc0135",
+       object_id != "sol0007",
+       object_id != "gff0001",
+       object_id != "cel0001")
 
 signbase_years <- signbase_full_clean %>% 
   drop_na(date_bp_max_min) %>% 
@@ -53,25 +58,105 @@ signbase_years_cal$object_id <- signbase_years$object_id
 signbase_full_clean <- inner_join(signbase_full_clean, signbase_years_cal)
 
 signbase_full_clean <- signbase_full_clean %>% 
-  filter(MedianBP > 36000) %>% 
-  filter(MedianBP < 39000)
-
+  filter(MedianBP > 37500) %>% 
+  filter(MedianBP < 42500)
 
 artifact_data <- signbase_full_clean %>% 
-  column_to_rownames("site_name") %>% 
-  dplyr::select(line:star)
+  arrange(object_id) %>% 
+  column_to_rownames("object_id") %>% 
+  dplyr::select(line:concenline)
 
 artifact_data <- artifact_data %>% 
   mutate(row_sums = rowSums(.)) %>% 
   filter(row_sums >= 2) %>% 
   dplyr::select(-row_sums)
 
+x_df <- 
+  artifact_data %>% 
+  rownames_to_column("object_id") %>% 
+  left_join(signbase_full_clean %>% 
+              distinct(object_id, .keep_all = TRUE) %>% 
+              select(object_id,
+                     country),
+            by = "object_id")
+
+countries <- x_df$country
+
+types <- list("1" = c("bla0005", "vhc0002", "vhc0022"),
+               "2" = c("hss0002",  "hfc0046",  "hfc0044", "hfc0028","laf0005",
[…]

### 2025-02-11 Extra/time-periods.R
looking at how many objects we lose when focusing on time periods

@@ -4,7 +4,7 @@ library(rcarbon)
 library(readxl)
 signbase_full <- read_csv("data/signBase_Version1.0.csv")
 
-signbase_full_clean <- signbase_full %>% 
+signbase_full_clean <- signbase_full %>%  # 511 rows
   filter(site_name != "Willendorf",
          site_name != "Riparo di Fontana Nuova",
          site_name != "Muralovka",
@@ -15,7 +15,7 @@ signbase_full_clean <- signbase_full %>%
          site_name != "Šandalja II") %>% 
   dplyr::select(-other, -rectangle)
 
-signbase_years <- signbase_full_clean %>% 
+signbase_years <- signbase_full_clean %>%  # 446
   drop_na(date_bp_max_min) %>% 
   mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+\\/\\-", "±")) %>% 
   mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+", "±")) %>% 
@@ -41,6 +41,7 @@ signbase_years <- signbase_full_clean %>%
           date_bp_max_error) %>% 
   mutate(date_bp_max_age = parse_number(date_bp_max_age),
          date_bp_max_error = parse_number(date_bp_max_error))
+
 signbase_years_cal <- 
   rcarbon::calibrate(signbase_years$date_bp_max_age,
                      signbase_years$date_bp_max_error,
@@ -55,7 +56,7 @@ signbase_full_clean <- signbase_full_clean %>%
   inner_join(signbase_years %>% 
               select(object_id, MedianBP))
 
-signbase_full_clean <- signbase_full_clean %>% 
+signbase_full_clean <- signbase_full_clean %>%  # 446
   mutate(time_period = case_when((MedianBP > 42999) ~ "transitional",
                                  (MedianBP > 39899) ~ "proto_aurignacian",
                                  (MedianBP > 36999) ~ "early_aurignacian",

### 2025-02-12 Extra/time-periods.R
pipeline

@@ -136,19 +136,19 @@ ggplot(Europe) +
   theme_minimal() +
   facet_wrap(~time_period)
 
-ser_data_trans <- trans_unique_data %>% 
+trans_artifact_data <- trans_unique_data %>% 
   column_to_rownames("site_name") %>% 
   dplyr::select(line:star)
 
-ser_data_proto <- proto_unique_data %>% 
+proto_artifact_data <- proto_unique_data %>% 
   column_to_rownames("site_name") %>% 
   dplyr::select(line:star)
 
-ser_data_early <- early_unique_data %>% 
+early_artifact_data <- early_unique_data %>% 
   column_to_rownames("site_name") %>% 
   dplyr::select(line:star)
 
-ser_data_evolved <- evolved_unique_data%>% 
+evolved_artifact_data <- evolved_unique_data%>% 
   column_to_rownames("site_name") %>% 
   dplyr::select(line:star)
 
@@ -284,6 +284,10 @@ mplot(
   country_colors = country_colors
 )
 
+
+
+to_investigate<- signbase_full %>% 
+  filter(is.na(date_bp_max_min))
 
 
 

### 2025-02-12 cluster_pipeline.R

@@ -1,0 +1,213 @@
+library(ggpubr)
+library(phangorn)
+library(ggraph)
+library(statnet)
+library(ecodist)
+library(ape)
+library(ggrepel)
+library(vegan)
+library(cowplot)
+library(igraph)
+
+mplot <- function(x, ..., fill_colour = "black", title = NULL, row_labels = NULL, col_labels = NULL, country_colors = NULL) {
+  d <- data.frame(
+    x = rep(1:ncol(x), each = nrow(x)),
+    y = rep(nrow(x):1, ncol(x)),
+    z = as.integer(x)
+  )
+  
+  if (!is.null(country_colors)) {
+    d$country_color <- rep(country_colors, ncol(x))
+  }
+  
+  if (is.null(col_labels)) col_labels <- colnames(x)
+  if (is.null(row_labels)) row_labels <- rev(rownames(x))
+  
+  p <- ggplot(d, aes(x, y)) +
+    geom_tile(col = "grey50", linewidth = 0.5, aes(fill = ifelse(z == 1, country_color, "white"))) +
+    scale_fill_identity() +  # Use the fill values directly
+    coord_equal(expand = FALSE) +
+    scale_x_continuous(
+      breaks = 1:ncol(x),
+      labels = col_labels,
+      position = "top"
+    ) +
+    scale_y_continuous(
+      breaks = 1:nrow(x),
+      labels = row_labels
+    ) +
+    theme(
+      axis.text = element_text(size = rel(0.5)),
+      axis.ticks = element_blank(),
+      legend.position = "none",
+      plot.title = element_text(hjust = 0.25)
+    ) +
+    labs(x = "Sign Type", y = "Site Name", fill = "Count", title = title) +
+    rotate_x_text(angle = 90, hjust = NULL, vjust = NULL)
+  
+  p
+}
+
+increase_focus <- function(x, country = NULL) {
+  mcp <- apply(
+    x, 
+    MARGIN = 1, 
+    FUN = \(z) { mean(which(z == 1)) }, 
+    simplify = FALSE
+  ) |> unlist()
+  row_order <- order(mcp, na.last = TRUE)
+  x <- x[row_order, ]
+  
+  if (!is.null(country)) {
+    country <- country[row_order]
+  }
+  
+  mrp <- apply(
+    x, 
+    MARGIN = 2, 
+    FUN = \(z) { mean(which(z == 1)) }, 
+    simplify = FALSE
+  ) |> unlist()
[…]

### 2025-02-13 Extra/eco-regions-signbase.R
pipeline created, time periods investigated

@@ -43,31 +43,7 @@ month_precip_files <- list.files(path = "large-files/cclgmpr_2-5m/", full.names 
 month_precip <- rast(month_precip_files) %>% 
   crop(rast_ext)
 
-mean_month_temp_rast <- ((month_avg_min_tmp + month_avg_max_tmp)/2)
 
-koppen_geiger_zones <- kgc(p = month_precip, 
-                           tmin = month_avg_min_tmp, 
-                           tmax = month_avg_max_tmp, 
-                           tmean = mean_month_temp_rast)
-
-koppen_geiger_zones_df <- as.data.frame(koppen_geiger_zones, xy = TRUE)
-
-koppen_geiger_zones_df <- koppen_geiger_zones_df %>% 
-  mutate(mean = as.factor(mean))
-
-ggplot(Europe) + 
-  geom_raster(data = koppen_geiger_zones_df,
-              aes(x = x,
-                  y = y,
-                  fill = mean)) +
-  geom_sf(data = signbase_sf) +
-  coord_sf(xlim = c(-10,30),
-           ylim = c(35,53), 
-           expand = FALSE) +
-  facet_wrap(~group)
-
-
-## pca on climate data
 
 clim_df <- as.data.frame(month_avg_min_tmp, xy= TRUE) %>% 
   st_as_sf(coords = c("x", "y"),
@@ -137,11 +113,11 @@ x$PC1
 ggplot(Europe) +
   geom_spatraster(data = clim_pca_rast)+
   scale_fill_grass_b(breaks = c(-17, -16, -15, -14, -13, -12, -11, -10, -9, -8, -7, -6, -5, -4, -3, -2, -1, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13)) +
-  geom_sf(data = signbase_sf) +
+  geom_sf(data = signbase_sf, aes(color = group)) +
   coord_sf(xlim = c(-10,30),
            ylim = c(35,53), 
            expand = FALSE) +
-  facet_wrap(~group) +
+  facet_wrap(~time_period) +
   theme()
 
 
@@ -151,7 +127,7 @@ signbase_sf <- signbase_sf %>%
   mutate(PCA = PCA$PC1)
 
 ggplot(signbase_sf) +
-  aes(x = group, y = PCA) +
+  aes(x = time_period, y = PCA, fill = group) +
   geom_boxplot()
 
 

### 2025-02-13 Extra/time-periods.R

@@ -15,6 +15,10 @@ signbase_full_clean <- signbase_full %>%  # 511 rows
          site_name != "Šandalja II") %>% 
   dplyr::select(-other, -rectangle)
 
+
+to_examine <- signbase_full_clean %>% 
+  filter(is.na(date_bp_max_min))
+
 signbase_years <- signbase_full_clean %>%  # 446
   drop_na(date_bp_max_min) %>% 
   mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+\\/\\-", "±")) %>% 
@@ -54,13 +58,14 @@ signbase_years$MedianBP <- signbase_years_cal$MedianBP
 
 signbase_full_clean <- signbase_full_clean %>% 
   inner_join(signbase_years %>% 
-              select(object_id, MedianBP))
+              dplyr::select(object_id, MedianBP))
 
 signbase_full_clean <- signbase_full_clean %>%  # 446
   mutate(time_period = case_when((MedianBP > 42999) ~ "transitional",
                                  (MedianBP > 39899) ~ "proto_aurignacian",
                                  (MedianBP > 36999) ~ "early_aurignacian",
                                  (MedianBP > 31999) ~ "evolved_aurignacian"))
+  
 
 
 ggplot(signbase_full_clean) +
@@ -108,12 +113,109 @@ evolved_unique_data <- signbase_full_clean %>%
   mutate(time_period = "evolved_aurignacian")
 
 
+trans_artifact_data <- trans_unique_data %>% 
+  column_to_rownames("site_name") %>% 
+  dplyr::select(line:star)
+
+
+proto_artifact_data <- proto_unique_data %>% 
+  column_to_rownames("site_name") %>% 
+  dplyr::select(line:star)
+
+early_artifact_data <- early_unique_data %>% 
+  column_to_rownames("site_name") %>% 
+  dplyr::select(line:star)
+
+evolved_artifact_data <- evolved_unique_data %>% 
+  column_to_rownames("site_name") %>% 
+  dplyr::select(line:star)
+
+
+##collective plots
+
+library(cowplot)
+
+seriation_trans <- produce_clusters(trans_artifact_data, trans_unique_data, method = "seriation")
+seriation_proto <- produce_clusters(proto_artifact_data, proto_unique_data, method = "seriation")
+seriation_early <- produce_clusters(early_artifact_data, early_unique_data, method = "seriation")
[…]

### 2025-02-13 cluster_pipeline.R

@@ -25,7 +25,7 @@ mplot <- function(x, ..., fill_colour = "black", title = NULL, row_labels = NULL
   
   p <- ggplot(d, aes(x, y)) +
     geom_tile(col = "grey50", linewidth = 0.5, aes(fill = ifelse(z == 1, country_color, "white"))) +
-    scale_fill_identity() +  # Use the fill values directly
+    scale_fill_identity() +  
     coord_equal(expand = FALSE) +
     scale_x_continuous(
       breaks = 1:ncol(x),
@@ -42,7 +42,7 @@ mplot <- function(x, ..., fill_colour = "black", title = NULL, row_labels = NULL
       legend.position = "none",
       plot.title = element_text(hjust = 0.25)
     ) +
-    labs(x = "Sign Type", y = "Site Name", fill = "Count", title = title) +
+    labs(x = "", y = "") +
     rotate_x_text(angle = 90, hjust = NULL, vjust = NULL)
   
   p
@@ -95,7 +95,9 @@ concentrate <- function(x, country = NULL, max_iter = 100) {
 }
 
 
-produce_clusters <- function(artifact_data, artifact_data_full, method = "seriation"){
+produce_clusters <- function(artifact_data, artifact_data_unique, method = "seriation"){
+  artifact_data <- artifact_data %>% 
+    select_if(~ !is.numeric(.) || sum(.) != 0)
   jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
   dm <- as.matrix(jac)
   
@@ -103,7 +105,8 @@ produce_clusters <- function(artifact_data, artifact_data_full, method = "seriat
   artifact_matrix <- as.matrix(artifact_data %>% 
                                  mutate(across(everything(), ~replace(., . > 1, 1))))
   rownames(artifact_matrix) <- rownames(artifact_data)
-  x_df <- artifact_data_full %>% 
+  
+  x_df <- artifact_data_unique %>% 
     left_join(signbase_full_clean,
               by = "site_name")
   countries <- x_df$country
@@ -133,17 +136,21 @@ produce_clusters <- function(artifact_data, artifact_data_full, method = "seriat
       geom_edge_link(aes(width = weight, 
                          alpha=weight), 
                      edge_colour = "black", 
-                     show.legend=T) +
+                     show.legend= F,
+                    ) +
       scale_edge_width(range=c(0.5,1.5)) +
       scale_edge_colour_gradient(low = "#CCCCCC",
                                  high = "#000000") +
       geom_node_point(alpha=1, size=5) +
[…]

### 2025-02-14 Extra/eco-regions-signbase.R
changes

@@ -131,3 +131,7 @@ ggplot(signbase_sf) +
   geom_boxplot()
 
 
+ggplot(signbase_sf) +
+  aes(x = time_period, y = elevation, fill = group) +
+  geom_boxplot()
+

### 2025-02-14 Extra/time-periods.R

@@ -16,9 +16,6 @@ signbase_full_clean <- signbase_full %>%  # 511 rows
   dplyr::select(-other, -rectangle)
 
 
-to_examine <- signbase_full_clean %>% 
-  filter(is.na(date_bp_max_min))
-
 signbase_years <- signbase_full_clean %>%  # 446
   drop_na(date_bp_max_min) %>% 
   mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+\\/\\-", "±")) %>% 
@@ -267,11 +264,7 @@ proto_strength_plot <- strength_function(proto_artifact_data, proto_unique_data,
 early_strength_plot <- strength_function(early_artifact_data, early_unique_data, type = "dispersion")
 evolved_strength_plot <- strength_function(evolved_artifact_data, evolved_unique_data, type = "dispersion")
 
-
-
-
-
-##look at individual signs over time
+##look at individual sign distribution across periods
 
 full_data_long <- time_unique_data_full %>% 
   pivot_longer(cols = line:star) %>% 
@@ -281,5 +274,6 @@ ggplot(full_data_long) +
   aes(x = name, fill = group) +
   geom_bar() +
   rotate_x_text(angle = 90, hjust = NULL, vjust = NULL) + 
-  facet_wrap(~time_period)
+  facet_wrap(~time_period, nrow = 4)
+
   

### 2025-02-14 Extra/eco-regions-signbase.R
improved pipeline

@@ -1,32 +1,14 @@
 ## elevation data
 library(terra)
 library(tidyterra)
-rast_ext <- ext(-10, 30, 35, 53)
 library(geodata)
+
+rast_ext <- ext(-10, 30, 35, 53)
+
 elev_rast <- elevation_global(res = 0.5, path = "/Users/sophiecooper/Independent Research Study/SignBase/large-files") %>% 
   crop(rast_ext)
 
-signbase_sf$elevation <- extract(elev_rast, signbase_sf)
-
-signbase_sf <- signbase_sf %>% 
-  mutate(elevation = elevation$wc2.1_30s_elev)
-
-aspect_data <- terrain(elev_rast, v = "aspect", unit = "degrees", neighbors=8)
 slope_data <- terrain(elev_rast, v = "slope", unit = "degrees", neighbors=8)
-
-signbase_sf$aspect <- extract(aspect_data, signbase_sf)
-signbase_sf$slope <- extract(slope_data, signbase_sf)
-
-signbase_sf <- signbase_sf %>% 
-  mutate(aspect = aspect$aspect)
-
-signbase_sf <- signbase_sf %>% 
-  mutate(slope = slope$slope)
-
-##climate data
-
-## creation of Koggen Geiger climate zones
-library(climetrics)
 
 month_avg_min_tmp_files <- list.files(path = "large-files/cclgmtn_2-5m/", full.names = TRUE)
 
@@ -42,8 +24,6 @@ month_precip_files <- list.files(path = "large-files/cclgmpr_2-5m/", full.names 
 
 month_precip <- rast(month_precip_files) %>% 
   crop(rast_ext)
-
-
 
 clim_df <- as.data.frame(month_avg_min_tmp, xy= TRUE) %>% 
   st_as_sf(coords = c("x", "y"),
@@ -61,15 +41,11 @@ clim_df$precipitation <- extract(month_precip, clim_df)
 clim_df <- clim_df %>% 
   st_drop_geometry() %>% 
   drop_na() %>% 
-  unnest(c(max_monthly_temp, precipitation, elevation, slope), names_repair = "universal") 
+  unnest(c(max_monthly_temp, precipitation, elevation, slope), names_repair = "universal") %>% 
+  mutate(northing = y)
 
-clim_xy <- clim_df %>% 
-  dplyr::select(x, y)
 
-clim_df <- clim_df%>% 
-  dplyr::select(-x)
-
[…]

### 2025-02-16 Extra/eco-regions-signbase.R
permanova pipeline

@@ -69,10 +69,6 @@ clim_pca <- raster.pca(clim_rast, n = 1)
 
 clim_pca_rast <- clim_pca$rasters
 
-clim_pca_df<- as.data.frame(clim_pca_rast, xy = TRUE) %>% 
-  mutate(PC1 = as.factor(PC1))
-
-clim_pca_fill <- clim_pca$pca.object$x
 
 ggplot(Europe) +
   geom_spatraster(data = clim_pca_rast)+
@@ -81,6 +77,12 @@ ggplot(Europe) +
   coord_sf(xlim = c(-10,30),
            ylim = c(35,53), 
            expand = FALSE) +
+  geom_text_repel(data = signbase_sf,
+                  aes(x = longitude ,
+                      y = latitude,
+                      label = site_name),
+                  max.overlaps = 30,
+                  size = 2) +
   facet_wrap(~time_period) +
   theme()
 
@@ -90,17 +92,71 @@ signbase_sf$PCA <- extract(clim_pca_rast, signbase_sf)
 signbase_sf <- signbase_sf %>% 
   mutate(PCA = PCA$PC1)
 
+library(ggbeeswarm)
 ggplot(signbase_sf) +
-  aes(x = time_period, y = PCA, fill = group) +
-  geom_boxplot()
+  aes(x = time_period, 
+      y = PCA,
+      fill = group,
+      colour = group) +
+  geom_boxplot(outliers = FALSE,
+               alpha = 0.1) +
+  geom_quasirandom(size = 4,
+                   dodge.width = 0.5,
+                   alpha = 0.3) +
+  theme_minimal()
 
-signbase_sf$elevation <- extract(elev_rast, signbase_sf)
+
+## Permanova test for environmental zones - how much is the impact on group?
+
+signbase_sf$clim <- terra::extract(clim_rast, signbase_sf)
 
 signbase_sf <- signbase_sf %>% 
-  mutate(elevation = elevation$wc2.1_30s_elev)
+  unnest(clim)
 
+min_temp_df <- signbase_sf %>% 
+  dplyr::select(min_temp_jan:min_temp_sep) %>% 
+  st_drop_geometry() %>% 
+  mutate(avg = rowSums(.)/12)
+
+signbase_sf$min_temp_avg <- min_temp_df$avg
+
[…]

### 2025-02-16 Extra/time-periods.R

@@ -13,8 +13,9 @@ signbase_full_clean <- signbase_full %>%  # 511 rows
          site_name != "El Salitre",
          site_name != "Grotte De La Princesse Pauline",
          site_name != "Šandalja II") %>% 
-  dplyr::select(-other, -rectangle)
-
+  dplyr::select(-other, -rectangle) %>% 
+  mutate(longitude = ifelse(site_name == "Riparo Bombrini", 7.437500, longitude)) %>% 
+  mutate(latitude = ifelse(site_name == "Riparo Bombrini", 43.77083, latitude))
 
 signbase_years <- signbase_full_clean %>%  # 446
   drop_na(date_bp_max_min) %>% 
@@ -207,7 +208,8 @@ time_unique_data_full <-rbind(trans_unique_data, proto_unique_data,
                               early_unique_data, evolved_unique_data)
 
 time_unique_data_full$time_period <- factor(time_unique_data_full$time_period, 
-                                            levels = c("transitional", "proto_aurignacian", "early_aurignacian", "evolved_aurignacian"))
+                                            levels = c("transitional", "proto_aurignacian", 
+                                                       "early_aurignacian", "evolved_aurignacian"))
 
 library(rnaturalearth)
 library(rnaturalearthdata)
@@ -239,7 +241,8 @@ ggplot(Europe) +
 
 
 ##testing overall dataset
-time_unique_data_test <- time_unique_data_full 
+time_unique_data_test <- time_unique_data_full %>% 
+  mutate(group = time_period)
 
 time_unique_data_artifact_test <- time_unique_data_full %>% 
   dplyr::select(line:star)
@@ -275,5 +278,7 @@ ggplot(full_data_long) +
   geom_bar() +
   rotate_x_text(angle = 90, hjust = NULL, vjust = NULL) + 
   facet_wrap(~time_period, nrow = 4)
+
+
 
   

### 2025-02-16 cluster_pipeline.R
permanova pipeline

@@ -8,6 +8,7 @@ library(ggrepel)
 library(vegan)
 library(cowplot)
 library(igraph)
+library(tidyverse)
 
 mplot <- function(x, ..., fill_colour = "black", title = NULL, row_labels = NULL, col_labels = NULL, country_colors = NULL) {
   d <- data.frame(
@@ -256,6 +257,34 @@ trans_network <- function(artifact_data){
     return(final_plot)
     }
 
+
+perm_function <- function(period = "none", variable = PCA){
+  if(period == "none"){perm_df <- signbase_sf}
+  
+  if(period == "transitional"){perm_df <- signbase_sf %>% 
+    filter(time_period == "transitional")}
+  
+  if(period == "proto_aurignacian"){perm_df <- signbase_sf %>% 
+    filter(time_period == "proto_aurignacian")}
+  
+  if(period == "early_aurignacian"){perm_df <- signbase_sf %>% 
+    filter(time_period == "early_aurignacian")}
+  
+  if(period == "evolved_aurignacian"){perm_df <- signbase_sf %>% 
+    filter(time_period == "evolved_aurignacian")}
+  
+  perm_jac <- perm_df %>% 
+    st_drop_geometry() %>% 
+    dplyr::select(id, {{variable}}) %>% 
+    column_to_rownames("id") %>% 
+    vegdist("jaccard")
+  
+  perm_result <- adonis2(perm_jac~as.factor(perm_df$group), 
+                         method = "jaccard",
+                         sqrt.dist = TRUE)
+  
+  return(perm_result)
+}
 
 
 

### 2025-02-18 endofquarterpaper.qmd
permanova pipeline
extra

@@
 ---
-title: "Geometric Signs on Paleolithic Mobile Objects as Cultural Group Markers"
+title: "Geometric Signs on Paleolithic Mobile Objects as Cultural Group Markers Across Aurignacian Time Periods"
 author: Sophie Cooper
@@
 Considered to be the starting point of the Upper Paleolithic, the Aurignacian period (ca.
-43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].
+43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].The Aurignacian can be divided up into four chronological periods: Transitional, Proto-Aurignacian, Early Aurignacian, and Evolved Aurignacian [@Banks_dErrico_Zilhao_2013, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021].
 
-In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe during the Aurignacian period.
+In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe across the subsections of the Aurignacian period.
 The basis of this analysis is the idea that variations in signs can serve as an indicator of group membership and cultural similarity.
 This approach is supported by ethnographic and historical evidence [@texier2010howiesons; @Hodder1979; @Hodder1977; @Vanhaeren_2005] as well as similar archaeological studies of ornamental and parietal art [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d; @McDonald_Harper_2016].
-In this study, we address the following questions: Are cultural groups evident in the distribution of portable geometric signs, and if so, what can they tell us about the cultural diversity and movement of Upper Paleolithic humans?
+In this study, we address the following questions: How does sign distribution change across periods of the Aurignacian, what do the patterns tell us about cultural diversity patterns and movement of Upper Paleolithic humans throughout the Aurignacian?
 
@@
 The clusters identified are geographically discrete and sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.
-Availability of raw materials was considered ~in regards to~ the potential explanation that observed patterns are determined by ~such~.
-~It was concluded that, due to observed patterns in the data that countered that explanation~, [BM: They found that] the differences in personal ornamentation could not be explained by material availability.
+The possibility that observed patterns were influenced by availability of raw materials was considered, and ultimately discarded as an explanation for the patterns.
 
 Following the established precedent of using ornamentation and signs to identify cultural groups, we look to see if designs on mobile objects can be applicable here, and what it can add to the research being done.
-The additional dimension of mobility in this study allows us to potentially identify more geographically dispersed cultural groups, which might help visualize the migration patterns of early modern humans.
+The additional dimension of mobility and comparison of in this study allows us to potentially identify more geographically dispersed cultural groups, which might help visualize the migration patterns of early modern humans.
 Previous studies have primarily excluded mobile objects, and tend to find more geographically compact cultural groups as a result.
@@
 
-In this paper, we test the hypothesis that cultural groups are discernible in a big data set of geometric signs on mobile objects.
-Firstly, we identify clusters using seriation analysis, network analysis, and PCoA.
+In this paper, we test the hypothesis that cultural groups are discernible in a data set of geometric signs on mobile objects.
+Firstly, we identify clusters within each time period using seriation and network analysis.
 We then test the strength of the clusters using modularity, perMANOVA, and PERMDISP2 tests.
-Lastly, we use mantel tests to evaluate the null hypothesis of isolation-by-distance.
+Lastly, we use mantel tests to evaluate the null hypothesis of isolation-by-distance,
 
@@
 The results were plotted with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths.
-
-Principal coordinate decomposition analysis (PCoA) analyzes the eigenvalue decomposition of the Jaccard matrix and arranges the sites by dissimilarity [@Legendre_Legendre_2012].
-This provides a visual of sign similarity as visualized by spacing and arrangement.
-The PCoA in this paper is calibrated to be plotted on a scatterplot, as opposed to a biplot.
 

### 2025-02-18 Extra/eco-regions-signbase.R

@@ -141,7 +141,6 @@ signbase_sf$precip_avg <- precip_df$avg
 
 
 
-
 clim_pca_df<- as.data.frame(clim_pca_rast, xy = TRUE) %>% 
   mutate(PCA = PC1)
 
@@ -156,7 +155,13 @@ signbase_sf <- signbase_sf %>%
 ggplot(signbase_sf) +
   aes(x = PCA, y = PC2,  color = group) +
   geom_point() +
-  facet_wrap(~time_period) 
+  stat_ellipse(mapping = aes(color = time_period))
 
 
 
+proto_mantel <- perm_function(period = "proto_aurignacian", variable = min_temp_jan:northing, method = "mantel")
+
+
+proto_perm <- perm_function(period = c("early_aurignacian" | "transitional"), variable = PCA, method = "permanova")
+
+                            

### 2025-02-18 cluster_pipeline.R

@@ -258,32 +258,46 @@ trans_network <- function(artifact_data){
     }
 
 
-perm_function <- function(period = "none", variable = PCA){
+
+## mantel function tells us whether changes in sign makeup is correlated with chosen variable change. 
+## Permanova tells us how similar in chosen variable members of group are
+
+perm_function <- function(period = "none", variable = PCA, method = "permanova"){
+  set.seed(500)
+  
   if(period == "none"){perm_df <- signbase_sf}
   
-  if(period == "transitional"){perm_df <- signbase_sf %>% 
-    filter(time_period == "transitional")}
-  
-  if(period == "proto_aurignacian"){perm_df <- signbase_sf %>% 
-    filter(time_period == "proto_aurignacian")}
-  
-  if(period == "early_aurignacian"){perm_df <- signbase_sf %>% 
-    filter(time_period == "early_aurignacian")}
-  
-  if(period == "evolved_aurignacian"){perm_df <- signbase_sf %>% 
-    filter(time_period == "evolved_aurignacian")}
-  
-  perm_jac <- perm_df %>% 
-    st_drop_geometry() %>% 
-    dplyr::select(id, {{variable}}) %>% 
-    column_to_rownames("id") %>% 
-    vegdist("jaccard")
-  
-  perm_result <- adonis2(perm_jac~as.factor(perm_df$group), 
+  else{perm_df <- signbase_sf %>% 
+    filter(time_period == {{period}})}
+
+  if(method == "permanova"){
+  var_jac <- perm_df %>% 
+      st_drop_geometry() %>% 
+      column_to_rownames("id") %>% 
+      dplyr::select({{variable}}) %>% 
+      vegdist("jaccard")
+    
+  result <- adonis2(var_jac~as.factor(perm_df$group), 
                          method = "jaccard",
-                         sqrt.dist = TRUE)
+                         sqrt.dist = TRUE)}
   
-  return(perm_result)
+  if(method == "mantel"){
+    sign_jac <- perm_df %>% 
+      st_drop_geometry() %>% 
+      column_to_rownames("id") %>% 
+      dplyr::select(line:star) %>% 
+      vegdist("jaccard", binary = TRUE)
+      
+    var_sim <- perm_df %>% 
+      st_drop_geometry() %>% 
+      column_to_rownames("id") %>% 
+      dplyr::select({{variable}}) %>% 
+      dist()
+    
+    result <- vegan::mantel(sign_jac,  var_sim, permutations = 1000)
+  }
+  
+  return(result)
 }
 
 

### 2025-02-28 Extra/eco-regions-signbase.R
edits

@@ -15,9 +15,9 @@ month_avg_min_tmp_files <- list.files(path = "large-files/cclgmtn_2-5m/", full.n
 month_avg_min_tmp <- rast(month_avg_min_tmp_files) %>% 
   crop(rast_ext)
 
-month_avy_max_tmp_files <- list.files(path = "large-files/cclgmtx_2-5m/", full.names = TRUE)
+month_avg_max_tmp_files <- list.files(path = "large-files/cclgmtx_2-5m/", full.names = TRUE)
 
-month_avg_max_tmp <- rast(month_avy_max_tmp_files) %>% 
+month_avg_max_tmp <- rast(month_avg_max_tmp_files) %>% 
   crop(rast_ext)
 
 month_precip_files <- list.files(path = "large-files/cclgmpr_2-5m/", full.names = TRUE)
@@ -136,6 +136,8 @@ signbase_sf$precip_avg <- precip_df$avg
 
 
 
+proto_mantel <- perm_function(period = "proto_aurignacian", variable = min_temp_jan:northing, method = "mantel")
+
 
 
 
@@ -155,13 +157,9 @@ signbase_sf <- signbase_sf %>%
 ggplot(signbase_sf) +
   aes(x = PCA, y = PC2,  color = group) +
   geom_point() +
-  stat_ellipse(mapping = aes(color = time_period))
+  geom_text_repel(aes(label = site_name)) +
+  facet_wrap(~time_period)
 
 
-
-proto_mantel <- perm_function(period = "proto_aurignacian", variable = min_temp_jan:northing, method = "mantel")
-
-
-proto_perm <- perm_function(period = c("early_aurignacian" | "transitional"), variable = PCA, method = "permanova")
 
                             

### 2025-02-28 endofquarterpaper.qmd
edits

@@
 # Introduction
+One distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile.
+During this period, we also see a rise in organized settlements and more complex social organizations across Europe [@Boyd_Silk_2020].
+Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand these shifting patterns of social organization and diversity
 
-A distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile.
-Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the grouping patterns of early modern humans [@McDonald_Harper_2016; @Pereira_Manen_2023; @Fuentes_Lucas_Robert_2019].
-During this period, we also see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe [@Boyd_Silk_2020].
+ <!-- change citations maybe -->
+[@McDonald_Harper_2016; @Pereira_Manen_2023; @Fuentes_Lucas_Robert_2019].
+<!-- add more about aurignacian -->
+
 Considered to be the starting point of the Upper Paleolithic, the Aurignacian period (ca.
@@
 This approach is supported by ethnographic and historical evidence [@texier2010howiesons; @Hodder1979; @Hodder1977; @Vanhaeren_2005] as well as similar archaeological studies of ornamental and parietal art [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d; @McDonald_Harper_2016].
-In this study, we address the following questions: How does sign distribution change across periods of the Aurignacian, what do the patterns tell us about cultural diversity patterns and movement of Upper Paleolithic humans throughout the Aurignacian?
+In this study, we address the following questions: How does sign distribution change across periods of the Aurignacian?
+How do the groupings vary in diversity, size, connectivity, and boundaries?
+What do the patterns tell us about the cultural groupings and movement of Upper Paleolithic humans throughout the Aurignacian?
 
@@
 
+<!-- remove section on mobility? -->
+
 Following the established precedent of using ornamentation and signs to identify cultural groups, we look to see if designs on mobile objects can be applicable here, and what it can add to the research being done.
@@
 
-Our use of geometric signs in this study is supported by @Sauvet_2018, in which used quadrilateral geometric parietal signs are used to identify Upper Paleolithic cultural groups in the Cantabrian region of Spain.
+<!--  -->
+
+Our use of geometric signs in this study is supported by @Sauvet_2018, in which quadrilateral geometric parietal signs are used to identify Upper Paleolithic cultural groups in the Cantabrian region of Spain.
 The authors identified and classified the signs based on variations in form, then performed a factorial correspondence analysis and ascending hierarchical clustering upon the data in order to identify two clusters, which they argued represents distinct and separate groups of humans.
@@
 
+<!-- add more about aurignacian time period - what was going on, what was the criteria -->
+
+The Aurignacian period occurred roughly 43,000 - 30,000 years BP, and is largely characterized by material culture indicators, as well as an ongoing biological shift from Neanderthals to early modern humans.
+The Aurignacian can be further divided into subphases, again mainly distinguished by shifts in material culture [@Tartar_2015], Proto, Early, Evolved, and Late Aurignacian.
+@Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main periods, Aur-P1 (Proto/Early), whose typological observations corresponded with the time period of 43–37 kyr BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods kyr BP and 37-32 kyr BP.
+@Banks_dErrico_Zilhao_2013 used Bayesian modeling methods to further divide Aur-P1 into the Proto and Early Aurignacian, placing the Proto-Aurignacian at 41.5 - 39.9 k cal BP, and the early Aurignacian at 39.8 - 37.8 k cal BP.
+They found that the Proto-Aurignacian occurs during the Greenland Interstadials 9 and 10, periods of climatic amelioration, while the Early Aurignacian began with the Henrich Stadial, which is characterized by dry and arid conditions, and ended with Greenland Interstadial 8.
+
+<!-- change hypothesis -->
 In this paper, we test the hypothesis that cultural groups are discernible in a data set of geometric signs on mobile objects.
@@
 We then test the strength of the clusters using modularity, perMANOVA, and PERMDISP2 tests.
-Lastly, we use mantel tests to evaluate the null hypothesis of isolation-by-distance,
+Lastly, we use mantel tests to evaluate the null hypothesis of isolation-by-distance.
 
@@
 ## clean the data
+library(tidyverse)
+library(magrittr)
+library(rcarbon)
+library(readxl)
+
 library(tidyverse)
@@
 nsigntypes <- ncol(signbase_full)
-signbase_full_clean <- signbase_full %>% 
+
+signbase_full_clean <- signbase_full %>%  # 511 rows
   filter(site_name != "Willendorf",
@@
          site_name != "Šandalja II") %>% 
-  dplyr::select(-other, -rectangle)
-nsites_clean <- signbase_full_clean%>% 
-  pull(site_name ) %>% 
-  unique() %>% 
-  length()
-nartifacts_clean <- nrow(signbase_full_clean)
-ncountries_clean <- signbase_full_clean %>% 
-  pull(country) %>% 
-  unique() %>% 
-  length()
-nsigntypes_clean <- nsigntypes - 1
-signbase_years <- signbase_full_clean %>% 
+  dplyr::select(-other, -rectangle) %>% 
+  mutate(longitude = ifelse(site_name == "Riparo Bombrini", 7.437500, longitude)) %>% 
+  mutate(latitude = ifelse(site_name == "Riparo Bombrini", 43.77083, latitude))
+
+signbase_years <- signbase_full_clean %>%  # 446
   drop_na(date_bp_max_min) %>% 
@@
          date_bp_max_error = parse_number(date_bp_max_error))
+
 signbase_years_cal <- 
@@
   data_frame()
+
+signbase_years$MedianBP <- signbase_years_cal$MedianBP
+
+signbase_full_clean <- signbase_full_clean %>% 
+  inner_join(signbase_years %>% 
+              dplyr::select(object_id, MedianBP))
+
+signbase_full_clean <- signbase_full_clean %>%
+  mutate(time_period = case_when((MedianBP > 42999) ~ "transitional",
+                                 (MedianBP > 39899) ~ "proto_aurignacian",
+                                 (MedianBP > 36999) ~ "early_aurignacian",
+                                 (MedianBP > 31999) ~ "evolved_aurignacian"))
+
+nsites_clean <- signbase_full_clean%>% 
+  pull(site_name ) %>% 
+  unique() %>% 
+  length()
+nartifacts_clean <- nrow(signbase_full_clean)
+ncountries_clean <- signbase_full_clean %>% 
+  pull(country) %>% 
+  unique() %>% 
+  length()
+nsigntypes_clean <- nsigntypes - 1
+
 pretty_print_dates <- 
@@
 ```{r}
-groups <- list("1" = c("Abri Pataud", "El Rascaño", 
-                       "Fumane", "La Viña",
-                       "Lhotka", "Pod Hradem",
-                       "Riparo Bombrini", "Sirgenstein Cave"),
-               "2" = c("Aurignac", "Gargas",
-                       "Grotte de la Verpillière I","Hohlenstein-Stadel",
-                       "Istállóskő Cave", "Mladeč",
-                       "d'Engihoul"),
-               "3" = c("Breitenbach", "Peskő Cave",
-                       "Vindija Cave", "Wildscheuer"),
-               "4" = c("Kvasice", "Labeko Koba", 
-                       "Les Rois"),
-               "5" = c("Le Trou Du Renard", "Abri Lartet/Gorge d'Enfer",
-                       "Tuto de Camalhot", "Castanet", 
-                       "Gatzarria", "Abri du Poisson"),
-               "6" = c("Bockstein-Törle", "Grotte de Goyet",
-                        "Trou Magrite", "Brassempouy"),
-               "7" = c( "Göpfelsteinhöhle",  "Slatinice", 
-                        "Grotte du Renne"),
-               "8" = c( "El Castillo", "Trou al'Wesse", 
-                        "Maisières-Canal"),
-               "9" = c("Vogelherd","Hohle Fels",
-                       "Geissenklösterle", "Le Terme Pialat", 
-                       "Les Cottés", "Spy",  "Grottes de Fonds-de-Forêt"),
-               "10" = c("Nová Dědina", "Menton/Grottes du Grimaldi",
-                        "Solutré", "Žlutava", 
-                        "Hornos de la Pena", "Otaslavice"),
-               "11" = c("Cellier", "Blanchard", 
-                        "La Ferrassie"),
-               "12" = c("La Souquette", "Shelter Birów IV",
-                        "Abri de Laussel"))
-group_df <- 
-  enframe(groups, 
-          name = "group", 
-          value = "site_name") %>%
-  unnest(cols = site_name)
-
 lat_long_df <- signbase_full_clean %>% 
@@
 
-signbase_unique_groups <- signbase_full_clean %>% 
+trans_unique_data <- signbase_full_clean %>% 
+  filter(time_period == "transitional") %>% 
   mutate(longitude = as.character(longitude),
@@
   summarize(across(where(is.numeric), sum)) %>% 
-  left_join(group_df) %>% 
-  left_join(lat_long_df)
+  left_join(lat_long_df) %>% 
+  mutate(time_period = "transitional")
 
-artifact_data <- signbase_unique_groups %>% 
+proto_unique_data <- signbase_full_clean %>% 
+  filter(time_period == "proto_aurignacian") %>% 
+  mutate(longitude = as.character(longitude),
+         latitude = as.character(latitude)) %>% 
+  group_by(site_name) %>%
+  summarize(across(where(is.numeric), sum)) %>% 
+  left_join(lat_long_df) %>% 
+  mutate(time_period = "proto_aurignacian")
+
+early_unique_data <- signbase_full_clean %>% 
+  filter(time_period == "early_aurignacian") %>% 
+  mutate(longitude = as.character(longitude),
+         latitude = as.character(latitude)) %>% 
+  group_by(site_name) %>%
+  summarize(across(where(is.numeric), sum)) %>% 
+  left_join(lat_long_df) %>% 
+  mutate(time_period = "early_aurignacian")
+
+evolved_unique_data <- signbase_full_clean %>% 
+  filter(time_period == "evolved_aurignacian") %>% 
+  mutate(longitude = as.character(longitude),
+         latitude = as.character(latitude)) %>% 
+  group_by(site_name) %>%
+  summarize(across(where(is.numeric), sum)) %>% 
+  left_join(lat_long_df) %>% 
+  mutate(time_period = "evolved_aurignacian")
+
+trans_artifact_data <- trans_unique_data %>% 
   column_to_rownames("site_name") %>% 
   dplyr::select(line:star)
+
+proto_artifact_data <- proto_unique_data %>% 
+  column_to_rownames("site_name") %>% 
+  dplyr::select(line:star)
+
+early_artifact_data <- early_unique_data %>% 
+  column_to_rownames("site_name") %>% 
+  dplyr::select(line:star)
+
+evolved_artifact_data <- evolved_unique_data %>% 
+  column_to_rownames("site_name") %>% 
+  dplyr::select(line:star)
+
+trans_unique_data <- trans_unique_data %>% 
+  mutate(group = case_when((site_name == "Pod Hradem") ~ "1",
+                           (site_name == "El Castillo" | site_name == "Hohle Fels") ~ "2"))
+
+proto_unique_data <- proto_unique_data %>% 
+  mutate(group = case_when((site_name == "Labeko Koba" | 
+                              site_name == "Gatzarria" | 
+                              site_name == "Hohlenstein-Stadel" | 
+                              site_name == "Mladeč" | 
+                              site_name == "Abri Pataud" | 
+                              site_name == "Fumane") ~ "1",
+                           (site_name == "Geissenklösterle" |
+                              site_name == "Grotte du Renne" |
+                              site_name == "Vogelherd" |
+                              site_name == "Spy" |
+                              site_name == "La Ferrassie" |
+                              site_name == "Les Cottés") ~ "2"))
+
+early_unique_data <- early_unique_data %>% 
+  mutate(group = case_when((site_name == "Grottes de Fonds-de-Forêt" |
+                              site_name == "Riparo Bombrini"|
+                              site_name == "Grotte de la Verpillière I"|
+                              site_name == "Vindija Cave" |
+                              site_name == "Trou al'Wesse") ~ "1",
+                           (site_name == "Hohle Fels" | 
+                              site_name == "Solutré" |
+                              site_name == "Castanet"|
+                              site_name == "Cellier" | 
+                              site_name == "Blanchard") ~ "2"))
+
+evolved_unique_data <- evolved_unique_data %>% 
+  mutate(group = case_when((site_name == "Les Rois" |
+                              site_name == "Gargas" |
+                              site_name == "La Viña"|
+                              site_name ==  "Sirgenstein Cave") ~ "1",
+                           (site_name == "Bockstein-Törle"|
+                              site_name == "Vogelherd" |
+                              site_name == "Hohle Fels"|
+                              site_name == "Trou Magrite") ~ "2"))
+
 ```
@@
 library(sf)
+library(ggpubr)
+time_unique_data_full <-rbind(trans_unique_data, proto_unique_data, 
+                              early_unique_data, evolved_unique_data)
 
-library(ggpubr)
-
-abundance_data <- signbase_full_clean %>% 
-  pivot_longer(cols= line:star,
-               values_to = "sign_total") %>% 
-  filter(sign_total != 0) %>% 
-  mutate(longitude = as.character(longitude),
-         latitude = as.character(latitude)) %>% 
-  group_by(site_name) %>%
-  summarize(across(where(is.numeric), sum))
-
-signbase_sf <- 
-  st_as_sf(signbase_unique_groups, 
-           coords = c("longitude", "latitude"),
+time_unique_data_full$time_period <- factor(time_unique_data_full$time_period, 
+                                            levels = c("transitional", "proto_aurignacian", 
+                                                       "early_aurignacian", "evolved_aurignacian"))
+signbase_sf <- time_unique_data_full  %>% 
+  st_as_sf(coords = c("longitude", "latitude"),
            remove = FALSE,
@@
 
-artifact_map <- ggplot(Europe) +
+ggplot(Europe) +
   geom_sf() +
@@
   theme_minimal() +
-  labs(size = "Sign Count")
-
-date_graph <- ggplot(signbase_years_cal) +
-  aes(x = MedianBP) +
-  geom_bar(width = 150) +
-  labs(x = "Median calibrated BP")
-
-abundance_graph <- ggplot(abundance_data) +
-  aes(x = sign_total, y = reorder(site_name, sign_total)) +
-  geom_col() +
-  rotate_x_text(angle = 90, hjust = NULL, vjust = NULL) +
-  labs(x = "Site name",
-       y = "Sign count") +
-  theme(axis.text.y = element_text(size = 4.25))
-
-library(cowplot)
-plot_grid(date_graph, abundance_graph, artifact_map,
-          labels = "AUTO", align = "v",
-          nrow = 3, scale = c(0.75, 0.75, 1.25))
+  facet_wrap(~time_period)
 
@@
 All our tests were run with presence/absence data, in order to keep the data from being affected by outliers like Vogelherd.
-We ran a Mantel test between an abundance distance matrix and and a Euclidean distance matrix in order to make sure that the exclusion of abundance would not significantly impact our results, and found that it was not statistically significant [@Lycett_2019] (P = `r p_value_abundance_correlation`, Mantel R = `r r_value_abundance_correlation`)
-
-The artifacts in question are determined to have been made in the Paleolithic era, specifically in the Aurignacien technocomplex.
+We ran a Mantel test between an abundance distance matrix and and a Euclidean distance matrix in order to make sure that the exclusion of abundance would not significantly impact our results, and found that it was not statistically significant [@Lycett_2019] (P = `r p_value_abundance_correlation`, Mantel R = `r r_value_abundance_correlation`).
 The calibrated radiocarbon dates of the artifacts range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years.
-As seen in @fig-site-map, the dates have a few outliers, but tend to cluster between 35000 - 42500 BP.
+We divided the data up into four periods:
 

### 2025-02-28 cluster_pipeline.R

@@ -262,7 +262,7 @@ trans_network <- function(artifact_data){
 ## mantel function tells us whether changes in sign makeup is correlated with chosen variable change. 
 ## Permanova tells us how similar in chosen variable members of group are
 
-perm_function <- function(period = "none", variable = PCA, method = "permanova"){
+perm_function <- function(period = "none", variable = PCA, method = "permanova", signbase_sf = signbase_sf){
   set.seed(500)
   
   if(period == "none"){perm_df <- signbase_sf}

### 2025-03-03 endofquarterpaper.qmd
edits

@@
 ---
-
 # Introduction
+<!-- swap around sentences-->
 One distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile.
-During this period, we also see a rise in organized settlements and more complex social organizations across Europe [@Boyd_Silk_2020].
+During this period, we also see a rise in more complex social organizations across Europe [@Boyd_Silk_2020] coupled with biological evolution.
 Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand these shifting patterns of social organization and diversity
@@
 [@McDonald_Harper_2016; @Pereira_Manen_2023; @Fuentes_Lucas_Robert_2019].
-<!-- add more about aurignacian -->
+
+<!-- mention upper paleolithic in last sentence of above paragraph -->
 
 Considered to be the starting point of the Upper Paleolithic, the Aurignacian period (ca.
-43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].The Aurignacian can be divided up into four chronological periods: Transitional, Proto-Aurignacian, Early Aurignacian, and Evolved Aurignacian [@Banks_dErrico_Zilhao_2013, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021].
+43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].
 
-In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe across the subsections of the Aurignacian period.
-The basis of this analysis is the idea that variations in signs can serve as an indicator of group membership and cultural similarity.
+In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe across the subperiods of the Aurignacian period.
+
+<!-- The basis of this analysis is the idea that variations in signs can serve as an indicator of group membership and cultural similarity.-->
+
 This approach is supported by ethnographic and historical evidence [@texier2010howiesons; @Hodder1979; @Hodder1977; @Vanhaeren_2005] as well as similar archaeological studies of ornamental and parietal art [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d; @McDonald_Harper_2016].
-In this study, we address the following questions: How does sign distribution change across periods of the Aurignacian?
-How do the groupings vary in diversity, size, connectivity, and boundaries?
-What do the patterns tell us about the cultural groupings and movement of Upper Paleolithic humans throughout the Aurignacian?
+
+<!-- read above literature to confirm relevance-->
 
@@
 
-The purpose of this study is to contribute to our understanding of the patterns of cultural and genetic diversity among the early modern humans of the Upper Paleolithic.
-Previous works have performed cluster analysis methods upon variations in artefacts to represent and understand patterns of groups and cultural diversity, and have primarily relied upon personal ornamentation and parietal art, with few taking mobile objects into account.
+Social organization dynamics have previously been explored using cluster analysis methods upon variations in artefacts to represent and understand patterns of groups and cultural diversity, and have primarily relied upon personal ornamentation and parietal art.
+
 For example, multivariate analyses of Gravettian beads suggests an east to west cline of nine cultural groups [@baker2024evidence].
+
 The data set consisted of 134 discrete beads types from both burial and occupation sites, as distinguished by shape and material.
-Seriation analysis, PCoA, and network analysis were used to identify the groups.
-The geographically discrete clusters identified encompassed the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and Western Mediterranean regions.
+Seriation analysis, PCoA, and network analysis were used to identify the groups from the bead data.
+
+The geographically discrete groups identified encompassed the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and Western Mediterranean regions.
+
+<!-- reread paper, make sure that last sentence accurately describes results  -->
 The isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance, is tested using a Mantel test, determining that geographic distance cannot solely account for the bead distribution.
+
+<!-- Add connector between baker and vanharen paragraphs, add more detail about Vanharen methods-->
 
@@
 
-<!-- remove section on mobility? -->
+<!-- Rephrase opening sentence to be less obvious -->
+The use of geometric signs to identify cultural groups is supported by @Sauvet_2018, in which quadrilateral geometric parietal signs are used to identify Upper Paleolithic cultural groups in the Cantabrian region of Spain.
+The authors identified and classified the signs based on variations in form, then performed a factorial correspondence analysis and ascending hierarchical clustering upon the data in order to identify two clusters, which they argued represents distinct and separate groups of humans.
+In their study, @Sauvet_2018 made a case for the relevance of signs as identity markers, pointing to both the relevant qualities of signs, which are similar to those explained above, and the case studies throughout history where they served a similar purpose.
 
-Following the established precedent of using ornamentation and signs to identify cultural groups, we look to see if designs on mobile objects can be applicable here, and what it can add to the research being done.
+<!-- remove section on mobility -->
+
+Following the <!-- Baker et al.
+and Vanharen et al work - reference them more directly --> of using ornamentation and signs to identify cultural groups, we look to see if designs on mobile objects can be applicable here, and what it can add to the research being done.
 The additional dimension of mobility and comparison of in this study allows us to potentially identify more geographically dispersed cultural groups, which might help visualize the migration patterns of early modern humans.
@@
 
-<!--  -->
+The Aurignacian period occurred roughly 43,000 - 30,000 years BP, and is largely characterized by material culture indicators, as well as an ongoing biological shift from Neanderthals to early modern humans.
+<!-- add specific details about cultural markers, both for aurignacian and for each time period (create table in excel with name, date range, artifact indicators, environmental context, sources in the caption)-->
 
-Our use of geometric signs in this study is supported by @Sauvet_2018, in which quadrilateral geometric parietal signs are used to identify Upper Paleolithic cultural groups in the Cantabrian region of Spain.
-The authors identified and classified the signs based on variations in form, then performed a factorial correspondence analysis and ascending hierarchical clustering upon the data in order to identify two clusters, which they argued represents distinct and separate groups of humans.
-In their study, @Sauvet_2018 made a case for the relevance of signs as identity markers, pointing to both the relevant qualities of signs, which are similar to those explained above, and the case studies throughout history where they served a similar purpose.
+The Aurignacian can be further divided into subphases, again mainly distinguished by shifts in material culture [@Tartar_2015], Proto, Early, Evolved, and Late Aurignacian.
 
-<!-- add more about aurignacian time period - what was going on, what was the criteria -->
-
-The Aurignacian period occurred roughly 43,000 - 30,000 years BP, and is largely characterized by material culture indicators, as well as an ongoing biological shift from Neanderthals to early modern humans.
-The Aurignacian can be further divided into subphases, again mainly distinguished by shifts in material culture [@Tartar_2015], Proto, Early, Evolved, and Late Aurignacian.
 @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main periods, Aur-P1 (Proto/Early), whose typological observations corresponded with the time period of 43–37 kyr BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods kyr BP and 37-32 kyr BP.
@@
 <!-- change hypothesis -->
-In this paper, we test the hypothesis that cultural groups are discernible in a data set of geometric signs on mobile objects.
+
+In this study, we aim to address the following questions: How does sign distribution change across periods of the Aurignacian?
+How do the groupings in sign distibution vary in diversity, size, connectivity, and boundaries?
+What do the patterns tell us about dynamics of social organization throughout the Aurignacian?
 Firstly, we identify clusters within each time period using seriation and network analysis.
@@
 library(ggpubr)
+
+ggplot(signbase_full_clean) +
+  aes(x = MedianBP, fill = time_period) +
+  geom_bar(width = 150)
+
 time_unique_data_full <-rbind(trans_unique_data, proto_unique_data, 
@@
 
+Dates provided by Signbase.
+
 ## Testing for Cultural Groupings
@@
 ## Testing for Cultural Groupings
+```{r}
+library(ggpubr)
+library(phangorn)
+library(ggraph)
+library(statnet)
+library(ecodist)
+library(ape)
+library(ggrepel)
+library(vegan)
+library(cowplot)
+library(igraph)
 
-```{r}
-#| label: fig-seriation
-#| fig-cap: "Seriation analysis of the data"
-## seriation
-library(ggpubr)
-artifact_matrix <- as.matrix(artifact_data %>% 
- mutate(across(everything(), ~replace(., . > 1, 1))))
-rownames(artifact_matrix) <- rownames(artifact_data)
-
-# Define the mplot function
 mplot <- function(x, ..., fill_colour = "black", title = NULL, row_labels = NULL, col_labels = NULL, country_colors = NULL) {
-  # Create the data frame
   d <- data.frame(
@@
 
-  # Add country colors to the data frame if provided
   if (!is.null(country_colors)) {
@@
 
-  # Use custom row/column labels if provided, otherwise use defaults
   if (is.null(col_labels)) col_labels <- colnames(x)
@@
 
-  # Plot
   p <- ggplot(d, aes(x, y)) +
     geom_tile(col = "grey50", linewidth = 0.5, aes(fill = ifelse(z == 1, country_color, "white"))) +
-    scale_fill_identity() +  # Use the fill values directly
+    scale_fill_identity() +  
     coord_equal(expand = FALSE) +
@@
     ) +
-    labs(x = "Sign Type", y = "Site Name", fill = "Count", title = title) +
+    labs(x = "", y = "") +
     rotate_x_text(angle = 90, hjust = NULL, vjust = NULL)
@@
 
-# Define the increase_focus function
 increase_focus <- function(x, country = NULL) {
-  # Reorder rows by the mean position of 1s in each row
   mcp <- apply(
@@
 
-  # Reorder the country vector if provided
   if (!is.null(country)) {
@@
 
-  # Reorder columns by the mean position of 1s in each column
   mrp <- apply(
@@
 
-  # Return the reordered matrix and country vector
   list(matrix = x, country = country)
@@
 
-# Define the concentrate function
 concentrate <- function(x, country = NULL, max_iter = 100) {
@@
 
-# get country names, for example
-x_df <- 
-  artifact_data %>% 
-  rownames_to_column("site_name") %>% 
-  left_join(signbase_unique_groups %>% 
-              select(site_name,
-                     group),
+produce_clusters <- function(artifact_data, artifact_data_unique, method = "seriation"){
+  artifact_data <- artifact_data %>% 
+    select_if(~ !is.numeric(.) || sum(.) != 0)
+  jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
+  dm <- as.matrix(jac)
+
+  if(method == "seriation"){
+  artifact_matrix <- as.matrix(artifact_data %>% 
+                                 mutate(across(everything(), ~replace(., . > 1, 1))))
+  rownames(artifact_matrix) <- rownames(artifact_data)
+
+  x_df <- artifact_data_unique %>% 
+    left_join(signbase_full_clean,
               by = "site_name")
-
-countries <- x_df$group
-
-# Generate a named vector of colors, one for each unique country
+  countries <- x_df$country
   unique_countries <- unique(countries)
-color_palette <-  RColorBrewer::brewer.pal(length(unique_countries), "Set3")   # Generate distinct colors
+  color_palette <-  RColorBrewer::brewer.pal(length(unique_countries), "Set3")
   row_colors <- setNames(color_palette, unique_countries)
-countries_colors <- row_colors[countries]
-
-# Concentrate the matrix and reorder the country vector
   result <- concentrate(artifact_matrix, country = countries)
@@
   reordered_countries <- result$country
-
-# Map countries to colors
+  countries_colors <- row_colors[countries]
   country_colors <- row_colors[reordered_countries]
 
-# Plot the concentrated matrix with country colors
-mplot(
-  concentrated_matrix,
-  country_colors = country_colors
-)
-```
+  final_plot <- mplot(concentrated_matrix,
+    country_colors = country_colors)}
 
-```{r}
-#| label: fig-neighbor-net
-#| fig-cap: "Neighbor-Net analysis of the data"
-library(phangorn)
-library(ggraph)
-library(statnet)
-jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
-dm <- as.matrix(jac)
+  if(method == "network"){
   disim <- 1 - dm
@@
                        names.eval='weight')
-disim_net %v% 'vertex.names' <- 
-  row.names(artifact_data)
-
+  disim_net %v% 'vertex.names' <- row.names(artifact_data)
   set.seed(500)
 
-ggraph(disim_net, layout="fr") +
+  final_plot <- ggraph(disim_net, layout="fr") +
       geom_edge_link(aes(width = weight, 
@@
                      edge_colour = "black", 
-                 show.legend=T) +
+                     show.legend= F,
+                    ) +
       scale_edge_width(range=c(0.5,1.5)) +
@@
       geom_node_text(aes(label=row.names(artifact_data)), 
-                 size=6, 
+                     size=3, 
                      family= "Arial", 
@@
       theme(text= element_text(size = 15), 
-        panel.background = element_rect(fill = 'white'))
+            panel.background = element_rect(fill = 'white')) +
+    border(color = "black", size = 0.5)
+
+  }
+  return(final_plot)
+}
+
+strength_function <- function(artifact_data, group_data, type = "table"){
+  artifact_data <- artifact_data %>% 
+    select_if(~ !is.numeric(.) || sum(.) != 0)
+
+  set.seed(500)
+  jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
+  dm <- as.matrix(jac)
+
+  if(type == "table"){
+  modul_matrix <-
+    graph_from_adjacency_matrix(1 - dm, mode = "undirected")
+  mem <- as_membership(parse_number(group_data$group))
+  as_group <- modularity(modul_matrix, membership = mem)
+
+  perman <- adonis2(jac~as.factor(group_data$group), 
+                    method = "jaccard",
+                    sqrt.dist = TRUE)
+  perman <- perman %>% 
+    data_frame() %>% 
+    slice_head()
+  perman_r <- perman$R2
+  perman_p <- perman$`Pr(>F)`
+
+  site_distances <- dist(cbind(group_data$longitude, group_data$latitude))
+  mantel_distance <- vegan::mantel(jac, site_distances, permutations = 1000)
+  p_value_distance_correlation <- round(mantel_distance$signif, 3)
+  r_value_distance_correlation <- round(mantel_distance$statistic, 3)
+
+  final_table <- data_frame("modularity" = as_group, 
+                            "permanova_r" = perman_r,
+                            "permanova_p" = perman_p,
+                            "distance_mantel_r" = r_value_distance_correlation, 
+                            "distance_mantel_p" = p_value_distance_correlation)
+
+  return(final_table)}
+
+  if(type == "dispersion"){
+    set.seed(500)
+    dispersion <- betadisper(jac, group = group_data$group)
+    disp_plot<- plot(dispersion, hull = FALSE, ellipse=TRUE)
+    return(disp_plot)}
+}
+
+trans_network <- function(artifact_data){
+  artifact_data <- artifact_data %>% 
+    select_if(~ !is.numeric(.) || sum(.) != 0)
+    jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
+    dm <- as.matrix(jac)
+    disim <- 1 - dm
+    disim[disim<0.1] <- 0
+    disim_net <- network(disim,
+                         directed=F,
+                         ignore.eval=F,
+                         names.eval='weight')
+    disim_net %v% 'vertex.names' <- row.names(artifact_data)
+    set.seed(500)
+    final_plot <- ggraph(disim_net, layout="fr") +
+      geom_edge_link(aes(width = weight, 
+                         alpha=weight), 
+                       edge_colour = "black", 
+                       show.legend=F) +
+      scale_edge_width(range=c(0.5,1.5)) +
+      scale_edge_colour_gradient(low = "#CCCCCC",
+                                   high = "#000000") +
+      geom_node_point(alpha=1, size=5) +
+      geom_node_text(aes(label=row.names(artifact_data)), 
+                     size=3, 
+                     family= "Arial", 
+                     repel=T, max.overlaps = 100) +
+      theme(text= element_text(size = 15), 
+            panel.background = element_rect(fill = 'white')) +
+      border(color = "black", size = 0.5)
+    return(final_plot)
+    }
+
+## mantel function tells us whether changes in sign makeup is correlated with chosen variable change. 
+## Permanova tells us how similar in chosen variable members of group are
+
+perm_function <- function(period = "none", variable = PCA, method = "permanova", signbase_sf = signbase_sf){
+  set.seed(500)
+
+  if(period == "none"){perm_df <- signbase_sf}
+
+  else{perm_df <- signbase_sf %>% 
+    filter(time_period == {{period}})}
+
+  if(method == "permanova"){
+  var_jac <- perm_df %>% 
+      st_drop_geometry() %>% 
+      column_to_rownames("id") %>% 
+      dplyr::select({{variable}}) %>% 
+      vegdist("jaccard")
+
+  result <- adonis2(var_jac~as.factor(perm_df$group), 
+                         method = "jaccard",
+                         sqrt.dist = TRUE)}
+
+  if(method == "mantel"){
+    sign_jac <- perm_df %>% 
+      st_drop_geometry() %>% 
+      column_to_rownames("id") %>% 
+      dplyr::select(line:star) %>% 
+      vegdist("jaccard", binary = TRUE)
+
+    var_sim <- perm_df %>% 
+      st_drop_geometry() %>% 
+      column_to_rownames("id") %>% 
+      dplyr::select({{variable}}) %>% 
+      dist()
+
+    result <- vegan::mantel(sign_jac,  var_sim, permutations = 1000)
+  }
+
+  return(result)
+}
 ```
 
+### Seriation
 ```{r}
-#| label: fig-pcoa
-#| fig-cap: "PCoA analysis of the data"
-library(ecodist)
-library(ape)
-pcoa <- pcoa(jac)
-pcoa_df <- data.frame(pcoa1 = pcoa$vectors[,1], 
-                      pcoa2 = pcoa$vectors[,2])
-pcoa_df <- pcoa_df %>% 
-  mutate(site_name = rownames(artifact_data))
-library(ggrepel)
-ggplot(pcoa_df) +
-  aes(x=pcoa1, 
-      y=pcoa2) +
-  geom_point() +
-  labs(x = "PC1",
-       y = "PC2",) +
-  theme(title = element_text(size = 10)) +
-  geom_text_repel(aes(label = site_name), 
-                  size = 2,
-                  max.overlaps = 20)
+#| label: fig-seriation
+#| fig-cap: "Seriation analysis of the data"
+library(cowplot)
+
+seriation_trans <- produce_clusters(trans_artifact_data, trans_unique_data, method = "seriation")
+seriation_proto <- produce_clusters(proto_artifact_data, proto_unique_data, method = "seriation")
+seriation_early <- produce_clusters(early_artifact_data, early_unique_data, method = "seriation")
+seriation_evolved <- produce_clusters(evolved_artifact_data, evolved_unique_data, method = "seriation")
+
+plot_grid(seriation_trans,seriation_proto, seriation_early , seriation_evolved, 
+          ncol = 2, align = "hv", 
+          labels = c("Transitional", "Proto", "Early", "Evolved"), label_size = 10)
+
+```
+
+### Network Analysis
+```{r}
+#| label: fig-neighbor-net
+#| fig-cap: "Neighbor-Net analysis of the data"
+network_trans <- trans_network(trans_artifact_data)
+network_proto <- produce_clusters(proto_artifact_data, proto_unique_data, method = "network")
+network_early <- produce_clusters(early_artifact_data, early_unique_data, method = "network")
+network_evolved <- produce_clusters(evolved_artifact_data, evolved_unique_data, method = "network")
+
+plot_grid(network_trans, network_proto, network_early, network_evolved, 
+          ncol = 2, align = "hv", labels = c("Transitional", "Proto", "Early", "Evolved"), 
+          label_size = 7, label_colour = "black", label_y = 1.02, label_x = -0.03)
 ```
@@
 group_table <- data_frame("Group" = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11"), "Site Names" = c(group_1, group_2, group_3, group_4, group_5, group_6, group_7, group_8, group_9, group_10, group_11))
+
 kable(group_table)

### 2025-03-03 Extra/time-periods.R

@@ -1,147 +1,3 @@
-library(tidyverse)
-library(magrittr)
-library(rcarbon)
-library(readxl)
-signbase_full <- read_csv("data/signBase_Version1.0.csv")
-
-signbase_full_clean <- signbase_full %>%  # 511 rows
-  filter(site_name != "Willendorf",
-         site_name != "Riparo di Fontana Nuova",
-         site_name != "Muralovka",
-         site_name != "Shanidar Cave",
-         site_name != "Hayonim Cave",
-         site_name != "El Salitre",
-         site_name != "Grotte De La Princesse Pauline",
-         site_name != "Šandalja II") %>% 
-  dplyr::select(-other, -rectangle) %>% 
-  mutate(longitude = ifelse(site_name == "Riparo Bombrini", 7.437500, longitude)) %>% 
-  mutate(latitude = ifelse(site_name == "Riparo Bombrini", 43.77083, latitude))
-
-signbase_years <- signbase_full_clean %>%  # 446
-  drop_na(date_bp_max_min) %>% 
-  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+\\/\\-", "±")) %>% 
-  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+", "±")) %>% 
-  separate(date_bp_max_min,
-           sep = " - ",
-           c("date_bp_max", "date_bp_min"),
-           remove = FALSE) %>% 
-  mutate(date_bp_max = ifelse(str_detect(date_bp_max, "\\/"),
-                              str_extract(date_bp_max, ".*?(?=\\/)"),
-                              date_bp_max)) %>% 
-  mutate(date_bp_min = ifelse(str_detect(date_bp_min, "\\/"),
-                              str_extract(date_bp_min, ".*?(?=\\/)"),
-                              date_bp_min)) %>% 
-  separate(date_bp_max,
-           sep = "±",
-           c("date_bp_max_age",
-             "date_bp_max_error")) %>% 
-  separate(date_bp_min,
-           sep = "±",
-           c("date_bp_min_age",
-             "date_bp_min_error")) %>% 
-  drop_na(date_bp_max_age,
-          date_bp_max_error) %>% 
-  mutate(date_bp_max_age = parse_number(date_bp_max_age),
-         date_bp_max_error = parse_number(date_bp_max_error))
-
-signbase_years_cal <- 
-  rcarbon::calibrate(signbase_years$date_bp_max_age,
-                     signbase_years$date_bp_max_error,
[…]

### 2025-03-03 endofquarterpaper.qmd
edits

@@
 ---
-# Introduction
-<!-- swap around sentences-->
-One distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile.
-During this period, we also see a rise in more complex social organizations across Europe [@Boyd_Silk_2020] coupled with biological evolution.
-Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand these shifting patterns of social organization and diversity
-
- <!-- change citations maybe -->
-[@McDonald_Harper_2016; @Pereira_Manen_2023; @Fuentes_Lucas_Robert_2019].
-
-<!-- mention upper paleolithic in last sentence of above paragraph -->
-
+# Introduction and Background
+The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020] coupled with biological evolution.
+Another distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile.
+Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence, @Sauvet_2018, @Kuhn_Stiner_2007, @Vanhaeren_d].
 Considered to be the starting point of the Upper Paleolithic, the Aurignacian period (ca.
 43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].
+In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe across the phases of the Aurignacian period.
 
-In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe across the subperiods of the Aurignacian period.
+Social organization dynamics have previously been explored using cluster analysis methods upon variations in artefacts to represent and understand patterns of cultural groups.
+For example, multivariate analyses of variations in Gravettian beads suggests an east to west cline of nine cultural groups [@baker2024evidence].
+Seriation analysis, PCoA, and network analysis were used to identify the groups from the bead data, which consisted of 134 discrete bead variations from both burial and occupation sites, as distinguished by shape and material.
+The geographically discrete groups identified encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and Western Mediterranean regions.
+The isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance, was tested using a Mantel test, finding that geographic distance cannot solely account for the bead distribution, proving the influence of cultural and social dynamics on the groups found.
 
-<!-- The basis of this analysis is the idea that variations in signs can serve as an indicator of group membership and cultural similarity.-->
+In another study to find cultural groups, similar analyses were performed upon personal ornament types from the Aurignacian period, with the variation in types serving as a proxy for linguistic diversity [@Vanhaeren_d].
+Seriation, correspondence, and GIS analyses were performed upon a personal ornament data set, which consisted of 157 distinct ornament types from Eastern Aurignacian sites.
+The analyses identified fourteen geographically cohesive groups sharing similar ornament type associations, which sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.
+The possibility that the observed groups were influenced by availability of raw materials was considered, and ultimately discarded as an explanation for the patterns.
 
-This approach is supported by ethnographic and historical evidence [@texier2010howiesons; @Hodder1979; @Hodder1977; @Vanhaeren_2005] as well as similar archaeological studies of ornamental and parietal art [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d; @McDonald_Harper_2016].
+Additionally, @Sauvet_2018 used quadrilateral geometric parietal signs to identify Upper Paleolithic cultural groups in the Cantabrian region of Spain.
+The authors identified and classified the signs based on variations in form, then performed a factorial correspondence analysis and ascending hierarchical clustering upon the data.
+Two clusters were identified from those methods, representing two distinct cultural groups.
+@Sauvet_2018 makes a case for the relevance of geometric signs as identity markers in identifying cultural groups, pointing to the relevant qualities of signs, as well as historical and ethnographic case studies of signs serving as markers of group membership case studies.
 
-<!-- read above literature to confirm relevance-->
-
-# Background
-
-Social organization dynamics have previously been explored using cluster analysis methods upon variations in artefacts to represent and understand patterns of groups and cultural diversity, and have primarily relied upon personal ornamentation and parietal art.
-
-For example, multivariate analyses of Gravettian beads suggests an east to west cline of nine cultural groups [@baker2024evidence].
-
-The data set consisted of 134 discrete beads types from both burial and occupation sites, as distinguished by shape and material.
-Seriation analysis, PCoA, and network analysis were used to identify the groups from the bead data.
-
-The geographically discrete groups identified encompassed the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and Western Mediterranean regions.
-
-<!-- reread paper, make sure that last sentence accurately describes results  -->
-The isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance, is tested using a Mantel test, determining that geographic distance cannot solely account for the bead distribution.
-
-<!-- Add connector between baker and vanharen paragraphs, add more detail about Vanharen methods-->
-
-Similar analyses (seriation, correspondence and GIS) were also performed upon personal ornament types from the Aurignacian period as a proxy for linguistic diversity.
-The data set consisted of 157 distinct ornament types, and the analyses identified fourteen geographically cohesive clusters sharing similar ornament type associations [@Vanhaeren_d].
-The clusters identified are geographically discrete and sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.
-The possibility that observed patterns were influenced by availability of raw materials was considered, and ultimately discarded as an explanation for the patterns.
-
-<!-- Rephrase opening sentence to be less obvious -->
-The use of geometric signs to identify cultural groups is supported by @Sauvet_2018, in which quadrilateral geometric parietal signs are used to identify Upper Paleolithic cultural groups in the Cantabrian region of Spain.
-The authors identified and classified the signs based on variations in form, then performed a factorial correspondence analysis and ascending hierarchical clustering upon the data in order to identify two clusters, which they argued represents distinct and separate groups of humans.
-In their study, @Sauvet_2018 made a case for the relevance of signs as identity markers, pointing to both the relevant qualities of signs, which are similar to those explained above, and the case studies throughout history where they served a similar purpose.
-
-<!-- remove section on mobility -->
-
-Following the <!-- Baker et al.
-and Vanharen et al work - reference them more directly --> of using ornamentation and signs to identify cultural groups, we look to see if designs on mobile objects can be applicable here, and what it can add to the research being done.
-The additional dimension of mobility and comparison of in this study allows us to potentially identify more geographically dispersed cultural groups, which might help visualize the migration patterns of early modern humans.
-Previous studies have primarily excluded mobile objects, and tend to find more geographically compact cultural groups as a result.
-
+Following the work done by @baker2024evidence, @Vanhaeren_d, and @Sauvet_2018, in using variation in ornamentation and signs to identify cultural groups, we look to see how geometric sign design changes across the sub-periods of the Aurignacian period.
 The Aurignacian period occurred roughly 43,000 - 30,000 years BP, and is largely characterized by material culture indicators, as well as an ongoing biological shift from Neanderthals to early modern humans.
-<!-- add specific details about cultural markers, both for aurignacian and for each time period (create table in excel with name, date range, artifact indicators, environmental context, sources in the caption)-->
-
-The Aurignacian can be further divided into subphases, again mainly distinguished by shifts in material culture [@Tartar_2015], Proto, Early, Evolved, and Late Aurignacian.
-
+Chronology of the Aurignacian is primarly determined by correspondance with those material culture markers, which include carinated scrapers, small-flake tool production, and split-based bone and ivory tools [@Chu_Richter_2020].
+The Aurignacian can be divided into phases, again mainly distinguished by shifts in material culture; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015].
 @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main periods, Aur-P1 (Proto/Early), whose typological observations corresponded with the time period of 43–37 kyr BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods kyr BP and 37-32 kyr BP.
@@
 
-<!-- change hypothesis -->
+<!-- add specific details about cultural markers, both for aurignacian and for each time period (create table in excel with name, date range, artifact indicators, environmental context, sources in the caption)-->
 
-In this study, we aim to address the following questions: How does sign distribution change across periods of the Aurignacian?
-How do the groupings in sign distibution vary in diversity, size, connectivity, and boundaries?
+```{r}
+library(knitr)
+
+[@Jöris_Street_2008, @Banks_dErrico_Zilhao_2013, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021, @Sanz-Royo_Terlato_Marín-Arroyo_2024, @Chu_Richter_2020], 
+```
+
+In this study, we aim to address the following questions: How does sign distribution change across phases of the Aurignacian?
+How do the groupings in sign distribution vary in diversity, size, connectivity, and boundaries?
 What do the patterns tell us about dynamics of social organization throughout the Aurignacian?
@@
 ## clean the data
-library(tidyverse)
-library(magrittr)
-library(rcarbon)
-library(readxl)
-
 library(tidyverse)
@@
 
-signbase_years <- signbase_full_clean %>%  # 446
+signbase_years <- signbase_full_clean %>%
   drop_na(date_bp_max_min) %>% 
@@
 
-nsites_clean <- signbase_full_clean%>% 
+signbase_full_clean <- signbase_full_clean %>% 
+  mutate(sign_total = rowSums((signbase_full_clean [,25:53]))) %>% 
+  filter(sign_total > 0)
+
+nsites_clean <- signbase_full_clean %>% 
   pull(site_name ) %>% 
@@
 
+time_unique_data_full <-rbind(trans_unique_data, proto_unique_data, 
+                              early_unique_data, evolved_unique_data)
+
+time_unique_data_full$time_period <- factor(time_unique_data_full$time_period, 
+                                            levels = c("transitional", "proto_aurignacian", 
+                                                       "early_aurignacian", "evolved_aurignacian"))
+
 ```
@@
 library(vegan)
-library(ade4)
 
 set.seed(200)
+
+artifact_data <- time_unique_data_full %>% 
+  column_to_rownames("id") %>% 
+  select(line:star)
+
 euc_data <- artifact_data %>% 
-  rownames_to_column("site_name") %>% 
-  group_by(site_name) %>% 
-  mutate(sign_count = sum(line:star)) %>% 
-  column_to_rownames("site_name") %>% 
+  mutate(sign_count = rowSums(.)) %>% 
   dplyr::select(sign_count)
@@
 euc_dist <- dist(x = euc_data, method = "euclidean")
-jac <- vegan::vegdist(artifact_data, method = "jaccard", binary = TRUE)
-euc_mantel <- vegan::mantel(jac, euc_dist, permutations = 1000)
+jac_full <- vegan::vegdist(artifact_data, method = "jaccard", binary = TRUE)
+
+euc_mantel <- vegan::mantel(euc_dist, jac_full, permutations = 1000)
 p_value_abundance_correlation <- round(euc_mantel$signif, 3)
@@
   geom_bar(width = 150)
-
-time_unique_data_full <-rbind(trans_unique_data, proto_unique_data, 
-                              early_unique_data, evolved_unique_data)
 
@@
   facet_wrap(~time_period)
-
 ```
@@
 library(ggrepel)
-library(vegan)
 library(cowplot)
@@
   site_distances <- dist(cbind(group_data$longitude, group_data$latitude))
-  mantel_distance <- vegan::mantel(jac, site_distances, permutations = 1000)
+  mantel_distance <- vegan::mantel(site_distances, jac, permutations = 1000)
   p_value_distance_correlation <- round(mantel_distance$signif, 3)
@@
 #| fig-cap: "Table of the sites by group"
-library(knitr)
-group_1 <- c("Abri Pataud, El Rascaño, Fumane, La Viña, Lhotka, Pod Hradem, Riparo Bombrini, Sirgenstein Cave, Šandalja II")
-group_2 <- c("Kvasice, Labeko Koba, Les Rois, Slatinice, Göpfelsteinhöhle")
-group_3 <- c("Aurignac, d'Engihoul, Istállóskő Cave, Hohlenstein-Stadel, Gargas, Grotte de la Verpillière I, Mladeč, Brassempouy, Gatzarria")
-group_4 <- c("Wildscheuer, Breitenbach, Peskő Cave, Vindija Cave, El Castillo")
-group_5 <- c("Tuto de Camalhot, Abri Lartet/Gorge d'Enfer, Castanet, Le Trou Du Renard, Grottes de Fonds-de-Forêt")
-group_6 <- c("Nová Dědina, Menton/Grottes du Grimaldi, Grotte du Renne, Solutré, Žlutava, Spy")
-group_7 <- c("Trou Magrite, Grotte De La Princesse Pauline, Grotte de Goyet, Bockstein-Törle")
-group_8 <- c("Trou al'Wesse, Maisières-Canal, Les Cottés, Le Terme Pialat")
-group_9 <- c("Vogelherd, Hohle Fels, Geissenklösterle")
-group_10 <- c("Cellier, Blanchard, La Ferrassie, Hornos de la Pena, Otaslavice")
-group_11 <- c("La Souquette, Shelter Birów IV, Abri de Laussel, Abri du Poisson")
-group_table <- data_frame("Group" = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11"), "Site Names" = c(group_1, group_2, group_3, group_4, group_5, group_6, group_7, group_8, group_9, group_10, group_11))
+group_table <- read_excel("data/site-groups.xlsx")
 
@@
 
-signbase_sf_group <- 
-  st_as_sf(signbase_unique_groups, 
-           coords = c("longitude", "latitude"),
-           remove = FALSE,
-           crs = 4326)
 ggplot(Europe) +
   geom_sf() +
-  geom_sf(data = signbase_sf_group,
-          aes(colour = as.factor(group))) +
+  geom_sf(data = signbase_sf, aes(color = group)) +
   coord_sf(xlim = c(-10,30), 
@@
            expand = FALSE) +
-  geom_text_repel(data = signbase_sf_group,
+  geom_text_repel(data = signbase_sf,
                   aes(x = longitude ,
@@
                       label = site_name),
-                  max.overlaps = 100,
+                  max.overlaps = 30,
                   size = 2) +
   theme_minimal() +
-  facet_wrap(~group) +
-  labs(color = "Group Number")
+  facet_wrap(~time_period)
 
@@
 #| fig-cap: "Bar plot of signs by group"
-graph_data <- artifact_data %>% 
-  mutate(across(everything(), ~replace(., . > 1, 1))) %>% 
- # concentrate() %>% 
-  rownames_to_column("site_name") %>% 
-  left_join(group_df)
 
-sign_type_data <- graph_data %>% 
-  pivot_longer(cols= notch:hashtag,
-               names_to = "sign_type",
-               values_to = "count") %>% 
-  dplyr::filter(count != 0) %>% 
-  mutate(group = parse_number(group))
+full_data_long <- time_unique_data_full %>% 
+  pivot_longer(cols = line:star) %>% 
+  filter(value != 0)
 
-x_order <- graph_data %>% 
-  dplyr::select(-site_name,
-         -group) %>% 
-  colnames()
-
-ggplot(sign_type_data) +
-  aes(x = sign_type) +
+ggplot(full_data_long) +
+  aes(x = name, fill = group) +
   geom_bar() +
   rotate_x_text(angle = 90, hjust = NULL, vjust = NULL) + 
-  labs(x = x_order) +
-  scale_x_discrete(limits = c(x_order)) +
-  scale_y_continuous(labels = NULL) +
-  facet_wrap(~group, nrow = 11)
+  facet_wrap(~time_period, nrow = 4)
 
@@
 ##Modularity test
-library(igraph)
-set.seed(500)
-modul_matrix <-
-  graph_from_adjacency_matrix(1 - dm, mode = "undirected")
-mem <- as_membership(parse_number(signbase_unique_groups$group))
-as_group <- modularity(modul_matrix, membership = mem)
+trans_strength <- strength_function(trans_artifact_data, trans_unique_data, type = "table")
+proto_strength <- strength_function(proto_artifact_data, proto_unique_data, type = "table")
+early_strength <- strength_function(early_artifact_data, early_unique_data, type = "table")
+evolved_strength <- strength_function(evolved_artifact_data, evolved_unique_data, type = "table")
+
+full_table <- rbind(trans_strength, proto_strength, early_strength, evolved_strength) %>% 
+  mutate(time_period = c("transitional", "proto", "early", "evolved"), .before = 1)
+
+modularity_table <- full_table %>% 
+  dplyr::select(time_period, modularity)
+
+kable(modularity_table)
 ```
@@
 ##permanova test
-perman <- adonis2(jac~as.factor(signbase_unique_groups$group), 
-                  method = "jaccard",
-                  sqrt.dist = TRUE)
-perman <- perman %>% 
-  data_frame() %>% 
-  slice_head()
-perman_r <- perman$R2
-perman_p <- perman$`Pr(>F)`
+permanova_table <- full_table %>% 
+  select(time_period, permanova_r, permanova_p)
+
+kable(permanova_table)
 
@@
 
-set.seed(500)
-dispersion <- betadisper(jac, group = signbase_unique_groups$group)
-plot(dispersion, hull = FALSE, ellipse=TRUE)
+trans_strength_plot <- strength_function(trans_artifact_data, trans_unique_data, type = "dispersion")
+proto_strength_plot <- strength_function(proto_artifact_data, proto_unique_data, type = "dispersion")
+early_strength_plot <- strength_function(early_artifact_data, early_unique_data, type = "dispersion")
+evolved_strength_plot <- strength_function(evolved_artifact_data, evolved_unique_data, type = "dispersion")
 
+plot_grid(trans_strength_plot$, proto_strength_plot, early_strength_plot , evolved_strength_plot, 
+          ncol = 2, align = "hv", labels = c("Transitional", "Proto", "Early", "Evolved"), 
+          label_size = 7, label_colour = "black", label_y = 1.02, label_x = -0.03)
 ```
@@
 ## Mantel test for geography vs signs
-set.seed(200)
-site_distances <- dist(cbind(signbase_unique_groups$longitude, signbase_unique_groups$latitude))
-mantel_distance <- vegan::mantel(jac, site_distances, permutations = 1000)
-p_value_distance_correlation <- round(mantel_distance$signif, 3)
-r_value_distance_correlation <- round(mantel_distance$statistic, 3)
-```
+mantel_table <- full_table %>% 
+  dplyr::select(time_period, distance_mantel_r, distance_mantel_p)
 
-```{r}
-#| label: fig-distance-mantel
-#| fig-cap: "Mantel correlogram showing the results of a mantel test run between a geographic distance matrix and a dissimilarity matrix for site sign types"
-distance_corr <- mantel.correlog(jac, site_distances, n.class=50, break.pts=NULL, cutoff=FALSE, r.type="pearson", nperm=999, mult="holm", progressive=TRUE)
-plot(distance_corr)
+kable(mantel_table)
+
 ```

### 2025-03-03 Extra/time-periods.R
edits

@@ -1,3 +1,147 @@
+library(tidyverse)
+library(magrittr)
+library(rcarbon)
+library(readxl)
+signbase_full <- read_csv("data/signBase_Version1.0.csv")
+
+signbase_full_clean <- signbase_full %>%  # 511 rows
+  filter(site_name != "Willendorf",
+         site_name != "Riparo di Fontana Nuova",
+         site_name != "Muralovka",
+         site_name != "Shanidar Cave",
+         site_name != "Hayonim Cave",
+         site_name != "El Salitre",
+         site_name != "Grotte De La Princesse Pauline",
+         site_name != "Šandalja II") %>% 
+  dplyr::select(-other, -rectangle) %>% 
+  mutate(longitude = ifelse(site_name == "Riparo Bombrini", 7.437500, longitude)) %>% 
+  mutate(latitude = ifelse(site_name == "Riparo Bombrini", 43.77083, latitude))
+
+signbase_years <- signbase_full_clean %>%  # 446
+  drop_na(date_bp_max_min) %>% 
+  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+\\/\\-", "±")) %>% 
+  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+", "±")) %>% 
+  separate(date_bp_max_min,
+           sep = " - ",
+           c("date_bp_max", "date_bp_min"),
+           remove = FALSE) %>% 
+  mutate(date_bp_max = ifelse(str_detect(date_bp_max, "\\/"),
+                              str_extract(date_bp_max, ".*?(?=\\/)"),
+                              date_bp_max)) %>% 
+  mutate(date_bp_min = ifelse(str_detect(date_bp_min, "\\/"),
+                              str_extract(date_bp_min, ".*?(?=\\/)"),
+                              date_bp_min)) %>% 
+  separate(date_bp_max,
+           sep = "±",
+           c("date_bp_max_age",
+             "date_bp_max_error")) %>% 
+  separate(date_bp_min,
+           sep = "±",
+           c("date_bp_min_age",
+             "date_bp_min_error")) %>% 
+  drop_na(date_bp_max_age,
+          date_bp_max_error) %>% 
+  mutate(date_bp_max_age = parse_number(date_bp_max_age),
+         date_bp_max_error = parse_number(date_bp_max_error))
+
+signbase_years_cal <- 
+  rcarbon::calibrate(signbase_years$date_bp_max_age,
[…]

### 2025-03-04 endofquarterpaper.qmd
edits

@@
 The Aurignacian period occurred roughly 43,000 - 30,000 years BP, and is largely characterized by material culture indicators, as well as an ongoing biological shift from Neanderthals to early modern humans.
-Chronology of the Aurignacian is primarly determined by correspondance with those material culture markers, which include carinated scrapers, small-flake tool production, and split-based bone and ivory tools [@Chu_Richter_2020].
-The Aurignacian can be divided into phases, again mainly distinguished by shifts in material culture; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015].
+Chronology of the Aurignacian is primarly determined by correspondance with those material culture markers, which include carinated scrapers, small-flake tool and split-based bone and ivory tools [@Chu_Richter_2020].
+The Aurignacian can be divided into phases, again mainly distinguished by shifts in material culture; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015], as can be seen in @fig-aur-table.
 @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main periods, Aur-P1 (Proto/Early), whose typological observations corresponded with the time period of 43–37 kyr BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods kyr BP and 37-32 kyr BP.
@@
 
-<!-- add specific details about cultural markers, both for aurignacian and for each time period (create table in excel with name, date range, artifact indicators, environmental context, sources in the caption)-->
+```{r}
+#| label: fig-aur-table
+#| fig-cap: "Table of data on Aurignacian periods [@Banks_dErrico_Zilhao_2013;  @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021; @Sanz-Royo_Terlato_Marín-Arroyo_2024, @Chu_Richter_2020; @Fletcher_Sánchez],"
+library(knitr)
+library(tidyverse)
+library(readxl)
 
-```{r}
-library(knitr)
+aur_table <- read_excel("data/aurignacian-periods.xlsx")
 
-[@Jöris_Street_2008, @Banks_dErrico_Zilhao_2013, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021, @Sanz-Royo_Terlato_Marín-Arroyo_2024, @Chu_Richter_2020], 
+kable(aur_table)
+
 ```
@@
 What do the patterns tell us about dynamics of social organization throughout the Aurignacian?
-Firstly, we identify clusters within each time period using seriation and network analysis.
+First, we identify clusters within each time period using seriation and network analysis.
 We then test the strength of the clusters using modularity, perMANOVA, and PERMDISP2 tests.
-Lastly, we use mantel tests to evaluate the null hypothesis of isolation-by-distance.
+Lastly, we use mantel tests to evaluate the null hypothesis of isolation-by-distance for each time period.
 
@@
 ## clean the data
-library(tidyverse)
 library(magrittr)
 library(rcarbon)
-library(readxl)
 signbase_full <- read_csv("data/signBase_Version1.0.csv")
@@
 signbase_full_clean <- signbase_full_clean %>%
-  mutate(time_period = case_when((MedianBP > 42999) ~ "transitional",
-                                 (MedianBP > 39899) ~ "proto_aurignacian",
-                                 (MedianBP > 36999) ~ "early_aurignacian",
+  mutate(time_period = case_when((MedianBP > 41499) ~ "transitional",
+                                 (MedianBP > 39799) ~ "proto_aurignacian",
+                                 (MedianBP > 37899) ~ "early_aurignacian",
                                  (MedianBP > 31999) ~ "evolved_aurignacian"))
@@
 trans_unique_data <- trans_unique_data %>% 
-  mutate(group = case_when((site_name == "Pod Hradem") ~ "1",
-                           (site_name == "El Castillo" | site_name == "Hohle Fels") ~ "2"))
+  mutate(group = case_when((site_name == "Pod Hradem"|
+                              site_name == "Fumane" |
+                              site_name == "Labeko Koba") ~ "1",
+                           (site_name == "El Castillo" | 
+                              site_name == "Hohle Fels" | 
+                              site_name == "Geissenklösterle") ~ "2"))
 
 proto_unique_data <- proto_unique_data %>% 
-  mutate(group = case_when((site_name == "Labeko Koba" | 
-                              site_name == "Gatzarria" | 
+  mutate(group = case_when((site_name == "Gatzarria" | 
                               site_name == "Hohlenstein-Stadel" | 
                               site_name == "Mladeč" | 
-                              site_name == "Abri Pataud" | 
-                              site_name == "Fumane") ~ "1",
+                              site_name == "Abri Pataud") ~ "1",
                            (site_name == "Geissenklösterle" |
@@
                               site_name == "Hohle Fels"|
-                              site_name == "Trou Magrite") ~ "2"))
+                              site_name == "Trou Magrite"|
+                              site_name == "Vindija Cave") ~ "2"))
 
@@
 #| label: fig-site-map
-#| fig-cap: "Map of Palaeolithic sites containing signs, sign abundance is represented by point size, data from Dutkiewicz et al. (2020)"
-#| 
+#| fig-cap: "Map of data collection sites, divided by time period, and graph of )"
+
 library(rnaturalearth)
@@
 
-The modularity test, when run on the above groupings, resulted in a score of `r as_group`.
-The perMANOVA test produced an R-squared value of `r perman_r` and a P value of `r perman_p`, which tells us that the groupings are statistically significant, and that 52% of the variability in the sign distribution can be attributed to group membership.
+The modularity test, when run on the above groupings, resulted in a score of .
+The perMANOVA test produced an R-squared value of  and a P value of , which tells us that the groupings are statistically significant, and that 52% of the variability in the sign distribution can be attributed to group membership.
 The multivariate homogeneity test of group dispersions produced @fig-perm-disp, showing that the overlap between groups is fairly minimal.
@@
 
-plot_grid(trans_strength_plot$, proto_strength_plot, early_strength_plot , evolved_strength_plot, 
-          ncol = 2, align = "hv", labels = c("Transitional", "Proto", "Early", "Evolved"), 
-          label_size = 7, label_colour = "black", label_y = 1.02, label_x = -0.03)
 ```
@@
 
-In terms of geographic correlation, there is no significant statistical correlation between the geographic distance matrix and the sign type dissimilarity matrix (P = `r p_value_distance_correlation`, Mantel R = `r r_value_distance_correlation`).
-Again, the results are displayed in a mantel correlogram (@fig-distance-mantel), which demonstrates that similarity and geographic distance do not experience a clearly correlated relationship.
+In terms of geographic correlation, there is no significant statistical correlation between the geographic distance matrix and the sign type dissimilarity matrix (P = , Mantel R = ).
 
@@
 Through our various analyses, we saw that the isolation-by-distance framework could not account for the distribution of sign types.
-Specifically, our Mantel test (@fig-distance-mantel) found no statistically significant relationship between geographic distance and sign variation.
+Specifically, our Mantel test () found no statistically significant relationship between geographic distance and sign variation.
 Therefore, we must look to other factors to explain the distribution.
-We also saw distinct groupings in the data, as demonstrated in @fig-seriation, @fig-neighbor-net, and @fig-pcoa.
-We divided the sites into approximately ten groups based on sign types with a modularity score of `r as_group`.
-@fig-group-map shows groups with sites both clustered geographically speaking, and more distant from each other, potentially showing us cultural groups and patterns of movement.
+We also saw distinct groupings in the data, as demonstrated in @fig-seriation, @fig-neighbor-net, and.
+We divided the sites into approximately ten groups based on sign types with a modularity score of .
+
+shows groups with sites both clustered geographically speaking, and more distant from each other, potentially showing us cultural groups and patterns of movement.
 

### 2025-03-04 endofquarterpaper.qmd
edits

@@
 #| label: fig-aur-table
-#| fig-cap: "Table of data on Aurignacian periods [@Banks_dErrico_Zilhao_2013;  @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021; @Sanz-Royo_Terlato_Marín-Arroyo_2024, @Chu_Richter_2020; @Fletcher_Sánchez],"
+#| fig-cap: "Table of data on Aurignacian periods (Sources: @Banks_dErrico_Zilhao_2013;  @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021; @Chu_Richter_2020; @Fletcher_Sánchez),"
 library(knitr)
@@
 library(ggpubr)
+library(cowplot)
 
-ggplot(signbase_full_clean) +
+date_plot <- ggplot(signbase_full_clean) +
   aes(x = MedianBP, fill = time_period) +
-  geom_bar(width = 150)
+  geom_bar(width = 150) +
+  scale_x_reverse() +
+  scale_fill_discrete(labels = c("early", "evolved", "proto", "transitional"))
+  theme(legend.key.size = rel(0.1), axis.title = element_text(size = 0.25))
 
-time_unique_data_full$time_period <- factor(time_unique_data_full$time_period, 
-                                            levels = c("transitional", "proto_aurignacian", 
-                                                       "early_aurignacian", "evolved_aurignacian"))
 signbase_sf <- time_unique_data_full  %>% 
@@
 
-ggplot(Europe) +
+site_map <- ggplot(Europe) +
   geom_sf() +
@@
   facet_wrap(~time_period)
+
+plot_grid(date_plot, site_map, nrow = 2, scale = c(0.5, 1.5) )
 ```

### 2025-03-04 endofquarterpaper.qmd
edits

@@
 #| label: fig-aur-table
-#| fig-cap: "Table of data on Aurignacian periods (Sources: @Banks_dErrico_Zilhao_2013;  @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021; @Chu_Richter_2020; @Fletcher_Sánchez),"
+#| fig-cap: "Table of data on Aurignacian periods [@Banks_dErrico_Zilhao_2013;  @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021; @Chu_Richter_2020; @Fletcher_Sánchez]"
 library(knitr)
@@
                                             levels = c("transitional", "proto_aurignacian", 
-                                                       "early_aurignacian", "evolved_aurignacian"))
+                                                       "early_aurignacian", "evolved_aurignacian"), 
+                                            labels = c("transitional", "proto-aurignacian",
+                                                       "early aurignacian", "evolved aurignacian"))
 
@@
 #| label: fig-site-map
-#| fig-cap: "Map of data collection sites, divided by time period, and graph of )"
+#| fig-cap: "Graph of artifact count by time period and map of sites, divided by time period"
 
@@
   scale_x_reverse() +
-  scale_fill_discrete(labels = c("early", "evolved", "proto", "transitional"))
-  theme(legend.key.size = rel(0.1), axis.title = element_text(size = 0.25))
+  scale_fill_discrete(labels = c("early", "evolved", "proto", "transitional")) +
+  labs(fill = "Time period") +
+  theme(legend.key.size = unit(0.5, "cm"))
 
@@
            crs = 4326)
+
 world <- ne_countries(scale = "medium", returnclass = "sf")
+
 Europe <- world[which(world$continent == "Europe"),]
@@
   theme_minimal() +
-  facet_wrap(~time_period)
+  facet_wrap(~time_period) +
+  theme(axis.text = element_blank(),
+        axis.title = element_blank())
 
-plot_grid(date_plot, site_map, nrow = 2, scale = c(0.5, 1.5) )
+plot_grid(date_plot, site_map, nrow = 2, align = "hv", rel_heights = c(1, 2))
 ```
@@
 
-These exclusions resulted in `r nartifacts_clean` artifacts found at `r nsites_clean` sites in `r ncountries_clean` countries, encapsulating `r nsigntypes_clean` sign types, as seen in @fig-site-map.
+These exclusions resulted in `r nartifacts_clean` artifacts found at `r nsites_clean` sites in `r ncountries_clean` countries, encapsulating `r nsigntypes_clean` sign types.
 All our tests were run with presence/absence data, in order to keep the data from being affected by outliers like Vogelherd.
-We ran a Mantel test between an abundance distance matrix and and a Euclidean distance matrix in order to make sure that the exclusion of abundance would not significantly impact our results, and found that it was not statistically significant [@Lycett_2019] (P = `r p_value_abundance_correlation`, Mantel R = `r r_value_abundance_correlation`).
+We ran a Mantel test between an abundance distance matrix and and a Euclidean distance matrix in order to make sure that the exclusion of abundance would not significantly impact our results, and found that it was not quite statistically significant [@Lycett_2019] (P = `r p_value_abundance_correlation`, Mantel R = `r r_value_abundance_correlation`).
+
+Maximum and minimum BP dates were provided by Signbase, and calibration was done through the rcarbon package in RStudio.
 The calibrated radiocarbon dates of the artifacts range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years.
-We divided the data up into four periods:
-
-Dates provided by Signbase.
+We divided the data up into four periods in accordance with the date ranges provided in @fig-aur-table.
 
@@
 Assuming that this framework is true, we would hypothetically see patterns of cultural features of Upper Paleolithic early modern humans that are dispersed according to geography.
-Conversely, if geographic distance cannot account for the distribution of the sign types, we would have to assume that other factors are at play, namely, movement of groups or the creation of social networks.
-
-Mantel tests are used here to determine whether there is a correlation between sign distribution and geography.
+Mantel tests are used here for each time period to determine the correlation between sign distribution and geography.
 The Mantel test is run between a geographic distance matrix and a Jaccard dissimilarity matrix.
@@
     theme(
-      axis.text = element_text(size = rel(0.5)),
+      axis.text = element_text(size = rel(1)),
       axis.ticks = element_blank(),
       legend.position = "none",
-      plot.title = element_text(hjust = 0.25)
+      plot.title = element_text(hjust = 1)
     ) +
@@
               by = "site_name")
-  countries <- x_df$country
+  countries <- x_df$group
   unique_countries <- unique(countries)
@@
   final_table <- data_frame("modularity" = as_group, 
-                            "permanova_r" = perman_r,
-                            "permanova_p" = perman_p,
-                            "distance_mantel_r" = r_value_distance_correlation, 
-                            "distance_mantel_p" = p_value_distance_correlation)
+                            "permanova r" = perman_r,
+                            "permanova p" = perman_p,
+                            "mantel r" = r_value_distance_correlation, 
+                            "mantel p" = p_value_distance_correlation)
 
@@
 ### Seriation
-```{r}
+```{r fig.height= 9, fig.width= 13}
 #| label: fig-seriation
 #| fig-cap: "Seriation analysis of the data"
-library(cowplot)
 
@@
           ncol = 2, align = "hv", 
-          labels = c("Transitional", "Proto", "Early", "Evolved"), label_size = 10)
+          labels = c("Transitional", "Proto", "Early", "Evolved"), label_size = 15, 
+          scale = c(0.85, 1, 1, 1))
 
@@
 ### Network Analysis
-```{r}
+```{r fig.height= 5.5, fig.width= 8}
 #| label: fig-neighbor-net
-#| fig-cap: "Neighbor-Net analysis of the data"
+#| fig-cap: "Network analysis of the data"
 network_trans <- trans_network(trans_artifact_data)
@@
           ncol = 2, align = "hv", labels = c("Transitional", "Proto", "Early", "Evolved"), 
-          label_size = 7, label_colour = "black", label_y = 1.02, label_x = -0.03)
+          label_size = 7, label_colour = "black", label_y = 1.02, label_x = -0.01)
 ```
 
-Combining the results of the above tests allowed us to find 11 groups of sites, as shown in @fig-group-table.
+Combining the results of the above tests allowed us to find two groups of sites within each time period, as shown in @fig-group-table.
 The geographic distributions of the groups are shown in @fig-group-map, and the sign makeup of each group is shown in @fig-group-sign.
@@
 
-```{r fig.height= 5, fig.width=10}
+```{r fig.height= 7, fig.width=10}
 #| label: fig-group-map
-#| fig-cap: "Map of sites faceted and colored by group number"
+#| fig-cap: "Map of sites faceted by time period and colored by group number"
 
@@
   theme_minimal() +
-  facet_wrap(~time_period)
+  facet_wrap(~time_period) +
+  theme(axis.text = element_blank(),
+        axis.title = element_blank())
 
@@
 #| label: fig-group-sign
-#| fig-cap: "Bar plot of signs by group"
+#| fig-cap: "Bar plot of signs by time period, colored by group"
 
-full_data_long <- time_unique_data_full %>% 
+full_data_long <- signbase_sf %>% 
   pivot_longer(cols = line:star) %>% 
@@
 full_table <- rbind(trans_strength, proto_strength, early_strength, evolved_strength) %>% 
-  mutate(time_period = c("transitional", "proto", "early", "evolved"), .before = 1)
+  mutate(`time period` = c("transitional", "proto", "early", "evolved"), .before = 1)
 
 modularity_table <- full_table %>% 
-  dplyr::select(time_period, modularity)
+  dplyr::select(`time period`, modularity)
 
@@
 permanova_table <- full_table %>% 
-  select(time_period, permanova_r, permanova_p)
+  dplyr::select(`time period`, `permanova r`, `permanova p`)
 
 kable(permanova_table)
-
 ```
 
-The modularity test, when run on the above groupings, resulted in a score of .
-The perMANOVA test produced an R-squared value of  and a P value of , which tells us that the groupings are statistically significant, and that 52% of the variability in the sign distribution can be attributed to group membership.
-The multivariate homogeneity test of group dispersions produced @fig-perm-disp, showing that the overlap between groups is fairly minimal.
+The r value from the perMANOVA tests tell us that for each of the time periods, around 20% of the variation in sign content can be explained by group membership, with the proto-aurignacian having the highest score of 27%, and the early aurignacian having the lowest score of 21%.
+The perMANOVA p value quantifies the randomness of the groupings, with 0.05 as the threshold for definitive non-randomness.
+The only time period to not meet that threshold is the transitional period, with a p-score of 0.3.
 
@@
 mantel_table <- full_table %>% 
-  dplyr::select(time_period, distance_mantel_r, distance_mantel_p)
+  dplyr::select(`time period`, `mantel r`, `mantel p`)
 

### 2025-03-05 chron-mantel.R
new mantel?

@@ -1,0 +1,13 @@
+chron_df <- signbase_full_clean %>% 
+  column_to_rownames("id") %>% 
+  dplyr::select(MedianBP) %>% 
+  dist()
+
+chron_sign_df <- signbase_full_clean %>% 
+  column_to_rownames("id") %>% 
+  dplyr::select(line:star) %>% 
+  vegdist("jaccard")
+
+chron_mantel <- vegan::mantel(chron_df, chron_sign_df, permutations = 1000)
+chron_mantel_plot <- mantel.correlog(chron_sign_df, chron_df, n.class=50, break.pts=NULL, cutoff=FALSE, r.type="pearson", nperm=999, mult="holm", progressive=TRUE)
+plot(chron_mantel_plot)

### 2025-03-05 endofquarterpaper.qmd
new mantel?
new edits

@@
 ---
-title: "Geometric Signs on Paleolithic Mobile Objects as Cultural Group Markers Across Aurignacian Time Periods"
+title: "Changes in Groupings of Geometric Signs on Paleolithic Mobile Objects as a Model of Social Dynamics Across Aurignacian Time Periods"
 author: Sophie Cooper
@@
 ---
+
 # Introduction and Background
+
 The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020] coupled with biological evolution.
@@
 Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence, @Sauvet_2018, @Kuhn_Stiner_2007, @Vanhaeren_d].
-Considered to be the starting point of the Upper Paleolithic, the Aurignacian period (ca.
+Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca.
 43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].
@@
 Social organization dynamics have previously been explored using cluster analysis methods upon variations in artefacts to represent and understand patterns of cultural groups.
-For example, multivariate analyses of variations in Gravettian beads suggests an east to west cline of nine cultural groups [@baker2024evidence].
+For example, multivariate analyses of variations in beads from the Gravettian era suggests an east to west cline of nine cultural groups [@baker2024evidence].
 Seriation analysis, PCoA, and network analysis were used to identify the groups from the bead data, which consisted of 134 discrete bead variations from both burial and occupation sites, as distinguished by shape and material.
@@
 
-In another study to find cultural groups, similar analyses were performed upon personal ornament types from the Aurignacian period, with the variation in types serving as a proxy for linguistic diversity [@Vanhaeren_d].
-Seriation, correspondence, and GIS analyses were performed upon a personal ornament data set, which consisted of 157 distinct ornament types from Eastern Aurignacian sites.
+Similar analyses were performed upon personal ornament types from the Aurignacian technocomplex, with the variation in types serving as a proxy for linguistic diversity [@Vanhaeren_d].
+Seriation, correspondence, and GIS analyses were used to identify groups in a personal ornament data set consisting of 157 distinct ornament types from Eastern Aurignacian sites.
 The analyses identified fourteen geographically cohesive groups sharing similar ornament type associations, which sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.
@@
 
-Additionally, @Sauvet_2018 used quadrilateral geometric parietal signs to identify Upper Paleolithic cultural groups in the Cantabrian region of Spain.
-The authors identified and classified the signs based on variations in form, then performed a factorial correspondence analysis and ascending hierarchical clustering upon the data.
+Additionally, @Sauvet_2018 performed quadrilateral geometric parietal signs to identify Upper Paleolithic cultural groups in the Cantabrian region of Spain.
+Factorial correspondence analysis and ascending hierarchical clustering based on 45 identified morphological characteristics were used to divide the signs into groups.
 Two clusters were identified from those methods, representing two distinct cultural groups.
@@
 
-Following the work done by @baker2024evidence, @Vanhaeren_d, and @Sauvet_2018, in using variation in ornamentation and signs to identify cultural groups, we look to see how geometric sign design changes across the sub-periods of the Aurignacian period.
+Following the work done by @baker2024evidence, @Vanhaeren_d, and @Sauvet_2018, in using variation in ornamentation and signs to identify cultural groups, we look to see how geometric sign design changes across the phases of the Aurignacian.
 The Aurignacian period occurred roughly 43,000 - 30,000 years BP, and is largely characterized by material culture indicators, as well as an ongoing biological shift from Neanderthals to early modern humans.
-Chronology of the Aurignacian is primarly determined by correspondance with those material culture markers, which include carinated scrapers, small-flake tool and split-based bone and ivory tools [@Chu_Richter_2020].
+Chronology of the Aurignacian is primarily determined by correspondence with those material culture markers, which include carinated scrapers, small-flake tool and split-based bone and ivory tools [@Chu_Richter_2020].
 The Aurignacian can be divided into phases, again mainly distinguished by shifts in material culture; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015], as can be seen in @fig-aur-table.
-@Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main periods, Aur-P1 (Proto/Early), whose typological observations corresponded with the time period of 43–37 kyr BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods kyr BP and 37-32 kyr BP.
+@Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main phases, Aur-P1 (Proto/Early), whose typological observations corresponded with the time period of 43–37 kyr BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods kyr BP and 37-32 kyr BP.
 @Banks_dErrico_Zilhao_2013 used Bayesian modeling methods to further divide Aur-P1 into the Proto and Early Aurignacian, placing the Proto-Aurignacian at 41.5 - 39.9 k cal BP, and the early Aurignacian at 39.8 - 37.8 k cal BP.
@@
                                  (MedianBP > 39799) ~ "proto_aurignacian",
-                                 (MedianBP > 37899) ~ "early_aurignacian",
+                                 (MedianBP > 37799) ~ "early_aurignacian",
                                  (MedianBP > 31999) ~ "evolved_aurignacian"))
@@
 ```{r}
+
 lat_long_df <- signbase_full_clean %>% 
@@
 
+signbase_full_clean$time_period <- factor(signbase_full_clean$time_period, 
+                                            levels = c("transitional", "proto_aurignacian", 
+                                                       "early_aurignacian", "evolved_aurignacian"), 
+                                            labels = c("transitional", "proto-aurignacian",
+                                                       "early aurignacian", "evolved aurignacian"))
+
 date_plot <- ggplot(signbase_full_clean) +
@@
   scale_x_reverse() +
-  scale_fill_discrete(labels = c("early", "evolved", "proto", "transitional")) +
   labs(fill = "Time period") +
@@
 Our data for this paper comes from SignBase, an open-access database collection of geometric signs on mobile objects from Paleolithic-era Europe, which was launched in 2020, with the explicit goal being to enable quantitative analyses of abstract geographical representation [@dutkiewicz2020].
-The creators of the database identified around 30 distinct geometric sign types, and recorded their occurrences on mobile artefacts from the Paleolithic.
+The creators of the database identified distinct geometric sign types, and recorded their occurrences on mobile artefacts from the Paleolithic.
 Identification and categorization of the sign types had already been performed by the creators of the SignBase database, the methods for which are described in @dutkiewicz2020, and we adhered to those categories and placements.
 
-The data set consists of `r nartifacts` artifacts found at `r nsites` sites in `r ncountries` countries.
-The database records `r nsigntypes` geometric sign type found on these artifacts.
+The data set consists of `r nartifacts` artifacts found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign type found on these artifacts.
 We excluded signs if they fell under the "other" category, as the wide range of possibilities that could qualify as "other" means that it would be difficult to establish a definitive connection.
 We also excluded all data from Willendorf (only one sign present), and Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave (due to their extreme geographic distance from the other sites).
-After those sites were excluded, all signs under the "rectangle" category were also removed as a side effect of that exclusion.
 Lastly, data from Grotte De La Princesse Pauline and Šandalja II was excluded, due to being dated much earlier than the rest of the artifacts.
@@
 All our tests were run with presence/absence data, in order to keep the data from being affected by outliers like Vogelherd.
-We ran a Mantel test between an abundance distance matrix and and a Euclidean distance matrix in order to make sure that the exclusion of abundance would not significantly impact our results, and found that it was not quite statistically significant [@Lycett_2019] (P = `r p_value_abundance_correlation`, Mantel R = `r r_value_abundance_correlation`).
+We ran a Mantel test between an abundance distance matrix and and a Euclidean distance matrix in order to make sure that the exclusion of abundance would not significantly impact our results [@Lycett_2019], and found that it was not quite statistically significant (P = `r p_value_abundance_correlation`, Mantel R = `r r_value_abundance_correlation`).
 
-Maximum and minimum BP dates were provided by Signbase, and calibration was done through the rcarbon package in RStudio.
+Maximum and minimum BP dates were provided by Signbase, and calibration was done through the rcarbon package in RStudio [@Crema_Bevan_2021].
 The calibrated radiocarbon dates of the artifacts range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years.
@@
 The Mantel test is run between a geographic distance matrix and a Jaccard dissimilarity matrix.
-The test seeks to determine the level of correspondence between the matrices through permuational evaluation of the null distribution [@Mantel_Smouse], producing a p-score as a result.
+The test seeks to determine the level of correspondence between the matrices through permuational evaluation of the null distribution [@Mantel_Smouse], producing a Mantel-R score and a p-score as a result.
 
@@
 ## Testing for Cultural Groupings
+
 ```{r}
@@
 ### Seriation
+
 ```{r fig.height= 9, fig.width= 13}
@@
 ### Network Analysis
+
 ```{r fig.height= 5.5, fig.width= 8}
@@
 
-Combining the results of the above tests allowed us to find two groups of sites within each time period, as shown in @fig-group-table.
+Combining the results of the above tests allowed us to find two main groups of sites within each time period, as shown in @fig-group-table.
 The geographic distributions of the groups are shown in @fig-group-map, and the sign makeup of each group is shown in @fig-group-sign.
@@
 
-The r value from the perMANOVA tests tell us that for each of the time periods, around 20% of the variation in sign content can be explained by group membership, with the proto-aurignacian having the highest score of 27%, and the early aurignacian having the lowest score of 21%.
+The R value from the perMANOVA tests tell us that for each of the time periods, around 20% of the variation in sign content can be explained by group membership, with the proto-aurignacian having the highest score of 27%, and the early Aurignacian having the lowest score of 21%.
 The perMANOVA p value quantifies the randomness of the groupings, with 0.05 as the threshold for definitive non-randomness.
@@
 
-In terms of geographic correlation, there is no significant statistical correlation between the geographic distance matrix and the sign type dissimilarity matrix (P = , Mantel R = ).
+The relevance of the isolation-by-distance hypothesis also varies from period to period.
 
@@
 
-In this paper, we set out to ask if identifiable and relevant patterns existed in the data, and if so, what could they tell us about the patterns of diversity and movement among Upper Paleolithic humans?
-
-Through our various analyses, we saw that the isolation-by-distance framework could not account for the distribution of sign types.
-Specifically, our Mantel test () found no statistically significant relationship between geographic distance and sign variation.
-Therefore, we must look to other factors to explain the distribution.
-We also saw distinct groupings in the data, as demonstrated in @fig-seriation, @fig-neighbor-net, and.
-We divided the sites into approximately ten groups based on sign types with a modularity score of .
-
-shows groups with sites both clustered geographically speaking, and more distant from each other, potentially showing us cultural groups and patterns of movement.
+In this paper, we set out to ask what differences in the patterns of groupings among geometric sign types found on mobile objects between phases of the Aurignacian could tell us about the social dynamics of that period.
 
@@
 # Conclusion
+
+# References

### 2025-03-05 endofquarterpaper.qmd
intro edits

@@
 
-# Introduction and Background
+# Introduction
 
@@
 Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca.
-43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].
-In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe across the phases of the Aurignacian period.
+43 - 30 k BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020].
+In this paper, we investugate the social organization dynamics by analyzing the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe across four phases of the Aurignacian period.
 
-Social organization dynamics have previously been explored using cluster analysis methods upon variations in artefacts to represent and understand patterns of cultural groups.
-For example, multivariate analyses of variations in beads from the Gravettian era suggests an east to west cline of nine cultural groups [@baker2024evidence].
-Seriation analysis, PCoA, and network analysis were used to identify the groups from the bead data, which consisted of 134 discrete bead variations from both burial and occupation sites, as distinguished by shape and material.
-The geographically discrete groups identified encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and Western Mediterranean regions.
-The isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance, was tested using a Mantel test, finding that geographic distance cannot solely account for the bead distribution, proving the influence of cultural and social dynamics on the groups found.
+Social organization dynamics of the Upper Paleolithic have previously been explored using statistical analyses of artifacts to understand patterns of cultural groups.
+Multivariate analyses of beads from the Gravettian technocomplex suggests an east to west cline of nine cultural groups [@baker2024evidence].
+Seriation analysis, PCoA, and network analysis were used to identify the groups from 134 discrete bead types recovered from both burial and occupation sites.
+To validate their claims of cultural groups, @baker2024evidence used a Mantel test to evaluate an isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance, finding that geographic distance alone did not solely account for the bead distribution.
+The nine geographically discrete groups encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and western Mediterranean regions.
 
-Similar analyses were performed upon personal ornament types from the Aurignacian technocomplex, with the variation in types serving as a proxy for linguistic diversity [@Vanhaeren_d].
-Seriation, correspondence, and GIS analyses were used to identify groups in a personal ornament data set consisting of 157 distinct ornament types from Eastern Aurignacian sites.
-The analyses identified fourteen geographically cohesive groups sharing similar ornament type associations, which sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.
-The possibility that the observed groups were influenced by availability of raw materials was considered, and ultimately discarded as an explanation for the patterns.
+Similar analyses of personal ornament types from the Aurignacian technocomplex identified fourteen geographically cohesive groups [@Vanhaeren_d].
+Drawing on ethnographic studies that show how body decoration and ornamentation indicates ethno-linguistic identity, @Vanhaeren_d used Aurignacian personal ornamentation as a proxy for ethno-linguistic diversity.
+They analyzed 157 distinct ornament types from 98 Aurignacian sites in Europe and the Near East.
+Geographically cohesive groups sharing similar ornament type associations were identified by seriation, correspondence, and GIS analyses.
+These groups sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.
+@Vanhaeren_d interpret the variation in personal ornament types as evidence of long-lasting cultural differences and as robustly establishing the ethno-linguistic diversity of the Aurignacian.
 
-Additionally, @Sauvet_2018 performed quadrilateral geometric parietal signs to identify Upper Paleolithic cultural groups in the Cantabrian region of Spain.
-Factorial correspondence analysis and ascending hierarchical clustering based on 45 identified morphological characteristics were used to divide the signs into groups.
-Two clusters were identified from those methods, representing two distinct cultural groups.
-@Sauvet_2018 makes a case for the relevance of geometric signs as identity markers in identifying cultural groups, pointing to the relevant qualities of signs, as well as historical and ethnographic case studies of signs serving as markers of group membership case studies.
+Cultural difference has also been identified from geometric signs in Upper Paleolithic parietal art from the Cantabrian region of Spain [@Sauvet_2018].
+Forty-two quadrilateral geometric signs from four cave sites, likely associated with the Magdalenian period, spanning a distance of 30 kilometers were recorded.
+Multivariate analysis of 45 morphological characteristics of the signs revealed two clusters.
+To interpret these clusters, @Sauvet_2018 draw on case studies of geometric signs as identity markers, including South African engraved ostrich shells from 60 k BP, maker's marks on Gallo-Roman pottery, and 19th century French craft guild members' personal marks.
+Using these case studies as analogies, @Sauvet_2018 interpret the Magdalenian geometric signs as identity markers, and the two clusters as representative of cultural groups.
 
-Following the work done by @baker2024evidence, @Vanhaeren_d, and @Sauvet_2018, in using variation in ornamentation and signs to identify cultural groups, we look to see how geometric sign design changes across the phases of the Aurignacian.
-The Aurignacian period occurred roughly 43,000 - 30,000 years BP, and is largely characterized by material culture indicators, as well as an ongoing biological shift from Neanderthals to early modern humans.
-Chronology of the Aurignacian is primarily determined by correspondence with those material culture markers, which include carinated scrapers, small-flake tool and split-based bone and ivory tools [@Chu_Richter_2020].
+In this study, we extend the use of geometric signs as identity markers to investigate cultural groups in the Aurignacian period.
+The Aurignacian period spans roughly 43 - 30 k BP, and is largely characterized by material culture indicators, such as carinated scrapers, small-flake tools and split-based bone and ivory tools, as well as an ongoing evolutionary shift from Neanderthals to early modern humans [@Chu_Richter_2020].
 The Aurignacian can be divided into phases, again mainly distinguished by shifts in material culture; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015], as can be seen in @fig-aur-table.
-@Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main phases, Aur-P1 (Proto/Early), whose typological observations corresponded with the time period of 43–37 kyr BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods kyr BP and 37-32 kyr BP.
-@Banks_dErrico_Zilhao_2013 used Bayesian modeling methods to further divide Aur-P1 into the Proto and Early Aurignacian, placing the Proto-Aurignacian at 41.5 - 39.9 k cal BP, and the early Aurignacian at 39.8 - 37.8 k cal BP.
-They found that the Proto-Aurignacian occurs during the Greenland Interstadials 9 and 10, periods of climatic amelioration, while the Early Aurignacian began with the Henrich Stadial, which is characterized by dry and arid conditions, and ended with Greenland Interstadial 8.
+@Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main phases, Aur-P1 (Proto/Early), which corresponds with the time period of 43–37 k BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods 37-32 k BP.
+@Banks_dErrico_Zilhao_2013 used Bayesian modeling methods to further divide Aur-P1 into the Proto and Early Aurignacian, placing the Proto-Aurignacian at 41.5 - 39.9 k BP, and the early Aurignacian at 39.8 - 37.8 k BP.
+They found that the Proto-Aurignacian occurs during the Greenland Interstadials (GI) 9 and 10, periods of climatic amelioration, while the Early Aurignacian began with the Henrich Stadial, which is characterized by dry and arid conditions, and ends with GI 8.
 
@@
 
-In this study, we aim to address the following questions: How does sign distribution change across phases of the Aurignacian?
-How do the groupings in sign distribution vary in diversity, size, connectivity, and boundaries?
-What do the patterns tell us about dynamics of social organization throughout the Aurignacian?
-First, we identify clusters within each time period using seriation and network analysis.
+We hypothesize that these periods, and the environmental changes across the Aurignacian, may correspond to changes in social dynamics of cultural groups, as represented by patterns in the distribution of geometric signs.
+We aim to address the following questions: How do grouping dynamics of geometric signs change across the four phases of the Aurignacian?
+How do the groupings vary in diversity, size, connectivity, and boundaries?
+Lastly, what do the patterns tell us about dynamics of social organization throughout the Aurignacian?
+We answer these questions by first identifying clusters within each time period using seriation and network analyses.
 We then test the strength of the clusters using modularity, perMANOVA, and PERMDISP2 tests.
-Lastly, we use mantel tests to evaluate the null hypothesis of isolation-by-distance for each time period.
+Lastly, we use Mantel tests to evaluate the null hypothesis of isolation-by-distance for each time period.
 

### 2025-03-06 endofquarterpaper.qmd
edits

@@
 ---
-title: "Changes in Groupings of Geometric Signs on Paleolithic Mobile Objects as a Model of Social Dynamics Across Aurignacian Time Periods"
+title: "Changes in Groupings of Geometric Signs as a Model of Social Dynamics Across Aurignacian Periods"
 author: Sophie Cooper
@@
 r_value_abundance_correlation <- round(euc_mantel$statistic, 3)
+
+##mantel chronological
+chron_df <- signbase_full_clean %>% 
+  column_to_rownames("id") %>% 
+  dplyr::select(MedianBP) %>% 
+  dist( method = "euclidean")
+
+chron_sign_df <- signbase_full_clean %>% 
+  column_to_rownames("id") %>% 
+  dplyr::select(line:star) %>% 
+  vegdist("jaccard")
+
+chron_mantel <- vegan::mantel(chron_df, chron_sign_df, permutations = 1000)
+
+chron_p <- round(chron_mantel$signif, 3)
+chron_r <- round(chron_mantel$statistic, 3)
+
 ```
@@
 We also excluded all data from Willendorf (only one sign present), and Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave (due to their extreme geographic distance from the other sites).
-Lastly, data from Grotte De La Princesse Pauline and Šandalja II was excluded, due to being dated much earlier than the rest of the artifacts.
-
-These exclusions resulted in `r nartifacts_clean` artifacts found at `r nsites_clean` sites in `r ncountries_clean` countries, encapsulating `r nsigntypes_clean` sign types.
-All our tests were run with presence/absence data, in order to keep the data from being affected by outliers like Vogelherd.
-We ran a Mantel test between an abundance distance matrix and and a Euclidean distance matrix in order to make sure that the exclusion of abundance would not significantly impact our results [@Lycett_2019], and found that it was not quite statistically significant (P = `r p_value_abundance_correlation`, Mantel R = `r r_value_abundance_correlation`).
+Lastly, data from Grotte De La Princesse Pauline and Šandalja II was excluded, due to being dated much earlier than the rest of the artifacts.These exclusions resulted in `r nartifacts_clean` artifacts found at `r nsites_clean` sites in `r ncountries_clean` countries, encapsulating `r nsigntypes_clean` sign types.
 
@@
 We divided the data up into four periods in accordance with the date ranges provided in @fig-aur-table.
+
+All our tests were run with presence/absence data, in order to keep the data from being affected by outliers like Vogelherd.
+We ran a Mantel test between an Euclidean abundance distance matrix and a jaccard distance matrix of the sign types in order to make sure that the exclusion of abundance would not significantly impact our results [@Lycett_2019], and found that it was not quite statistically significant (P = `r p_value_abundance_correlation`, Mantel R = `r r_value_abundance_correlation`).
+Another Mantel test was run between a euclidean distance matrix for MedianBP and the jaccard sign type matrix in order to determine whether there was a definitive correlation between distance in time and sign types.
+The Mantel test found that there was a statistically significant correlation (p = `r chron_p`, Mantel-R = `r chron_r`).
 
@@
 
-  x_df <- artifact_data_unique %>% 
-    left_join(signbase_full_clean,
-              by = "site_name")
-  countries <- x_df$group
+  countries <- artifact_data_unique$group
   unique_countries <- unique(countries)
@@
 
-strength_function <- function(artifact_data, group_data, type = "table"){
+strength_function <- function(artifact_data, group_data){
   artifact_data <- artifact_data %>% 
@@
   dm <- as.matrix(jac)
-
-  if(type == "table"){
   modul_matrix <-
@@
 
-  return(final_table)}
-
-  if(type == "dispersion"){
-    set.seed(500)
-    dispersion <- betadisper(jac, group = group_data$group)
-    disp_plot<- plot(dispersion, hull = FALSE, ellipse=TRUE)
-    return(disp_plot)}
+  return(final_table)
 }
@@
 ##Modularity test
-trans_strength <- strength_function(trans_artifact_data, trans_unique_data, type = "table")
-proto_strength <- strength_function(proto_artifact_data, proto_unique_data, type = "table")
-early_strength <- strength_function(early_artifact_data, early_unique_data, type = "table")
-evolved_strength <- strength_function(evolved_artifact_data, evolved_unique_data, type = "table")
+trans_strength <- strength_function(trans_artifact_data, trans_unique_data)
+proto_strength <- strength_function(proto_artifact_data, proto_unique_data)
+early_strength <- strength_function(early_artifact_data, early_unique_data)
+evolved_strength <- strength_function(evolved_artifact_data, evolved_unique_data)
 
@@
 
-modularity_table <- full_table %>% 
-  dplyr::select(`time period`, modularity)
+strength_table <- full_table %>% 
+  dplyr::select(`time period`, modularity, `permanova r`, `permanova p`)
 
-kable(modularity_table)
-```
-
-```{r}
-##permanova test
-permanova_table <- full_table %>% 
-  dplyr::select(`time period`, `permanova r`, `permanova p`)
-
-kable(permanova_table)
+kable(full_table)
 ```
@@
 The only time period to not meet that threshold is the transitional period, with a p-score of 0.3.
-
-```{r fig.height= 8, fig.width=5}
-#| label: fig-perm-disp
-#| fig-cap: "Multivariate homogeneity of group dispersions"
-
-trans_strength_plot <- strength_function(trans_artifact_data, trans_unique_data, type = "dispersion")
-proto_strength_plot <- strength_function(proto_artifact_data, proto_unique_data, type = "dispersion")
-early_strength_plot <- strength_function(early_artifact_data, early_unique_data, type = "dispersion")
-evolved_strength_plot <- strength_function(evolved_artifact_data, evolved_unique_data, type = "dispersion")
-
-```
 

### 2025-03-06 chron-mantel.R

@@ -1,13 +1,0 @@
-chron_df <- signbase_full_clean %>% 
-  column_to_rownames("id") %>% 
-  dplyr::select(MedianBP) %>% 
-  dist()
-
-chron_sign_df <- signbase_full_clean %>% 
-  column_to_rownames("id") %>% 
-  dplyr::select(line:star) %>% 
-  vegdist("jaccard")
-
-chron_mantel <- vegan::mantel(chron_df, chron_sign_df, permutations = 1000)
-chron_mantel_plot <- mantel.correlog(chron_sign_df, chron_df, n.class=50, break.pts=NULL, cutoff=FALSE, r.type="pearson", nperm=999, mult="holm", progressive=TRUE)
-plot(chron_mantel_plot)

### 2025-03-08 endofquarterpaper.qmd
edited results section

@@
 ---
-
 # Introduction
-
 The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020] coupled with biological evolution.
@@
 We answer these questions by first identifying clusters within each time period using seriation and network analyses.
-We then test the strength of the clusters using modularity, perMANOVA, and PERMDISP2 tests.
+We then test the strength of the clusters using modularity and perMANOVA tests.
 Lastly, we use Mantel tests to evaluate the null hypothesis of isolation-by-distance for each time period.
@@
                               site_name == "Fumane" |
-                              site_name == "Labeko Koba") ~ "1",
+                              site_name == "Geissenklösterle") ~ "1",
                            (site_name == "El Castillo" | 
                               site_name == "Hohle Fels" | 
-                              site_name == "Geissenklösterle") ~ "2"))
+                              site_name == "Labeko Koba") ~ "2"))
 
@@
                               site_name == "Spy" |
-                              site_name == "La Ferrassie" |
+                              site_name == "La Ferrassie"|
                                site_name == "Les Cottés") ~ "2"))
@@
                               site_name == "La Viña"|
-                              site_name ==  "Sirgenstein Cave") ~ "1",
+                              site_name ==  "Sirgenstein Cave"|
+                              site_name == "Vindija Cave") ~ "1",
                            (site_name == "Bockstein-Törle"|
@@
                               site_name == "Hohle Fels"|
-                              site_name == "Trou Magrite"|
-                              site_name == "Vindija Cave") ~ "2"))
+                              site_name == "Trou Magrite") ~ "2"))
 
@@
                                                        "early_aurignacian", "evolved_aurignacian"), 
-                                            labels = c("transitional", "proto-aurignacian",
-                                                       "early aurignacian", "evolved aurignacian"))
+                                            labels = c("Transitional", "Proto-Aurignacian",
+                                                       "Early Aurignacian", "Evolved Aurignacian"))
 
@@
 
+Our data for this paper comes from SignBase, an open-access database collection of geometric signs on mobile objects from Paleolithic-era Europe, which was launched in 2020, with the explicit goal being to enable quantitative analyses of abstract geographical representation [@dutkiewicz2020].
+The creators of the database identified distinct geometric sign types, and recorded their occurrences on mobile artefacts from the Paleolithic.
+Identification and categorization of the sign types had already been performed by the creators of the SignBase database, the methods for which are described in @dutkiewicz2020, and we adhered to those categories and placements.
+Maximum and minimum BP dates were also provided by Signbase, and calibration was done through the rcarbon package in RStudio [@Crema_Bevan_2021].
+
+The data set consists of `r nartifacts` artifacts found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign type found on these artifacts.
+We excluded signs if they fell under the "other" category, as the wide range of potential possibilities means that it would be difficult to establish a definitive connection.
+We also excluded all data from Willendorf (only one sign present), and Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave (due to their extreme geographic distance from the other sites).
+Lastly, data from Grotte De La Princesse Pauline and Šandalja II was excluded, due to being dated much earlier than the rest of the artifacts.These exclusions resulted in `r nartifacts_clean` artifacts found at `r nsites_clean` sites in `r ncountries_clean` countries, encapsulating `r nsigntypes_clean` sign types.
+The calibrated radiocarbon dates of the cleaned data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years.
+We divided the data up into four periods in accordance with the date ranges provided in @fig-aur-table.
+
 ```{r fig.height= 10, fig.width=10}
@@
 
-Our data for this paper comes from SignBase, an open-access database collection of geometric signs on mobile objects from Paleolithic-era Europe, which was launched in 2020, with the explicit goal being to enable quantitative analyses of abstract geographical representation [@dutkiewicz2020].
-The creators of the database identified distinct geometric sign types, and recorded their occurrences on mobile artefacts from the Paleolithic.
-Identification and categorization of the sign types had already been performed by the creators of the SignBase database, the methods for which are described in @dutkiewicz2020, and we adhered to those categories and placements.
-
-The data set consists of `r nartifacts` artifacts found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign type found on these artifacts.
-We excluded signs if they fell under the "other" category, as the wide range of possibilities that could qualify as "other" means that it would be difficult to establish a definitive connection.
-We also excluded all data from Willendorf (only one sign present), and Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave (due to their extreme geographic distance from the other sites).
-Lastly, data from Grotte De La Princesse Pauline and Šandalja II was excluded, due to being dated much earlier than the rest of the artifacts.These exclusions resulted in `r nartifacts_clean` artifacts found at `r nsites_clean` sites in `r ncountries_clean` countries, encapsulating `r nsigntypes_clean` sign types.
-
-Maximum and minimum BP dates were provided by Signbase, and calibration was done through the rcarbon package in RStudio [@Crema_Bevan_2021].
-The calibrated radiocarbon dates of the artifacts range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years.
-We divided the data up into four periods in accordance with the date ranges provided in @fig-aur-table.
-
-All our tests were run with presence/absence data, in order to keep the data from being affected by outliers like Vogelherd.
-We ran a Mantel test between an Euclidean abundance distance matrix and a jaccard distance matrix of the sign types in order to make sure that the exclusion of abundance would not significantly impact our results [@Lycett_2019], and found that it was not quite statistically significant (P = `r p_value_abundance_correlation`, Mantel R = `r r_value_abundance_correlation`).
-Another Mantel test was run between a euclidean distance matrix for MedianBP and the jaccard sign type matrix in order to determine whether there was a definitive correlation between distance in time and sign types.
-The Mantel test found that there was a statistically significant correlation (p = `r chron_p`, Mantel-R = `r chron_r`).
+All our tests were run with presence/absence data.
+We ran a Mantel test between an Euclidean abundance distance matrix and a Jaccard distance matrix of the sign types in order to make sure that the exclusion of abundance would not significantly impact our results [@Lycett_2019], and found that it was not quite statistically significant (p-value = `r p_value_abundance_correlation`, Mantel statistic R = `r r_value_abundance_correlation`).
+Another Mantel test was run between a euclidean distance matrix for MedianBP and the Jaccard sign type matrix in order to determine whether there was a definitive correlation between distance in time and sign types.
+The Mantel test found that there was a statistically significant correlation (p-value = `r chron_p`, Mantel statistic R = `r chron_r`).
 
@@
 Seriation in this paper was performed using code designed to mimic the Browner-Kile seriation algorithm [@Brower_Kile_1988], in which the data is presented in a matrix diagram, and a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal.
-This is done so by calculating the mean row and column positions of presences, and arranging the rows and columns by respectively.
-
+This is done so by calculating the mean row and column positions of presences, and arranging the rows and columns respectively.
 Network analysis represents the relationship between nodes [@Mills_2017].
@@
 
-The modularity test quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006, @Newman_Girvan_2004].
+The modularity test is community structure network detection method, and quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006, @Newman_Girvan_2004].
 The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings.
-The PerMANOVA test quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017], with the results expressed by an R-statistic, and a P-score.
-
-PERMDISP is a multivariate extension of Levene's test [@Anderson_2006, @Anderson_Ellingsen_McArdle_2006], for the analysis of multivariate homogeneity of group dispersion.
-In this study, we perform the PERMDISP test upon the Jaccard matrix as a measure of compositional dissimilarity.
-We did so using the "betadisper" function from the RStudio package 'vegan', which mimics the PERMDISP2 procedure [@Vegan2001].
-In this code, the non-euclidean distance matrix is reduced to principal coordinates to embed them within a Euclidean space.
-The results are used to produce a visual of the level of overlap between the groups in terms of sign content.
+The perMANOVA test quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017], with the results expressed by an R-statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, and a statistical significance score (p \< 0.05), which quantifies whether the groups are significantly different.
 
@@
 Assuming that this framework is true, we would hypothetically see patterns of cultural features of Upper Paleolithic early modern humans that are dispersed according to geography.
-Mantel tests are used here for each time period to determine the correlation between sign distribution and geography.
+Mantel tests are used for each time period to determine the correlation between sign distribution and geography, accomplished with use of the vegan package in RStudio [@Vegan2001].
 The Mantel test is run between a geographic distance matrix and a Jaccard dissimilarity matrix.
-The test seeks to determine the level of correspondence between the matrices through permuational evaluation of the null distribution [@Mantel_Smouse], producing a Mantel-R score and a p-score as a result.
+The test seeks to determine the level of correspondence between the matrices through permuational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations) producing a Mantel R statistic and a p-value as a result.
+The r statistic falls between -1 and 1, with a value of -1 meaning a strong negative correlation and a value of 1 meaning a strong positive correlation, while a value of 0 means no correlation.
+The p-value quantifies statistical signifiance of correlation between the two matrices, with p \< 0.05 serving as the threshold for statistical significance.
 
@@
 ## Testing for Cultural Groupings
-
 ```{r}
@@
   disim <- 1 - dm
-  disim[disim<0.5] <- 0
+  disim[disim<0.2] <- 0
   disim_net <- network(disim,
@@
     }
-
-## mantel function tells us whether changes in sign makeup is correlated with chosen variable change. 
-## Permanova tells us how similar in chosen variable members of group are
-
-perm_function <- function(period = "none", variable = PCA, method = "permanova", signbase_sf = signbase_sf){
-  set.seed(500)
-
-  if(period == "none"){perm_df <- signbase_sf}
-
-  else{perm_df <- signbase_sf %>% 
-    filter(time_period == {{period}})}
-
-  if(method == "permanova"){
-  var_jac <- perm_df %>% 
-      st_drop_geometry() %>% 
-      column_to_rownames("id") %>% 
-      dplyr::select({{variable}}) %>% 
-      vegdist("jaccard")
-
-  result <- adonis2(var_jac~as.factor(perm_df$group), 
-                         method = "jaccard",
-                         sqrt.dist = TRUE)}
-
-  if(method == "mantel"){
-    sign_jac <- perm_df %>% 
-      st_drop_geometry() %>% 
-      column_to_rownames("id") %>% 
-      dplyr::select(line:star) %>% 
-      vegdist("jaccard", binary = TRUE)
-
-    var_sim <- perm_df %>% 
-      st_drop_geometry() %>% 
-      column_to_rownames("id") %>% 
-      dplyr::select({{variable}}) %>% 
-      dist()
-
-    result <- vegan::mantel(sign_jac,  var_sim, permutations = 1000)
-  }
-
-  return(result)
-}
 ```
@@
 ### Seriation
-
 ```{r fig.height= 9, fig.width= 13}
@@
 
+For each of the time periods, seriation identifies at least two groups, one with a minimal range of signs, and one with a broader range. Diversity serves as a major dividing factor here.
+
 ### Network Analysis
@@
 
-Combining the results of the above tests allowed us to find two main groups of sites within each time period, as shown in @fig-group-table.
-The geographic distributions of the groups are shown in @fig-group-map, and the sign makeup of each group is shown in @fig-group-sign.
+Network analysis shows similar groups to the ones identified in @fig-seriation, but it also shows evolving interconnectivity between groups.
+In the Transitional period, the network analysis shows two clearly defined groups, one consisting of Geissenklösterle, Fumane, and Pod Hradem, and the other consisting of Hohle Fels, El Castillo, and Labeko Koba, with no connections between the groups.
+The Proto-Aurignacian also shows two clearly defined groups, one consisting of Geissenklösterle, Spy, Vogelherd, and Grotte du Renne, and the other consisting of Mladeč, Hohlenstein-Stadel, Gatzarria, and Abri Pataud.
+However, it also displays some sites that are less clearly connected, with La Ferrassie being squarely in between both groups with equal connections to both, and Les Cottés lacking a connection to any of the groups.
+The Early Aurignacian shows more strongly interconnected groups - although the shape still suggests two main groups - with Grotte de la Verpillière I in particular having multiple strong connections to both of the groups.
+Lastly, the Evolved Aurignacian again shows two main groups, although group innerconnectivity has decreased, one consisting of Hohle Fels, Trou Magrite, Vogelherd, and Bockstein-Törle, and the other consisting of Gargas, La Viña, Sirgenstein Cave, Vindija Cave, and Les Rois.
+While interconnectivity between groups is still present, with connections between Trou Magrite, Gargas, and Vindija Cave, and Bockstein-Törle and Gargas, the strength of interconnectivity is weaker than in the Early Aurignacian.
 
 ## Groups
+
+The results of the above tests allowed us to find two main groups of sites within each time period, as shown in @fig-group-table.
+The geographic distributions of the groups are shown in @fig-group-map, and the sign makeup of each group is shown in @fig-group-sign.
 
@@
 full_table <- rbind(trans_strength, proto_strength, early_strength, evolved_strength) %>% 
-  mutate(`time period` = c("transitional", "proto", "early", "evolved"), .before = 1)
+  mutate(`time period` = c("Transitional", "Proto-Aurignacian",
+                                                       "Early Aurignacian", "Evolved Aurignacian"), .before = 1)
 
@@
 
-kable(full_table)
+kable(strength_table)
+
+proto_modularity <- proto_strength %>% 
+  dplyr::select(modularity)
+evolved_modularity <- evolved_strength %>% 
+  dplyr::select(modularity)
+
+trans_perm_r <- trans_strength %>% 
+  dplyr::select(`permanova r`)
+early_perm_r <- early_strength %>% 
+  dplyr::select(`permanova r`)
+
+trans_perm_p <- trans_strength %>% 
+  dplyr::select(`permanova p`)
 ```
 
-The R value from the perMANOVA tests tell us that for each of the time periods, around 20% of the variation in sign content can be explained by group membership, with the proto-aurignacian having the highest score of 27%, and the early Aurignacian having the lowest score of 21%.
-The perMANOVA p value quantifies the randomness of the groupings, with 0.05 as the threshold for definitive non-randomness.
-The only time period to not meet that threshold is the transitional period, with a p-score of 0.3.
+The modularity score quantifies the strength of the groupings, showing that the variations between time periods are minimal.
+The results tell us that the Proto-Aurignacian exhibits the strongest groupings, with a score of `r proto_modularity`, and that the Evolved Aurignacian exhibits the weakest groupings, with a score of `r evolved_modularity`.
+There is a slight increase in modularity between the Transitional period and the Proto-Aurignacian, before an even slighter decrease into the Early Aurignacian, followed by a larger decrease with the Evolved Aurignacian.
+
+The r statistic from the perMANOVA tests tell us the proportion of sign variation that can be explained by group membership.
+The Transitional period has the highest score of `r trans_perm_r` while the Early Aurignacian has the lowest score of `r early_perm_r`.
+We see a large decline in scores between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian periods.
+The perMANOVA p score quantifies the statistical significance of the difference, with 0.05 as the threshold.
+The only time period to not meet that threshold is the Transitional period, with a p-score of `r trans_perm_p`.
+the Proto and Early Aurignacian periods remain the same, with a slight increase in the Evolved Aurignacian period, although it still meets the threshold for statistical significance.
 
@@
 
+proto_mantel_r <- proto_strength %>% 
+  dplyr::select(`mantel r`)
+
+proto_mantel_p <- proto_strength %>% 
+  dplyr::select(`permanova p`)
+
+early_mantel_r <- proto_strength %>% 
+  dplyr::select(`mantel r`)
+
+early_mantel_p <- proto_strength %>% 
+  dplyr::select(`permanova p`)
+
 ```
@@
 The relevance of the isolation-by-distance hypothesis also varies from period to period.
+The p score quantifies the correlation, with a score of 0.05 or lower meaning that there is a statistically significant relationship between sign type distribution and geographical distance.
+None of the periods meet that threshold, meaning that the isolation-by-distance hypothesis is not applicable to any of them.
+However, we still see changes in the p score between periods, suggesting a higher level of correlation with geographic distance in some, and a lower correlation in others.
+The Proto-Aurignacian period demonstrates the least amount of correlation (Mantel R = `r proto_mantel_r`, p = `r proto_mantel_p `), while the Early Aurignacian exhibits the highest amount of correlation (Mantel R = `r early_mantel_r`, p = `r early_mantel_p `).
 
@@
 In this paper, we set out to ask what differences in the patterns of groupings among geometric sign types found on mobile objects between phases of the Aurignacian could tell us about the social dynamics of that period.
+@fig-seriation shows that each period can be divided into two main cultural groups, with diversity serving as a major dividing factor.
 
-One explanation for the sign type distribution could be that the groupings of the data represent cultural groups, as indicated by sign type patterns, and the geographic distribution between sites represent group mobility.
 @Whallon_2006 proposes a framework for understanding hunter-gatherer mobility, arguing that mobility occurs among hunter-gatherers for multiple reasons.
@@
 
-However, this preliminary research has been able to produce some conclusions to be reasonably confident in.
-Firstly, given that we know definitively that distribution and variation of sign types is not the result of geographic distance, we can conclude that the data is likely relevant to understanding patterns of cultural diversity.
-Secondly, we have found clusters of sites that correspond in sign type, possibly indicating mobility of specific cultural groups.
-These conclusions contribute to, and broaden our current understanding of cultural and biological diversity among Upper Paleolithic Homo sapiens in Europe.
-
 # Conclusion

### 2025-03-11 endofquarterpaper.qmd
edit

@@
 ---
+
 # Introduction
+
 The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020] coupled with biological evolution.
@@
 43 - 30 k BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020].
-In this paper, we investugate the social organization dynamics by analyzing the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe across four phases of the Aurignacian period.
+In this paper, we investigate the social organization dynamics by analyzing the distribution and variation of geometric sign types found on mobile artifacts from the Upper Paleolithic in Europe across four phases of the Aurignacian period.
 
@@
 
-Our data for this paper comes from SignBase, an open-access database collection of geometric signs on mobile objects from Paleolithic-era Europe, which was launched in 2020, with the explicit goal being to enable quantitative analyses of abstract geographical representation [@dutkiewicz2020].
-The creators of the database identified distinct geometric sign types, and recorded their occurrences on mobile artefacts from the Paleolithic.
+Our data was sourced from SignBase, an open-access database collection of geometric signs on mobile objects from Upper Paleolithic-era Europe.
 Identification and categorization of the sign types had already been performed by the creators of the SignBase database, the methods for which are described in @dutkiewicz2020, and we adhered to those categories and placements.
-Maximum and minimum BP dates were also provided by Signbase, and calibration was done through the rcarbon package in RStudio [@Crema_Bevan_2021].
+The data set consists of `r nartifacts` artifacts found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign types.
+We excluded signs if they fell under the "other" category, due to the wide range of potential possibilities.
+We also excluded all data from Willendorf (only one sign present), Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave (due to their extreme geographic distance from the other sites).
+Lastly, data from Grotte De La Princesse Pauline and Šandalja II was excluded, due to being dated much earlier than the rest of the artifacts.These exclusions resulted in `r nartifacts_clean` artifacts found at `r nsites_clean` sites in `r ncountries_clean` countries, encapsulating `r nsigntypes_clean` sign types.
+A map of the site distribution for each time period is shown in @fig-site-map.
 
-The data set consists of `r nartifacts` artifacts found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign type found on these artifacts.
-We excluded signs if they fell under the "other" category, as the wide range of potential possibilities means that it would be difficult to establish a definitive connection.
-We also excluded all data from Willendorf (only one sign present), and Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave (due to their extreme geographic distance from the other sites).
-Lastly, data from Grotte De La Princesse Pauline and Šandalja II was excluded, due to being dated much earlier than the rest of the artifacts.These exclusions resulted in `r nartifacts_clean` artifacts found at `r nsites_clean` sites in `r ncountries_clean` countries, encapsulating `r nsigntypes_clean` sign types.
-The calibrated radiocarbon dates of the cleaned data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years.
-We divided the data up into four periods in accordance with the date ranges provided in @fig-aur-table.
-
-```{r fig.height= 10, fig.width=10}
+```{r}
 #| label: fig-site-map
-#| fig-cap: "Graph of artifact count by time period and map of sites, divided by time period"
+#| fig-cap: "Map of sites, divided by time period"
 
@@
 library(ggpubr)
-library(cowplot)
-
-signbase_full_clean$time_period <- factor(signbase_full_clean$time_period, 
-                                            levels = c("transitional", "proto_aurignacian", 
-                                                       "early_aurignacian", "evolved_aurignacian"), 
-                                            labels = c("Transitional", "Proto-Aurignacian",
-                                                       "Early Aurignacian", "Evolved Aurignacian"))
-
-date_plot <- ggplot(signbase_full_clean) +
-  aes(x = MedianBP, fill = time_period) +
-  geom_bar(width = 150) +
-  scale_x_reverse() +
-  labs(fill = "Time period") +
-  theme(legend.key.size = unit(0.5, "cm"))
 
@@
 
-site_map <- ggplot(Europe) +
+ggplot(Europe) +
   geom_sf() +
@@
         axis.title = element_blank())
+```
 
-plot_grid(date_plot, site_map, nrow = 2, align = "hv", rel_heights = c(1, 2))
+Maximum and minimum BP dates were also provided by Signbase, and calibration was done through the rcarbon package in RStudio [@Crema_Bevan_2021].
+We divided the data up into four chronological periods in accordance with the date ranges recorded in @fig-aur-table.
+The calibrated radiocarbon dates of the cleaned data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years.
+Artifact count by time period is shown in @fig-time-count.
+
+```{r fig.height= 5, fig.width=10}
+#| label: fig-time-count
+#| fig-cap: "Graph of artifact count by time period"
+
+signbase_full_clean$time_period <- factor(signbase_full_clean$time_period, 
+                                            levels = c("transitional", "proto_aurignacian", 
+                                                       "early_aurignacian", "evolved_aurignacian"), 
+                                            labels = c("Transitional", "Proto-Aurignacian",
+                                                       "Early Aurignacian", "Evolved Aurignacian"))
+
+ggplot(signbase_full_clean) +
+  aes(x = MedianBP, fill = time_period) +
+  geom_bar(width = 150) +
+  scale_x_reverse() +
+  labs(fill = "Time period") +
+  theme(legend.key.size = unit(0.5, "cm"))
+
 ```
@@
 
-Seriation analysis is the arrangement of data into a linear order to reveal patterns, and is a common technique in analyzing sets of archaeological data [@Hahsler_Hornik_Buchta].
+Seriation analysis is the arrangement of data into a linear order to reveal patterns [@Hahsler_Hornik_Buchta].
 Seriation in this paper was performed using code designed to mimic the Browner-Kile seriation algorithm [@Brower_Kile_1988], in which the data is presented in a matrix diagram, and a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal.
@@
 
-The modularity score quantifies the strength of the groupings, showing that the variations between time periods are minimal.
+The modularity score quantifies the strength of the groupings, showing that the variations in grouping strength between time periods are minimal.
 The results tell us that the Proto-Aurignacian exhibits the strongest groupings, with a score of `r proto_modularity`, and that the Evolved Aurignacian exhibits the weakest groupings, with a score of `r evolved_modularity`.
@@
 However, we still see changes in the p score between periods, suggesting a higher level of correlation with geographic distance in some, and a lower correlation in others.
-The Proto-Aurignacian period demonstrates the least amount of correlation (Mantel R = `r proto_mantel_r`, p = `r proto_mantel_p `), while the Early Aurignacian exhibits the highest amount of correlation (Mantel R = `r early_mantel_r`, p = `r early_mantel_p `).
+The Proto-Aurignacian period demonstrates the least amount of correlation (Mantel R = `r proto_mantel_r`, p = `r proto_mantel_p`), while the Early Aurignacian exhibits the highest amount of correlation (Mantel R = `r early_mantel_r`, p = `r early_mantel_p`).
 
@@
 
-In this paper, we set out to ask what differences in the patterns of groupings among geometric sign types found on mobile objects between phases of the Aurignacian could tell us about the social dynamics of that period.
-@fig-seriation shows that each period can be divided into two main cultural groups, with diversity serving as a major dividing factor.
+Using geometric sign types as a proxy for cultural difference, we aimed to answer the following questions: How grouping dynamics of geometric signs change across the four phases of the Aurignacian?
+How do the groupings vary in diversity, size, connectivity, and boundaries?
+Lastly, what do the patterns tell us about dynamics of social organization throughout the Aurignacian?
 
-@Whallon_2006 proposes a framework for understanding hunter-gatherer mobility, arguing that mobility occurs among hunter-gatherers for multiple reasons.
-The first two reasons, referred to by Whallon as logistical and residential mobility, occur when groups move in order to procure a new supply of resources once the more immediate ones have been depleted.
-Another type of mobility is network mobility, in which groups, or subsets of groups, move in order to make contact with other groups.
-This occurs in order to form a "safety net", in times of scarce resources, so that groups are assured of a positive reception if they are forced to move.
-Lastly, and similarly, is informational mobility, in which groups, or individuals, move with the primary goal of gathering information, typically about the environmental conditions and resources in other areas.
-Similarly, this operates as a "safety net" of sorts, in that these groups then posses knowledge on where to move in times of resource scarcity.
-One point of note is that, according to @Whallon_2006, decorative objects may be exchanged as a method of network formation, which may account for the presence of some signs at some sites.
+Firstly, the results as displayed in @fig-seriation show that each period can be divided into two main cultural groups, with diversity serving as a major dividing factor.
+@fig-neighbor-net shows that group inter- and inner-connectivity varies from period to period, with weak inner group connections and no between-group connections in the Transitional period, to stronger inner connections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to slightly weaker inter-group connections in the Evolved Aurignacian.
+The modularity tests tell us that the groupings are strongest in the Transitional period and weakest in the Early Aurignacian, and the perMANOVA tests also tell us that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian experiences the least.
+The Mantel tests tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian experiences the highest correlation.
+The results
 
-Further research will need to be done on this topic in order to understand these patterns.
-In particular, the ecological and biogeographical conditions at the sites will need to be analyzed for potential connections, as mobility is heavily environmentally-motivated.
-Additionally, further analysis of the sign type categories as created by @dutkiewicz2020, as it is possible, when it comes to the simpler signs, for identical creation to occur independently.
+One explanation for these changing social dynamics could be the biological and environmental change established during these periods.
 

### 2025-03-14 endofquarterpaper.qmd

@@
 ---
-title: "Changes in Groupings of Geometric Signs as a Model of Social Dynamics Across Aurignacian Periods"
+title: "Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases"
 author: Sophie Cooper
@@
 bibliography: references.bib
-zotero: "Signbase-study"
 link-citations: true
@@
 43 - 30 k BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020].
-In this paper, we investigate the social organization dynamics by analyzing the distribution and variation of geometric sign types found on mobile artifacts from the Upper Paleolithic in Europe across four phases of the Aurignacian period.
+In this paper, we investigate the social organization dynamics by analyzing the distribution and variation of geometric sign types found on mobile objects from the Upper Paleolithic in Europe across four phases of the Aurignacian period.
 
-Social organization dynamics of the Upper Paleolithic have previously been explored using statistical analyses of artifacts to understand patterns of cultural groups.
+Social organization dynamics of the Upper Paleolithic have previously been explored using statistical analyses of objects to understand patterns of cultural groups.
 Multivariate analyses of beads from the Gravettian technocomplex suggests an east to west cline of nine cultural groups [@baker2024evidence].
@@
 ```{r}
-#| label: fig-aur-table
-#| fig-cap: "Table of data on Aurignacian periods [@Banks_dErrico_Zilhao_2013;  @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021; @Chu_Richter_2020; @Fletcher_Sánchez]"
+#| label: tbl-aur-table
+#| tbl-cap: "Table of data on Aurignacian periods [@Banks_dErrico_Zilhao_2013;  @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021; @Chu_Richter_2020; @Fletcher_Sánchez]"
+#| tbl-cap-location: bottom
+
 library(knitr)
@@
 
-We hypothesize that these periods, and the environmental changes across the Aurignacian, may correspond to changes in social dynamics of cultural groups, as represented by patterns in the distribution of geometric signs.
+We hypothesize that these phases, and the environmental changes across the Aurignacian, may correspond to changes in social dynamics of cultural groups, as represented by patterns in the distribution of geometric signs.
 We aim to address the following questions: How do grouping dynamics of geometric signs change across the four phases of the Aurignacian?
@@
 Lastly, what do the patterns tell us about dynamics of social organization throughout the Aurignacian?
-We answer these questions by first identifying clusters within each time period using seriation and network analyses.
+We answer these questions by first identifying clusters within each phase using seriation and network analyses.
 We then test the strength of the clusters using modularity and perMANOVA tests.
-Lastly, we use Mantel tests to evaluate the null hypothesis of isolation-by-distance for each time period.
+Lastly, we use Mantel tests to evaluate the null hypothesis of isolation-by-distance for each phase.
 
-# Methods and Materials
+# Materials and Methods
 
@@
 
-trans_unique_data <- signbase_full_clean %>% 
+trans_df<- signbase_full_clean %>% 
   filter(time_period == "transitional") %>% 
   mutate(longitude = as.character(longitude),
-         latitude = as.character(latitude)) %>% 
+         latitude = as.character(latitude))
+
+trans_object_num <- trans_df %>% 
+  group_by(site_name) %>%
+  summarise(nobjects = n())
+
+trans_unique_data <- trans_df %>% 
   group_by(site_name) %>% 
@@
   left_join(lat_long_df) %>% 
+  left_join(trans_object_num) %>% 
   mutate(time_period = "transitional")
 
-proto_unique_data <- signbase_full_clean %>% 
+proto_df <- signbase_full_clean %>% 
   filter(time_period == "proto_aurignacian") %>% 
   mutate(longitude = as.character(longitude),
-         latitude = as.character(latitude)) %>% 
+         latitude = as.character(latitude))
+
+proto_object_num <- proto_df %>% 
+  group_by(site_name) %>%
+  summarise(nobjects = n())
+
+proto_unique_data <- proto_df %>% 
   group_by(site_name) %>%
@@
   left_join(lat_long_df) %>% 
+  left_join(proto_object_num) %>% 
   mutate(time_period = "proto_aurignacian")
 
-early_unique_data <- signbase_full_clean %>% 
+early_df <- signbase_full_clean %>% 
   filter(time_period == "early_aurignacian") %>% 
   mutate(longitude = as.character(longitude),
-         latitude = as.character(latitude)) %>% 
+         latitude = as.character(latitude))
+
+early_object_df <- early_df %>% 
+  group_by(site_name) %>%
+  summarise(nobjects = n())
+
+early_unique_data <- early_df %>% 
   group_by(site_name) %>%
@@
   left_join(lat_long_df) %>% 
+  left_join(early_object_df) %>% 
   mutate(time_period = "early_aurignacian")
 
-evolved_unique_data <- signbase_full_clean %>% 
+evolved_df <- signbase_full_clean %>% 
   filter(time_period == "evolved_aurignacian") %>% 
   mutate(longitude = as.character(longitude),
-         latitude = as.character(latitude)) %>% 
+         latitude = as.character(latitude))
+
+evolved_object_df <- evolved_df %>% 
+  group_by(site_name) %>%
+  summarise(nobjects = n())
+
+evolved_unique_data  <- evolved_df %>% 
   group_by(site_name) %>%
@@
   left_join(lat_long_df) %>% 
+  left_join(evolved_object_df) %>% 
   mutate(time_period = "evolved_aurignacian")
@@
   column_to_rownames("site_name") %>% 
-  dplyr::select(line:star)
+  dplyr::select(line:star) %>% 
+  select_if(~ !is.numeric(.) || sum(.) != 0)
 
@@
                               site_name == "Grotte de la Verpillière I"|
-                              site_name == "Vindija Cave" |
-                              site_name == "Trou al'Wesse") ~ "1",
+                              site_name == "Vindija Cave") ~ "1",
                            (site_name == "Hohle Fels" | 
@@
                               site_name == "Cellier" | 
-                              site_name == "Blanchard") ~ "2"))
+                              site_name == "Blanchard"|
+                              site_name == "Trou al'Wesse") ~ "2"))
 
@@
 
+trans_site_count <- nrow(trans_unique_data)
+trans_sign_count <- ncol(trans_artifact_data)
+proto_site_count <- nrow(proto_unique_data)
+proto_sign_count <- ncol(proto_artifact_data)
+early_site_count <- nrow(early_unique_data)
+early_sign_count <- ncol(early_artifact_data)
+evolved_site_count <- nrow(evolved_unique_data)
+evolved_sign_count <- ncol(evolved_artifact_data)
+
+nvogel <- nrow(signbase_full_clean %>% 
+                 filter(site_name == "Vogelherd"))
+
+artifact_count_df <- signbase_full_clean %>% 
+  group_by(site_name) %>% 
+  summarise(artifact_count = n())
+
+artifact_count_median <- median(artifact_count_df$artifact_count)
+
 ```
@@
   column_to_rownames("id") %>% 
-  select(line:star)
+  dplyr::select(line:star)
 
-euc_data <- artifact_data %>% 
-  mutate(sign_count = rowSums(.)) %>% 
-  dplyr::select(sign_count)
+object_euc_data <- time_unique_data_full%>%
+  dplyr::select(id, nobjects) %>% 
+  dist(method = "euclidean")
 
-euc_dist <- dist(x = euc_data, method = "euclidean")
 jac_full <- vegan::vegdist(artifact_data, method = "jaccard", binary = TRUE)
 
-euc_mantel <- vegan::mantel(euc_dist, jac_full, permutations = 1000)
-p_value_abundance_correlation <- round(euc_mantel$signif, 3)
-r_value_abundance_correlation <- round(euc_mantel$statistic, 3)
+object_count_mantel <- vegan::mantel(object_euc_data, jac_full, permutations = 1000)
+p_value_object_abundance_correlation <- round(object_count_mantel$signif, 3)
+r_value_object_abundance_correlation <- round(object_count_mantel$statistic, 3)
+
+sign_euc_data <- artifact_data %>% 
+  mutate(sign_count = rowSums(.)) %>% 
+  dplyr::select(sign_count) %>% 
+  dist(method = "euclidean")
+
+sign_count_mantel <- vegan::mantel(sign_euc_data, jac_full, permutations = 1000)
+p_value_sign_abundance_correlation <- round(sign_count_mantel$signif, 3)
+r_value_sign_abundance_correlation <- round(sign_count_mantel$statistic, 3)
 
@@
 
+##chron permanova
+perman_chron <- adonis2(jac_full~as.factor(time_unique_data_full$time_period), 
+                    method = "jaccard",
+                    sqrt.dist = TRUE)
+perman_chron <- perman_chron %>% 
+    data_frame() %>% 
+    slice_head()
+
 ```
@@
 
-Our data was sourced from SignBase, an open-access database collection of geometric signs on mobile objects from Upper Paleolithic-era Europe.
-Identification and categorization of the sign types had already been performed by the creators of the SignBase database, the methods for which are described in @dutkiewicz2020, and we adhered to those categories and placements.
-The data set consists of `r nartifacts` artifacts found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign types.
+Our data comes from SignBase 1.0, an open-access collection of geometric signs on mobile objects from Upper Paleolithic Europe.
+We used the identification and categorization of the sign types provided by SignBase's curators [@dutkiewicz2020].
+The data set consists of `r nartifacts` objects found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign types.
 We excluded signs if they fell under the "other" category, due to the wide range of potential possibilities.
-We also excluded all data from Willendorf (only one sign present), Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave (due to their extreme geographic distance from the other sites).
-Lastly, data from Grotte De La Princesse Pauline and Šandalja II was excluded, due to being dated much earlier than the rest of the artifacts.These exclusions resulted in `r nartifacts_clean` artifacts found at `r nsites_clean` sites in `r ncountries_clean` countries, encapsulating `r nsigntypes_clean` sign types.
-A map of the site distribution for each time period is shown in @fig-site-map.
+We also excluded seven sites from the sample (only one sign present: Willendorf; extreme geographic distance from the other sites: Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave; dated much later than the rest of the assemblages: Grotte De La Princesse Pauline and Šandalja II).
+These exclusions resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map).
 
-```{r}
+```{r fig.height= 9, fig.width=10}
 #| label: fig-site-map
-#| fig-cap: "Map of sites, divided by time period"
+#| fig-cap: "Map of sites containing geometric signs on mobile objects from Upper Paleolithic Europe, sampled from the SignBase dataset, divided by Aurignacian phases"
 
@@
 library(ggpubr)
+library(cowplot)
+
+signbase_full_clean$time_period <- factor(signbase_full_clean$time_period, 
+                                            levels = c("transitional", "proto_aurignacian", 
+                                                       "early_aurignacian", "evolved_aurignacian"), 
+                                            labels = c("Transitional", "Proto-Aurignacian",
+                                                       "Early Aurignacian", "Evolved Aurignacian"))
+
+signcount_barplot <- ggplot(signbase_full_clean) +
+  aes(x = MedianBP, fill = time_period) +
+  geom_bar(width = 150) +
+  scale_x_reverse() +
+  labs(fill = "Time period") +
+  theme(legend.key.size = unit(0.5, "cm")) +
+  scale_fill_discrete(guide = "none")
 
@@
 
-ggplot(Europe) +
+sign_map <- ggplot(Europe) +
   geom_sf() +
-  geom_sf(data = signbase_sf) +
+  geom_sf(data = signbase_sf, aes(color = time_period)) +
+  scale_colour_discrete(guide = "none")+
   coord_sf(xlim = c(-10,30), 
@@
                   max.overlaps = 30,
-                  size = 2) +
+                  size = 2, 
+                  bg.color = "white",
+                  bg.r = 0.25) +
   theme_minimal() +
-  facet_wrap(~time_period) +
+  facet_wrap(~time_period, ) +
   theme(axis.text = element_blank(),
-        axis.title = element_blank())
-```
+        axis.title = element_blank(),
+        strip.text = element_text(size = 10))
 
-Maximum and minimum BP dates were also provided by Signbase, and calibration was done through the rcarbon package in RStudio [@Crema_Bevan_2021].
-We divided the data up into four chronological periods in accordance with the date ranges recorded in @fig-aur-table.
-The calibrated radiocarbon dates of the cleaned data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years.
-Artifact count by time period is shown in @fig-time-count.
-
-```{r fig.height= 5, fig.width=10}
-#| label: fig-time-count
-#| fig-cap: "Graph of artifact count by time period"
-
-signbase_full_clean$time_period <- factor(signbase_full_clean$time_period, 
-                                            levels = c("transitional", "proto_aurignacian", 
-                                                       "early_aurignacian", "evolved_aurignacian"), 
-                                            labels = c("Transitional", "Proto-Aurignacian",
-                                                       "Early Aurignacian", "Evolved Aurignacian"))
-
-ggplot(signbase_full_clean) +
-  aes(x = MedianBP, fill = time_period) +
-  geom_bar(width = 150) +
-  scale_x_reverse() +
-  labs(fill = "Time period") +
-  theme(legend.key.size = unit(0.5, "cm"))
+plot_grid(sign_map, signcount_barplot, ncol=1, align = "v", scale = c(1, 0.8))
 
@@
 
-All our tests were run with presence/absence data.
-We ran a Mantel test between an Euclidean abundance distance matrix and a Jaccard distance matrix of the sign types in order to make sure that the exclusion of abundance would not significantly impact our results [@Lycett_2019], and found that it was not quite statistically significant (p-value = `r p_value_abundance_correlation`, Mantel statistic R = `r r_value_abundance_correlation`).
-Another Mantel test was run between a euclidean distance matrix for MedianBP and the Jaccard sign type matrix in order to determine whether there was a definitive correlation between distance in time and sign types.
-The Mantel test found that there was a statistically significant correlation (p-value = `r chron_p`, Mantel statistic R = `r chron_r`).
+Ages of the objects were determined by calibrating the uncalibrated ages provided by SignBase, using the rcarbon package [@Crema_Bevan_2021].
+The calibrated radiocarbon dates of the sampled data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years.
+We assigned each artifact to a chronological phase following the date ranges recorded in @tbl-aur-table.
+The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` signs, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` signs, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` signs, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` signs.
 
-## Testing for Cultural Groupings
+SignBase 1.0 is a record of objects noting only the presence or absence of each sign, without recording the quantity or sequence of individual signs on a given object.
+The distribution of objects across sites is highly skewed, with Vogelherd having `r nvogel`, but the median number of objects is `r artifact_count_median`.
+To minimize the influence of this skewed distribution, we used presence/absence data for each sign at each site, reducing the effect of variation in object count between sites.
+To make sure that the exclusion of abundance would not significantly impact our results, we ran a Mantel tests both between a Euclidean object abundance distance matrix and a Jaccard distance matrix of the sign types [@Lycett_2019] (p = `r p_value_object_abundance_correlation`, Mantel statistic R = `r r_value_object_abundance_correlation`), and a Eudclidean sign abundance distance matrix and a Jaccard sign type distance matrix (p = `r p_value_sign_abundance_correlation`, Mantel statistic R = `r r_value_sign_abundance_correlation`) and found that neither were statistically significant.
+Another Mantel test was run between a Euclidean distance matrix for dates and the Jaccard sign type matrix in order to determine whether there was a definitive correlation between distance in time and distribution of sign types, finding a statistically significant correlation (p = `r chron_p`, Mantel statistic R = `r chron_r`).
+This confirms that a meaningful change in sign types occurs over time.
+
+{run permanova test for chronology}
+
+## Cultural Group Detection Using Seriation and Network Analysis
 
 Seriation analysis is the arrangement of data into a linear order to reveal patterns [@Hahsler_Hornik_Buchta].
-Seriation in this paper was performed using code designed to mimic the Browner-Kile seriation algorithm [@Brower_Kile_1988], in which the data is presented in a matrix diagram, and a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal.
-This is done so by calculating the mean row and column positions of presences, and arranging the rows and columns respectively.
-Network analysis represents the relationship between nodes [@Mills_2017].
-In this network analysis, each node is a site, and each edge is a measure of sign similarity, with the direct value represented by edge weight.
-The results were plotted with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths.
+We used the Browner-Kile seriation algorithm [@Brower_Kile_1988] from the seriation package [@seriation_pckge], in which a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal.
+Network analysis represents and quantifies the relationship between nodes [@Mills_2017], in this case, sites.
+In our network analysis, each node is a site, and each edge is a measure of sign type distribution similarity.
+We input the Jaccard similarity matrix for each phase, computed the network statistics, and plotted the results with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths.
+Network analysis was performed using the vegan [@Vegan2001], statnet [@statnet_package] and igraph [@igraph_package] packages.
 
-## Testing Strength of Cultural Groupings
+## Modularity and PerMANOVA for Cultural Group Strength
 
-The modularity test is community structure network detection method, and quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006, @Newman_Girvan_2004].
+To evaluate the strength of the groupings found in the seriation and network analyses, we used modularity and perMANOVA tests.
+The modularity test is a community structure network detection method, and quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006; @Newman_Girvan_2004].
 The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings.
-The perMANOVA test quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017], with the results expressed by an R-statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, and a statistical significance score (p \< 0.05), which quantifies whether the groups are significantly different.
+PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017].
+The results are expressed by an R-statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, and a p-value, which quantifies whether the groups are significantly different (we use an alpha value of 0.05 for the threshold of statistical significance).
+The modularity test and perMANOVA were performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.
 
-## Testing Isolation-by-Distance Hypothesis
+## Effects of Geographic Distance Using Mantel Tests
 
-The isolation-by-distance framework proposes that cultural difference can be primarily explained by geographic distance.
-Assuming that this framework is true, we would hypothetically see patterns of cultural features of Upper Paleolithic early modern humans that are dispersed according to geography.
-Mantel tests are used for each time period to determine the correlation between sign distribution and geography, accomplished with use of the vegan package in RStudio [@Vegan2001].
-The Mantel test is run between a geographic distance matrix and a Jaccard dissimilarity matrix.
-The test seeks to determine the level of correspondence between the matrices through permuational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations) producing a Mantel R statistic and a p-value as a result.
-The r statistic falls between -1 and 1, with a value of -1 meaning a strong negative correlation and a value of 1 meaning a strong positive correlation, while a value of 0 means no correlation.
-The p-value quantifies statistical signifiance of correlation between the two matrices, with p \< 0.05 serving as the threshold for statistical significance.
+The isolation-by-distance framework proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019].
+If cultural difference can be explained solely by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases.
+To test this, we determined the correlation between sign distribution and geography using Mantel tests, via the vegan package [@Vegan2001].
+The Mantel tests were run with a geographic distance matrix and a Jaccard similarity matrix.
+The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value.
+The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation.
+The p-value quantifies statistical significance of the correlation between the two matrices, with p = 0.05 serving as the threshold for statistical significance.
 
@@
 
-## Testing for Cultural Groupings
+## Cultural Group Detection
 
 ```{r}
-library(ggpubr)
-library(phangorn)
 library(ggraph)
 library(statnet)
-library(ecodist)
 library(ape)
-library(ggrepel)
-library(cowplot)
 library(igraph)
@@
 strength_function <- function(artifact_data, group_data){
-  artifact_data <- artifact_data %>% 
-    select_if(~ !is.numeric(.) || sum(.) != 0)
 
@@
 trans_network <- function(artifact_data){
-  artifact_data <- artifact_data %>% 
-    select_if(~ !is.numeric(.) || sum(.) != 0)
     jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
@@
 
-### Seriation
-
 ```{r fig.height= 9, fig.width= 13}
@@
 
-For each of the time periods, seriation identifies at least two groups, one with a minimal range of signs, and one with a broader range. Diversity serves as a major dividing factor here.
-
-### Network Analysis
+For each of the phases, seriation identifies at least two groups, one with a minimal range of signs, and one with a broader range.
+We also see that the notch and line signs are the most prevalent signs across all time periods.
+In the Transitional phase, we see some variety within groups, with the first consisting of Fumane, Pod Hradem, and Geissenklösterle, two of which only have the "notch" sign, and one of which has the notch and circumnotch, and the other consisting of Labeko Koba, Hohle Fels, and El Castillo, which have a broader diversity of signs.
+In the Proto-Aurignacian, we see an expansion in both sites and range of signs, as well as movement to new sites, with little overlap in site names.
+The first group consists of Abri Pataud, Gatzarria, Hohlenstein-Stadel, and Mladeč, three of which only have the notch and line signs, and one which also has a dashline sign, and the other consists of Geissenklösterle, Spy, Vogelherd, Grotte du Renne, La Ferrassie, and Les Cottés.
+In the second group, we see fairly consistent patterning of a diverse range of signs, although Les Cottés varies a little.
+In the Early Aurignacian, we again see movement to new sites with no overlap from the Proto-Aurignacian, as well as increased diversity of patterns within the groups.
+The first group consists of Grottes de Fonds-de-Forêt (radnotch and notch), Riparo Bombrini (notch), and Grotte de la Verpillière I (notch and line), The second consists of Solutré, Hohle Fels, Castanet, Cellier, Blanchard, and Trou al'Wesse, which contain a wide range of signs in a variety of combinations.
+Lastly, in the Evolved Aurignacian, we see slightly more consistent groups, again with movement to new sites with no overlap from the Early Aurignacian.
+The first group consists of Les Rois, Gargas, La Viña, Sirgenstein Cave, and Vindija Cave, covering the line, notch, and and obline signs.
+The second consists of Trou Magrite, Bockstein-Törle, Vogelherd, and Hohle Fels.
+Interestingly, this is the only phase to have more sites in the group with a minimal range of signs, than in the group with a more diverse range.
 
@@
 
-Network analysis shows similar groups to the ones identified in @fig-seriation, but it also shows evolving interconnectivity between groups.
-In the Transitional period, the network analysis shows two clearly defined groups, one consisting of Geissenklösterle, Fumane, and Pod Hradem, and the other consisting of Hohle Fels, El Castillo, and Labeko Koba, with no connections between the groups.
-The Proto-Aurignacian also shows two clearly defined groups, one consisting of Geissenklösterle, Spy, Vogelherd, and Grotte du Renne, and the other consisting of Mladeč, Hohlenstein-Stadel, Gatzarria, and Abri Pataud.
+Network analysis shows similar groups to the ones identified in @fig-seriation, but it also shows evolving inter-connectivity between groups.
+In the Transitional phase, the network analysis shows two clearly defined groups, one consisting of Geissenklösterle, Fumane, and Pod Hradem, and the other consisting of Hohle Fels, El Castillo, and Labeko Koba, with no connections between the groups.
+The Proto-Aurignacian phase also shows two clearly defined groups, one consisting of Geissenklösterle, Spy, Vogelherd, and Grotte du Renne, and the other consisting of Mladeč, Hohlenstein-Stadel, Gatzarria, and Abri Pataud.
 However, it also displays some sites that are less clearly connected, with La Ferrassie being squarely in between both groups with equal connections to both, and Les Cottés lacking a connection to any of the groups.
-The Early Aurignacian shows more strongly interconnected groups - although the shape still suggests two main groups - with Grotte de la Verpillière I in particular having multiple strong connections to both of the groups.
-Lastly, the Evolved Aurignacian again shows two main groups, although group innerconnectivity has decreased, one consisting of Hohle Fels, Trou Magrite, Vogelherd, and Bockstein-Törle, and the other consisting of Gargas, La Viña, Sirgenstein Cave, Vindija Cave, and Les Rois.
-While interconnectivity between groups is still present, with connections between Trou Magrite, Gargas, and Vindija Cave, and Bockstein-Törle and Gargas, the strength of interconnectivity is weaker than in the Early Aurignacian.
+The Early Aurignacian shows more strongly inter-connected groups - although the shape still suggests two main groups - with Grotte de la Verpillière I in particular having multiple strong connections to both of the groups.
+Lastly, the Evolved Aurignacian again shows two main groups, although group inner-connectivity has decreased, one consisting of Hohle Fels, Trou Magrite, Vogelherd, and Bockstein-Törle, and the other consisting of Gargas, La Viña, Sirgenstein Cave, Vindija Cave, and Les Rois.
+While inter-connectivity between groups is still present, with connections between Trou Magrite, Gargas, and Vindija Cave, and Bockstein-Törle and Gargas, the strength of inter-connectivity is weaker than in the Early Aurignacian.
 
-## Groups
-
-The results of the above tests allowed us to find two main groups of sites within each time period, as shown in @fig-group-table.
+In identifying the groups, we referenced both the seriation and network analyses.
+The results of the above tests allowed us to find two main groups of sites within each phase, as shown in @tbl-group-table.
 The geographic distributions of the groups are shown in @fig-group-map, and the sign makeup of each group is shown in @fig-group-sign.
@@
 ```{r}
-#| label: fig-group-table
-#| fig-cap: "Table of the sites by group"
+#| label: tbl-group-table
+#| tbl-cap: "Table of the sites by group per phase"
+#| tbl-cap-location: bottom
 group_table <- read_excel("data/site-groups.xlsx")
@@
                   max.overlaps = 30,
-                  size = 2) +
+                  size = 2,
+                  bg.color = "white",
+                  bg.r = 0.25) +
   theme_minimal() +
@@
 
-## Testing Strength of Cultural Groupings
+## Cultural Group Strength
 
 ```{r}
+#| label: tbl-strength-table
+#| tbl-cap: "Table of the results per phase for the modularity test and perMANOVA"
+#| tbl-cap-location: bottom
 ##Modularity test
@@
 
-The modularity score quantifies the strength of the groupings, showing that the variations in grouping strength between time periods are minimal.
+The modularity score quantifies the strength of the groupings, showing that the variations in grouping strength between phases are fairly minimal.
 The results tell us that the Proto-Aurignacian exhibits the strongest groupings, with a score of `r proto_modularity`, and that the Evolved Aurignacian exhibits the weakest groupings, with a score of `r evolved_modularity`.
@@
 
-The r statistic from the perMANOVA tests tell us the proportion of sign variation that can be explained by group membership.
+The R statistic from the perMANOVA tests tell us the proportion of sign variation that can be explained by group membership.
 The Transitional period has the highest score of `r trans_perm_r` while the Early Aurignacian has the lowest score of `r early_perm_r`.
@@
 The perMANOVA p score quantifies the statistical significance of the difference, with 0.05 as the threshold.
-The only time period to not meet that threshold is the Transitional period, with a p-score of `r trans_perm_p`.
-the Proto and Early Aurignacian periods remain the same, with a slight increase in the Evolved Aurignacian period, although it still meets the threshold for statistical significance.
+The only time period to not meet that threshold is the Transitional period (p = `r trans_perm_p`).
+There is a slight increase between the Proto and Early Aurignacian, and a slightly large increase in the Evolved Aurignacian period, although they all still meets the threshold for statistical significance.
 
-## Testing Isolation-by-Distance Hypothesis
+## Effect of Geographic Distance
 
 ```{r}
+#| label: tbl-mantel-table
+#| tbl-cap: "Table of the results per phase for the Mantel test between a Jaccard matrix of sign type diversity and a geographic distance matrix"
+#| tbl-cap-location: bottom
 ## Mantel test for geography vs signs
@@
 proto_mantel_p <- proto_strength %>% 
-  dplyr::select(`permanova p`)
+  dplyr::select(`mantel p`)
 
-early_mantel_r <- proto_strength %>% 
+early_mantel_r <- early_strength %>% 
   dplyr::select(`mantel r`)
 
-early_mantel_p <- proto_strength %>% 
-  dplyr::select(`permanova p`)
+early_mantel_p <- early_strength %>% 
+  dplyr::select(`mantel p`)
 
@@
 
-Using geometric sign types as a proxy for cultural difference, we aimed to answer the following questions: How grouping dynamics of geometric signs change across the four phases of the Aurignacian?
-How do the groupings vary in diversity, size, connectivity, and boundaries?
-Lastly, what do the patterns tell us about dynamics of social organization throughout the Aurignacian?
+The first aim of our study was to determine how the grouping dynamics of geometric signs changed across the four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and boundaries.
+The results as displayed in @fig-seriation show that each phase can be divided into two main cultural groups, with diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.
+@fig-neighbor-net shows that group inter-connectivity and inner-connectivity varies from phase to phase, with weak inner group connections and no between-group connections in the Transitional phase, to stronger inner connections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
+The modularity tests, as shown in @tbl-strength-table, tell us that the groupings are strongest in the Transitional period and weakest in the Early Aurignacian, and the perMANOVA tests tell us that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian experiences the least.
+The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian experiences the highest correlation.
 
-Firstly, the results as displayed in @fig-seriation show that each period can be divided into two main cultural groups, with diversity serving as a major dividing factor.
-@fig-neighbor-net shows that group inter- and inner-connectivity varies from period to period, with weak inner group connections and no between-group connections in the Transitional period, to stronger inner connections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to slightly weaker inter-group connections in the Evolved Aurignacian.
-The modularity tests tell us that the groupings are strongest in the Transitional period and weakest in the Early Aurignacian, and the perMANOVA tests also tell us that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian experiences the least.
-The Mantel tests tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian experiences the highest correlation.
-The results
+The second aim of our study was to determine what these patterns tell us about dynamics of social organization throughout the Aurignacian.
+The results suggest that the Transitional phase was sparsely populated, with minimal interaction between loosely connected groups.
+For the Proto-Aurignacian, the results suggest a large expansion in social activity, as well as strongly inner-connected groups with some inter-connectivity present, with geographic distance having the least influence on sign distribution.
+The Early Aurignacian experiences a large surge in inter-connectivity, as well as inner-connectivity strength, with a great deal of connection between groups and an increase in the correlation with geographic distance.
+The Evolved Aurignacian experiences a decline in connection strength, although it maintains degrees of inter-connectivity.
 
-One explanation for these changing social dynamics could be the biological and environmental change established during these periods.
+These results are supported by the existing data on the environmental and biological change established during these phases.
+@Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of 1Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Aur-P1 settlement being reached around 41 k BP.
+This could provide an answer as to why the Transitional phase appears sparsely populated with little to no inter-connectivity, and the Proto-Aurignacian appears to be more densely populated with stronger cultural groups.
+Additionally @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations.
+This could be a reason as to why we see a large increase in both inter-connectivity and inner-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance.
+Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Aur-P2, theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions.
+@Shao_Wegener_Klein_Schmidt_Weniger_2024 also find an increase in expansion marking the beginning of Aur-P2.
+This could explain why we see weaker connections in the Evolved Aurignacian, as expansion results in less interconnection.
 
@@
 
+The observed social dynamics of the phases of the Aurignacian technocomplex correspond with the existing data on the environmental and social conditions of the Upper Paleolithic.
+This tells us that the usage of artwork as a proxy for cultural difference is an effective method for understanding the social dynamics of prehistoric humans.
+
 # References

### 2025-03-14 endofquarterpaper.qmd

@@
 Our data comes from SignBase 1.0, an open-access collection of geometric signs on mobile objects from Upper Paleolithic Europe.
+SignBase 1.0 is a record of objects noting only the presence or absence of each sign, without recording the quantity or sequence of individual signs on a given object.
 We used the identification and categorization of the sign types provided by SignBase's curators [@dutkiewicz2020].
@@
 #| label: fig-site-map
-#| fig-cap: "Map of sites containing geometric signs on mobile objects from Upper Paleolithic Europe, sampled from the SignBase dataset, divided by Aurignacian phases"
+#| fig-cap: "A: Map of sites containing geometric signs on mobile objects from Upper Paleolithic Europe, sampled from the SignBase dataset, divided by Aurignacian phases. B: "
 
@@
 
-plot_grid(sign_map, signcount_barplot, ncol=1, align = "v", scale = c(1, 0.8))
+plot_grid(sign_map, signcount_barplot, ncol=1, align = "hv", scale = c(1, 0.8), labels = "AUTO")
 
@@
 
+The distribution of objects across sites is highly skewed, with the median number of objects as `r artifact_count_median`, but Vogelherd has `r nvogel`.
+To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites.
+A Mantel test between a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types indicated that there was no significant correlation between object abundance and sign type distribution (Mantel statistic R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).
+
 Ages of the objects were determined by calibrating the uncalibrated ages provided by SignBase, using the rcarbon package [@Crema_Bevan_2021].
 The calibrated radiocarbon dates of the sampled data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years.
-We assigned each artifact to a chronological phase following the date ranges recorded in @tbl-aur-table.
-The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` signs, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` signs, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` signs, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` signs.
-
-SignBase 1.0 is a record of objects noting only the presence or absence of each sign, without recording the quantity or sequence of individual signs on a given object.
-The distribution of objects across sites is highly skewed, with Vogelherd having `r nvogel`, but the median number of objects is `r artifact_count_median`.
-To minimize the influence of this skewed distribution, we used presence/absence data for each sign at each site, reducing the effect of variation in object count between sites.
-To make sure that the exclusion of abundance would not significantly impact our results, we ran a Mantel tests both between a Euclidean object abundance distance matrix and a Jaccard distance matrix of the sign types [@Lycett_2019] (p = `r p_value_object_abundance_correlation`, Mantel statistic R = `r r_value_object_abundance_correlation`), and a Eudclidean sign abundance distance matrix and a Jaccard sign type distance matrix (p = `r p_value_sign_abundance_correlation`, Mantel statistic R = `r r_value_sign_abundance_correlation`) and found that neither were statistically significant.
-Another Mantel test was run between a Euclidean distance matrix for dates and the Jaccard sign type matrix in order to determine whether there was a definitive correlation between distance in time and distribution of sign types, finding a statistically significant correlation (p = `r chron_p`, Mantel statistic R = `r chron_r`).
-This confirms that a meaningful change in sign types occurs over time.
-
-{run permanova test for chronology}
+We assigned each object to a chronological phase following the date ranges recorded in @tbl-aur-table.
+The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` sign types, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` sign types, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` sign types, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` sign types.
+A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (Mantel statistic R = `r chron_r`, p = `r chron_p`).
+This confirms that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.
 
@@
 Seriation analysis is the arrangement of data into a linear order to reveal patterns [@Hahsler_Hornik_Buchta].
-We used the Browner-Kile seriation algorithm [@Brower_Kile_1988] from the seriation package [@seriation_pckge], in which a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal.
-Network analysis represents and quantifies the relationship between nodes [@Mills_2017], in this case, sites.
+We used the Brower-Kile [-@Brower_Kile_1988] seriation algorithm from the seriation package [@seriation_pckge], in which a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal.
+To investigate connectivity among clusters of sites in the seriation solution, we used network analysis to represent and quantify the relationship between sites [@Mills_2017].
 In our network analysis, each node is a site, and each edge is a measure of sign type distribution similarity.
@@
 PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017].
-The results are expressed by an R-statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, and a p-value, which quantifies whether the groups are significantly different (we use an alpha value of 0.05 for the threshold of statistical significance).
+The results are expressed by an R-statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, and a p-value, which quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
 The modularity test and perMANOVA were performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.
@@
 The isolation-by-distance framework proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019].
-If cultural difference can be explained solely by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases.
-To test this, we determined the correlation between sign distribution and geography using Mantel tests, via the vegan package [@Vegan2001].
-The Mantel tests were run with a geographic distance matrix and a Jaccard similarity matrix.
+If cultural difference can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases.
+To test this, we determined the correlation between sign distribution and geography using Mantel tests, run with a geographic distance matrix and a Jaccard similarity matrix.
 The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value.
 The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation.
-The p-value quantifies statistical significance of the correlation between the two matrices, with p = 0.05 serving as the threshold for statistical significance.
+The p-value quantifies the probability of obtaining the observed correlation if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
+The Mantel tests were performed using the vegan package [@Vegan2001]
 
@@
 
-For each of the phases, seriation identifies at least two groups, one with a minimal range of signs, and one with a broader range.
-We also see that the notch and line signs are the most prevalent signs across all time periods.
-In the Transitional phase, we see some variety within groups, with the first consisting of Fumane, Pod Hradem, and Geissenklösterle, two of which only have the "notch" sign, and one of which has the notch and circumnotch, and the other consisting of Labeko Koba, Hohle Fels, and El Castillo, which have a broader diversity of signs.
-In the Proto-Aurignacian, we see an expansion in both sites and range of signs, as well as movement to new sites, with little overlap in site names.
-The first group consists of Abri Pataud, Gatzarria, Hohlenstein-Stadel, and Mladeč, three of which only have the notch and line signs, and one which also has a dashline sign, and the other consists of Geissenklösterle, Spy, Vogelherd, Grotte du Renne, La Ferrassie, and Les Cottés.
-In the second group, we see fairly consistent patterning of a diverse range of signs, although Les Cottés varies a little.
-In the Early Aurignacian, we again see movement to new sites with no overlap from the Proto-Aurignacian, as well as increased diversity of patterns within the groups.
-The first group consists of Grottes de Fonds-de-Forêt (radnotch and notch), Riparo Bombrini (notch), and Grotte de la Verpillière I (notch and line), The second consists of Solutré, Hohle Fels, Castanet, Cellier, Blanchard, and Trou al'Wesse, which contain a wide range of signs in a variety of combinations.
-Lastly, in the Evolved Aurignacian, we see slightly more consistent groups, again with movement to new sites with no overlap from the Early Aurignacian.
-The first group consists of Les Rois, Gargas, La Viña, Sirgenstein Cave, and Vindija Cave, covering the line, notch, and and obline signs.
-The second consists of Trou Magrite, Bockstein-Törle, Vogelherd, and Hohle Fels.
-Interestingly, this is the only phase to have more sites in the group with a minimal range of signs, than in the group with a more diverse range.
+For each of the phases, seriation identifies at least two groups, one with a restricted range of sign types, and one with a broader range.
+The notch and line signs are the most prevalent sign types across all time periods.
+In the Transitional phase, there is a small range of sign types.
+The restricted group consists of Fumane, Pod Hradem, and Geissenklösterle, two of which only have a notch, and one of which has a notch and circumnotch.
+The broader group consists of Labeko Koba, Hohle Fels, and El Castillo.
+In the Proto-Aurignacian we see an expansion in both the number of sites and range of signs, with no overlap in sites from the Transitional phase, except for Geissenklösterle, which is present in both phases.
+The restricted group consists of Abri Pataud, Gatzarria, Hohlenstein-Stadel, and Mladeč, three of which only have notches and lines, and one also has a dashline.
+The broader group consists of Geissenklösterle, Spy, Vogelherd, Grotte du Renne, La Ferrassie, and Les Cottés, where we see a more consistent pattern of a more diverse range of signs, compared to the broader group of the Transitional phase.
+In the Early Aurignacian, we see appearance of completely new sites with no overlap with the Proto-Aurignacian sites, as well as increased diversity of patterns within the groups.
+The restricted group consists of Grottes de Fonds-de-Forêt (radnotch and notch), Riparo Bombrini (notch), and Grotte de la Verpillière I (notch and line), The broader group consists of Solutré, Hohle Fels, Castanet, Cellier, Blanchard, and Trou al'Wesse, which contain a wide range of signs in a variety of combinations.
+Lastly, in the Evolved Aurignacian, we see slightly more consistent groups, again with new sites with no overlap with the Early Aurignacian sites, except for Hohle Fels, which also appears in the Transitional phase.
+Vogelherd also reappears, having appeared earlier in the Proto-Aurignacian.The restricted group consists of Les Rois, Gargas, La Viña, Sirgenstein Cave, and Vindija Cave, covering the line, notch, and and obline signs.
+The broader group consists of Trou Magrite, Bockstein-Törle, Vogelherd, and Hohle Fels.
+Overall, across the four phases, the number of sites in the restricted group tends to be less than or equal to the number of sites in the broader group, with the exception of the final Evolved phase.
 

### 2025-03-14 endofquarterpaper.qmd
update table caption

@@
 #| label: tbl-aur-table
-#| tbl-cap: "Table of data on Aurignacian periods [@Banks_dErrico_Zilhao_2013;  @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021; @Chu_Richter_2020; @Fletcher_Sánchez]"
+#| tbl-cap: "Summary of chronological, cultural and environmental characteristics of Aurignacian periods [@Banks_dErrico_Zilhao_2013;  @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021; @Chu_Richter_2020; @Fletcher_Sánchez]"
 #| tbl-cap-location: bottom

### 2025-03-15 endofquarterpaper.qmd
fix some cosmetic issues

@@
 The Aurignacian period spans roughly 43 - 30 k BP, and is largely characterized by material culture indicators, such as carinated scrapers, small-flake tools and split-based bone and ivory tools, as well as an ongoing evolutionary shift from Neanderthals to early modern humans [@Chu_Richter_2020].
-The Aurignacian can be divided into phases, again mainly distinguished by shifts in material culture; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015], as can be seen in @fig-aur-table.
+The Aurignacian can be divided into phases which are mainly distinguished by shifts in material culture; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015], as can be seen in @tbl-aur-table.
 @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main phases, Aur-P1 (Proto/Early), which corresponds with the time period of 43–37 k BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods 37-32 k BP.
@@
 kable(aur_table)
-
 ```
@@
                       format = "f")
+
 date_min <- signbase_years_cal %>% 
@@
 
+# BM: a little simpler and less typings using %in% 
 trans_unique_data <- trans_unique_data %>% 
-  mutate(group = case_when((site_name == "Pod Hradem"|
-                              site_name == "Fumane" |
-                              site_name == "Geissenklösterle") ~ "1",
-                           (site_name == "El Castillo" | 
-                              site_name == "Hohle Fels" | 
-                              site_name == "Labeko Koba") ~ "2"))
+  mutate(group = case_when(site_name %in% c("Pod Hradem", "Fumane", "Geissenklösterle") ~ "1",
+                           site_name %in% c("El Castillo", "Hohle Fels", "Labeko Koba") ~ "2"))
 
@@
 artifact_count_median <- median(artifact_count_df$artifact_count)
-
 ```
@@
 
-object_euc_data <- time_unique_data_full%>%
-  dplyr::select(id, nobjects) %>% 
+# BM: not sure why the id column was also selected here? I moved it to rownames
+object_euc_data <- time_unique_data_full %>%
+    column_to_rownames("id") %>% 
+    dplyr::select(nobjects) %>% 
     dist(method = "euclidean")
 
-jac_full <- vegan::vegdist(artifact_data, method = "jaccard", binary = TRUE)
+# BM: Ensure binary data as input to jaccard distance
+artifact_data_bin <- 
+artifact_data %>% 
+  mutate_all(~ as.numeric(. > 0))
+
+jac_full <- vegan::vegdist(artifact_data_bin, method = "jaccard", binary = TRUE)
 
@@
 
-##mantel chronological
+## mantel chronological
 chron_df <- signbase_full_clean %>% 
@@
   dplyr::select(line:star) %>% 
+  # BM: Ensure binary data as input to jaccard distance
+  mutate_all(~ as.numeric(. > 0)) %>% 
   vegdist("jaccard")
@@
 
-##chron permanova
-perman_chron <- adonis2(jac_full~as.factor(time_unique_data_full$time_period), 
+## chron permanova
+perman_chron <- adonis2(jac_full ~ as.factor(time_unique_data_full$time_period), 
                     method = "jaccard",
                     sqrt.dist = TRUE)
+
 perman_chron <- perman_chron %>% 
@@
 
-Our data comes from SignBase 1.0, an open-access collection of geometric signs on mobile objects from Upper Paleolithic Europe.
-SignBase 1.0 is a record of objects noting only the presence or absence of each sign, without recording the quantity or sequence of individual signs on a given object.
-We used the identification and categorization of the sign types provided by SignBase's curators [@dutkiewicz2020].
+Our data come from SignBase 1.0, an open-access collection of geometric signs on mobile objects from Upper Paleolithic Europe.
+SignBase 1.0 is a record of objects noting only the presence or absence of each sign on the object, without details of the quantity or sequence of individual signs on a given object.
+We used the identification of sign types provided by SignBase's curators [@dutkiewicz2020].
 The data set consists of `r nartifacts` objects found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign types.
@@
 #| label: fig-site-map
-#| fig-cap: "A: Map of sites containing geometric signs on mobile objects from Upper Paleolithic Europe, sampled from the SignBase dataset, divided by Aurignacian phases. B: "
+#| fig-cap: "A: Map of sites in each Aurignacian phase that contain geometric signs on mobile objects.
+B: Abundance of signs over time, phases indicated by colours.
+C.
+Distribution of object count by site"
+#| fig-height: 10
 
@@
 
-signbase_full_clean$time_period <- factor(signbase_full_clean$time_period, 
-                                            levels = c("transitional", "proto_aurignacian", 
-                                                       "early_aurignacian", "evolved_aurignacian"), 
-                                            labels = c("Transitional", "Proto-Aurignacian",
-                                                       "Early Aurignacian", "Evolved Aurignacian"))
+signbase_full_clean$time_period <-
+  factor(
+    signbase_full_clean$time_period,
+    levels = c(
+      "transitional",
+      "proto_aurignacian",
+      "early_aurignacian",
+      "evolved_aurignacian"
+    ),
+    labels = c(
+      "Transitional",
+      "Proto-Aurignacian",
+      "Early Aurignacian",
+      "Evolved Aurignacian"
+    )
+  )
 
@@
   theme(legend.key.size = unit(0.5, "cm")) +
-  scale_fill_discrete(guide = "none")
+  theme_minimal() +
+  scale_fill_discrete(guide = "none") +
+  xlab("Years cal BP")
+
+objectcount_barplot <- 
+ggplot(artifact_count_df) +
+  aes(reorder(site_name, artifact_count),
+      artifact_count) +
+  geom_col() +
+  coord_flip() +
+    theme_minimal() +
+    xlab("") +
+    ylab("Object count")
 
@@
 
-plot_grid(sign_map, signcount_barplot, ncol=1, align = "hv", scale = c(1, 0.8), labels = "AUTO")
+maps_bar_plots <- 
+plot_grid(sign_map,
+          plot_grid(signcount_barplot,
+                    objectcount_barplot,
+                    rel_widths = c(1, 0.5),
+                    labels = c("B", "C")), 
+          ncol = 1, 
+          scale = c(1, 1),
+          rel_heights = c(1, 0.5),
+          align = "vh", 
+          labels = "A",
+          axis = "lr")
+
+ggsave(maps_bar_plots, 
+      filename =  "map-and_hist-plot.png",
+       h = 10, w = 10,
+       bg = "white")
+
+knitr::include_graphics("map-and_hist-plot.png")
 
@@
 To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites.
-A Mantel test between a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types indicated that there was no significant correlation between object abundance and sign type distribution (Mantel statistic R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).
+A Mantel test between a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).
 
@@
 The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` sign types, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` sign types, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` sign types, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` sign types.
-A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (Mantel statistic R = `r chron_r`, p = `r chron_p`).
+A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (R = `r chron_r`, p = `r chron_p`).
 This confirms that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.
@@
 PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017].
-The results are expressed by an R-statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, and a p-value, which quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
+The results are expressed by an R statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, and a p-value, which quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
 The modularity test and perMANOVA were performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.
@@
 
-mplot <- function(x, ..., fill_colour = "black", title = NULL, row_labels = NULL, col_labels = NULL, country_colors = NULL) {
+mplot <- function(x, ..., fill_colour = "black", title = NULL, row_labels = NULL, col_labels = NULL, palette_colors = NULL) {
   d <- data.frame(
@@
 
-  if (!is.null(country_colors)) {
-    d$country_color <- rep(country_colors, ncol(x))
+  if (!is.null(palette_colors)) {
+    d$palette_colors <- rep(palette_colors, ncol(x))
   }
@@
   p <- ggplot(d, aes(x, y)) +
-    geom_tile(col = "grey50", linewidth = 0.5, aes(fill = ifelse(z == 1, country_color, "white"))) +
+    geom_tile(col = "grey50", 
+              linewidth = 0.5, 
+              aes(fill = ifelse(z == 1, palette_colors, "white"))) +
     scale_fill_identity() +  
@@
     labs(x = "", y = "") +
-    rotate_x_text(angle = 90, hjust = NULL, vjust = NULL)
+    rotate_x_text(angle = 90, hjust = 0, vjust = 0.5)
 
@@
 
-increase_focus <- function(x, country = NULL) {
+increase_focus <- function(x, group = NULL) {
   mcp <- apply(
@@
 
-  if (!is.null(country)) {
-    country <- country[row_order]
+  if (!is.null(group)) {
+    group <- group[row_order]
   }
@@
 
-  list(matrix = x, country = country)
+  list(matrix = x, group = group)
 }
 
-concentrate <- function(x, country = NULL, max_iter = 100) {
+concentrate <- function(x, group = NULL, max_iter = 100) {
   old <- x
@@
   while (not_identical && iter < max_iter) {
-    result <- increase_focus(old, country)
+    result <- increase_focus(old, group)
     new <- result$matrix
-    country <- result$country
+    group <- result$group
     not_identical <- !identical(old, new)
@@
 
-  list(matrix = new, country = country)
+  list(matrix = new, group = group)
 }
 
-produce_clusters <- function(artifact_data, artifact_data_unique, method = "seriation"){
+produce_clusters <- function(artifact_data, artifact_data_unique, method = "seriation", color_palette = c("#8DD3C7", "#FFFFB3")){
+
   artifact_data <- artifact_data %>% 
-    select_if(~ !is.numeric(.) || sum(.) != 0)
+    select_if(~ !is.numeric(.) || sum(.) != 0) %>% 
+    mutate_all(~ as.numeric(. > 0))
+
   jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
@@
   if(method == "seriation"){
-  artifact_matrix <- as.matrix(artifact_data %>% 
-                                 mutate(across(everything(), ~replace(., . > 1, 1))))
+  artifact_matrix <- as.matrix(artifact_data)
   rownames(artifact_matrix) <- rownames(artifact_data)
 
-  countries <- artifact_data_unique$group
-  unique_countries <- unique(countries)
-  color_palette <-  RColorBrewer::brewer.pal(length(unique_countries), "Set3")
-  row_colors <- setNames(color_palette, unique_countries)
-  result <- concentrate(artifact_matrix, country = countries)
+  group <- artifact_data_unique$group
+  unique_group <- unique(group)
+ # color_palette <-  RColorBrewer::brewer.pal(length(unique_group), "Set3")
+  row_colors <- setNames(color_palette, unique_group)
+  result <- concentrate(artifact_matrix, group = group)
   concentrated_matrix <- result$matrix
-  reordered_countries <- result$country
-  countries_colors <- row_colors[countries]
-  country_colors <- row_colors[reordered_countries]
+  reordered_group <- result$group
+  group_colors <- row_colors[reordered_group]
 
   final_plot <- mplot(concentrated_matrix,
-    country_colors = country_colors)}
+    palette_colors = group_colors)
+  }
 
@@
   set.seed(500)
-  jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
+  jac <- vegdist(artifact_data %>% mutate_all(~ as.numeric(. > 0)), "jaccard", binary = TRUE)
   dm <- as.matrix(jac)
@@
 
-  final_table <- data_frame("modularity" = as_group, 
-                            "permanova r" = perman_r,
-                            "permanova p" = perman_p,
-                            "mantel r" = r_value_distance_correlation, 
-                            "mantel p" = p_value_distance_correlation)
+  final_table <- data_frame("Modularity" = as_group, 
+                            "PerMANOVA r" = perman_r,
+                            "PerMANOVA p" = perman_p,
+                            "Mantel R" = r_value_distance_correlation, 
+                            "Mantel p" = p_value_distance_correlation)
 
@@
 trans_network <- function(artifact_data){
-    jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
+    jac <- vegdist(artifact_data %>%  mutate_all(~ as.numeric(. > 0)), "jaccard", binary = TRUE)
     dm <- as.matrix(jac)
@@
 
-seriation_trans <- produce_clusters(trans_artifact_data, trans_unique_data, method = "seriation")
-seriation_proto <- produce_clusters(proto_artifact_data, proto_unique_data, method = "seriation")
-seriation_early <- produce_clusters(early_artifact_data, early_unique_data, method = "seriation")
-seriation_evolved <- produce_clusters(evolved_artifact_data, evolved_unique_data, method = "seriation")
+seriation_trans <- 
+  produce_clusters(trans_artifact_data, 
+                   trans_unique_data, 
+                   method = "seriation")
 
-plot_grid(seriation_trans,seriation_proto, seriation_early , seriation_evolved, 
-          ncol = 2, align = "hv", 
+seriation_proto <- 
+  produce_clusters(proto_artifact_data, 
+                   proto_unique_data, 
+                   # BM: not sure why this one gets the 
+                   # colour differently from the others
+                   color_palette = c("#FFFFB3", "#8DD3C7"),
+                   method = "seriation")
+
+seriation_early <- 
+  produce_clusters(early_artifact_data, 
+                   early_unique_data, 
+                   method = "seriation")
+
+seriation_evolved <- 
+  produce_clusters(evolved_artifact_data, 
+                   evolved_unique_data, 
+                   method = "seriation")
+
+plot_grid(seriation_trans,
+          seriation_proto, 
+          seriation_early , 
+          seriation_evolved, 
+          ncol = 2, 
+          align = "hv", 
           labels = c("Transitional", "Proto", "Early", "Evolved"), label_size = 15, 
-          scale = c(0.85, 1, 1, 1))
+          scale = c(1, 1, 1, 1))
 
@@
 Overall, across the four phases, the number of sites in the restricted group tends to be less than or equal to the number of sites in the broader group, with the exception of the final Evolved phase.
+<!-- BM: I reckon we need to say something about the signs also, not just the sites -->
 
@@
 
-```{r fig.height= 10, fig.width=10}
-#| label: fig-group-sign
-#| fig-cap: "Bar plot of signs by time period, colored by group"
-
-full_data_long <- signbase_sf %>% 
-  pivot_longer(cols = line:star) %>% 
-  filter(value != 0)
-
-ggplot(full_data_long) +
-  aes(x = name, fill = group) +
-  geom_bar() +
-  rotate_x_text(angle = 90, hjust = NULL, vjust = NULL) + 
-  facet_wrap(~time_period, nrow = 4)
-
-```
-
 ## Cultural Group Strength
@@
 #| tbl-cap-location: bottom
-##Modularity test
+
+## Modularity test
 trans_strength <- strength_function(trans_artifact_data, trans_unique_data)
@@
 full_table <- rbind(trans_strength, proto_strength, early_strength, evolved_strength) %>% 
-  mutate(`time period` = c("Transitional", "Proto-Aurignacian",
-                                                       "Early Aurignacian", "Evolved Aurignacian"), .before = 1)
+  mutate(`Aurignacian phase` = c("Transitional", 
+                           "Proto-Aurignacian",
+                           "Early Aurignacian", 
+                           "Evolved Aurignacian"), 
+         .before = 1)
 
 strength_table <- full_table %>% 
-  dplyr::select(`time period`, modularity, `permanova r`, `permanova p`)
+  dplyr::select(`Aurignacian phase`, Modularity, `PerMANOVA r`, `PerMANOVA p`) %>% 
+  mutate(across(where(is.numeric), ~ round(., 3)))
 
@@
 proto_modularity <- proto_strength %>% 
-  dplyr::select(modularity)
+  dplyr::select(Modularity)
 evolved_modularity <- evolved_strength %>% 
-  dplyr::select(modularity)
+  dplyr::select(Modularity)
 
 trans_perm_r <- trans_strength %>% 
-  dplyr::select(`permanova r`)
+  dplyr::select(`PerMANOVA r`)
 early_perm_r <- early_strength %>% 
-  dplyr::select(`permanova r`)
+  dplyr::select(`PerMANOVA r`)
 
 trans_perm_p <- trans_strength %>% 
-  dplyr::select(`permanova p`)
+  dplyr::select(`PerMANOVA p`)
 ```
@@
 mantel_table <- full_table %>% 
-  dplyr::select(`time period`, `mantel r`, `mantel p`)
+  dplyr::select(`Aurignacian phase`, `Mantel R`, `Mantel p`)
 
@@
 proto_mantel_r <- proto_strength %>% 
-  dplyr::select(`mantel r`)
+  dplyr::select(`Mantel R`)
 
 proto_mantel_p <- proto_strength %>% 
-  dplyr::select(`mantel p`)
+  dplyr::select(`Mantel p`)
 
 early_mantel_r <- early_strength %>% 
-  dplyr::select(`mantel r`)
+  dplyr::select(`Mantel R`)
 
 early_mantel_p <- early_strength %>% 
-  dplyr::select(`mantel p`)
+  dplyr::select(`Mantel p`)
 

### 2025-03-15 endofquarterpaper.qmd
editing up to results: Cultural Group Strength

@@
 
-Network analysis shows similar groups to the ones identified in @fig-seriation, but it also shows evolving inter-connectivity between groups.
-In the Transitional phase, the network analysis shows two clearly defined groups, one consisting of Geissenklösterle, Fumane, and Pod Hradem, and the other consisting of Hohle Fels, El Castillo, and Labeko Koba, with no connections between the groups.
-The Proto-Aurignacian phase also shows two clearly defined groups, one consisting of Geissenklösterle, Spy, Vogelherd, and Grotte du Renne, and the other consisting of Mladeč, Hohlenstein-Stadel, Gatzarria, and Abri Pataud.
-However, it also displays some sites that are less clearly connected, with La Ferrassie being squarely in between both groups with equal connections to both, and Les Cottés lacking a connection to any of the groups.
-The Early Aurignacian shows more strongly inter-connected groups - although the shape still suggests two main groups - with Grotte de la Verpillière I in particular having multiple strong connections to both of the groups.
-Lastly, the Evolved Aurignacian again shows two main groups, although group inner-connectivity has decreased, one consisting of Hohle Fels, Trou Magrite, Vogelherd, and Bockstein-Törle, and the other consisting of Gargas, La Viña, Sirgenstein Cave, Vindija Cave, and Les Rois.
-While inter-connectivity between groups is still present, with connections between Trou Magrite, Gargas, and Vindija Cave, and Bockstein-Törle and Gargas, the strength of inter-connectivity is weaker than in the Early Aurignacian.
-
-In identifying the groups, we referenced both the seriation and network analyses.
-The results of the above tests allowed us to find two main groups of sites within each phase, as shown in @tbl-group-table.
-The geographic distributions of the groups are shown in @fig-group-map, and the sign makeup of each group is shown in @fig-group-sign.
-
-```{r}
-#| label: tbl-group-table
-#| tbl-cap: "Table of the sites by group per phase"
-#| tbl-cap-location: bottom
-group_table <- read_excel("data/site-groups.xlsx")
-
-kable(group_table)
-
-```
-
-```{r fig.height= 7, fig.width=10}
-#| label: fig-group-map
-#| fig-cap: "Map of sites faceted by time period and colored by group number"
-
-ggplot(Europe) +
-  geom_sf() +
-  geom_sf(data = signbase_sf, aes(color = group)) +
-  coord_sf(xlim = c(-10,30), 
-           ylim = c(35,53), 
-           expand = FALSE) +
-  geom_text_repel(data = signbase_sf,
-                  aes(x = longitude ,
-                      y = latitude,
-                      label = site_name),
-                  max.overlaps = 30,
-                  size = 2,
-                  bg.color = "white",
-                  bg.r = 0.25) +
-  theme_minimal() +
-  facet_wrap(~time_period) +
-  theme(axis.text = element_blank(),
-        axis.title = element_blank())
-
-```
+The results of the network analysis supports the restricted-broad groupings identified in @fig-seriation, and provides additional insights into the evolution of connectivity between groups.
+In the Transitional phase, the network analysis reveals two distinct and disconnected groups.
+The restricted group consists of Geissenklösterle, Fumane, and Pod Hradem, and the broad group consisting of Hohle Fels, El Castillo, and Labeko Koba, with no connections between the groups.
+The Proto-Aurignacian phase also shows two clearly defined groups, the broad group consisting of Geissenklösterle, Spy, Vogelherd, and Grotte du Renne, and the restricted group consisting of Mladeč, Hohlenstein-Stadel, Gatzarria, and Abri Pataud.
+However, it also displays some connectivity between the two groups of sites, with La Ferrassie connected to both groups.
+Les Cottés stands out with no connections to either group.
+The Early Aurignacian shows more strongly inter-connected groups -- although the shape still suggests two main groups -- with Grotte de la Verpillière I in particular having multiple strong connections to both groups.
+Lastly, the Evolved Aurignacian again shows two main groups, although intra-group connectivity has decreased, with the broad group consisting of Hohle Fels, Trou Magrite, Vogelherd, and Bockstein-Törle, and the restricted group consisting of Gargas, La Viña, Sirgenstein Cave, Vindija Cave, and Les Rois.
+As for the Early Aurignacian,inter-connectivity between groups continues to be evident in the Evolved Aurignacian, with connections between Trou Magrite, Gargas, and Vindija Cave, and Bockstein-Törle and Gargas, however the strength of inter-connectivity is weaker than in the Early Aurignacian.
 
@@
 
+<!-- BM: I reckon here might be the place to make some comments about geographic distribution of the groups and boundaries, and reference the map below -->
+
+```{r fig.height= 7, fig.width=10}
+#| label: fig-group-map
+#| fig-cap: "Map of sites faceted by time period and colored by group number"
+
+ggplot(Europe) +
+  geom_sf() +
+  geom_sf(data = signbase_sf, aes(color = group)) +
+  coord_sf(xlim = c(-10,30), 
+           ylim = c(35,53), 
+           expand = FALSE) +
+  geom_text_repel(data = signbase_sf,
+                  aes(x = longitude ,
+                      y = latitude,
+                      label = site_name),
+                  max.overlaps = 30,
+                  size = 2,
+                  bg.color = "white",
+                  bg.r = 0.25) +
+  theme_minimal() +
+  facet_wrap(~time_period) +
+  theme(axis.text = element_blank(),
+        axis.title = element_blank())
+
+```
+
 # Discussion

### 2025-03-15 endofquarterpaper.qmd
stash

--- 2025-03-14 endofquarterpaper.qmd
@@
 #| label: tbl-aur-table
-#| tbl-cap: "Summary of chronological, cultural and environmental characteristics of Aurignacian periods [@Banks_dErrico_Zilhao_2013;  @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021; @Chu_Richter_2020; @Fletcher_Sánchez]"
+#| tbl-cap: "Table of data on Aurignacian periods [@Banks_dErrico_Zilhao_2013;  @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021; @Chu_Richter_2020; @Fletcher_Sánchez]"
 #| tbl-cap-location: bottom
@@
 The results as displayed in @fig-seriation show that each phase can be divided into two main cultural groups, with diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.
-@fig-neighbor-net shows that group inter-connectivity and inner-connectivity varies from phase to phase, with weak inner group connections and no between-group connections in the Transitional phase, to stronger inner connections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
-The modularity tests, as shown in @tbl-strength-table, tell us that the groupings are strongest in the Transitional period and weakest in the Early Aurignacian, and the perMANOVA tests tell us that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian experiences the least.
+
+@fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak inner group connections and no between-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
+The modularity tests, as shown in @tbl-strength-table, show that the groupings are strongest in the Transitional period and weakest in the Early Aurignacian, and the perMANOVA tests tell us that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian experiences the least.
 The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian experiences the highest correlation.
+
+Map shows that geographic boundaries do not readily correspond to the group distributions and Mantel test results
 
@@
 These results are supported by the existing data on the environmental and biological change established during these phases.
-@Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of 1Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Aur-P1 settlement being reached around 41 k BP.
-This could provide an answer as to why the Transitional phase appears sparsely populated with little to no inter-connectivity, and the Proto-Aurignacian appears to be more densely populated with stronger cultural groups.
+@Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Proto/Early Aurignacian (Aur-P1) settlement being reached around 41 k BP.
+This rapid expansion could be what stimulated the emergence of sign sue as a marker of group membership.
+
 Additionally @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations.
-This could be a reason as to why we see a large increase in both inter-connectivity and inner-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance.
-Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Aur-P2, theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions.
-@Shao_Wegener_Klein_Schmidt_Weniger_2024 also find an increase in expansion marking the beginning of Aur-P2.
-This could explain why we see weaker connections in the Evolved Aurignacian, as expansion results in less interconnection.
+This likely explains why we see a large increase in both inter-connectivity and inner-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance.
+Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian (Aur-P2) theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions.
+@Shao_Wegener_Klein_Schmidt_Weniger_2024 also find an increase in expansion marking the beginning of the Evolved Aurignacian.
+Geographic expansion resulting in less group interconnection could explain why we see weaker connections in the Evolved Aurignacian.
 

### 2025-03-16 endofquarterpaper.qmd
added paragrah analyzing Fig 4

--- 2025-03-15 endofquarterpaper.qmd
@@
 ```{r}
-
 lat_long_df <- signbase_full_clean %>% 
@@
 
-# BM: a little simpler and less typings using %in% 
 trans_unique_data <- trans_unique_data %>% 
@@
 proto_unique_data <- proto_unique_data %>% 
-  mutate(group = case_when((site_name == "Gatzarria" | 
-                              site_name == "Hohlenstein-Stadel" | 
-                              site_name == "Mladeč" | 
-                              site_name == "Abri Pataud") ~ "1",
-                           (site_name == "Geissenklösterle" |
-                              site_name == "Grotte du Renne" |
-                              site_name == "Vogelherd" |
-                              site_name == "Spy" |
-                              site_name == "La Ferrassie"|
-                               site_name == "Les Cottés") ~ "2"))
+  mutate(group = case_when(site_name %in% c("Gatzarria" , "Hohlenstein-Stadel", "Mladeč", "Abri Pataud") ~ "1",
+                           site_name %in%  c("Geissenklösterle", "Grotte du Renne", 
+                                             "Vogelherd", "Spy" ,"La Ferrassie", "Les Cottés") ~ "2"))
 
 early_unique_data <- early_unique_data %>% 
-  mutate(group = case_when((site_name == "Grottes de Fonds-de-Forêt" |
-                              site_name == "Riparo Bombrini"|
-                              site_name == "Grotte de la Verpillière I"|
-                              site_name == "Vindija Cave") ~ "1",
-                           (site_name == "Hohle Fels" | 
-                              site_name == "Solutré" |
-                              site_name == "Castanet"|
-                              site_name == "Cellier" | 
-                              site_name == "Blanchard"|
-                              site_name == "Trou al'Wesse") ~ "2"))
+  mutate(group = case_when(site_name %in%  c("Grottes de Fonds-de-Forêt", "Riparo Bombrini", 
+                                             "Grotte de la Verpillière I", "Vindija Cave") ~ "1",
+                           site_name %in% c("Hohle Fels", "Solutré", "Castanet", "Cellier", "Blanchard", "Trou al'Wesse") ~ "2"))
 
 evolved_unique_data <- evolved_unique_data %>% 
-  mutate(group = case_when((site_name == "Les Rois" |
-                              site_name == "Gargas" |
-                              site_name == "La Viña"|
-                              site_name ==  "Sirgenstein Cave"|
-                              site_name == "Vindija Cave") ~ "1",
-                           (site_name == "Bockstein-Törle"|
-                              site_name == "Vogelherd" |
-                              site_name == "Hohle Fels"|
-                              site_name == "Trou Magrite") ~ "2"))
+  mutate(group = case_when(site_name %in%  c("Les Rois", "Gargas", "La Viña", "Sirgenstein Cave", "Vindija Cave") ~ "1",
+                           site_name %in% c("Bockstein-Törle", "Vogelherd", "Hohle Fels", "Trou Magrite") ~ "2"))
 
@@
 
-Our data come from SignBase 1.0, an open-access collection of geometric signs on mobile objects from Upper Paleolithic Europe.
+Our data comes from SignBase 1.0, an open-access collection of geometric signs on mobile objects from Upper Paleolithic Europe.
 SignBase 1.0 is a record of objects noting only the presence or absence of each sign on the object, without details of the quantity or sequence of individual signs on a given object.
@@
 PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017].
-The results are expressed by an R statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, and a p-value, which quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
+The results are expressed by an R2 statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, a pseudo-F statistic, which compares the total sum of squared dissimilarities between groups to total sum of squared dissimilarities within groups, with larger pseudo F-statistics indicating more separation between groups, and a p-value, which quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
+
+<!-- SC: I think we need to elaborate on what the null hypothesis is here.
+This blog, https://archetypalecology.wordpress.com/2018/02/21/permutational-multivariate-analysis-of-variance-permanova-in-r-preliminary/, does a pretty good job of explaining what exactly the p-value means, which is the proportion of F-statistics for the permuted tests that are greater than or equal to the actual F-statistic.-->
+
 The modularity test and perMANOVA were performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.
@@
   as_group <- modularity(modul_matrix, membership = mem)
+  as_group <- round(as_group, 3)
 
@@
     slice_head()
-  perman_r <- perman$R2
-  perman_p <- perman$`Pr(>F)`
+  perman_r <- round(perman$R2, 3)
+  perman_p <- round(perman$`Pr(>F)`, 3)
+  perman_f <- round(perman$F, 3)
 
@@
   final_table <- data_frame("Modularity" = as_group, 
-                            "PerMANOVA r" = perman_r,
+                            "PerMANOVA R2" = perman_r,
+                            "PerMANOVA F" = perman_f,
                             "PerMANOVA p" = perman_p,
@@
 Overall, across the four phases, the number of sites in the restricted group tends to be less than or equal to the number of sites in the broader group, with the exception of the final Evolved phase.
+We also see that simpler signs, like line and notch, tend to be more common in both groups, but tend to primarily make up the restricted groups.
+
 <!-- BM: I reckon we need to say something about the signs also, not just the sites -->
@@
 Lastly, the Evolved Aurignacian again shows two main groups, although intra-group connectivity has decreased, with the broad group consisting of Hohle Fels, Trou Magrite, Vogelherd, and Bockstein-Törle, and the restricted group consisting of Gargas, La Viña, Sirgenstein Cave, Vindija Cave, and Les Rois.
-As for the Early Aurignacian,inter-connectivity between groups continues to be evident in the Evolved Aurignacian, with connections between Trou Magrite, Gargas, and Vindija Cave, and Bockstein-Törle and Gargas, however the strength of inter-connectivity is weaker than in the Early Aurignacian.
+As for the Early Aurignacian, inter-connectivity between groups continues to be evident in the Evolved Aurignacian, with connections between Trou Magrite, Gargas, and Vindija Cave, and Bockstein-Törle and Gargas, however the strength of inter-connectivity is weaker than in the Early Aurignacian.
+
+In identifying the groups, we referenced both the seriation and network analyses.
+The results of the above tests allowed us to find two main groups of sites within each phase, as shown in @tbl-group-table.
+The geographic distributions of the groups are shown in @fig-group-map, and the sign makeup of each group is shown in @fig-group-sign.
+
+```{r}
+ #| label: tbl-group-table
+ #| tbl-cap: "Table of the sites by group per phase"
+ #| tbl-cap-location: bottom
+ group_table <- read_excel("data/site-groups.xlsx")
+
+ kable(group_table)
+```
 
@@
 strength_table <- full_table %>% 
-  dplyr::select(`Aurignacian phase`, Modularity, `PerMANOVA r`, `PerMANOVA p`) %>% 
-  mutate(across(where(is.numeric), ~ round(., 3)))
+  dplyr::select(`Aurignacian phase`, Modularity, `PerMANOVA R2`, `PerMANOVA F`, `PerMANOVA p`)
 
@@
 trans_perm_r <- trans_strength %>% 
-  dplyr::select(`PerMANOVA r`)
+  dplyr::select(`PerMANOVA R2`)
 early_perm_r <- early_strength %>% 
-  dplyr::select(`PerMANOVA r`)
+  dplyr::select(`PerMANOVA R2`)
 
@@
   dplyr::select(`PerMANOVA p`)
+
+trans_perm_f <- trans_strength %>% 
+  dplyr::select(`PerMANOVA F`)
+early_perm_f <- early_strength %>% 
+  dplyr::select(`PerMANOVA F`)
 ```
@@
 
-The R statistic from the perMANOVA tests tell us the proportion of sign variation that can be explained by group membership.
+The R2 statistic from the perMANOVA tests tell us the proportion of sign variation that can be explained by group membership.
 The Transitional period has the highest score of `r trans_perm_r` while the Early Aurignacian has the lowest score of `r early_perm_r`.
 We see a large decline in scores between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian periods.
-The perMANOVA p score quantifies the statistical significance of the difference, with 0.05 as the threshold.
+The pseudo-F statistic quantifies the difference between groups, with the Transitional phase having the highest score of `trans_perm_f`, and the Early Aurignacian having the lowest score of `early_perm_f`.
+Similar to the R2 statistic, we see a decline in scores between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian period.
+The perMANOVA p score quantifies the statistical significance of the pseudo-F statistic, with 0.05 as the threshold.
 The only time period to not meet that threshold is the Transitional period (p = `r trans_perm_p`).
@@
 
-<!-- BM: I reckon here might be the place to make some comments about geographic distribution of the groups and boundaries, and reference the map below -->
-
 ```{r fig.height= 7, fig.width=10}
@@
 
+@fig-group-map shows the geographic distribution of sites as colored by group, demonstrating that geographic boundaries do not readily correspond to the group distributions.
+In the Transitional phase, the sites are widely spread out across Western Europe, and tend to be located fairly closely to the other sites in their groups.
+The exception is Hohle Fels, which, instead of being located further west with the rest of the sites in the broader group, is located further east, closer to the sites in the restricted group, and almost overlapping with Geissenklösterle.
+In the Proto-Aurignacian, we see a slightly more restricted distribution of sites westward but a spread northward, and more overlapping of sites between groups.
+For the restricted group, we interestingly see a divide across the center, with Mladeč and Hohlenstein-Stadel located close together further to the east, and Gatzarria and Abri Pataud located close together further to the west, with a large space between the two clusters.
+The broader sites tend to be more uniformly spread out across the geographic space, without much clustering close together.
+We again see some overlap between Abri Pataud and La Ferrassie, as well as between Geissenklösterle and Hohlenstein-Stadel.
+In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together, incluing a great deal of both intra- and inter-group proximity.
+In the Evolved Aurignacian, we see a wider distribution and more distance between sites.
+In particular, the sites in the reduced group are especially spread out across the space, while the broader sites are slightly more clustered.
+We also see a bit of a geographic divide between groups, with the reduced sites tending to be located further to the west, while the broader sites tend to be located further to the east.
+The exceptions is Sirgenstein Cave, which is located closer to the broader sites, almost overlapping with Vogelherd, Bockstein-Törle, and Hohle Fels.
+
 # Discussion
@@
 The results as displayed in @fig-seriation show that each phase can be divided into two main cultural groups, with diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.
-@fig-neighbor-net shows that group inter-connectivity and inner-connectivity varies from phase to phase, with weak inner group connections and no between-group connections in the Transitional phase, to stronger inner connections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
-The modularity tests, as shown in @tbl-strength-table, tell us that the groupings are strongest in the Transitional period and weakest in the Early Aurignacian, and the perMANOVA tests tell us that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian experiences the least.
+
+@fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
+The modularity tests, as shown in @tbl-strength-table, show that the groupings are strongest in the Transitional period and weakest in the Early Aurignacian, and the perMANOVA tests tell us that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian experiences the least.
 The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian experiences the highest correlation.
@@
 The results suggest that the Transitional phase was sparsely populated, with minimal interaction between loosely connected groups.
-For the Proto-Aurignacian, the results suggest a large expansion in social activity, as well as strongly inner-connected groups with some inter-connectivity present, with geographic distance having the least influence on sign distribution.
-The Early Aurignacian experiences a large surge in inter-connectivity, as well as inner-connectivity strength, with a great deal of connection between groups and an increase in the correlation with geographic distance.
+For the Proto-Aurignacian, the results suggest a large expansion in social activity, as well as strongly intra-connected groups with some inter-connectivity present, with geographic distance having the least influence on sign distribution.
+The Early Aurignacian experiences a large surge in inter-connectivity, as well as intra-connectivity strength, with a great deal of connection between groups and an increase in the correlation with geographic distance.
 The Evolved Aurignacian experiences a decline in connection strength, although it maintains degrees of inter-connectivity.
@@
 These results are supported by the existing data on the environmental and biological change established during these phases.
-@Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of 1Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Aur-P1 settlement being reached around 41 k BP.
-This could provide an answer as to why the Transitional phase appears sparsely populated with little to no inter-connectivity, and the Proto-Aurignacian appears to be more densely populated with stronger cultural groups.
+@Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Proto/Early Aurignacian (Aur-P1) settlement being reached around 41 k BP.
+This rapid expansion could be what stimulated the emergence of signs as a marker of group membership.
 Additionally @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations.
-This could be a reason as to why we see a large increase in both inter-connectivity and inner-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance.
-Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Aur-P2, theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions.
-@Shao_Wegener_Klein_Schmidt_Weniger_2024 also find an increase in expansion marking the beginning of Aur-P2.
-This could explain why we see weaker connections in the Evolved Aurignacian, as expansion results in less interconnection.
+This likely explains why we see a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance.
+Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian (Aur-P2) theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions.
+@Shao_Wegener_Klein_Schmidt_Weniger_2024 also find an increase in expansion marking the beginning of the Evolved Aurignacian.
+Geographic expansion resulting in less group interconnection could explain why we see weaker connections in the Evolved Aurignacian.
 

### 2025-03-21 endofquarterpaper.qmd

@@
 We then test the strength of the clusters using modularity and perMANOVA tests.
-Lastly, we use Mantel tests to evaluate the null hypothesis of isolation-by-distance for each phase.
+All of our programming in this study was conducted in RStudio.
 
@@
 artifact_count_median <- median(artifact_count_df$artifact_count)
+n_material <- c(unique(signbase_full_clean$material))
+
 ```
@@
 These exclusions resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map).
+
+```{r fig.height= 9, fig.width=10}
+```
 
@@
 To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites.
-A Mantel test between a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).
+We used Mantel tests with a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types.
+The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value.
+The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation.
+The p-value quantifies the probability of obtaining the observed correlation if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
+The Mantel tests were performed using the vegan package [@Vegan2001].
+The test indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).
 
@@
 This confirms that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.
-
-## Cultural Group Detection Using Seriation and Network Analysis
-
-Seriation analysis is the arrangement of data into a linear order to reveal patterns [@Hahsler_Hornik_Buchta].
-We used the Brower-Kile [-@Brower_Kile_1988] seriation algorithm from the seriation package [@seriation_pckge], in which a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal.
-To investigate connectivity among clusters of sites in the seriation solution, we used network analysis to represent and quantify the relationship between sites [@Mills_2017].
-In our network analysis, each node is a site, and each edge is a measure of sign type distribution similarity.
-We input the Jaccard similarity matrix for each phase, computed the network statistics, and plotted the results with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths.
-Network analysis was performed using the vegan [@Vegan2001], statnet [@statnet_package] and igraph [@igraph_package] packages.
-
-## Modularity and PerMANOVA for Cultural Group Strength
-
-To evaluate the strength of the groupings found in the seriation and network analyses, we used modularity and perMANOVA tests.
-The modularity test is a community structure network detection method, and quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006; @Newman_Girvan_2004].
-The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings.
-PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017].
-The results are expressed by an R2 statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, a pseudo-F statistic, which compares the total sum of squared dissimilarities between groups to total sum of squared dissimilarities within groups, with larger pseudo F-statistics indicating more separation between groups, and a p-value, which quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
-
-<!-- SC: I think we need to elaborate on what the null hypothesis is here.
-This blog, https://archetypalecology.wordpress.com/2018/02/21/permutational-multivariate-analysis-of-variance-permanova-in-r-preliminary/, does a pretty good job of explaining what exactly the p-value means, which is the proportion of F-statistics for the permuted tests that are greater than or equal to the actual F-statistic.-->
-
-The modularity test and perMANOVA were performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.
-
-## Effects of Geographic Distance Using Mantel Tests
-
-The isolation-by-distance framework proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019].
-If cultural difference can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases.
-To test this, we determined the correlation between sign distribution and geography using Mantel tests, run with a geographic distance matrix and a Jaccard similarity matrix.
-The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value.
-The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation.
-The p-value quantifies the probability of obtaining the observed correlation if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
-The Mantel tests were performed using the vegan package [@Vegan2001]
-
-# Results
-
-## Cultural Group Detection
 
@@
 
+If the distribution of sign types can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases.
+This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019].
+To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geography for each phase.
+We used Mantel tests with a geographic distance matrix and a Jaccard similarity matrix, with the results shown in @tbl-mantel-table.
+None of the phases have a statistically significant correlation between sign type distribution and geographical distance, so the isolation-by-distance hypothesis is rejected.
+
+```{r}
+#| label: tbl-mantel-table
+#| tbl-cap: "Table of the results per phase for the Mantel test between a Jaccard matrix of sign type diversity and a geographic distance matrix"
+#| tbl-cap-location: bottom
+## Mantel test for geography vs signs
+
+trans_strength <- strength_function(trans_artifact_data, trans_unique_data)
+ proto_strength <- strength_function(proto_artifact_data, proto_unique_data)
+ early_strength <- strength_function(early_artifact_data, early_unique_data)
+ evolved_strength <- strength_function(evolved_artifact_data, evolved_unique_data)
+
+ full_table <- rbind(trans_strength, proto_strength, early_strength, evolved_strength) %>% 
+   mutate(`Aurignacian phase` = c("Transitional", 
+                            "Proto-Aurignacian",
+                            "Early Aurignacian", 
+                            "Evolved Aurignacian"), 
+          .before = 1)
+
+mantel_table <- full_table %>% 
+  dplyr::select(`Aurignacian phase`, `Mantel R`, `Mantel p`)
+
+kable(mantel_table)
+
+```
+
+## Cultural Group Detection Using Seriation and Network Analysis
+
+Seriation analysis is the arrangement of data into a linear order to reveal patterns [@Hahsler_Hornik_Buchta].
+We used the Brower-Kile [-@Brower_Kile_1988] seriation algorithm from the seriation package [@seriation_pckge], in which a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal.
+To investigate connectivity among clusters of sites in the seriation solution, we used network analysis to represent and quantify the relationship between sites [@Mills_2017].
+In our network analysis, each node is a site, and each edge is a measure of sign type distribution similarity.
+We input the Jaccard similarity matrix for each phase, computed the network statistics, and plotted the results with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths.
+Network analysis was performed using the vegan [@Vegan2001], statnet [@statnet_package] and igraph [@igraph_package] packages.
+
+## Modularity and PerMANOVA for Cultural Group Strength
+
+To evaluate the strength of the groupings found in the seriation and network analyses, we used modularity and perMANOVA tests.
+The modularity test is a community structure network detection method, and quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006; @Newman_Girvan_2004].
+The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings.
+PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017].
+The results are expressed by an R^2^ statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, a pseudo-F statistic, and a p-value.
+The pseudo-F statistic compares the total sum of squared dissimilarities between groups to total sum of squared dissimilarities within groups.
+Larger pseudo-F statistics indicate more separation between groups.
+The p-value quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
+The modularity test and perMANOVA were performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.
+
+# Results
+
+## Cultural Group Detection
 ```{r fig.height= 9, fig.width= 13}
 #| label: fig-seriation
-#| fig-cap: "Seriation analysis of the data"
+#| fig-cap: "Seriation analysis of the data, with group membership indicated by color. Restricted group is in yellow, broader group is in blue."
 
@@
 
-For each of the phases, seriation identifies at least two groups, one with a restricted range of sign types, and one with a broader range.
-The notch and line signs are the most prevalent sign types across all time periods.
-In the Transitional phase, there is a small range of sign types.
+```{r}
+sign_type_df <- time_unique_data_full %>% 
+  pivot_longer(cols = line:star) %>% 
+  group_by(id) %>% 
+  filter(value != 0) %>% 
+  mutate(value = 1) %>% 
+  group_by(id) %>% 
+  summarise((sign_count = sum(value))) %>% 
+  mutate(sign_type_count = `(sign_count = sum(value))`, `(sign_count = sum(value))` = NULL)
+
+time_unique_data_full <- time_unique_data_full %>% 
+  left_join(sign_type_df, join_by(id))
+
+max_signs <- max(time_unique_data_full$sign_type_count)
+
+```
+
+For each of the phases, seriation identifies at least two groups, one with a restricted range of sign types (1-3 types), which tend to be simpler signs, and one with a broader range (up to `r max_signs` types), which include more complex signs.
+Simpler signs composed of fewer components, like line and notch, are the most prevalent across all time periods.
+Overall, across the four phases, the number of sites in the restricted group tends to be less than or equal to the number of sites in the broader group, with the exception of the final Evolved phase.
+
+In the Transitional phase, there is a relatively small range of sign types compared to other phases.
 The restricted group consists of Fumane, Pod Hradem, and Geissenklösterle, two of which only have a notch, and one of which has a notch and circumnotch.
-The broader group consists of Labeko Koba, Hohle Fels, and El Castillo.
+The broader group consists of Labeko Koba, Hohle Fels, and El Castillo, which vary in their sign composition and have only the line and oblique line (obline) signs in common.
+
 In the Proto-Aurignacian we see an expansion in both the number of sites and range of signs, with no overlap in sites from the Transitional phase, except for Geissenklösterle, which is present in both phases.
-The restricted group consists of Abri Pataud, Gatzarria, Hohlenstein-Stadel, and Mladeč, three of which only have notches and lines, and one also has a dashline.
+The restricted group consists of Abri Pataud, Hohlenstein-Stadel, and Mladeč, which only have notches and lines, and Gatzarria which also has a dashline.
 The broader group consists of Geissenklösterle, Spy, Vogelherd, Grotte du Renne, La Ferrassie, and Les Cottés, where we see a more consistent pattern of a more diverse range of signs, compared to the broader group of the Transitional phase.
+In addition to line and notch, the obline, radial notch (radnotch), cross, hatching, and circumferential notch (circumnotch) are also frequently found in the broader group.
+
 In the Early Aurignacian, we see appearance of completely new sites with no overlap with the Proto-Aurignacian sites, as well as increased diversity of patterns within the groups.
-The restricted group consists of Grottes de Fonds-de-Forêt (radnotch and notch), Riparo Bombrini (notch), and Grotte de la Verpillière I (notch and line), The broader group consists of Solutré, Hohle Fels, Castanet, Cellier, Blanchard, and Trou al'Wesse, which contain a wide range of signs in a variety of combinations.
+The restricted group consists of Grottes de Fonds-de-Forêt (radnotch and notch), Riparo Bombrini (notch), and Grotte de la Verpillière I (notch and line).
+The broader group consists of Solutré, Hohle Fels, Castanet, Cellier, Blanchard, and Trou al'Wesse, which contain a wide range of signs in a variety of combinations.
+In addition to line and notch, the obline, dot and vulva are also frequently found in the broader group.
+
 Lastly, in the Evolved Aurignacian, we see slightly more consistent groups, again with new sites with no overlap with the Early Aurignacian sites, except for Hohle Fels, which also appears in the Transitional phase.
-Vogelherd also reappears, having appeared earlier in the Proto-Aurignacian.The restricted group consists of Les Rois, Gargas, La Viña, Sirgenstein Cave, and Vindija Cave, covering the line, notch, and and obline signs.
+Vogelherd also reappears, having appeared earlier in the Proto-Aurignacian.The restricted group consists of Les Rois, Gargas, La Viña, Sirgenstein Cave, and Vindija Cave, with the line, notch, and and obline signs.
 The broader group consists of Trou Magrite, Bockstein-Törle, Vogelherd, and Hohle Fels.
-Overall, across the four phases, the number of sites in the restricted group tends to be less than or equal to the number of sites in the broader group, with the exception of the final Evolved phase.
-We also see that simpler signs, like line and notch, tend to be more common in both groups, but tend to primarily make up the restricted groups.
-
-<!-- BM: I reckon we need to say something about the signs also, not just the sites -->
+Along with line, notch, and obline, the hatching, and oblique notch (obnotch) signs are also frequently found in the broader group.
 
@@
 
-In identifying the groups, we referenced both the seriation and network analyses.
-The results of the above tests allowed us to find two main groups of sites within each phase, as shown in @tbl-group-table.
-The geographic distributions of the groups are shown in @fig-group-map, and the sign makeup of each group is shown in @fig-group-sign.
-
-```{r}
- #| label: tbl-group-table
- #| tbl-cap: "Table of the sites by group per phase"
- #| tbl-cap-location: bottom
- group_table <- read_excel("data/site-groups.xlsx")
-
- kable(group_table)
-```
+In identifying the groups, we referenced both the seriation and network analyses. The results of the above tests allowed us to find two main groups of sites within each phase (@fig-seriation).
 
@@
 ## Modularity test
-trans_strength <- strength_function(trans_artifact_data, trans_unique_data)
-proto_strength <- strength_function(proto_artifact_data, proto_unique_data)
-early_strength <- strength_function(early_artifact_data, early_unique_data)
-evolved_strength <- strength_function(evolved_artifact_data, evolved_unique_data)
-
-full_table <- rbind(trans_strength, proto_strength, early_strength, evolved_strength) %>% 
-  mutate(`Aurignacian phase` = c("Transitional", 
-                           "Proto-Aurignacian",
-                           "Early Aurignacian", 
-                           "Evolved Aurignacian"), 
-         .before = 1)
 
@@
 
-The modularity score quantifies the strength of the groupings, showing that the variations in grouping strength between phases are fairly minimal.
-The results tell us that the Proto-Aurignacian exhibits the strongest groupings, with a score of `r proto_modularity`, and that the Evolved Aurignacian exhibits the weakest groupings, with a score of `r evolved_modularity`.
-There is a slight increase in modularity between the Transitional period and the Proto-Aurignacian, before an even slighter decrease into the Early Aurignacian, followed by a larger decrease with the Evolved Aurignacian.
-
-The R2 statistic from the perMANOVA tests tell us the proportion of sign variation that can be explained by group membership.
+The modularity score quantifies the strength of the groupings, showing that the variations in grouping strength between phases are minimal.
+The Proto-Aurignacian exhibits the strongest groupings, with a score of `r proto_modularity`, and the Evolved Aurignacian has the weakest groupings, with a score of `r evolved_modularity`.
+There is a slight increase in modularity between the Transitional period and the Proto-Aurignacian, before an even slighter decrease into the Early Aurignacian, followed by a larger increase with the Evolved Aurignacian.
+
+The perMANOVA R^2^ statistic shows the proportion of sign variation that can be explained by group membership.
 The Transitional period has the highest score of `r trans_perm_r` while the Early Aurignacian has the lowest score of `r early_perm_r`.
 We see a large decline in scores between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian periods.
-The pseudo-F statistic quantifies the difference between groups, with the Transitional phase having the highest score of `trans_perm_f`, and the Early Aurignacian having the lowest score of `early_perm_f`.
-Similar to the R2 statistic, we see a decline in scores between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian period.
-The perMANOVA p score quantifies the statistical significance of the pseudo-F statistic, with 0.05 as the threshold.
-The only time period to not meet that threshold is the Transitional period (p = `r trans_perm_p`).
-There is a slight increase between the Proto and Early Aurignacian, and a slightly large increase in the Evolved Aurignacian period, although they all still meets the threshold for statistical significance.
-
-## Effect of Geographic Distance
-
-```{r}
-#| label: tbl-mantel-table
-#| tbl-cap: "Table of the results per phase for the Mantel test between a Jaccard matrix of sign type diversity and a geographic distance matrix"
-#| tbl-cap-location: bottom
-## Mantel test for geography vs signs
-mantel_table <- full_table %>% 
-  dplyr::select(`Aurignacian phase`, `Mantel R`, `Mantel p`)
-
-kable(mantel_table)
-
-proto_mantel_r <- proto_strength %>% 
-  dplyr::select(`Mantel R`)
-
-proto_mantel_p <- proto_strength %>% 
-  dplyr::select(`Mantel p`)
-
-early_mantel_r <- early_strength %>% 
-  dplyr::select(`Mantel R`)
-
-early_mantel_p <- early_strength %>% 
-  dplyr::select(`Mantel p`)
-
-```
-
-The relevance of the isolation-by-distance hypothesis also varies from period to period.
-The p score quantifies the correlation, with a score of 0.05 or lower meaning that there is a statistically significant relationship between sign type distribution and geographical distance.
-None of the periods meet that threshold, meaning that the isolation-by-distance hypothesis is not applicable to any of them.
-However, we still see changes in the p score between periods, suggesting a higher level of correlation with geographic distance in some, and a lower correlation in others.
-The Proto-Aurignacian period demonstrates the least amount of correlation (Mantel R = `r proto_mantel_r`, p = `r proto_mantel_p`), while the Early Aurignacian exhibits the highest amount of correlation (Mantel R = `r early_mantel_r`, p = `r early_mantel_p`).
+The pseudo-F statistic quantifies the difference between groups, with the Transitional phase having the highest value of `r trans_perm_f`, and the Early Aurignacian having the lowest value of `r early_perm_f`.
+Similar to the R^2^ statistic, we see a decline in values between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian period.
+The perMANOVA p-value indicates a statistically significant difference in the distribution of signs between the restricted and broad groups for the Proto- , Early, and Evolved Aurignacian phases, but not the Transitional phase (p = `r trans_perm_p`).
+
+# Discussion
+
+The first aim of our study was to determine how the grouping dynamics of geometric signs changed across the four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and boundaries.
+The results as displayed in @fig-seriation show that each phase can be divided into two main cultural groups, with diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.
+
+@fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
+The modularity tests, as shown in @tbl-strength-table, show that the groupings are strongest in the Transitional period and weakest in the Early Aurignacian, and the perMANOVA tests tell us that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian experiences the least.
+The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian experiences the highest correlation.
 
@@
   theme(axis.text = element_blank(),
-        axis.title = element_blank())
-
-```
-
-@fig-group-map shows the geographic distribution of sites as colored by group, demonstrating that geographic boundaries do not readily correspond to the group distributions.
-In the Transitional phase, the sites are widely spread out across Western Europe, and tend to be located fairly closely to the other sites in their groups.
+        axis.title = element_blank(),
+        strip.text = element_text(size = 10))
+
+```
+
+The geographic structure of the groups changes from phase to phase (@fig-group-map).
+In the Transitional phase, sites are widely spread out across Western Europe, and tend to be located closely to the other sites in their groups.
 The exception is Hohle Fels, which, instead of being located further west with the rest of the sites in the broader group, is located further east, closer to the sites in the restricted group, and almost overlapping with Geissenklösterle.
-In the Proto-Aurignacian, we see a slightly more restricted distribution of sites westward but a spread northward, and more overlapping of sites between groups.
-For the restricted group, we interestingly see a divide across the center, with Mladeč and Hohlenstein-Stadel located close together further to the east, and Gatzarria and Abri Pataud located close together further to the west, with a large space between the two clusters.
-The broader sites tend to be more uniformly spread out across the geographic space, without much clustering close together.
-We again see some overlap between Abri Pataud and La Ferrassie, as well as between Geissenklösterle and Hohlenstein-Stadel.
-In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together, incluing a great deal of both intra- and inter-group proximity.
+In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups.
+In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together.
 In the Evolved Aurignacian, we see a wider distribution and more distance between sites.
-In particular, the sites in the reduced group are especially spread out across the space, while the broader sites are slightly more clustered.
-We also see a bit of a geographic divide between groups, with the reduced sites tending to be located further to the west, while the broader sites tend to be located further to the east.
-The exceptions is Sirgenstein Cave, which is located closer to the broader sites, almost overlapping with Vogelherd, Bockstein-Törle, and Hohle Fels.
-
-# Discussion
-
-The first aim of our study was to determine how the grouping dynamics of geometric signs changed across the four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and boundaries.
-The results as displayed in @fig-seriation show that each phase can be divided into two main cultural groups, with diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.
-
-@fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
-The modularity tests, as shown in @tbl-strength-table, show that the groupings are strongest in the Transitional period and weakest in the Early Aurignacian, and the perMANOVA tests tell us that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian experiences the least.
-The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian experiences the highest correlation.
 
@@
 
+It is also possible that the established precedent of aggregation sites could explain some of our results.
+The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for various subsistence or social reasons.
+Aggregation sites are places where those fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg].
+Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @Soffer_Adovasio_2010; @Bourdier_2013; @Svoboda_2013].
+It is therefore possible that the sites with multiple connections to both groups, such as La Ferrassie in the the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.
+However, without examining environmental data, this remains only a hypothesis at this time.
+
+Our work adds to a growing awareness of diversity within the Upper Paleolithic.
+Traditionally, the focus on the transition from Neanderthals to early modern humans is envisioned as a point of clear rupture with the Middle Paleolithic, thus leading some researchers to over-emphasize homogeneity [@Teyssandier_2008; @3567b0e7-61f1-3f3c-8f91-c78399bc0a5a].
+Previous studies have complicated this perception, investigating those patterns of diversity and heterogeneity within the Upper Paleolithic [@Vanhaeren_d; @fbdeb6f7-9945-3e64-82d6-7c3ab51f8115; @97387c23-a43b-34e6-95ef-0a248850bcf3; @c35eb420-0df4-3425-adac-d15b922564bd].
+This study contributes to that awareness, suggesting permeable, yet distinct cultural patterning and boundaries, as well as patterns of exchange and movement between groups.
+
+This study also adds to the growing body of methods for the identification of cultural groups, as well as change in social dynamics over time.
+We have shown that seriation and network analysis can be successfully used to identify cultural groups from geometric signs on mobile objects, and modularity and perMANOVA tests can be used to evaluate the strength of those groupings.
+
+One limitation of the study is the applicability of some of the simpler signs composed from fewer elements, which tend to primarily make up the signs found the in the restricted groups.
+It is worth considering whether occurrence of those simpler signs at different sites is strong evidence of cultural connection, as the simplicity of the signs means an increase in the possibility that they could have been independently invented, without any sort of cultural connection of exchange.
+Ultimately, we chose to evaluate them as evidence of cultural connection, however, we believe that this study might be refined in the future through the usage of the more specific data on the order and patterns in which signs occurred, which was recently released covering sites in Germany in SignBase 2.0.
+
+Another potential point of contention is the subdivision of each phase into only two cultural groups, which some may argue serves to simplify complex patterns of diversity.
+We feel that these patterns do not represent discrete, consistent groups to which all the sites within the group belong, but rather represent metagroups or behavioral types on the ends of a spectrum.
+This is in accordance with the cutural phylogeny models proposed by @boyd2013cultural, in which cultural groups are defined by presence or absence of certain cultural variants.
+In this case, restricted sign variability vs.
+broader sign variability is the cultural variant that we are tracking in this study, and the two group division represents two opposite types of the cultural variant.
+
 # Conclusion

### 2025-03-21 endofquarterpaper.qmd
replace , with ; in citation set in intro

@@
 Another distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile.
-Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence, @Sauvet_2018, @Kuhn_Stiner_2007, @Vanhaeren_d].
+Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d].
 Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca.
@@
 ## Cultural Group Detection
+
 ```{r fig.height= 9, fig.width= 13}

### 2025-03-25 endofquarterpaper.qmd
stash

--- 2025-03-21 endofquarterpaper.qmd
@@
 These exclusions resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map).
-
-```{r fig.height= 9, fig.width=10}
-```
 
@@
 The results as displayed in @fig-seriation show that each phase can be divided into two main cultural groups, with diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.
-
 @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
@@
 
+One factor that may influence the geographic structure of sites is the possibility that some sites functioned as aggregation sites.
+The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for various subsistence or social reasons.
+Aggregation sites are the places where those fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg].
+Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
+It is therefore possible that the sites with multiple connections to both groups, such as La Ferrassie in the the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignroacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.
+However, without examining environmental data, this remains only a hypothesis at this time.
+
 The second aim of our study was to determine what these patterns tell us about dynamics of social organization throughout the Aurignacian.
@@
 This rapid expansion could be what stimulated the emergence of signs as a marker of group membership.
-Additionally @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations.
+This is supported by other research finding similar evidence of artistic pro
+
+<!-- Add evidence about scholars seeing expansion in signs/artwork following population expansions in other regions/time periods-->
+
+Additionally, @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations.
 This likely explains why we see a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance.
 Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian (Aur-P2) theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions.
-@Shao_Wegener_Klein_Schmidt_Weniger_2024 also find an increase in expansion marking the beginning of the Evolved Aurignacian.
-Geographic expansion resulting in less group interconnection could explain why we see weaker connections in the Evolved Aurignacian.
-
-It is also possible that the established precedent of aggregation sites could explain some of our results.
-The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for various subsistence or social reasons.
-Aggregation sites are places where those fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg].
-Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @Soffer_Adovasio_2010; @Bourdier_2013; @Svoboda_2013].
-It is therefore possible that the sites with multiple connections to both groups, such as La Ferrassie in the the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.
-However, without examining environmental data, this remains only a hypothesis at this time.
+@Shao_Wegener_Klein_Schmidt_Weniger_2024 also finds an increase in expansion marking the beginning of the Evolved Aurignacian.
+Geographic expansion resulting in less group interconnection could explain why we see weaker, more dispersed connections in the Evolved Aurignacian.
 
 Our work adds to a growing awareness of diversity within the Upper Paleolithic.
-Traditionally, the focus on the transition from Neanderthals to early modern humans is envisioned as a point of clear rupture with the Middle Paleolithic, thus leading some researchers to over-emphasize homogeneity [@Teyssandier_2008; @3567b0e7-61f1-3f3c-8f91-c78399bc0a5a].
-Previous studies have complicated this perception, investigating those patterns of diversity and heterogeneity within the Upper Paleolithic [@Vanhaeren_d; @fbdeb6f7-9945-3e64-82d6-7c3ab51f8115; @97387c23-a43b-34e6-95ef-0a248850bcf3; @c35eb420-0df4-3425-adac-d15b922564bd].
-This study contributes to that awareness, suggesting permeable, yet distinct cultural patterning and boundaries, as well as patterns of exchange and movement between groups.
+Traditionally, the focus on the transition from Neanderthals to early modern humans is envisioned as a point of clear rupture with the Middle Paleolithic, thus leading some researchers to over-emphasize homogeneity [@Teyssandier_2008; @tey2006].
+Previous studies have complicated this perception, investigating those patterns of diversity and heterogeneity within the Upper Paleolithic [@Vanhaeren_d; @belfer-cohen2003; @bordes1970; @bordes1963].
+This study contributes to that awareness, suggesting permeable, yet distinct cultural variation and boundaries, as well as patterns of exchange and movement between groups.
+
+This work also adds to our knowledge on connectivity within the Upper Paleolithic.
+Our network analysis and perMANOVA test results show evolving patterns of connectivity between groups and sites, suggesting dynamic patterns of social exchange and connection.
 
@@
 We feel that these patterns do not represent discrete, consistent groups to which all the sites within the group belong, but rather represent metagroups or behavioral types on the ends of a spectrum.
-This is in accordance with the cutural phylogeny models proposed by @boyd2013cultural, in which cultural groups are defined by presence or absence of certain cultural variants.
+This is in accordance with the cultural phylogeny models proposed by @boyd2013cultural, in which cultural groups are not discreet, bounded entities, but rather defined by presence or absence of certain cultural variants which move between groups.
 In this case, restricted sign variability vs.

### 2025-03-25 Extra/modularity-test.R

@@ -1,0 +1,31 @@
+jac <- vegdist(trans_artifact_data %>%  mutate_all(~ as.numeric(. > 0)), "jaccard", binary = TRUE)
+dm <- as.matrix(jac)
+disim <- 1 - dm
+disim[disim<0.1] <- 0
+disim_net <- network(disim,
+                     directed=F,
+                     ignore.eval=F,
+                     names.eval='weight')
+
+phyl <- nj(dm)
+
+mem <- as_membership((trans_unique_data$group))
+
+
+
+
+library(geomorph)
+res <- modularity.test(dm, partition.gp = mem, CI = FALSE)
+plot(res)
+
+data(pupfish) 
+Y.gpa <- gpagen(pupfish$coords, print.progress = FALSE)   
+#landmarks on the body and operculum
+land.gps <- rep('a',56); land.gps[39:48] <- 'b'
+
+MT <- modularity.test(Y.gpa$coords, land.gps, CI = FALSE)
+summary(MT) # Test summary
+plot(MT)
+
+
+

### 2025-03-27 endofquarterpaper.qmd
editing discussion

@@
 artifact_count_median <- median(artifact_count_df$artifact_count)
+
 n_material <- c(unique(signbase_full_clean$material))
@@
 
-The geographic structure of the groups changes from phase to phase (@fig-group-map).
+The geographic boundaries of the groups change from phase to phase (@fig-group-map).
 In the Transitional phase, sites are widely spread out across Western Europe, and tend to be located closely to the other sites in their groups.
@@
 Aggregation sites are the places where those fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg].
-Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
+Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
+
+.
 It is therefore possible that the sites with multiple connections to both groups, such as La Ferrassie in the the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignroacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.
@@
 This rapid expansion could be what stimulated the emergence of signs as a marker of group membership.
-This is supported by other research finding similar evidence of artistic pro
-
-<!-- Add evidence about scholars seeing expansion in signs/artwork following population expansions in other regions/time periods-->
-
+Other studies have also found a similar associations between quantitity and diversity of art, and population density increases in other regions and time periods [@smith1992Design; @hays1993symbol; @McDonald_Veth_Lilley_2006; @Lourandos_1985; @barton_clark_cohen1994; @mcdonald2012social; @bernardini2005reconsidering; @wiessner1984reconsidering; @smith1992].
 Additionally, @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations.
@@
 
-Our work adds to a growing awareness of diversity within the Upper Paleolithic.
+Our work adds to a growing awareness of diversity and interconnection within the Upper Paleolithic.
 Traditionally, the focus on the transition from Neanderthals to early modern humans is envisioned as a point of clear rupture with the Middle Paleolithic, thus leading some researchers to over-emphasize homogeneity [@Teyssandier_2008; @tey2006].
-Previous studies have complicated this perception, investigating those patterns of diversity and heterogeneity within the Upper Paleolithic [@Vanhaeren_d; @belfer-cohen2003; @bordes1970; @bordes1963].
-This study contributes to that awareness, suggesting permeable, yet distinct cultural variation and boundaries, as well as patterns of exchange and movement between groups.
-
-This work also adds to our knowledge on connectivity within the Upper Paleolithic.
-Our network analysis and perMANOVA test results show evolving patterns of connectivity between groups and sites, suggesting dynamic patterns of social exchange and connection.
-
-This study also adds to the growing body of methods for the identification of cultural groups, as well as change in social dynamics over time.
-We have shown that seriation and network analysis can be successfully used to identify cultural groups from geometric signs on mobile objects, and modularity and perMANOVA tests can be used to evaluate the strength of those groupings.
-
-One limitation of the study is the applicability of some of the simpler signs composed from fewer elements, which tend to primarily make up the signs found the in the restricted groups.
+Previous studies have complicated this perception, investigating those patterns of diversity and heterogeneity within the Upper Paleolithic [@Belfer-Cohen_Goring-Morris_2002; @Maher_Conkey_2019; @Vanhaeren_d; @belfer-cohen2003; @bordes1970; @Shemer_Barzilai_Marder_2024; @Schwendler_2012; @bordes1963].
+This study contributes to that awareness, suggesting permeable, yet distinct cultural variation and boundaries.
+
+This study adds to our knowledge of social networks within the Aurignacian period, as our network analysis and perMANOVA test results suggest dynamic patterns of social exchange and connection.
+Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996].
+@Foley_Gamble_2009 places the emergence of inter-group complexity at around 30 k BP, however, our analyses show evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of inter-connectivity than previously predicted.
+We hope that dividing the data up into smaller time periods serves to rectify some of the issues in low temporal resolution that tend to affect results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023].
+Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019], as well as the mobility associated with hunter-gatherers [@Whallon_2006; @Padilla-Iglesias_Bischoff_2024].
+
+One limitation of the study is the inclusion of some of the simpler signs composed from fewer elements, which tend to primarily make up the signs found the in the restricted groups.
 It is worth considering whether occurrence of those simpler signs at different sites is strong evidence of cultural connection, as the simplicity of the signs means an increase in the possibility that they could have been independently invented, without any sort of cultural connection of exchange.
-Ultimately, we chose to evaluate them as evidence of cultural connection, however, we believe that this study might be refined in the future through the usage of the more specific data on the order and patterns in which signs occurred, which was recently released covering sites in Germany in SignBase 2.0.
+Ultimately, we chose to include them as evidence of cultural connection, as a greater number of included sign types serves to establish stronger connections [@Gravel-Miguel_Coward_2023].
+However, we believe that this study might be refined in the future through the usage of the more specific data on the order and patterns in which signs occurred, which was recently released covering sites in Germany in SignBase 2.0.
 

### 2025-03-27 endofquarterpaper.qmd
more edits

@@
 artifact_count_median <- median(artifact_count_df$artifact_count)
-
-n_material <- c(unique(signbase_full_clean$material))
 
@@
 Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
-
-.
 It is therefore possible that the sites with multiple connections to both groups, such as La Ferrassie in the the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignroacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.
@@
 The Evolved Aurignacian experiences a decline in connection strength, although it maintains degrees of inter-connectivity.
+
+This study adds to our knowledge of social networks within the Aurignacian period, as our results suggest dynamic patterns of social exchange and connection.
+Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996].
+@Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our network analysis shows evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted.
+We hope that dividing the data up into smaller time periods serves to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023].
+Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019], as well as the mobility associated with hunter-gatherers [@Whallon_2006; @Padilla-Iglesias_Bischoff_2024].
 
@@
 
-Our work adds to a growing awareness of diversity and interconnection within the Upper Paleolithic.
-Traditionally, the focus on the transition from Neanderthals to early modern humans is envisioned as a point of clear rupture with the Middle Paleolithic, thus leading some researchers to over-emphasize homogeneity [@Teyssandier_2008; @tey2006].
+Our work also adds to a growing awareness of diversity within the Upper Paleolithic.
+Traditionally, the focus on the transition from Neanderthals to early modern humans is envisioned as a point of clear rupture with the Middle Paleolithic, thus leading some researchers to over-emphasize homogeneity [@Teyssandier_2008; @tey2006] and de-emphasize diversity.
 Previous studies have complicated this perception, investigating those patterns of diversity and heterogeneity within the Upper Paleolithic [@Belfer-Cohen_Goring-Morris_2002; @Maher_Conkey_2019; @Vanhaeren_d; @belfer-cohen2003; @bordes1970; @Shemer_Barzilai_Marder_2024; @Schwendler_2012; @bordes1963].
 This study contributes to that awareness, suggesting permeable, yet distinct cultural variation and boundaries.
-
-This study adds to our knowledge of social networks within the Aurignacian period, as our network analysis and perMANOVA test results suggest dynamic patterns of social exchange and connection.
-Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996].
-@Foley_Gamble_2009 places the emergence of inter-group complexity at around 30 k BP, however, our analyses show evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of inter-connectivity than previously predicted.
-We hope that dividing the data up into smaller time periods serves to rectify some of the issues in low temporal resolution that tend to affect results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023].
-Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019], as well as the mobility associated with hunter-gatherers [@Whallon_2006; @Padilla-Iglesias_Bischoff_2024].
 

### 2025-04-08 endofquarterpaper.qmd

@@
 
-The distribution of objects across sites is highly skewed, with the median number of objects as `r artifact_count_median`, but Vogelherd has `r nvogel`.
+The distribution of objects across sites is highly skewed (@fig-site-map C), with a median number of `r artifact_count_median` objects, but Vogelherd has `r nvogel`.
 To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites.
@@
 The p-value quantifies the probability of obtaining the observed correlation if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
-The Mantel tests were performed using the vegan package [@Vegan2001].
-The test indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).
+The Mantel tests, performed using the vegan package [@Vegan2001], indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).
 
@@
 The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` sign types, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` sign types, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` sign types, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` sign types.
-A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (R = `r chron_r`, p = `r chron_p`).
-This confirms that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.
+A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (R = `r chron_r`, p = `r chron_p`), confirming that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.
 
@@
 The Proto-Aurignacian exhibits the strongest groupings, with a score of `r proto_modularity`, and the Evolved Aurignacian has the weakest groupings, with a score of `r evolved_modularity`.
-There is a slight increase in modularity between the Transitional period and the Proto-Aurignacian, before an even slighter decrease into the Early Aurignacian, followed by a larger increase with the Evolved Aurignacian.
+There is an increase in modularity between the Transitional period and the Proto-Aurignacian, before an decrease into the Early Aurignacian, followed by an increase with the Evolved Aurignacian.
 
@@
 
-This study adds to our knowledge of social networks within the Aurignacian period, as our results suggest dynamic patterns of social exchange and connection.
-Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996].
-@Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our network analysis shows evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted.
-We hope that dividing the data up into smaller time periods serves to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023].
-Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019], as well as the mobility associated with hunter-gatherers [@Whallon_2006; @Padilla-Iglesias_Bischoff_2024].
-
-These results are supported by the existing data on the environmental and biological change established during these phases.
+These results are consistent with existing data on the environmental and biological change established during these phases.
 @Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Proto/Early Aurignacian (Aur-P1) settlement being reached around 41 k BP.
 This rapid expansion could be what stimulated the emergence of signs as a marker of group membership.
-Other studies have also found a similar associations between quantitity and diversity of art, and population density increases in other regions and time periods [@smith1992Design; @hays1993symbol; @McDonald_Veth_Lilley_2006; @Lourandos_1985; @barton_clark_cohen1994; @mcdonald2012social; @bernardini2005reconsidering; @wiessner1984reconsidering; @smith1992].
+Other studies have also found a similar associations in other regions and time period between increases in quantity and diversity of art, and increases in population density [@smith1992Design; @hays1993symbol; @McDonald_Veth_Lilley_2006; @Lourandos_1985; @barton_clark_cohen1994; @mcdonald2012social; @bernardini2005reconsidering; @wiessner1984reconsidering; @smith1992].
 Additionally, @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations.
@@
 
+The results also add to our knowledge of social networks within the Aurignacian period, suggesting dynamic patterns of social exchange and connection.
+Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996].
+@Foley_Gamble_2009 places the emergence of intergroup social networks at around 30 k BP, however, our network analysis shows evidence of intergroup social networks beginning at around 41 k BP, suggesting an earlier emergence than previously predicted.
+Dividing the data up into smaller time periods potentially serves to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023; @Daems_Coco_Gillreath-Brown_Kafetzaki_2024].
+Additionally, by focusing on change over time, we can see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019], as well as the mobility associated with hunter-gatherers [@Whallon_2006; @Padilla-Iglesias_Bischoff_2024].
+
 Our work also adds to a growing awareness of diversity within the Upper Paleolithic.
@@
 Another potential point of contention is the subdivision of each phase into only two cultural groups, which some may argue serves to simplify complex patterns of diversity.
-We feel that these patterns do not represent discrete, consistent groups to which all the sites within the group belong, but rather represent metagroups or behavioral types on the ends of a spectrum.
-This is in accordance with the cultural phylogeny models proposed by @boyd2013cultural, in which cultural groups are not discreet, bounded entities, but rather defined by presence or absence of certain cultural variants which move between groups.
+However, these patterns should not be viewed as representing discrete, consistent groups to which all the sites within the group belong, but rather as representing opposite ends of a cultural variant spectrum.
+We adopt the cultural phylogeny models proposed by @boyd2013cultural, in which cultural groups are not discreet, bounded entities, but rather defined by presence or absence of certain cultural variants which move between groups.
 In this case, restricted sign variability vs.

### 2025-04-11 endofquarterpaper.qmd
stash

@@
 
-The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020] coupled with biological evolution.
-Another distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile.
-Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d].
+The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020], coupled with a biological transition from Neanderthals to early modern humans.
+Archaeologists have frequently relied upon variations in artwork, the large rise of which is a distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d].
 Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca.
@@
 The nine geographically discrete groups encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and western Mediterranean regions.
+
+Similar analyses of personal ornament types from the Aurignacian technocomplex identified fourteen geographically cohesive groups [@Vanhaeren_d].
+Drawing on ethnographic studies that show how body decoration and ornamentation indicates ethno-linguistic identity, @Vanhaeren_d used Aurignacian personal ornamentation as a proxy for ethno-linguistic diversity.
+The variation in personal ornament types is interpreted as evidence of long-lasting cultural differences and as robustly establishing the ethno-linguistic diversity of the Aurignacian.
+157 distinct ornament types from 98 Aurignacian sites in Europe and the Near East were analyzed, identifying geographically cohesive groups sharing similar ornament type associations by seriation, correspondence, and GIS analyses.
+These groups sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.
 
@@
                                                        "Early Aurignacian", "Evolved Aurignacian"))
-
+```
+
+```{r}
 trans_site_count <- nrow(trans_unique_data)
@@
 
-nvogel <- nrow(signbase_full_clean %>% 
-                 filter(site_name == "Vogelherd"))
-
 artifact_count_df <- signbase_full_clean %>% 
@@
 
-artifact_count_median <- median(artifact_count_df$artifact_count)
-
+sign_material_df <- signbase_full_clean %>% 
+  group_by(material) %>% 
+  summarise(count = n())%>% 
+  arrange(desc(count))
+
+iv_count <- sign_material_df %>% 
+  filter(material == "ivory") %>% 
+  pull(count)
+
+bone_count <-sign_material_df %>% 
+  filter(material == "bone") %>% 
+  pull(count)
+
+lime_count <- sign_material_df %>% 
+  filter(material == "limestone") %>% 
+  pull(count)
+
+ant_count <- sign_material_df %>% 
+  filter(material == "antler") %>% 
+  pull(count)
+
+tooth_count <- sign_material_df %>% 
+  filter(material == "animal tooth") %>% 
+  pull(count)
+
+ochre_count <- sign_material_df %>% 
+  filter(material == "ochre") %>% 
+  pull(count)
+
+undet_count <- sign_material_df %>% 
+  filter(material == "undetermined") %>% 
+  pull(count)
+
+stone_row <- tibble("material" = c("stone"), "count" = c(1))
+
+sign_material_df <- sign_material_df %>% 
+  rbind(stone_row) %>% 
+  filter(material != "stone indet.")
+
+misc_string <- sign_material_df %>% 
+  filter(count == 1) %>% 
+  dplyr::select(material) %>% 
+  c()
+
+misc_string <- misc_string$material
+
+object_type_df <- signbase_full_clean %>% 
+  group_by(object_type) %>% 
+  summarise(count = n()) %>% 
+  arrange(desc(count))
 ```
@@
   dplyr::select(MedianBP) %>% 
-  dist( method = "euclidean")
+  dist(method = "euclidean")
 
@@
   dplyr::select(line:star) %>% 
-  # BM: Ensure binary data as input to jaccard distance
   mutate_all(~ as.numeric(. > 0)) %>% 
@@
 These exclusions resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map).
+`r iv_count` of the objects are made of ivory, `r bone_count` objects are made of bone, `r lime_count` objects are made of limestone, `r ant_count` objects are made of antler, `r tooth_count` objects are made of animal teeth, `r ochre_count` objects are made of ochre, and `r undet_count` objects are made of undetermined material.
+The rest of the objects are made of `r combine_words(misc_string)`, with one made of each material.
 
@@
 
-The distribution of objects across sites is highly skewed (@fig-site-map C), with a median number of `r artifact_count_median` objects, but Vogelherd has `r nvogel`.
+The distribution of objects across sites is highly skewed, as shown in @fig-site-map C.
 To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites.
-We used Mantel tests with a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types.
+We used Mantel tests with a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types in order to make sure that the results would not be significantly impacted by the exlcusion of object abundance.
 The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value.
@@
 This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019].
-To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geography for each phase.
-We used Mantel tests with a geographic distance matrix and a Jaccard similarity matrix, with the results shown in @tbl-mantel-table.
-None of the phases have a statistically significant correlation between sign type distribution and geographical distance, so the isolation-by-distance hypothesis is rejected.
+To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geographic distance for each phase using Mantel tests.
+The results as shown in @tbl-mantel-table show that none of the phases have a statistically significant correlation between sign type distribution and geographical distance, so the isolation-by-distance hypothesis is rejected.
 
@@
 To investigate connectivity among clusters of sites in the seriation solution, we used network analysis to represent and quantify the relationship between sites [@Mills_2017].
-In our network analysis, each node is a site, and each edge is a measure of sign type distribution similarity.
-We input the Jaccard similarity matrix for each phase, computed the network statistics, and plotted the results with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths.
-Network analysis was performed using the vegan [@Vegan2001], statnet [@statnet_package] and igraph [@igraph_package] packages.
+Each node is a site, and each edge is a measure of sign type distribution similarity, as computed by the Jaccard similarity algorothm.
+Results were plotted with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths, and network analysis was performed using the vegan [@Vegan2001], statnet [@statnet_package] and igraph [@igraph_package] packages.
 
@@
 
-The first aim of our study was to determine how the grouping dynamics of geometric signs changed across the four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and boundaries.
-The results as displayed in @fig-seriation show that each phase can be divided into two main cultural groups, with diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.
+The first aim of our study was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and boundaries.
+The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.
 @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
@@
 
-These results are consistent with existing data on the environmental and biological change established during these phases.
+These results are consistent with existing data on the environmental and population change established during these phases.
 @Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Proto/Early Aurignacian (Aur-P1) settlement being reached around 41 k BP.
@@
 Other studies have also found a similar associations in other regions and time period between increases in quantity and diversity of art, and increases in population density [@smith1992Design; @hays1993symbol; @McDonald_Veth_Lilley_2006; @Lourandos_1985; @barton_clark_cohen1994; @mcdonald2012social; @bernardini2005reconsidering; @wiessner1984reconsidering; @smith1992].
+
 Additionally, @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations.
@@
 Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996].
-@Foley_Gamble_2009 places the emergence of intergroup social networks at around 30 k BP, however, our network analysis shows evidence of intergroup social networks beginning at around 41 k BP, suggesting an earlier emergence than previously predicted.
-Dividing the data up into smaller time periods potentially serves to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023; @Daems_Coco_Gillreath-Brown_Kafetzaki_2024].
-Additionally, by focusing on change over time, we can see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019], as well as the mobility associated with hunter-gatherers [@Whallon_2006; @Padilla-Iglesias_Bischoff_2024].
-
-Our work also adds to a growing awareness of diversity within the Upper Paleolithic.
-Traditionally, the focus on the transition from Neanderthals to early modern humans is envisioned as a point of clear rupture with the Middle Paleolithic, thus leading some researchers to over-emphasize homogeneity [@Teyssandier_2008; @tey2006] and de-emphasize diversity.
-Previous studies have complicated this perception, investigating those patterns of diversity and heterogeneity within the Upper Paleolithic [@Belfer-Cohen_Goring-Morris_2002; @Maher_Conkey_2019; @Vanhaeren_d; @belfer-cohen2003; @bordes1970; @Shemer_Barzilai_Marder_2024; @Schwendler_2012; @bordes1963].
-This study contributes to that awareness, suggesting permeable, yet distinct cultural variation and boundaries.
-
-One limitation of the study is the inclusion of some of the simpler signs composed from fewer elements, which tend to primarily make up the signs found the in the restricted groups.
-It is worth considering whether occurrence of those simpler signs at different sites is strong evidence of cultural connection, as the simplicity of the signs means an increase in the possibility that they could have been independently invented, without any sort of cultural connection of exchange.
-Ultimately, we chose to include them as evidence of cultural connection, as a greater number of included sign types serves to establish stronger connections [@Gravel-Miguel_Coward_2023].
-However, we believe that this study might be refined in the future through the usage of the more specific data on the order and patterns in which signs occurred, which was recently released covering sites in Germany in SignBase 2.0.
-
-Another potential point of contention is the subdivision of each phase into only two cultural groups, which some may argue serves to simplify complex patterns of diversity.
-However, these patterns should not be viewed as representing discrete, consistent groups to which all the sites within the group belong, but rather as representing opposite ends of a cultural variant spectrum.
-We adopt the cultural phylogeny models proposed by @boyd2013cultural, in which cultural groups are not discreet, bounded entities, but rather defined by presence or absence of certain cultural variants which move between groups.
-In this case, restricted sign variability vs.
-broader sign variability is the cultural variant that we are tracking in this study, and the two group division represents two opposite types of the cultural variant.
+@Foley_Gamble_2009 place the emergence of inter-group social networks at around 30 k BP, however, our network analysis shows evidence of inter-group social networks beginning at around 41 k BP, suggesting an earlier emergence than previously predicted.
+Dividing the data into smaller time periods potentially serves to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023; @Daems_Coco_Gillreath-Brown_Kafetzaki_2024].
+Additionally, by focusing on change over time, we can see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019]
+
+One possible limitation is the potential ambiguity that arises with some of the sign types.
+While @Von_Petzinger_2009 argues in favor of Aurignacian geometric signs as symbolic expression, potential ambiguity still remains over whether function is consistent across all sign type occurrences, particularly in regards to the simpler sign types composed from fewer elements.
+We included the more ambiguous sign types because similar sign occurrence may still serve as evidence of cultural connectivity regardless of function, and additionally, greater quantities of included sign types serves to establish stronger connections [@Gravel-Miguel_Coward_2023].
+Potential ambiguity may be addressed in the future through the usage of the data recently released in SignBase 2.0 covering sites in Germany, which contains information on the specific order and patterns in which signs occurred.
 
@@
 
-The observed social dynamics of the phases of the Aurignacian technocomplex correspond with the existing data on the environmental and social conditions of the Upper Paleolithic.
-This tells us that the usage of artwork as a proxy for cultural difference is an effective method for understanding the social dynamics of prehistoric humans.
+This study both adds to our understanding of Aurignacian social dynamics and suggests potential new directions for future research.
+The observed social dynamics produced by our methods align with the existing data on the environmental and population conditions of the Upper Paleolithic.
+We feel that our research opens up new opportunities for the application of these methods to data on artwork variation in other regions and time periods.
+Our research also suggests an earlier emergence of dynamic inter-connectivity than previously predicted, as we see weaker inter-connections beginning to emerge by 41 k BP, and stronger inter-connections by 38 k BP.
+Future research directions might involve the analysis of environmental conditions and their interactions with social dynamics, in order to determine whether some of the sites that we have pointed to as potential aggregation sites function as such.
+Another potential future research direction would be the usage of the more specific data released in SignBase 2.0 in order to counter potential sign type ambiguity.
 

### 2025-04-11 endofquarterpaper.qmd
edit

@@
 The nine geographically discrete groups encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and western Mediterranean regions.
-
-Similar analyses of personal ornament types from the Aurignacian technocomplex identified fourteen geographically cohesive groups [@Vanhaeren_d].
-Drawing on ethnographic studies that show how body decoration and ornamentation indicates ethno-linguistic identity, @Vanhaeren_d used Aurignacian personal ornamentation as a proxy for ethno-linguistic diversity.
-The variation in personal ornament types is interpreted as evidence of long-lasting cultural differences and as robustly establishing the ethno-linguistic diversity of the Aurignacian.
-157 distinct ornament types from 98 Aurignacian sites in Europe and the Near East were analyzed, identifying geographically cohesive groups sharing similar ornament type associations by seriation, correspondence, and GIS analyses.
-These groups sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.
 

### 2025-04-14 endofquarterpaper.qmd
merge

@@
 
-The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020], coupled with a biological transition from Neanderthals to early modern humans.
-Archaeologists have frequently relied upon variations in artwork, the large rise of which is a distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d].
+The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020] coupled with biological evolution.
+Another distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile.
+Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d].
 Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca.
@@
                                                        "Early Aurignacian", "Evolved Aurignacian"))
-```
-
-```{r}
+
 trans_site_count <- nrow(trans_unique_data)
@@
 
+nvogel <- nrow(signbase_full_clean %>% 
+                 filter(site_name == "Vogelherd"))
+
 artifact_count_df <- signbase_full_clean %>% 
@@
 
-sign_material_df <- signbase_full_clean %>% 
-  group_by(material) %>% 
-  summarise(count = n())%>% 
-  arrange(desc(count))
-
-iv_count <- sign_material_df %>% 
-  filter(material == "ivory") %>% 
-  pull(count)
-
-bone_count <-sign_material_df %>% 
-  filter(material == "bone") %>% 
-  pull(count)
-
-lime_count <- sign_material_df %>% 
-  filter(material == "limestone") %>% 
-  pull(count)
-
-ant_count <- sign_material_df %>% 
-  filter(material == "antler") %>% 
-  pull(count)
-
-tooth_count <- sign_material_df %>% 
-  filter(material == "animal tooth") %>% 
-  pull(count)
-
-ochre_count <- sign_material_df %>% 
-  filter(material == "ochre") %>% 
-  pull(count)
-
-undet_count <- sign_material_df %>% 
-  filter(material == "undetermined") %>% 
-  pull(count)
-
-stone_row <- tibble("material" = c("stone"), "count" = c(1))
-
-sign_material_df <- sign_material_df %>% 
-  rbind(stone_row) %>% 
-  filter(material != "stone indet.")
-
-misc_string <- sign_material_df %>% 
-  filter(count == 1) %>% 
-  dplyr::select(material) %>% 
-  c()
-
-misc_string <- misc_string$material
-
-object_type_df <- signbase_full_clean %>% 
-  group_by(object_type) %>% 
-  summarise(count = n()) %>% 
-  arrange(desc(count))
+artifact_count_median <- median(artifact_count_df$artifact_count)
+
 ```
@@
   dplyr::select(MedianBP) %>% 
-  dist(method = "euclidean")
+  dist( method = "euclidean")
 
@@
   dplyr::select(line:star) %>% 
+  # BM: Ensure binary data as input to jaccard distance
   mutate_all(~ as.numeric(. > 0)) %>% 
@@
 These exclusions resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map).
-`r iv_count` of the objects are made of ivory, `r bone_count` objects are made of bone, `r lime_count` objects are made of limestone, `r ant_count` objects are made of antler, `r tooth_count` objects are made of animal teeth, `r ochre_count` objects are made of ochre, and `r undet_count` objects are made of undetermined material.
-The rest of the objects are made of `r combine_words(misc_string)`, with one made of each material.
 
@@
 
-The distribution of objects across sites is highly skewed, as shown in @fig-site-map C.
+The distribution of objects across sites is highly skewed, as seen in @fig-site-map C.
 To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites.
-We used Mantel tests with a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types in order to make sure that the results would not be significantly impacted by the exlcusion of object abundance.
+We used Mantel tests with a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types.
 The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value.
@@
 The p-value quantifies the probability of obtaining the observed correlation if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
-The Mantel tests, performed using the vegan package [@Vegan2001], indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).
+The Mantel tests were performed using the vegan package [@Vegan2001].
+The test indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).
 
@@
 The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` sign types, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` sign types, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` sign types, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` sign types.
-A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (R = `r chron_r`, p = `r chron_p`), confirming that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.
+A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (R = `r chron_r`, p = `r chron_p`).
+This confirms that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.
 
@@
 This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019].
-To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geographic distance for each phase using Mantel tests.
-The results as shown in @tbl-mantel-table show that none of the phases have a statistically significant correlation between sign type distribution and geographical distance, so the isolation-by-distance hypothesis is rejected.
+To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geography for each phase.
+We used Mantel tests with a geographic distance matrix and a Jaccard similarity matrix, with the results shown in @tbl-mantel-table.
+None of the phases have a statistically significant correlation between sign type distribution and geographical distance, so the isolation-by-distance hypothesis is rejected.
 
@@
 The Proto-Aurignacian exhibits the strongest groupings, with a score of `r proto_modularity`, and the Evolved Aurignacian has the weakest groupings, with a score of `r evolved_modularity`.
-There is an increase in modularity between the Transitional period and the Proto-Aurignacian, before an decrease into the Early Aurignacian, followed by an increase with the Evolved Aurignacian.
+There is a slight increase in modularity between the Transitional period and the Proto-Aurignacian, before an even slighter decrease into the Early Aurignacian, followed by a larger increase with the Evolved Aurignacian.
 
@@
 
-The results also add to our knowledge of social networks within the Aurignacian period, suggesting dynamic patterns of social exchange and connection.
+This study adds to our knowledge of social networks within the Aurignacian period, as our results suggest dynamic patterns of social exchange and connection.
 Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996].
-@Foley_Gamble_2009 place the emergence of inter-group social networks at around 30 k BP, however, our network analysis shows evidence of inter-group social networks beginning at around 41 k BP, suggesting an earlier emergence than previously predicted.
-Dividing the data into smaller time periods potentially serves to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023; @Daems_Coco_Gillreath-Brown_Kafetzaki_2024].
-Additionally, by focusing on change over time, we can see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019]
+@Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our network analysis shows evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted.
+We hope that dividing the data up into smaller time periods serves to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023].
+Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019].
 

### 2025-04-14 Extra/undergraduate-research-symposium-images.qmd
merge

--- 2025-04-14 endofquarterpaper.qmd
@@
----
-title: "Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases"
-author: Sophie Cooper
-format: html
-echo: false
-execute: 
-  warning: false
-bibliography: references.bib
-link-citations: true
----
-
-# Introduction
-
-The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020] coupled with biological evolution.
-Another distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile.
-Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d].
-Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca.
-43 - 30 k BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020].
-In this paper, we investigate the social organization dynamics by analyzing the distribution and variation of geometric sign types found on mobile objects from the Upper Paleolithic in Europe across four phases of the Aurignacian period.
-
-Social organization dynamics of the Upper Paleolithic have previously been explored using statistical analyses of objects to understand patterns of cultural groups.
-Multivariate analyses of beads from the Gravettian technocomplex suggests an east to west cline of nine cultural groups [@baker2024evidence].
-Seriation analysis, PCoA, and network analysis were used to identify the groups from 134 discrete bead types recovered from both burial and occupation sites.
-To validate their claims of cultural groups, @baker2024evidence used a Mantel test to evaluate an isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance, finding that geographic distance alone did not solely account for the bead distribution.
-The nine geographically discrete groups encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and western Mediterranean regions.
-
-Similar analyses of personal ornament types from the Aurignacian technocomplex identified fourteen geographically cohesive groups [@Vanhaeren_d].
-Drawing on ethnographic studies that show how body decoration and ornamentation indicates ethno-linguistic identity, @Vanhaeren_d used Aurignacian personal ornamentation as a proxy for ethno-linguistic diversity.
-They analyzed 157 distinct ornament types from 98 Aurignacian sites in Europe and the Near East.
-Geographically cohesive groups sharing similar ornament type associations were identified by seriation, correspondence, and GIS analyses.
-These groups sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.
-@Vanhaeren_d interpret the variation in personal ornament types as evidence of long-lasting cultural differences and as robustly establishing the ethno-linguistic diversity of the Aurignacian.
-
-Cultural difference has also been identified from geometric signs in Upper Paleolithic parietal art from the Cantabrian region of Spain [@Sauvet_2018].
-Forty-two quadrilateral geometric signs from four cave sites, likely associated with the Magdalenian period, spanning a distance of 30 kilometers were recorded.
-Multivariate analysis of 45 morphological characteristics of the signs revealed two clusters.
-To interpret these clusters, @Sauvet_2018 draw on case studies of geometric signs as identity markers, including South African engraved ostrich shells from 60 k BP, maker's marks on Gallo-Roman pottery, and 19th century French craft guild members' personal marks.
-Using these case studies as analogies, @Sauvet_2018 interpret the Magdalenian geometric signs as identity markers, and the two clusters as representative of cultural groups.
-
-In this study, we extend the use of geometric signs as identity markers to investigate cultural groups in the Aurignacian period.
-The Aurignacian period spans roughly 43 - 30 k BP, and is largely characterized by material culture indicators, such as carinated scrapers, small-flake tools and split-based bone and ivory tools, as well as an ongoing evolutionary shift from Neanderthals to early modern humans [@Chu_Richter_2020].
-The Aurignacian can be divided into phases which are mainly distinguished by shifts in material culture; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015], as can be seen in @tbl-aur-table.
-@Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main phases, Aur-P1 (Proto/Early), which corresponds with the time period of 43–37 k BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods 37-32 k BP.
-@Banks_dErrico_Zilhao_2013 used Bayesian modeling methods to further divide Aur-P1 into the Proto and Early Aurignacian, placing the Proto-Aurignacian at 41.5 - 39.9 k BP, and the early Aurignacian at 39.8 - 37.8 k BP.
-They found that the Proto-Aurignacian occurs during the Greenland Interstadials (GI) 9 and 10, periods of climatic amelioration, while the Early Aurignacian began with the Henrich Stadial, which is characterized by dry and arid conditions, and ends with GI 8.
-
 ```{r}
-#| label: tbl-aur-table
-#| tbl-cap: "Summary of chronological, cultural and environmental characteristics of Aurignacian periods [@Banks_dErrico_Zilhao_2013;  @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021; @Chu_Richter_2020; @Fletcher_Sánchez]"
-#| tbl-cap-location: bottom
-
-library(knitr)
-library(tidyverse)
-library(readxl)
-
-aur_table <- read_excel("data/aurignacian-periods.xlsx")
-
-kable(aur_table)
-```
-
-We hypothesize that these phases, and the environmental changes across the Aurignacian, may correspond to changes in social dynamics of cultural groups, as represented by patterns in the distribution of geometric signs.
-We aim to address the following questions: How do grouping dynamics of geometric signs change across the four phases of the Aurignacian?
-How do the groupings vary in diversity, size, connectivity, and boundaries?
-Lastly, what do the patterns tell us about dynamics of social organization throughout the Aurignacian?
-We answer these questions by first identifying clusters within each phase using seriation and network analyses.
-We then test the strength of the clusters using modularity and perMANOVA tests.
-All of our programming in this study was conducted in RStudio.
-
-# Materials and Methods
-
-```{r}
-## clean the data
-library(magrittr)
-library(rcarbon)
-signbase_full <- read_csv("data/signBase_Version1.0.csv")
-nsites <- signbase_full %>% 
-  pull(site_name ) %>% 
-  unique() %>% 
-  length()
-nartifacts <- nrow(signbase_full)
-ncountries <- signbase_full %>% 
-  pull(country) %>% 
-  unique() %>% 
-  length()
-nsigntypes <- ncol(signbase_full)
-
-signbase_full_clean <- signbase_full %>%  # 511 rows
-  filter(site_name != "Willendorf",
-         site_name != "Riparo di Fontana Nuova",
-         site_name != "Muralovka",
-         site_name != "Shanidar Cave",
-         site_name != "Hayonim Cave",
-         site_name != "El Salitre",
-         site_name != "Grotte De La Princesse Pauline",
-         site_name != "Šandalja II") %>% 
-  dplyr::select(-other, -rectangle) %>% 
-  mutate(longitude = ifelse(site_name == "Riparo Bombrini", 7.437500, longitude)) %>% 
-  mutate(latitude = ifelse(site_name == "Riparo Bombrini", 43.77083, latitude))
-
-signbase_years <- signbase_full_clean %>%
-  drop_na(date_bp_max_min) %>% 
-  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+\\/\\-", "±")) %>% 
-  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+", "±")) %>% 
-  separate(date_bp_max_min,
-           sep = " - ",
-           c("date_bp_max", "date_bp_min"),
-           remove = FALSE) %>% 
-  mutate(date_bp_max = ifelse(str_detect(date_bp_max, "\\/"),
-                              str_extract(date_bp_max, ".*?(?=\\/)"),
-                              date_bp_max)) %>% 
-  mutate(date_bp_min = ifelse(str_detect(date_bp_min, "\\/"),
-                              str_extract(date_bp_min, ".*?(?=\\/)"),
-                              date_bp_min)) %>% 
-  separate(date_bp_max,
-           sep = "±",
-           c("date_bp_max_age",
-             "date_bp_max_error")) %>% 
-  separate(date_bp_min,
-           sep = "±",
-           c("date_bp_min_age",
-             "date_bp_min_error")) %>% 
-  drop_na(date_bp_max_age,
-          date_bp_max_error) %>% 
-  mutate(date_bp_max_age = parse_number(date_bp_max_age),
-         date_bp_max_error = parse_number(date_bp_max_error))
-
-signbase_years_cal <- 
-  rcarbon::calibrate(signbase_years$date_bp_max_age,
-                     signbase_years$date_bp_max_error,
-                     verbose = FALSE) %>% 
-  summary() %>% 
-  data_frame()
-
-signbase_years$MedianBP <- signbase_years_cal$MedianBP
-
-signbase_full_clean <- signbase_full_clean %>% 
-  inner_join(signbase_years %>% 
-              dplyr::select(object_id, MedianBP))
-
-signbase_full_clean <- signbase_full_clean %>%
-  mutate(time_period = case_when((MedianBP > 41499) ~ "transitional",
-                                 (MedianBP > 39799) ~ "proto_aurignacian",
-                                 (MedianBP > 37799) ~ "early_aurignacian",
-                                 (MedianBP > 31999) ~ "evolved_aurignacian"))
-
-signbase_full_clean <- signbase_full_clean %>% 
-  mutate(sign_total = rowSums((signbase_full_clean [,25:53]))) %>% 
-  filter(sign_total > 0)
-
-nsites_clean <- signbase_full_clean %>% 
-  pull(site_name ) %>% 
-  unique() %>% 
-  length()
-nartifacts_clean <- nrow(signbase_full_clean)
-ncountries_clean <- signbase_full_clean %>% 
-  pull(country) %>% 
-  unique() %>% 
-  length()
-nsigntypes_clean <- nsigntypes - 1
-
-pretty_print_dates <- 
-  function(x) formatC(x, 
-                      decimal.mark=".", 
-                      big.mark=",", 
-                      digits = 0, 
-                      format = "f")
-
-date_min <- signbase_years_cal %>% 
-  pull(MedianBP) %>% 
-  min()
-date_max <- signbase_years_cal %>% 
-  pull(MedianBP) %>% 
-  max()
-date_range <- date_max - date_min
-date_min_pretty <- pretty_print_dates(date_min)
-date_max_pretty <- pretty_print_dates(date_max)
-date_range_pretty <- pretty_print_dates(date_range)
-```
-
-```{r}
-lat_long_df <- signbase_full_clean %>% 
-  dplyr::select(site_name, longitude, latitude) %>% 
-  distinct(site_name, .keep_all = TRUE)
-
-trans_df<- signbase_full_clean %>% 
-  filter(time_period == "transitional") %>% 
-  mutate(longitude = as.character(longitude),
-         latitude = as.character(latitude))
-
-trans_object_num <- trans_df %>% 
-  group_by(site_name) %>%
-  summarise(nobjects = n())
-
-trans_unique_data <- trans_df %>% 
-  group_by(site_name) %>% 
-  summarize(across(where(is.numeric), sum)) %>% 
-  left_join(lat_long_df) %>% 
-  left_join(trans_object_num) %>% 
-  mutate(time_period = "transitional")
-
-proto_df <- signbase_full_clean %>% 
-  filter(time_period == "proto_aurignacian") %>% 
-  mutate(longitude = as.character(longitude),
-         latitude = as.character(latitude))
-
-proto_object_num <- proto_df %>% 
-  group_by(site_name) %>%
-  summarise(nobjects = n())
-
-proto_unique_data <- proto_df %>% 
-  group_by(site_name) %>%
-  summarize(across(where(is.numeric), sum)) %>% 
-  left_join(lat_long_df) %>% 
-  left_join(proto_object_num) %>% 
-  mutate(time_period = "proto_aurignacian")
-
-early_df <- signbase_full_clean %>% 
-  filter(time_period == "early_aurignacian") %>% 
-  mutate(longitude = as.character(longitude),
-         latitude = as.character(latitude))
-
-early_object_df <- early_df %>% 
-  group_by(site_name) %>%
-  summarise(nobjects = n())
-
-early_unique_data <- early_df %>% 
-  group_by(site_name) %>%
-  summarize(across(where(is.numeric), sum)) %>% 
-  left_join(lat_long_df) %>% 
-  left_join(early_object_df) %>% 
-  mutate(time_period = "early_aurignacian")
-
-evolved_df <- signbase_full_clean %>% 
-  filter(time_period == "evolved_aurignacian") %>% 
-  mutate(longitude = as.character(longitude),
-         latitude = as.character(latitude))
-
-evolved_object_df <- evolved_df %>% 
-  group_by(site_name) %>%
-  summarise(nobjects = n())
-
-evolved_unique_data  <- evolved_df %>% 
-  group_by(site_name) %>%
-  summarize(across(where(is.numeric), sum)) %>% 
-  left_join(lat_long_df) %>% 
-  left_join(evolved_object_df) %>% 
-  mutate(time_period = "evolved_aurignacian")
-
-trans_artifact_data <- trans_unique_data %>% 
-  column_to_rownames("site_name") %>% 
-  dplyr::select(line:star) %>% 
-  select_if(~ !is.numeric(.) || sum(.) != 0)
-
-proto_artifact_data <- proto_unique_data %>% 
-  column_to_rownames("site_name") %>% 
-  dplyr::select(line:star) %>% 
-  select_if(~ !is.numeric(.) || sum(.) != 0)
-
-early_artifact_data <- early_unique_data %>% 
-  column_to_rownames("site_name") %>% 
-  dplyr::select(line:star) %>% 
-  select_if(~ !is.numeric(.) || sum(.) != 0)
-
-evolved_artifact_data <- evolved_unique_data %>% 
-  column_to_rownames("site_name") %>% 
-  dplyr::select(line:star) %>% 
-  select_if(~ !is.numeric(.) || sum(.) != 0)
-
-trans_unique_data <- trans_unique_data %>% 
-  mutate(group = case_when(site_name %in% c("Pod Hradem", "Fumane", "Geissenklösterle") ~ "1",
-                           site_name %in% c("El Castillo", "Hohle Fels", "Labeko Koba") ~ "2"))
-
-proto_unique_data <- proto_unique_data %>% 
-  mutate(group = case_when(site_name %in% c("Gatzarria" , "Hohlenstein-Stadel", "Mladeč", "Abri Pataud") ~ "1",
-                           site_name %in%  c("Geissenklösterle", "Grotte du Renne", 
-                                             "Vogelherd", "Spy" ,"La Ferrassie", "Les Cottés") ~ "2"))
-
-early_unique_data <- early_unique_data %>% 
-  mutate(group = case_when(site_name %in%  c("Grottes de Fonds-de-Forêt", "Riparo Bombrini", 
-                                             "Grotte de la Verpillière I", "Vindija Cave") ~ "1",
-                           site_name %in% c("Hohle Fels", "Solutré", "Castanet", "Cellier", "Blanchard", "Trou al'Wesse") ~ "2"))
-
-evolved_unique_data <- evolved_unique_data %>% 
-  mutate(group = case_when(site_name %in%  c("Les Rois", "Gargas", "La Viña", "Sirgenstein Cave", "Vindija Cave") ~ "1",
-                           site_name %in% c("Bockstein-Törle", "Vogelherd", "Hohle Fels", "Trou Magrite") ~ "2"))
-
-time_unique_data_full <-rbind(trans_unique_data, proto_unique_data, 
-                              early_unique_data, evolved_unique_data)
-
-time_unique_data_full$time_period <- factor(time_unique_data_full$time_period, 
-                                            levels = c("transitional", "proto_aurignacian", 
-                                                       "early_aurignacian", "evolved_aurignacian"), 
-                                            labels = c("Transitional", "Proto-Aurignacian",
-                                                       "Early Aurignacian", "Evolved Aurignacian"))
-
-trans_site_count <- nrow(trans_unique_data)
-trans_sign_count <- ncol(trans_artifact_data)
-proto_site_count <- nrow(proto_unique_data)
-proto_sign_count <- ncol(proto_artifact_data)
-early_site_count <- nrow(early_unique_data)
-early_sign_count <- ncol(early_artifact_data)
-evolved_site_count <- nrow(evolved_unique_data)
-evolved_sign_count <- ncol(evolved_artifact_data)
-
-nvogel <- nrow(signbase_full_clean %>% 
-                 filter(site_name == "Vogelherd"))
-
-artifact_count_df <- signbase_full_clean %>% 
-  group_by(site_name) %>% 
-  summarise(artifact_count = n())
-
-artifact_count_median <- median(artifact_count_df$artifact_count)
-
-```
-
-```{r}
-# mantel abundance
-library(vegan)
-
-set.seed(200)
-
-artifact_data <- time_unique_data_full %>% 
-  column_to_rownames("id") %>% 
-  dplyr::select(line:star)
-
-# BM: not sure why the id column was also selected here? I moved it to rownames
-object_euc_data <- time_unique_data_full %>%
-    column_to_rownames("id") %>% 
-    dplyr::select(nobjects) %>% 
-    dist(method = "euclidean")
-
-# BM: Ensure binary data as input to jaccard distance
-artifact_data_bin <- 
-artifact_data %>% 
-  mutate_all(~ as.numeric(. > 0))
-
-jac_full <- vegan::vegdist(artifact_data_bin, method = "jaccard", binary = TRUE)
-
-object_count_mantel <- vegan::mantel(object_euc_data, jac_full, permutations = 1000)
-p_value_object_abundance_correlation <- round(object_count_mantel$signif, 3)
-r_value_object_abundance_correlation <- round(object_count_mantel$statistic, 3)
-
-sign_euc_data <- artifact_data %>% 
-  mutate(sign_count = rowSums(.)) %>% 
-  dplyr::select(sign_count) %>% 
-  dist(method = "euclidean")
-
-sign_count_mantel <- vegan::mantel(sign_euc_data, jac_full, permutations = 1000)
-p_value_sign_abundance_correlation <- round(sign_count_mantel$signif, 3)
-r_value_sign_abundance_correlation <- round(sign_count_mantel$statistic, 3)
-
-## mantel chronological
-chron_df <- signbase_full_clean %>% 
-  column_to_rownames("id") %>% 
-  dplyr::select(MedianBP) %>% 
-  dist( method = "euclidean")
-
-chron_sign_df <- signbase_full_clean %>% 
-  column_to_rownames("id") %>% 
-  dplyr::select(line:star) %>% 
-  # BM: Ensure binary data as input to jaccard distance
-  mutate_all(~ as.numeric(. > 0)) %>% 
-  vegdist("jaccard")
-
-chron_mantel <- vegan::mantel(chron_df, chron_sign_df, permutations = 1000)
-
-chron_p <- round(chron_mantel$signif, 3)
-chron_r <- round(chron_mantel$statistic, 3)
-
-## chron permanova
-perman_chron <- adonis2(jac_full ~ as.factor(time_unique_data_full$time_period), 
-                    method = "jaccard",
-                    sqrt.dist = TRUE)
-
-perman_chron <- perman_chron %>% 
-    data_frame() %>% 
-    slice_head()
-
-```
-
-## Data
-
-Our data comes from SignBase 1.0, an open-access collection of geometric signs on mobile objects from Upper Paleolithic Europe.
-SignBase 1.0 is a record of objects noting only the presence or absence of each sign on the object, without details of the quantity or sequence of individual signs on a given object.
-We used the identification of sign types provided by SignBase's curators [@dutkiewicz2020].
-The data set consists of `r nartifacts` objects found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign types.
-We excluded signs if they fell under the "other" category, due to the wide range of potential possibilities.
-We also excluded seven sites from the sample (only one sign present: Willendorf; extreme geographic distance from the other sites: Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave; dated much later than the rest of the assemblages: Grotte De La Princesse Pauline and Šandalja II).
-These exclusions resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map).
-
-```{r fig.height= 9, fig.width=10}
-#| label: fig-site-map
-#| fig-cap: "A: Map of sites in each Aurignacian phase that contain geometric signs on mobile objects.
-B: Abundance of signs over time, phases indicated by colours.
-C.
-Distribution of object count by site"
-#| fig-height: 10
-
-library(rnaturalearth)
-library(rnaturalearthdata)
-library(ggrepel)
-library(sf)
-library(ggpubr)
-library(cowplot)
-
-signbase_full_clean$time_period <-
-  factor(
-    signbase_full_clean$time_period,
-    levels = c(
-      "transitional",
-      "proto_aurignacian",
-      "early_aurignacian",
-      "evolved_aurignacian"
-    ),
-    labels = c(
-      "Transitional",
-      "Proto-Aurignacian",
-      "Early Aurignacian",
-      "Evolved Aurignacian"
-    )
-  )
-
-signcount_barplot <- ggplot(signbase_full_clean) +
-  aes(x = MedianBP, fill = time_period) +
-  geom_bar(width = 150) +
-  scale_x_reverse() +
-  labs(fill = "Time period") +
-  theme(legend.key.size = unit(0.5, "cm")) +
-  theme_minimal() +
-  scale_fill_discrete(guide = "none") +
-  xlab("Years cal BP")
-
-objectcount_barplot <- 
-ggplot(artifact_count_df) +
-  aes(reorder(site_name, artifact_count),
-      artifact_count) +
-  geom_col() +
-  coord_flip() +
-    theme_minimal() +
-    xlab("") +
-    ylab("Object count")
-
-signbase_sf <- time_unique_data_full  %>% 
-  st_as_sf(coords = c("longitude", "latitude"),
-           remove = FALSE,
-           crs = 4326)
-
-world <- ne_countries(scale = "medium", returnclass = "sf")
-
-Europe <- world[which(world$continent == "Europe"),]
-
-sign_map <- ggplot(Europe) +
-  geom_sf() +
-  geom_sf(data = signbase_sf, aes(color = time_period)) +
-  scale_colour_discrete(guide = "none")+
-  coord_sf(xlim = c(-10,30), 
-           ylim = c(35,53), 
-           expand = FALSE) +
-  geom_text_repel(data = signbase_sf,
-                  aes(x = longitude ,
-                      y = latitude,
-                      label = site_name),
-                  max.overlaps = 30,
-                  size = 2, 
-                  bg.color = "white",
-                  bg.r = 0.25) +
-  theme_minimal() +
-  facet_wrap(~time_period, ) +
-  theme(axis.text = element_blank(),
-        axis.title = element_blank(),
-        strip.text = element_text(size = 10))
-
-maps_bar_plots <- 
-plot_grid(sign_map,
-          plot_grid(signcount_barplot,
-                    objectcount_barplot,
-                    rel_widths = c(1, 0.5),
-                    labels = c("B", "C")), 
-          ncol = 1, 
-          scale = c(1, 1),
-          rel_heights = c(1, 0.5),
-          align = "vh", 
-          labels = "A",
-          axis = "lr")
-
-ggsave(maps_bar_plots, 
-      filename =  "map-and_hist-plot.png",
-       h = 10, w = 10,
-       bg = "white")
-
-knitr::include_graphics("map-and_hist-plot.png")
-
-```
-
-The distribution of objects across sites is highly skewed, as seen in @fig-site-map C.
-To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites.
-We used Mantel tests with a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types.
-The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value.
-The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation.
-The p-value quantifies the probability of obtaining the observed correlation if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
-The Mantel tests were performed using the vegan package [@Vegan2001].
-The test indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).
-
-Ages of the objects were determined by calibrating the uncalibrated ages provided by SignBase, using the rcarbon package [@Crema_Bevan_2021].
-The calibrated radiocarbon dates of the sampled data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years.
-We assigned each object to a chronological phase following the date ranges recorded in @tbl-aur-table.
-The Transitional phase has `r trans_site_count` sites and `r trans_sign_count` sign types, the Proto-Aurignacian has `r proto_site_count` sites and `r proto_sign_count` sign types, the Early Aurignacian has `r early_site_count` sites and `r early_sign_count` sign types, and the Evolved Aurignacian has `r evolved_site_count` sites and `r evolved_sign_count` sign types.
-A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (R = `r chron_r`, p = `r chron_p`).
-This confirms that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.
-
-```{r}
-library(ggraph)
-library(statnet)
-library(ape)
-library(igraph)
-
  mplot <- function(x, ..., fill_colour = "black", title = NULL, row_labels = NULL, col_labels = NULL, palette_colors = NULL) {
@@
      theme(
-      axis.text = element_text(size = rel(1)),
+       axis.text = element_text(size = rel(0.3)),
        axis.ticks = element_blank(),
@@
 
-strength_function <- function(artifact_data, group_data){
-
-  set.seed(500)
-  jac <- vegdist(artifact_data %>% mutate_all(~ as.numeric(. > 0)), "jaccard", binary = TRUE)
-  dm <- as.matrix(jac)
-  modul_matrix <-
-    graph_from_adjacency_matrix(1 - dm, mode = "undirected")
-  mem <- as_membership(parse_number(group_data$group))
-  as_group <- modularity(modul_matrix, membership = mem)
-  as_group <- round(as_group, 3)
-
-  perman <- adonis2(jac~as.factor(group_data$group), 
-                    method = "jaccard",
-                    sqrt.dist = TRUE)
-  perman <- perman %>% 
-    data_frame() %>% 
-    slice_head()
-  perman_r <- round(perman$R2, 3)
-  perman_p <- round(perman$`Pr(>F)`, 3)
-  perman_f <- round(perman$F, 3)
-
-  site_distances <- dist(cbind(group_data$longitude, group_data$latitude))
-  mantel_distance <- vegan::mantel(site_distances, jac, permutations = 1000)
-  p_value_distance_correlation <- round(mantel_distance$signif, 3)
-  r_value_distance_correlation <- round(mantel_distance$statistic, 3)
-
-  final_table <- data_frame("Modularity" = as_group, 
-                            "PerMANOVA R2" = perman_r,
-                            "PerMANOVA F" = perman_f,
-                            "PerMANOVA p" = perman_p,
-                            "Mantel R" = r_value_distance_correlation, 
-                            "Mantel p" = p_value_distance_correlation)
-
-  return(final_table)
-}
-
-trans_network <- function(artifact_data){
-    jac <- vegdist(artifact_data %>%  mutate_all(~ as.numeric(. > 0)), "jaccard", binary = TRUE)
-    dm <- as.matrix(jac)
-    disim <- 1 - dm
-    disim[disim<0.1] <- 0
-    disim_net <- network(disim,
-                         directed=F,
-                         ignore.eval=F,
-                         names.eval='weight')
-    disim_net %v% 'vertex.names' <- row.names(artifact_data)
-    set.seed(500)
-    final_plot <- ggraph(disim_net, layout="fr") +
-      geom_edge_link(aes(width = weight, 
-                         alpha=weight), 
-                       edge_colour = "black", 
-                       show.legend=F) +
-      scale_edge_width(range=c(0.5,1.5)) +
-      scale_edge_colour_gradient(low = "#CCCCCC",
-                                   high = "#000000") +
-      geom_node_point(alpha=1, size=5) +
-      geom_node_text(aes(label=row.names(artifact_data)), 
-                     size=3, 
-                     family= "Arial", 
-                     repel=T, max.overlaps = 100) +
-      theme(text= element_text(size = 15), 
-            panel.background = element_rect(fill = 'white')) +
-      border(color = "black", size = 0.5)
-    return(final_plot)
-    }
 ```
 
-If the distribution of sign types can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases.
-This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019].
-To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geography for each phase.
-We used Mantel tests with a geographic distance matrix and a Jaccard similarity matrix, with the results shown in @tbl-mantel-table.
-None of the phases have a statistically significant correlation between sign type distribution and geographical distance, so the isolation-by-distance hypothesis is rejected.
-
 ```{r}
-#| label: tbl-mantel-table
-#| tbl-cap: "Table of the results per phase for the Mantel test between a Jaccard matrix of sign type diversity and a geographic distance matrix"
-#| tbl-cap-location: bottom
-## Mantel test for geography vs signs
-
-trans_strength <- strength_function(trans_artifact_data, trans_unique_data)
- proto_strength <- strength_function(proto_artifact_data, proto_unique_data)
- early_strength <- strength_function(early_artifact_data, early_unique_data)
- evolved_strength <- strength_function(evolved_artifact_data, evolved_unique_data)
-
- full_table <- rbind(trans_strength, proto_strength, early_strength, evolved_strength) %>% 
-   mutate(`Aurignacian phase` = c("Transitional", 
-                            "Proto-Aurignacian",
-                            "Early Aurignacian", 
-                            "Evolved Aurignacian"), 
-          .before = 1)
-
-mantel_table <- full_table %>% 
-  dplyr::select(`Aurignacian phase`, `Mantel R`, `Mantel p`)
-
-kable(mantel_table)
-
-```
-
-## Cultural Group Detection Using Seriation and Network Analysis
-
-Seriation analysis is the arrangement of data into a linear order to reveal patterns [@Hahsler_Hornik_Buchta].
-We used the Brower-Kile [-@Brower_Kile_1988] seriation algorithm from the seriation package [@seriation_pckge], in which a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal.
-To investigate connectivity among clusters of sites in the seriation solution, we used network analysis to represent and quantify the relationship between sites [@Mills_2017].
-Each node is a site, and each edge is a measure of sign type distribution similarity, as computed by the Jaccard similarity algorothm.
-Results were plotted with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths, and network analysis was performed using the vegan [@Vegan2001], statnet [@statnet_package] and igraph [@igraph_package] packages.
-
-## Modularity and PerMANOVA for Cultural Group Strength
-
-To evaluate the strength of the groupings found in the seriation and network analyses, we used modularity and perMANOVA tests.
-The modularity test is a community structure network detection method, and quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006; @Newman_Girvan_2004].
-The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings.
-PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017].
-The results are expressed by an R^2^ statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, a pseudo-F statistic, and a p-value.
-The pseudo-F statistic compares the total sum of squared dissimilarities between groups to total sum of squared dissimilarities within groups.
-Larger pseudo-F statistics indicate more separation between groups.
-The p-value quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
-The modularity test and perMANOVA were performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.
-
-# Results
-
-## Cultural Group Detection
-
-```{r fig.height= 9, fig.width= 13}
-#| label: fig-seriation
-#| fig-cap: "Seriation analysis of the data, with group membership indicated by color. Restricted group is in yellow, broader group is in blue."
-
 seriation_trans <- 
@@
 
-plot_grid(seriation_trans,
+x <- plot_grid(seriation_trans,
           seriation_proto, 
@@
           align = "hv", 
-          labels = c("Transitional", "Proto", "Early", "Evolved"), label_size = 15, 
-          scale = c(1, 1, 1, 1))
+          labels = c("Transitional", "Proto", "Early", "Evolved"), label_size = 5, 
+          scale = c(1.07, 1.07, 1.07, 1.07))
+
+ggsave(x, 
+      filename =  "seriation.png",
+       h = 2.3, w = 3.7,
+       bg = "white")
 
@@
 ```{r}
-sign_type_df <- time_unique_data_full %>% 
-  pivot_longer(cols = line:star) %>% 
-  group_by(id) %>% 
-  filter(value != 0) %>% 
-  mutate(value = 1) %>% 
-  group_by(id) %>% 
-  summarise((sign_count = sum(value))) %>% 
-  mutate(sign_type_count = `(sign_count = sum(value))`, `(sign_count = sum(value))` = NULL)
-
-time_unique_data_full <- time_unique_data_full %>% 
-  left_join(sign_type_df, join_by(id))
-
-max_signs <- max(time_unique_data_full$sign_type_count)
-
-```
-
-For each of the phases, seriation identifies at least two groups, one with a restricted range of sign types (1-3 types), which tend to be simpler signs, and one with a broader range (up to `r max_signs` types), which include more complex signs.
-Simpler signs composed of fewer components, like line and notch, are the most prevalent across all time periods.
-Overall, across the four phases, the number of sites in the restricted group tends to be less than or equal to the number of sites in the broader group, with the exception of the final Evolved phase.
-
-In the Transitional phase, there is a relatively small range of sign types compared to other phases.
-The restricted group consists of Fumane, Pod Hradem, and Geissenklösterle, two of which only have a notch, and one of which has a notch and circumnotch.
-The broader group consists of Labeko Koba, Hohle Fels, and El Castillo, which vary in their sign composition and have only the line and oblique line (obline) signs in common.
-
-In the Proto-Aurignacian we see an expansion in both the number of sites and range of signs, with no overlap in sites from the Transitional phase, except for Geissenklösterle, which is present in both phases.
-The restricted group consists of Abri Pataud, Hohlenstein-Stadel, and Mladeč, which only have notches and lines, and Gatzarria which also has a dashline.
-The broader group consists of Geissenklösterle, Spy, Vogelherd, Grotte du Renne, La Ferrassie, and Les Cottés, where we see a more consistent pattern of a more diverse range of signs, compared to the broader group of the Transitional phase.
-In addition to line and notch, the obline, radial notch (radnotch), cross, hatching, and circumferential notch (circumnotch) are also frequently found in the broader group.
-
-In the Early Aurignacian, we see appearance of completely new sites with no overlap with the Proto-Aurignacian sites, as well as increased diversity of patterns within the groups.
-The restricted group consists of Grottes de Fonds-de-Forêt (radnotch and notch), Riparo Bombrini (notch), and Grotte de la Verpillière I (notch and line).
-The broader group consists of Solutré, Hohle Fels, Castanet, Cellier, Blanchard, and Trou al'Wesse, which contain a wide range of signs in a variety of combinations.
-In addition to line and notch, the obline, dot and vulva are also frequently found in the broader group.
-
-Lastly, in the Evolved Aurignacian, we see slightly more consistent groups, again with new sites with no overlap with the Early Aurignacian sites, except for Hohle Fels, which also appears in the Transitional phase.
-Vogelherd also reappears, having appeared earlier in the Proto-Aurignacian.The restricted group consists of Les Rois, Gargas, La Viña, Sirgenstein Cave, and Vindija Cave, with the line, notch, and and obline signs.
-The broader group consists of Trou Magrite, Bockstein-Törle, Vogelherd, and Hohle Fels.
-Along with line, notch, and obline, the hatching, and oblique notch (obnotch) signs are also frequently found in the broader group.
-
-```{r fig.height= 5.5, fig.width= 8}
-#| label: fig-neighbor-net
-#| fig-cap: "Network analysis of the data"
 network_trans <- trans_network(trans_artifact_data)
@@
 
-plot_grid(network_trans, network_proto, network_early, network_evolved, 
-          ncol = 2, align = "hv", labels = c("Transitional", "Proto", "Early", "Evolved"), 
-          label_size = 7, label_colour = "black", label_y = 1.02, label_x = -0.01)
+y <- plot_grid(network_trans, network_proto, network_early, network_evolved, 
+          ncol = 2, align = "hv", labels = c("Transitional", "Proto", "Early", "Evolved"), label_size = 5)
+
+ggsave(y, 
+      filename =  "network.png",
+       h = 8, w = 10,
+       bg = "white")
 ```
-
-The results of the network analysis supports the restricted-broad groupings identified in @fig-seriation, and provides additional insights into the evolution of connectivity between groups.
-In the Transitional phase, the network analysis reveals two distinct and disconnected groups.
-The restricted group consists of Geissenklösterle, Fumane, and Pod Hradem, and the broad group consisting of Hohle Fels, El Castillo, and Labeko Koba, with no connections between the groups.
-The Proto-Aurignacian phase also shows two clearly defined groups, the broad group consisting of Geissenklösterle, Spy, Vogelherd, and Grotte du Renne, and the restricted group consisting of Mladeč, Hohlenstein-Stadel, Gatzarria, and Abri Pataud.
-However, it also displays some connectivity between the two groups of sites, with La Ferrassie connected to both groups.
-Les Cottés stands out with no connections to either group.
-The Early Aurignacian shows more strongly inter-connected groups -- although the shape still suggests two main groups -- with Grotte de la Verpillière I in particular having multiple strong connections to both groups.
-Lastly, the Evolved Aurignacian again shows two main groups, although intra-group connectivity has decreased, with the broad group consisting of Hohle Fels, Trou Magrite, Vogelherd, and Bockstein-Törle, and the restricted group consisting of Gargas, La Viña, Sirgenstein Cave, Vindija Cave, and Les Rois.
-As for the Early Aurignacian, inter-connectivity between groups continues to be evident in the Evolved Aurignacian, with connections between Trou Magrite, Gargas, and Vindija Cave, and Bockstein-Törle and Gargas, however the strength of inter-connectivity is weaker than in the Early Aurignacian.
-
-In identifying the groups, we referenced both the seriation and network analyses. The results of the above tests allowed us to find two main groups of sites within each phase (@fig-seriation).
-
-## Cultural Group Strength
-
-```{r}
-#| label: tbl-strength-table
-#| tbl-cap: "Table of the results per phase for the modularity test and perMANOVA"
-#| tbl-cap-location: bottom
-
-## Modularity test
-
-strength_table <- full_table %>% 
-  dplyr::select(`Aurignacian phase`, Modularity, `PerMANOVA R2`, `PerMANOVA F`, `PerMANOVA p`)
-
-kable(strength_table)
-
-proto_modularity <- proto_strength %>% 
-  dplyr::select(Modularity)
-evolved_modularity <- evolved_strength %>% 
-  dplyr::select(Modularity)
-
-trans_perm_r <- trans_strength %>% 
-  dplyr::select(`PerMANOVA R2`)
-early_perm_r <- early_strength %>% 
-  dplyr::select(`PerMANOVA R2`)
-
-trans_perm_p <- trans_strength %>% 
-  dplyr::select(`PerMANOVA p`)
-
-trans_perm_f <- trans_strength %>% 
-  dplyr::select(`PerMANOVA F`)
-early_perm_f <- early_strength %>% 
-  dplyr::select(`PerMANOVA F`)
-```
-
-The modularity score quantifies the strength of the groupings, showing that the variations in grouping strength between phases are minimal.
-The Proto-Aurignacian exhibits the strongest groupings, with a score of `r proto_modularity`, and the Evolved Aurignacian has the weakest groupings, with a score of `r evolved_modularity`.
-There is a slight increase in modularity between the Transitional period and the Proto-Aurignacian, before an even slighter decrease into the Early Aurignacian, followed by a larger increase with the Evolved Aurignacian.
-
-The perMANOVA R^2^ statistic shows the proportion of sign variation that can be explained by group membership.
-The Transitional period has the highest score of `r trans_perm_r` while the Early Aurignacian has the lowest score of `r early_perm_r`.
-We see a large decline in scores between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian periods.
-The pseudo-F statistic quantifies the difference between groups, with the Transitional phase having the highest value of `r trans_perm_f`, and the Early Aurignacian having the lowest value of `r early_perm_f`.
-Similar to the R^2^ statistic, we see a decline in values between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian period.
-The perMANOVA p-value indicates a statistically significant difference in the distribution of signs between the restricted and broad groups for the Proto- , Early, and Evolved Aurignacian phases, but not the Transitional phase (p = `r trans_perm_p`).
-
-# Discussion
-
-The first aim of our study was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and boundaries.
-The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.
-@fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
-The modularity tests, as shown in @tbl-strength-table, show that the groupings are strongest in the Transitional period and weakest in the Early Aurignacian, and the perMANOVA tests tell us that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian experiences the least.
-The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian experiences the highest correlation.
-
-```{r fig.height= 7, fig.width=10}
-#| label: fig-group-map
-#| fig-cap: "Map of sites faceted by time period and colored by group number"
-
-ggplot(Europe) +
-  geom_sf() +
-  geom_sf(data = signbase_sf, aes(color = group)) +
-  coord_sf(xlim = c(-10,30), 
-           ylim = c(35,53), 
-           expand = FALSE) +
-  geom_text_repel(data = signbase_sf,
-                  aes(x = longitude ,
-                      y = latitude,
-                      label = site_name),
-                  max.overlaps = 30,
-                  size = 2,
-                  bg.color = "white",
-                  bg.r = 0.25) +
-  theme_minimal() +
-  facet_wrap(~time_period) +
-  theme(axis.text = element_blank(),
-        axis.title = element_blank(),
-        strip.text = element_text(size = 10))
-
-```
-
-The geographic boundaries of the groups change from phase to phase (@fig-group-map).
-In the Transitional phase, sites are widely spread out across Western Europe, and tend to be located closely to the other sites in their groups.
-The exception is Hohle Fels, which, instead of being located further west with the rest of the sites in the broader group, is located further east, closer to the sites in the restricted group, and almost overlapping with Geissenklösterle.
-In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups.
-In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together.
-In the Evolved Aurignacian, we see a wider distribution and more distance between sites.
-
-One factor that may influence the geographic structure of sites is the possibility that some sites functioned as aggregation sites.
-The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for various subsistence or social reasons.
-Aggregation sites are the places where those fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg].
-Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
-It is therefore possible that the sites with multiple connections to both groups, such as La Ferrassie in the the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignroacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.
-However, without examining environmental data, this remains only a hypothesis at this time.
-
-The second aim of our study was to determine what these patterns tell us about dynamics of social organization throughout the Aurignacian.
-The results suggest that the Transitional phase was sparsely populated, with minimal interaction between loosely connected groups.
-For the Proto-Aurignacian, the results suggest a large expansion in social activity, as well as strongly intra-connected groups with some inter-connectivity present, with geographic distance having the least influence on sign distribution.
-The Early Aurignacian experiences a large surge in inter-connectivity, as well as intra-connectivity strength, with a great deal of connection between groups and an increase in the correlation with geographic distance.
-The Evolved Aurignacian experiences a decline in connection strength, although it maintains degrees of inter-connectivity.
-
-These results are consistent with existing data on the environmental and population change established during these phases.
-@Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Proto/Early Aurignacian (Aur-P1) settlement being reached around 41 k BP.
-This rapid expansion could be what stimulated the emergence of signs as a marker of group membership.
-Other studies have also found a similar associations in other regions and time period between increases in quantity and diversity of art, and increases in population density [@smith1992Design; @hays1993symbol; @McDonald_Veth_Lilley_2006; @Lourandos_1985; @barton_clark_cohen1994; @mcdonald2012social; @bernardini2005reconsidering; @wiessner1984reconsidering; @smith1992].
-
-Additionally, @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations.
-This likely explains why we see a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance.
-Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian (Aur-P2) theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions.
-@Shao_Wegener_Klein_Schmidt_Weniger_2024 also finds an increase in expansion marking the beginning of the Evolved Aurignacian.
-Geographic expansion resulting in less group interconnection could explain why we see weaker, more dispersed connections in the Evolved Aurignacian.
-
-This study adds to our knowledge of social networks within the Aurignacian period, as our results suggest dynamic patterns of social exchange and connection.
-Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996].
-@Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our network analysis shows evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted.
-We hope that dividing the data up into smaller time periods serves to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023].
-Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019].
-
-One possible limitation is the potential ambiguity that arises with some of the sign types.
-While @Von_Petzinger_2009 argues in favor of Aurignacian geometric signs as symbolic expression, potential ambiguity still remains over whether function is consistent across all sign type occurrences, particularly in regards to the simpler sign types composed from fewer elements.
-We included the more ambiguous sign types because similar sign occurrence may still serve as evidence of cultural connectivity regardless of function, and additionally, greater quantities of included sign types serves to establish stronger connections [@Gravel-Miguel_Coward_2023].
-Potential ambiguity may be addressed in the future through the usage of the data recently released in SignBase 2.0 covering sites in Germany, which contains information on the specific order and patterns in which signs occurred.
-
-# Conclusion
-
-This study both adds to our understanding of Aurignacian social dynamics and suggests potential new directions for future research.
-The observed social dynamics produced by our methods align with the existing data on the environmental and population conditions of the Upper Paleolithic.
-We feel that our research opens up new opportunities for the application of these methods to data on artwork variation in other regions and time periods.
-Our research also suggests an earlier emergence of dynamic inter-connectivity than previously predicted, as we see weaker inter-connections beginning to emerge by 41 k BP, and stronger inter-connections by 38 k BP.
-Future research directions might involve the analysis of environmental conditions and their interactions with social dynamics, in order to determine whether some of the sites that we have pointed to as potential aggregation sites function as such.
-Another potential future research direction would be the usage of the more specific data released in SignBase 2.0 in order to counter potential sign type ambiguity.
-
-# References

### 2025-04-14 Extra/modularity.R

@@ -1,0 +1,60 @@
+library(robin)
+
+artifact_data <- trans_artifact_data %>% 
+  select_if(~ !is.numeric(.) || sum(.) != 0) %>% 
+  mutate_all(~ as.numeric(. > 0))
+
+jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
+dm <- as.matrix(jac)
+
+disim <- 1 - dm
+disim[disim<0.2] <- 0
+disim_net <- network(disim,
+                     directed=F,
+                     ignore.eval=F,
+                     names.eval='weight')
+
+disim_net %v% 'vertex.names' <- row.names(artifact_data)
+
+modul_matrix <-
+  graph_from_adjacency_matrix(1 - dm, mode = "undirected")
+
+
+mem <- as_membership(parse_number(trans_unique_data$group))
+as_group <- modularity(modul_matrix, membership = mem)
+
+
+library(intergraph)
+
+ig <- asIgraph(disim_net)
+
+as_group <- modularity(ig, membership = mem)
+
+
+g <- graph.adjacency(
+  as.matrix(as.dist(cor(base::t(artifact_data), method="pearson"))),
+  mode="undirected",
+  weighted=TRUE,
+  diag=FALSE
+)
+
+g <- make_graph(edges = row.names(artifact_data), directed = FALSE)
+
+g$weight <- jac
+
+
+g <- graph.adjacency(
+  as.matrix(vegdist(artifact_data, "jaccard", binary = TRUE)),
+  mode="undirected",
+  weighted=TRUE,
+  diag=FALSE
+)
+
+
+g <- graph.adjacency(
+  as.matrix(as.dist(cor(t(artifact_data), method="pearson"))),
+  mode="undirected",
+  weighted=TRUE,
+  diag=FALSE
+)
+as_group <- modularity(g, membership = mem)

### 2025-04-14 endofquarterpaper.qmd

@@
 Drawing on ethnographic studies that show how body decoration and ornamentation indicates ethno-linguistic identity, @Vanhaeren_d used Aurignacian personal ornamentation as a proxy for ethno-linguistic diversity.
-They analyzed 157 distinct ornament types from 98 Aurignacian sites in Europe and the Near East.
-Geographically cohesive groups sharing similar ornament type associations were identified by seriation, correspondence, and GIS analyses.
+The variation in personal ornament types is interpreted as evidence of long-lasting cultural differences and as robustly establishing the ethno-linguistic diversity of the Aurignacian.
+157 distinct ornament types from 98 Aurignacian sites in Europe and the Near East were analyzed, identifying geographically cohesive groups sharing similar ornament type associations by seriation, correspondence, and GIS analyses.
 These groups sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.
-@Vanhaeren_d interpret the variation in personal ornament types as evidence of long-lasting cultural differences and as robustly establishing the ethno-linguistic diversity of the Aurignacian.
 
@@
 
-The first aim of our study was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and boundaries.
+Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and boundaries.
 The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.
 @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
-The modularity tests, as shown in @tbl-strength-table, show that the groupings are strongest in the Transitional period and weakest in the Early Aurignacian, and the perMANOVA tests tell us that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian experiences the least.
-The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian experiences the highest correlation.
+The modularity tests, as shown in @tbl-strength-table, show that the groupings are most distinct in the Transitional period and least distinct in the Early Aurignacian.
+The perMANOVA tests show that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.
+The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation.
 
@@
 The geographic boundaries of the groups change from phase to phase (@fig-group-map).
-In the Transitional phase, sites are widely spread out across Western Europe, and tend to be located closely to the other sites in their groups.
-The exception is Hohle Fels, which, instead of being located further west with the rest of the sites in the broader group, is located further east, closer to the sites in the restricted group, and almost overlapping with Geissenklösterle.
-In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups.
-In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together.
-In the Evolved Aurignacian, we see a wider distribution and more distance between sites.
+In the Transitional phase, sites are widely spread out across Western Europe.
+This wider distribution may be a factor in weak connections seen in @fig-neighbor-net.
+In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the stronger connectivity seen there.
+In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together, which may be associated with the increased connectivity in that phase.
+In the Evolved Aurignacian, we see a wider distribution and more distance between sites, which may be associated with the reduced connectivity seen in that phase.
 

### 2025-04-14 endofquarterpaper.qmd

@@
 
+## Grouping Dynamics Across the Four Phases
+
 Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and boundaries.
@@
 This wider distribution may be a factor in weak connections seen in @fig-neighbor-net.
-In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the stronger connectivity seen there.
+In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the slightly stronger connectivity seen there.
+This phase is also associated with an increase in sites and sign types, which may be correlated with the increased proximity and population density.
 In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together, which may be associated with the increased connectivity in that phase.
@@
 One factor that may influence the geographic structure of sites is the possibility that some sites functioned as aggregation sites.
-The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for various subsistence or social reasons.
-Aggregation sites are the places where those fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg].
+The aggregation/dispersion pattern is first defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for various subsistence or social reasons.
+Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg].
 Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
-It is therefore possible that the sites with multiple connections to both groups, such as La Ferrassie in the the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignroacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.
-However, without examining environmental data, this remains only a hypothesis at this time.
-
-The second aim of our study was to determine what these patterns tell us about dynamics of social organization throughout the Aurignacian.
+It is therefore possible that sites with multiple connections between groups [@fig-neighbor-net], such as La Ferrassie in the the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.
+
+## Dynamics of Social Organization
+
+The second aim was to determine what these patterns tell us about dynamics of social organization throughout the Aurignacian.
 The results suggest that the Transitional phase was sparsely populated, with minimal interaction between loosely connected groups.
-For the Proto-Aurignacian, the results suggest a large expansion in social activity, as well as strongly intra-connected groups with some inter-connectivity present, with geographic distance having the least influence on sign distribution.
-The Early Aurignacian experiences a large surge in inter-connectivity, as well as intra-connectivity strength, with a great deal of connection between groups and an increase in the correlation with geographic distance.
-The Evolved Aurignacian experiences a decline in connection strength, although it maintains degrees of inter-connectivity.
-
-These results are consistent with existing data on the environmental and population change established during these phases.
-@Shao_Wegener_Klein_Schmidt_Weniger_2024 tells us that from 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans, before being followed by rapid expansion from 43.25 - 41 k BP, with the maximum extent of Proto/Early Aurignacian (Aur-P1) settlement being reached around 41 k BP.
-This rapid expansion could be what stimulated the emergence of signs as a marker of group membership.
-Other studies have also found a similar associations in other regions and time period between increases in quantity and diversity of art, and increases in population density [@smith1992Design; @hays1993symbol; @McDonald_Veth_Lilley_2006; @Lourandos_1985; @barton_clark_cohen1994; @mcdonald2012social; @bernardini2005reconsidering; @wiessner1984reconsidering; @smith1992].
-
-Additionally, @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 at the very beginning of the Early Aurignacian resulted in an expansion into new environmental niches, which they hypothesize was associated with expansion of social networks both within and between populations.
-This likely explains why we see a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased correlation with geographic distance.
-Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian (Aur-P2) theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions.
+For the Proto-Aurignacian, the results suggest a large expansion in social interaction within groups, as well as some degree of interaction between groups, with geographic distance having the least influence on sign distribution [@tbl-mantel-table].
+The Early Aurignacian experiences a large surge in interaction between groups, as well as interaction within groups, and an increase in the correlation with geographic distance.
+The Evolved Aurignacian experiences a decline in between-group interaction, although it maintains within-group interaction.
+
+These results are consistent with previous work on the environmental and population change during these phases.
+From 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans [@Shao_Wegener_Klein_Schmidt_Weniger_2024].
+This was followed by rapid expansion from 43.25 - 41 k BP (Transitional/Proto), with the maximum extent of Proto/Early Aurignacian settlement being reached around 41 k BP.
+This rapid population expansion could be what stimulated the emergence of signs as a marker of group membership.
+Other studies have also found similar associations in other regions and time periods between increases in quantity and diversity of art, and increases in population density \[@smith1992Design; @hays1993symbol; @McDonald_Veth_Lilley_2006; @Lourandos_1985; @barton_clark_cohen1994; @mcdonald2012social; @bernardini2005reconsidering; @wiessner1984reconsidering; @smith1992\].
+
+Similarly, @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 (HS4) at the very beginning of the Early Aurignacian resulted in a expansion of people into new environmental niches, which they hypothesize was associated with an expansion of social networks both within and between populations.
+This correlates with our results which show a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased sign type correlation with geographic distance.
+Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions.
 @Shao_Wegener_Klein_Schmidt_Weniger_2024 also finds an increase in expansion marking the beginning of the Evolved Aurignacian.
-Geographic expansion resulting in less group interconnection could explain why we see weaker, more dispersed connections in the Evolved Aurignacian.
-
-This study adds to our knowledge of social networks within the Aurignacian period, as our results suggest dynamic patterns of social exchange and connection.
+This geographic expansion may explain our results, where we find weaker, more dispersed connections in the Evolved Aurignacian.
+
+\<--
+
+cite signbase 2.0–\>.
+
+# Conclusion
+
+This study adds to our knowledge of social networks within the Aurignacian period, as our results document dynamic patterns of social exchange and connection.
 Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996].
-@Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our network analysis shows evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted.
-We hope that dividing the data up into smaller time periods serves to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023].
+@Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our results show evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted.
+We attempted to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023] by dividing the data up into smaller time periods.
 Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019].
 
-One possible limitation is the potential ambiguity that arises with some of the sign types.
+One limitation is the potential ambiguity that arises with some of the sign types.
 While @Von_Petzinger_2009 argues in favor of Aurignacian geometric signs as symbolic expression, potential ambiguity still remains over whether function is consistent across all sign type occurrences, particularly in regards to the simpler sign types composed from fewer elements.
 We included the more ambiguous sign types because similar sign occurrence may still serve as evidence of cultural connectivity regardless of function, and additionally, greater quantities of included sign types serves to establish stronger connections [@Gravel-Miguel_Coward_2023].
-Potential ambiguity may be addressed in the future through the usage of the data recently released in SignBase 2.0 covering sites in Germany, which contains information on the specific order and patterns in which signs occurred.
-
-# Conclusion
-
-This study both adds to our understanding of Aurignacian social dynamics and suggests potential new directions for future research.
-The observed social dynamics produced by our methods align with the existing data on the environmental and population conditions of the Upper Paleolithic.
-We feel that our research opens up new opportunities for the application of these methods to data on artwork variation in other regions and time periods.
-Our research also suggests an earlier emergence of dynamic inter-connectivity than previously predicted, as we see weaker inter-connections beginning to emerge by 41 k BP, and stronger inter-connections by 38 k BP.
-Future research directions might involve the analysis of environmental conditions and their interactions with social dynamics, in order to determine whether some of the sites that we have pointed to as potential aggregation sites function as such.
-Another potential future research direction would be the usage of the more specific data released in SignBase 2.0 in order to counter potential sign type ambiguity.
+This ambiguity may be addressed in the future through the usage of the data recently released in SignBase 2.0 covering sites in Germany, which contains more detailed information on the specific order and patterns in which signs occurred on individual objects
+
+\<--
+
+cite signbase 2.0–\>.
 

### 2025-04-14 endofquarterpaper.qmd
update bad citation

@@
 This rapid population expansion could be what stimulated the emergence of signs as a marker of group membership.
-Other studies have also found similar associations in other regions and time periods between increases in quantity and diversity of art, and increases in population density \[@smith1992Design; @hays1993symbol; @McDonald_Veth_Lilley_2006; @Lourandos_1985; @barton_clark_cohen1994; @mcdonald2012social; @bernardini2005reconsidering; @wiessner1984reconsidering; @smith1992\].
+Other studies have also found similar associations in other regions and time periods between increases in quantity and diversity of art, and increases in population density [@smith1992Design; @hays1993symbol; @McDonald_Veth_Lilley_2006; @Lourandos_1985; @barton_clark_cohen1994; @mcdonald2012social; @bernardini2005reconsidering; @wiessner1984reconsidering; @smith1992].
 
@@
 
-\<--
-
-cite signbase 2.0–\>.
+<-- cite signbase 2.0 -->.
 

### 2025-04-14 endofquarterpaper.qmd
stash

--- 2025-04-14 endofquarterpaper.qmd
@@
 
-## Modularity and PerMANOVA for Cultural Group Strength
-
-To evaluate the strength of the groupings found in the seriation and network analyses, we used modularity and perMANOVA tests.
+## Modularity and PerMANOVA for Cultural Connection Strength
+
+To evaluate the strength of the connections of the groups found in the seriation and network analyses, we used modularity and perMANOVA tests.
 The modularity test is a community structure network detection method, and quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006; @Newman_Girvan_2004].
@@
 
-## Cultural Group Strength
+## Cultural  Strength
 

### 2025-04-29 Extra/undergraduate-research-symposium-images.qmd
stash

@@
 ```{r}
-seriation_trans <- 
-  produce_clusters(trans_artifact_data, 
+seriation_trans <- produce_clusters(trans_artifact_data, 
                                     trans_unique_data, 
@@
 
-seriation_proto <- 
-  produce_clusters(proto_artifact_data, 
+seriation_proto <- produce_clusters(proto_artifact_data,  
                                     proto_unique_data, 
-                   # BM: not sure why this one gets the 
-                   # colour differently from the others
                                     color_palette = c("#FFFFB3", "#8DD3C7"),
-                   method = "seriation")
+                                    method ="seriation")
 
-seriation_early <- 
-  produce_clusters(early_artifact_data, 
+seriation_early <- produce_clusters(early_artifact_data, 
                                     early_unique_data, 
@@
 
-seriation_evolved <- 
-  produce_clusters(evolved_artifact_data, 
+seriation_evolved <- produce_clusters(evolved_artifact_data, 
                                       evolved_unique_data, 
                                       method = "seriation")
+```
 
-x <- plot_grid(seriation_trans,
-          seriation_proto, 
-          seriation_early , 
-          seriation_evolved, 
-          ncol = 2, 
-          align = "hv", 
-          labels = c("Transitional", "Proto", "Early", "Evolved"), label_size = 5, 
-          scale = c(1.07, 1.07, 1.07, 1.07))
-
-ggsave(x, 
-      filename =  "seriation.png",
-       h = 2.3, w = 3.7,
-       bg = "white")
+```{r}
+network_trans <- produce_clusters(trans_artifact_data, trans_unique_data, method = "network")
+network_proto <- produce_clusters(proto_artifact_data, proto_unique_data, method = "network")
+network_early <- produce_clusters(early_artifact_data, early_unique_data, method = "network")
+network_evolved <- produce_clusters(evolved_artifact_data, evolved_unique_data, method = "network")
 
@@
 ```{r}
-network_trans <- trans_network(trans_artifact_data)
-network_proto <- produce_clusters(proto_artifact_data, proto_unique_data, method = "network")
-network_early <- produce_clusters(early_artifact_data, early_unique_data, method = "network")
-network_evolved <- produce_clusters(evolved_artifact_data, evolved_unique_data, method = "network")
+ser_net_trans <- plot_grid(seriation_trans, network_trans, align = "hv", labels = "AUTO", rel_widths = c(1.3, 1.4))
+ser_net_proto <- plot_grid(seriation_proto, network_proto, align = "hv", rel_widths = c(1.4, 1.4))
+ser_net_early <- plot_grid(seriation_early, network_early, align = "hv", rel_widths = c(1.65, 1.8))
+ser_net_evo <- plot_grid(seriation_evolved, network_evolved, align = "hv", rel_widths = c(1.4, 1.4))
 
-y <- plot_grid(network_trans, network_proto, network_early, network_evolved, 
-          ncol = 2, align = "hv", labels = c("Transitional", "Proto", "Early", "Evolved"), label_size = 5)
+final <- plot_grid(ser_net_trans, ser_net_proto, ser_net_early, ser_net_evo, align = "hv", ncol = 1, labels = c("Transitional", "Proto", "Early", "Evolved"))
 
-ggsave(y, 
-      filename =  "network.png",
-       h = 8, w = 10,
-       bg = "white")
+ggsave(final, 
+      filename =  "poster.png",
+       bg = "white", h = 9, w = 4.5)
 ```
+
+```{r}
+ser_net_trans <- plot_grid(seriation_trans, network_trans, align = "hv", labels = "AUTO", ncol = 1)
+ser_net_proto <- plot_grid(seriation_proto, network_proto, align = "hv", ncol = 1)
+ser_net_early <- plot_grid(seriation_early, network_early, align = "hv", ncol = 1)
+ser_net_evo <- plot_grid(seriation_evolved, network_evolved, align = "hv", ncol = 1)
+
+final <- plot_grid(ser_net_trans, ser_net_proto, ser_net_early, ser_net_evo, align = "hv", ncol = 4, labels = c("Transitional", "Proto", "Early", "Evolved"))
+
+ggsave(final, 
+      filename =  "poster2.png",
+       bg = "white", w = 12)
+
+```
+
+```{r}
+ser_plot <- plot_grid(seriation_trans, seriation_proto, seriation_early, seriation_early, align = "hv", ncol = 1, labels = c("Transitional", "Proto", "Early", "Evolved"), label_size = 9)
+net_plot <- plot_grid(network_trans, network_proto, network_early, network_evolved, align = "hv", ncol = 1)
+
+  final <- plot_grid(ser_plot, net_plot, align = "hv", ncol = 2, rel_widths = c(1.3, 1), labels = "AUTO", label_x = c(0, -0.07))
+
+ggsave(final, 
+      filename =  "poster3.png", bg = "white", w = 5, h = 7)
+```

### 2025-04-29 endofquarterpaper.qmd
stash
minor edits

@@
 
-trans_network <- function(artifact_data){
-    jac <- vegdist(artifact_data %>%  mutate_all(~ as.numeric(. > 0)), "jaccard", binary = TRUE)
-    dm <- as.matrix(jac)
-    disim <- 1 - dm
-    disim[disim<0.1] <- 0
-    disim_net <- network(disim,
-                         directed=F,
-                         ignore.eval=F,
-                         names.eval='weight')
-    disim_net %v% 'vertex.names' <- row.names(artifact_data)
-    set.seed(500)
-    final_plot <- ggraph(disim_net, layout="fr") +
-      geom_edge_link(aes(width = weight, 
-                         alpha=weight), 
-                       edge_colour = "black", 
-                       show.legend=F) +
-      scale_edge_width(range=c(0.5,1.5)) +
-      scale_edge_colour_gradient(low = "#CCCCCC",
-                                   high = "#000000") +
-      geom_node_point(alpha=1, size=5) +
-      geom_node_text(aes(label=row.names(artifact_data)), 
-                     size=3, 
-                     family= "Arial", 
-                     repel=T, max.overlaps = 100) +
-      theme(text= element_text(size = 15), 
-            panel.background = element_rect(fill = 'white')) +
-      border(color = "black", size = 0.5)
-    return(final_plot)
-    }
 ```
@@
 #| fig-cap: "Network analysis of the data"
-network_trans <- trans_network(trans_artifact_data)
+network_trans <- produce_clusters(trans_artifact_data, trans_unique_data, method = "network")
 network_proto <- produce_clusters(proto_artifact_data, proto_unique_data, method = "network")
@@
 
-## Cultural  Strength
+## Cultural Connection Strength
 
@@
 Similar to the R^2^ statistic, we see a decline in values between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian period.
-The perMANOVA p-value indicates a statistically significant difference in the distribution of signs between the restricted and broad groups for the Proto- , Early, and Evolved Aurignacian phases, but not the Transitional phase (p = `r trans_perm_p`).
+The perMANOVA p-value indicates a statistically significant difference in the distribution of signs between the restricted and broad groups for the Proto , Early, and Evolved Aurignacian phases, but not the Transitional phase (p = `r trans_perm_p`).
 
@@
 
-## Grouping Dynamics Across the Four Phases
+## Grouping Dynamics Across the Four Phases of the Aurignacian
 
@@
 
-## Dynamics of Social Organization
+## Social Organization Dynamics Throughout the Aurignacian
 
@@
 
-<-- cite signbase 2.0 -->.
-
 # Conclusion
@@
 We included the more ambiguous sign types because similar sign occurrence may still serve as evidence of cultural connectivity regardless of function, and additionally, greater quantities of included sign types serves to establish stronger connections [@Gravel-Miguel_Coward_2023].
-This ambiguity may be addressed in the future through the usage of the data recently released in SignBase 2.0 covering sites in Germany, which contains more detailed information on the specific order and patterns in which signs occurred on individual objects
-
-\<--
-
-cite signbase 2.0–\>.
+This ambiguity may be addressed in the future through the usage of the data recently released in SignBase 2.0 covering sites in Germany, which contains more detailed information on the specific order and patterns in which signs occurred on individual objects.
 

### 2025-04-29 endofquarterpaper.qmd
social structure paragraph

@@
 
+It is possible that variation in diversity of sign types among sites may be representative of variation in social arrangement or social context.
+Despite the tendency to conceive of hunter-gatherers as monolithic, previous studies have established a diverse range of hunter-gatherer social arrangements [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017].
+Ethnographic studies of hunter-gatherers were utilized by @Bloch_1977 to theorize a correlation between the amounts of social structure, instituted hierarchy, and ritual communication in a society.
+@Conkey_1985 built upon that theorized correlation, arguing that artwork and other forms of material culture can serve as a form of ritual communication, and that an increase in diversity may suggest more elaborate or complex social contexts, specifically in regards to the Paleolithic.
+It is therefore possible that changes in diversity from site to site, particularly the larger grouping structure of broader versus restricted signs, may be representative of two different types of social arrangment..
+
 ```{r fig.height= 7, fig.width=10}

### 2025-05-02 endofquarterpaper.qmd
edit

@@
 
+```{r}
+
+library(tabula)
+library(rnaturalearth)
+
+as_factor <- function(x, reverse = FALSE) {
+  lvl <- unique(x)
+  if (reverse) {
+    lvl <- rev(lvl)
+  }
+  factor(x, levels = lvl)
+}
+
+to_long <- function(from, factor = FALSE, reverse = FALSE) {
+  x <- data.frame(
+    row = as.vector(row(from, as.factor = factor)),
+    column = as.vector(col(from, as.factor = factor)),
+    value = as.vector(from),
+    stringsAsFactors = FALSE
+  )
+  if (factor) {
+    x$row <- as_factor(x$row, reverse = reverse)
+    x$column <- as_factor(x$column, reverse = reverse)
+  }
+  return(x)
+}
+
+autoplot.DiversityIndex <- function(object, ...) {
+  ## Prepare data
+  count <- cbind.data.frame(
+    label = object@labels,
+    x = object@size,
+    y = object@.Data
+  )
+
+  ## Simulated assemblages
+  gg_sim <- NULL
+  if (length(object@simulation) != 0) {
+    # Build a long table for ggplot2
+    refined <- object@simulation
+
+    sim_stacked <- to_long(refined[, -c(1)], factor = TRUE)
+    sim <- cbind.data.frame(
+      size = refined[, 1],
+      sim_stacked,
+      Estimate = ifelse(sim_stacked[["column"]] == "mean", "mean", "conf. int.")
+    )
+    gg_sim <- ggplot2::geom_path(
+      mapping = ggplot2::aes(
+        x = .data$size,
+        y = .data$value,
+        colour = .data$Estimate,
+        group = .data$column
+      ),
+      data = sim,
+      na.rm = TRUE,
+      inherit.aes = FALSE
+    )
+  }
+
+  y_lab <- switch (
+    class(object),
+    HeterogeneityIndex = "Heterogeneity",
+    EvennessIndex = "Evenness",
+    RichnessIndex = "Richness",
+    "Diversity"
+  )
+
+  ## ggplot
+  ggplot2::ggplot(data = count) +
+    ggplot2::aes(x = .data$x, y = .data$y, label = .data$label) +
+    ggplot2::geom_point() +
+    gg_sim +
+    ggplot2::scale_x_log10(name = "Sample size") +
+    ggplot2::scale_y_continuous(
+      name = sprintf("%s (%s)", y_lab, object@method)
+    )
+}
+
+plot_diversity_fn <- function(input_df){
+
+signs <-
+  input_df %>% 
+  select(
+         site_name,
+         # c(line:star)
+         # we are looking at the more visually complex signs
+         cross, 
+         v, 
+         star, 
+         zigzag, 
+         zigzagrow, 
+         grid, 
+         hashtag, 
+         hatching, 
+         anthropomorph,
+         zoomorph, 
+         vulva) %>% 
+  group_by(site_name) %>% 
+  summarise(across(where(is.numeric), ~ sum(.))) %>% 
+  column_to_rownames("site_name")
+
+diversity_index <- 
+  heterogeneity(signs, method = "shannon")
+
+diversity_sim <- simulate(diversity_index)
+
+autoplot.DiversityIndex(diversity_sim) +
+  geom_text_repel(aes(label = diversity_sim@labels),
+                  size = 3) +
+  guides(colour = "none") +
+  theme_minimal(base_size = 8)
+}
+
+signbase_full_clean_diversity_plot_by_period <- 
+time_unique_data_full %>% 
+  nest_by(time_period) %>% 
+  mutate(plot = list(plot_diversity_fn(data)))
+
+library(cowplot)
+plot_grid(plotlist =  signbase_full_clean_diversity_plot_by_period$plot,
+          labels = signbase_full_clean_diversity_plot_by_period$time_period,
+          label_size = 8)
+```
+
 ## Social Organization Dynamics Throughout the Aurignacian

### 2025-05-02 Extra/signbase-diversity.R

@@ -1,8 +1,3 @@
-
-
-#---------------------------------------------
-# BM from Aleah and Eva
-
 library(tabula)
 library(tidyverse)
 library(rnaturalearth)
@@ -33,6 +28,7 @@ signbase_transformed %>%
       fill = name) +
   geom_col() +
   coord_flip()
+
 
 ## Measure diversity by comparing to simulated assemblages
 set.seed(12345)
@@ -228,6 +224,5 @@ library(cowplot)
 plot_grid(plotlist =  signbase_full_clean_diversity_plot_by_period$plot,
           labels = signbase_full_clean_diversity_plot_by_period$time_period,
           label_size = 8)
-
 
 

### 2025-05-05 Extra/undergraduate-research-symposium-images.qmd
edits

@@
 ```
+
+```{r}
+sign_type_df <- signbase_full_clean %>% 
+  pivot_longer(cols = line:star) %>% 
+  filter(value != 0) %>% 
+  mutate(sign_type = name) %>% 
+  group_by(sign_type) %>% 
+  summarise(value = n()) %>% 
+  arrange(desc(value))
+
+sign_type_plot <- ggplot(sign_type_df) +
+  aes(reorder(sign_type, value), x = sign_type, y = value) +
+  geom_col() +
+  xlab("Sign type") +
+  ylab("Count")+
+  coord_flip()
+
+ggsave(sign_type_plot, 
+      filename =  "signtype.png",
+       bg = "white", w = 3.5)
+
+```

### 2025-05-05 endofquarterpaper.qmd
edits

@@
 
-It is possible that variation in diversity of sign types among sites may be representative of variation in social arrangement or social context.
-Despite the tendency to conceive of hunter-gatherers as monolithic, previous studies have established a diverse range of hunter-gatherer social arrangements [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017].
-Ethnographic studies of hunter-gatherers were utilized by @Bloch_1977 to theorize a correlation between the amounts of social structure, instituted hierarchy, and ritual communication in a society.
-@Conkey_1985 built upon that theorized correlation, arguing that artwork and other forms of material culture can serve as a form of ritual communication, and that an increase in diversity may suggest more elaborate or complex social contexts, specifically in regards to the Paleolithic.
-It is therefore possible that changes in diversity from site to site, particularly the larger grouping structure of broader versus restricted signs, may be representative of two different types of social arrangment..
+Variation in diversity of sign types among sites may be representative of variation in social arrangement or social context.
+Despite the tendency to conceive of hunter-gatherers as monolithic, previous studies have established a potential for variation in hunter-gatherer social arrangements [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017].
+@Bloch_1977 drew upon ethnographic studies of hunter-gatherers to theorize a correlation between the amounts of social structure, instituted hierarchy, and ritual communication within a society.
+@Conkey_1985 built upon that theorized correlation to argue that artwork and other forms of material culture can serve as a form of ritual communication, and that an increase in diversity within those material cultures may suggest more elaborate or complex social contexts.
+Changes in sign diversity from site to site, particularly the larger grouping structure of broader versus restricted signs, may be representative of two different social arrangement types, such as, for example, a more hierarchical structure versus a less hierarchical structure.
 
@@
   geom_text_repel(aes(label = diversity_sim@labels),
-                  size = 3) +
+                  size = 3, max.overlaps = 15) +
   guides(colour = "none") +
-  theme_minimal(base_size = 8)
+  theme_minimal(base_size = 8) +
+  border(color = "black", size = 0.5)
 }

### 2025-05-06 endofquarterpaper.qmd
edit

@@
 @Conkey_1985 built upon that theorized correlation to argue that artwork and other forms of material culture can serve as a form of ritual communication, and that an increase in diversity within those material cultures may suggest more elaborate or complex social contexts.
-Changes in sign diversity from site to site, particularly the larger grouping structure of broader versus restricted signs, may be representative of two different social arrangement types, such as, for example, a more hierarchical structure versus a less hierarchical structure.
+If the signs served functions of ritual communication, then changes in sign diversity from site to site, particularly the larger grouping structure of broader versus restricted signs, may be representative of two different social arrangement types, such as, for example, a more hierarchical structure versus a less hierarchical structure.
 
@@
 Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
-It is therefore possible that sites with multiple connections between groups [@fig-neighbor-net], such as La Ferrassie in the the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.
+It is therefore possible that sites with multiple connections between groups [@fig-neighbor-net], such as La Ferrassie in the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.
 
@@
   input_df %>% 
-  select(
-         site_name,
-         # c(line:star)
-         # we are looking at the more visually complex signs
-         cross, 
-         v, 
-         star, 
-         zigzag, 
-         zigzagrow, 
-         grid, 
-         hashtag, 
-         hatching, 
-         anthropomorph,
-         zoomorph, 
-         vulva) %>% 
+  select(site_name,
+         c(line:star)) %>% 
   group_by(site_name) %>% 

### 2025-05-06 Extra/undergraduate-research-symposium-images.qmd
edit

--- 2025-04-29 Extra/undergraduate-research-symposium-images.qmd
@@
 ```
+
+```{r}
+sign_type_df <- signbase_full_clean %>% 
+  pivot_longer(cols = line:star) %>% 
+  filter(value != 0) %>% 
+  mutate(sign_type = name) %>% 
+  group_by(sign_type) %>% 
+  summarise(value = n()) %>% 
+  arrange(desc(value))
+
+sign_type_plot <- ggplot(sign_type_df) +
+  aes(reorder(sign_type, value), x = sign_type, y = value) +
+  geom_col() +
+  xlab("Sign type") +
+  ylab("Count") +
+  theme(axis.title = element_text(size = 10)) +
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, ))
+
+ggsave(sign_type_plot, 
+      filename =  "signtype.png",
+       bg = "white", w = 5, h = 3)
+
+```
+
+```{r}
+library(rnaturalearth)
+
+as_factor <- function(x, reverse = FALSE) {
+  lvl <- unique(x)
+  if (reverse) {
+    lvl <- rev(lvl)
+  }
+  factor(x, levels = lvl)
+}
+
+to_long <- function(from, factor = FALSE, reverse = FALSE) {
+  x <- data.frame(
+    row = as.vector(row(from, as.factor = factor)),
+    column = as.vector(col(from, as.factor = factor)),
+    value = as.vector(from),
+    stringsAsFactors = FALSE
+  )
+  if (factor) {
+    x$row <- as_factor(x$row, reverse = reverse)
+    x$column <- as_factor(x$column, reverse = reverse)
+  }
+  return(x)
+}
+
+autoplot.DiversityIndex <- function(object, ...) {
+  ## Prepare data
+  count <- cbind.data.frame(
+    label = object@labels,
+    x = object@size,
+    y = object@.Data
+  )
+
+  ## Simulated assemblages
+  gg_sim <- NULL
+  if (length(object@simulation) != 0) {
+    # Build a long table for ggplot2
+    refined <- object@simulation
+
+    sim_stacked <- to_long(refined[, -c(1)], factor = TRUE)
+    sim <- cbind.data.frame(
+      size = refined[, 1],
+      sim_stacked,
+      Estimate = ifelse(sim_stacked[["column"]] == "mean", "mean", "conf. int.")
+    )
+    gg_sim <- ggplot2::geom_path(
+      mapping = ggplot2::aes(
+        x = .data$size,
+        y = .data$value,
+        colour = .data$Estimate,
+        group = .data$column
+      ),
+      data = sim,
+      na.rm = TRUE,
+      inherit.aes = FALSE
+    )
+  }
+
+  y_lab <- switch (
+    class(object),
+    HeterogeneityIndex = "Heterogeneity",
+    EvennessIndex = "Evenness",
+    RichnessIndex = "Richness",
+    "Diversity"
+  )
+
+  ## ggplot
+  ggplot2::ggplot(data = count) +
+    ggplot2::aes(x = .data$x, y = .data$y, label = .data$label) +
+    ggplot2::geom_point() +
+    gg_sim +
+    ggplot2::scale_x_log10(name = "Sample size") +
+    ggplot2::scale_y_continuous(
+      name = "Heterogeneity"
+    )
+}
+
+plot_diversity_fn <- function(input_df){
+
+signs <-
+  input_df %>% 
+  select(
+         site_name,
+         c(line:star)) %>% 
+  group_by(site_name) %>% 
+  summarise(across(where(is.numeric), ~ sum(.))) %>% 
+  column_to_rownames("site_name")
+
+diversity_index <- 
+  heterogeneity(signs, method = "shannon")
+
+diversity_sim <- simulate(diversity_index)
+
+autoplot.DiversityIndex(diversity_sim) +
+  geom_text_repel(aes(label = diversity_sim@labels),
+                  size = 3, max.overlaps = 15) +
+  guides(colour = "none") +
+  theme_minimal(base_size = 8) +
+  border(color = "black", size = 0.5)
+}
+
+signbase_full_clean_diversity_plot_by_period <- 
+time_unique_data_full %>% 
+  nest_by(time_period) %>% 
+  mutate(plot = list(plot_diversity_fn(data)))
+
+library(cowplot)
+plot_grid(plotlist =  signbase_full_clean_diversity_plot_by_period$plot,
+          labels = signbase_full_clean_diversity_plot_by_period$time_period,
+          label_size = 8, vjust = 2, hjust = -0.6)
+```

### 2025-05-12 endofquarterpaper.qmd
hunter-gatherer edits

@@
 
-Variation in diversity of sign types among sites may be representative of variation in social arrangement or social context.
-Despite the tendency to conceive of hunter-gatherers as monolithic, previous studies have established a potential for variation in hunter-gatherer social arrangements [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017].
-@Bloch_1977 drew upon ethnographic studies of hunter-gatherers to theorize a correlation between the amounts of social structure, instituted hierarchy, and ritual communication within a society.
-@Conkey_1985 built upon that theorized correlation to argue that artwork and other forms of material culture can serve as a form of ritual communication, and that an increase in diversity within those material cultures may suggest more elaborate or complex social contexts.
-If the signs served functions of ritual communication, then changes in sign diversity from site to site, particularly the larger grouping structure of broader versus restricted signs, may be representative of two different social arrangement types, such as, for example, a more hierarchical structure versus a less hierarchical structure.
+@
+
+Variation in diversity of sign types among sites may indicate variation in social structure.
+Despite the tendency to conceive of hunter-gatherers as monolithic, previous studies have established a potential for variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006].
+Additional research has illustrated how ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978;@Tilley_1982].
+Therefore, it is possible that greater diversity in geometric signs and other forms of material culture can serve as evidence of wider variety of formalized social roles [@Conkey_1985].
+This is especially relevant in regards to our larger grouping structure of broader versus restricted signs, as it may indicate two broad types of social structure, with the broader groups having a more formalized social structure, and the restricted groups having a less formalized social structure.
 
@@
 Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
+
+It is therefore possible that some of the sites shown in @fig-div-plot which have a more diverse range of sign types than predicted by sample size, such as Les Cottés in the Proto-Aurignacian, Solutré in the Early Aurignacaibn, and Bockstein-Törle and Vogelherd in the Evolved Aurignacian acted as aggregation sites, or served a similar purpose.
+
 It is therefore possible that sites with multiple connections between groups [@fig-neighbor-net], such as La Ferrassie in the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.
@@
 ```{r}
+#| label: fig-div-plot
+#| fig-cap: "Plots of expected versus actual diversity of signs for each site by time period.
+The blue line is the expected Heterogeneity value under the null model (simple increase in sample size means simple increase in diversity of sign types.
+The upper red line is the upper 95% confidence interval, if a site is near or above that line, it has a much higher diversity of signs than expected under the null model.The lower red line is the lower 95% confidence interval, if a site is near or below that line, it has a much lower diversity of signs than expected under the null model."
 
@@
 # Conclusion
-
 This study adds to our knowledge of social networks within the Aurignacian period, as our results document dynamic patterns of social exchange and connection.

### 2025-05-12 endofquarterpaper.qmd
stash

--- 2025-05-05 endofquarterpaper.qmd
@@
 author: Sophie Cooper
-format: html
+format: docx
 echo: false
@@
 
-Variation in diversity of sign types among sites may be representative of variation in social arrangement or social context.
-Despite the tendency to conceive of hunter-gatherers as monolithic, previous studies have established a potential for variation in hunter-gatherer social arrangements [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017].
-@Bloch_1977 drew upon ethnographic studies of hunter-gatherers to theorize a correlation between the amounts of social structure, instituted hierarchy, and ritual communication within a society.
-@Conkey_1985 built upon that theorized correlation to argue that artwork and other forms of material culture can serve as a form of ritual communication, and that an increase in diversity within those material cultures may suggest more elaborate or complex social contexts.
-Changes in sign diversity from site to site, particularly the larger grouping structure of broader versus restricted signs, may be representative of two different social arrangement types, such as, for example, a more hierarchical structure versus a less hierarchical structure.
+It is possible that variation in diversity of sign types among sites may be representative of variation in social arrangement or social context.
+Despite the tendency to conceive of hunter-gatherers as monolithic, previous studies have established a diverse range of hunter-gatherer social arrangements [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017].
+Ethnographic studies of hunter-gatherers were utilized by @Bloch_1977 to theorize a correlation between the amounts of social structure, instituted hierarchy, and ritual communication in a society.
+@Conkey_1985 built upon that theorized correlation, arguing that artwork and other forms of material culture can serve as a form of ritual communication, and that an increase in diversity may suggest more elaborate or complex social contexts, specifically in regards to the Paleolithic.
+It is therefore possible that changes in diversity from site to site, particularly the larger grouping structure of broader versus restricted signs, may be representative of two different types of social arrangement..
 
@@
   geom_text_repel(aes(label = diversity_sim@labels),
-                  size = 3, max.overlaps = 15) +
+                  size = 3) +
   guides(colour = "none") +
-  theme_minimal(base_size = 8) +
-  border(color = "black", size = 0.5)
+  theme_minimal(base_size = 8)
 }

### 2025-05-12 endofquarterpaper.qmd
resolve merge conflict

--- 2025-05-12 endofquarterpaper.qmd
@@
 
-@
-
 Variation in diversity of sign types among sites may indicate variation in social structure.
 Despite the tendency to conceive of hunter-gatherers as monolithic, previous studies have established a potential for variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006].
-Additional research has illustrated how ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978;@Tilley_1982].
-Therefore, it is possible that greater diversity in geometric signs and other forms of material culture can serve as evidence of wider variety of formalized social roles [@Conkey_1985].
-This is especially relevant in regards to our larger grouping structure of broader versus restricted signs, as it may indicate two broad types of social structure, with the broader groups having a more formalized social structure, and the restricted groups having a less formalized social structure.
+Additional research has established how variation in ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982].
+It is therefore possible that greater diversity in geometric signs serves as evidence of wider variety of formal social roles [@Conkey_1985].
+This is especially relevant in regards to our larger grouping structure of broader versus restricted signs, as it may indicate two broad types of social structure, with the broader groups having a more formal social structure, and the restricted groups having a less formal social structure.
 

### 2025-05-12 endofquarterpaper.qmd
stash

@@
 author: Sophie Cooper
-format: docx
+format: html
 echo: false

### 2025-05-12 endofquarterpaper.qmd

@@
 We aim to address the following questions: How do grouping dynamics of geometric signs change across the four phases of the Aurignacian?
-How do the groupings vary in diversity, size, connectivity, and boundaries?
+How do the groupings vary in diversity, size, connectivity, and distribution?
 Lastly, what do the patterns tell us about dynamics of social organization throughout the Aurignacian?
@@
 
-Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and boundaries.
+Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and distribution.
 The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.
@@
 The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation.
-
-Variation in diversity of sign types among sites may indicate variation in social structure.
-Despite the tendency to conceive of hunter-gatherers as monolithic, previous studies have established a potential for variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006].
-Additional research has established how variation in ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982].
-It is therefore possible that greater diversity in geometric signs serves as evidence of wider variety of formal social roles [@Conkey_1985].
-This is especially relevant in regards to our larger grouping structure of broader versus restricted signs, as it may indicate two broad types of social structure, with the broader groups having a more formal social structure, and the restricted groups having a less formal social structure.
 
@@
 
-The geographic boundaries of the groups change from phase to phase (@fig-group-map).
+The geographic distribution of the groups change from phase to phase (@fig-group-map).
 In the Transitional phase, sites are widely spread out across Western Europe.
@@
 One factor that may influence the geographic structure of sites is the possibility that some sites functioned as aggregation sites.
-The aggregation/dispersion pattern is first defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for various subsistence or social reasons.
+The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons.
 Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg].
@@
 It is therefore possible that sites with multiple connections between groups [@fig-neighbor-net], such as La Ferrassie in the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.
+
+Variation in diversity of sign types may indicate variation in social structure.
+Despite the tendency to conceive of hunter-gatherers as homogenous, previous studies have established a potential for variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006].
+Additional research has established how variation in ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982].
+It is therefore possible that greater diversity in geometric signs indicates a wider variety of formal social roles [@Conkey_1985].
+This is especially relevant in regards to our larger grouping structure of broader versus restricted signs, as it may indicate two broad types of social structure, with the broader groups having a more formal social structure, and the restricted groups having a less formal social structure.
 

### 2025-05-12 endofquarterpaper.qmd
merge

@@
 Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
-
-It is therefore possible that some of the sites shown in @fig-div-plot which have a more diverse range of sign types than predicted by sample size, such as Les Cottés in the Proto-Aurignacian, Solutré in the Early Aurignacaibn, and Bockstein-Törle and Vogelherd in the Evolved Aurignacian acted as aggregation sites, or served a similar purpose.
-
 It is therefore possible that sites with multiple connections between groups [@fig-neighbor-net], such as La Ferrassie in the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.
+
+It is therefore possible that some of the sites shown in @fig-div-plot which have a more diverse range of sign types than predicted by sample size, such as Les Cottés in the Proto-Aurignacian, Solutré in the Early Aurignacian, and Bockstein-Törle and Vogelherd in the Evolved Aurignacian acted as aggregation sites, or served a similar purpose.
 

### 2025-05-14 endofquarterpaper.qmd
modularity codee

@@
 
-If the distribution of sign types can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases.
+If the distribution of sign types can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases [.
 This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019].
@@
 Each node is a site, and each edge is a measure of sign type distribution similarity, as computed by the Jaccard similarity algorothm.
+To avoid cluttering the graph, all edges below the value of 0.2 were removed.
 Results were plotted with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths, and network analysis was performed using the vegan [@Vegan2001], statnet [@statnet_package] and igraph [@igraph_package] packages.
@@
 The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons.
-Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg].
+Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a diverse artifact assemblage [@conkley_agg].
 Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
-It is therefore possible that sites with multiple connections between groups [@fig-neighbor-net], such as La Ferrassie in the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.
-
-It is therefore possible that some of the sites shown in @fig-div-plot which have a more diverse range of sign types than predicted by sample size, such as Les Cottés in the Proto-Aurignacian, Solutré in the Early Aurignacian, and Bockstein-Törle and Vogelherd in the Evolved Aurignacian acted as aggregation sites, or served a similar purpose.
-
-Variation in diversity of sign types may indicate variation in social structure.
-Despite the tendency to conceive of hunter-gatherers as homogenous, previous studies have established a potential for variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006].
-Additional research has established how variation in ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982].
-It is therefore possible that greater diversity in geometric signs indicates a wider variety of formal social roles [@Conkey_1985].
-This is especially relevant in regards to our larger grouping structure of broader versus restricted signs, as it may indicate two broad types of social structure, with the broader groups having a more formal social structure, and the restricted groups having a less formal social structure.
+
+One potential identifier of aggregation sites comes from @fig-div-plot, which uses the tabula package [@Frerebeau_2025] to generate the Shannon-Weaver diversity index of sign types per site, and plot the expected diversity index for a given sample size against the actual diversity index per site [@Kintigh_1984].
+The plot identifies multiple sites that diverge from the diversity index predicted by sample size, including those with a higher index than expected, such as Hohle Fels in the Transitional phase, Spy, La Ferrassie, and Les Cottés in the Proto-Aurignacian, Solutré, Hohle Fels, and Grottes de Fonds-de-Forêt in the Early Aurignacian, and Vogelherd, Bockstein-Törle, and Vindija Cave in the Evolved Aurignacian.
+It is possible that those sites with a greater than expected diversity index acted as aggregation sites or served a similar purpose.
+@fig-neighbor-net also shows La Ferrassie in the Proto-Aurignacian as having multiple connections to sites in both groups, perhaps suggesting that it served as an inter-group aggregation site.
 
@@
     ggplot2::scale_x_log10(name = "Sample size") +
-    ggplot2::scale_y_continuous(
-      name = sprintf("%s (%s)", y_lab, object@method)
-    )
+    ggplot2::scale_y_continuous(name = "Diversity")
+
 }
@@
 
+@fig-div-plot shows a great deal of general divergence from the expected diversity index for the sample size.
+Another potential explanation for the observed variation in diversity is that it may indicate variation in social structure.
+Despite the tendency to conceive of hunter-gatherers as homogenous, previous studies have established a potential for variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006].
+Additional research has established how variation in ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982].
+It is therefore possible that greater diversity in geometric signs indicates a wider variety of formal social roles [@Conkey_1985].
+This is especially relevant in regards to our larger grouping structure of broader versus restricted signs, as it may indicate two broad types of social structure, with the broader groups having a more formal social structure, and the restricted groups having a less formal social structure.
+
 ## Social Organization Dynamics Throughout the Aurignacian
@@
 # Conclusion
+
 This study adds to our knowledge of social networks within the Aurignacian period, as our results document dynamic patterns of social exchange and connection.
@@
 While @Von_Petzinger_2009 argues in favor of Aurignacian geometric signs as symbolic expression, potential ambiguity still remains over whether function is consistent across all sign type occurrences, particularly in regards to the simpler sign types composed from fewer elements.
-We included the more ambiguous sign types because similar sign occurrence may still serve as evidence of cultural connectivity regardless of function, and additionally, greater quantities of included sign types serves to establish stronger connections [@Gravel-Miguel_Coward_2023].
+We included the more ambiguous sign types because similar sign occurrence may still serve as evidence of cultural connectivity regardless of function, even if it is random copying [@Herskind2023], and additionally, greater quantities of included sign types serves to establish stronger connections [@Gravel-Miguel_Coward_2023].
 This ambiguity may be addressed in the future through the usage of the data recently released in SignBase 2.0 covering sites in Germany, which contains more detailed information on the specific order and patterns in which signs occurred on individual objects.

### 2025-05-15 endofquarterpaper.qmd
stash

--- 2025-05-12 endofquarterpaper.qmd
@@
 Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
+
+It is therefore possible that some of the sites shown in @fig-div-plot which have a more diverse range of sign types than predicted by sample size, such as Les Cottés in the Proto-Aurignacian, Solutré in the Early Aurignacaibn, and Bockstein-Törle and Vogelherd in the Evolved Aurignacian acted as aggregation sites, or served a similar purpose.
+
 It is therefore possible that sites with multiple connections between groups [@fig-neighbor-net], such as La Ferrassie in the Proto-Aurignacian, Grotte de la Verpillière I in the Early Aurignacian, and Gargas in the Evolved Aurignacian acted as aggregation sites or served a similar purpose.
-
-It is therefore possible that some of the sites shown in @fig-div-plot which have a more diverse range of sign types than predicted by sample size, such as Les Cottés in the Proto-Aurignacian, Solutré in the Early Aurignacian, and Bockstein-Törle and Vogelherd in the Evolved Aurignacian acted as aggregation sites, or served a similar purpose.
 

### 2025-05-16 paper/paper.qmd
simplify project structure

--- 2025-05-14 endofquarterpaper.qmd
@@
 
-aur_table <- read_excel("data/aurignacian-periods.xlsx")
+aur_table <- read_excel(here::here("data/aurignacian-periods.xlsx"))
 
@@
 library(rcarbon)
-signbase_full <- read_csv("data/signBase_Version1.0.csv")
+signbase_full <- read_csv(here::here("data/signBase_Version1.0.csv"))
 nsites <- signbase_full %>% 
@@
 ggsave(maps_bar_plots, 
-      filename =  "map-and_hist-plot.png",
+      filename =  here::here("figures/map-and_hist-plot.png"),
        h = 10, w = 10,
@@
 
-knitr::include_graphics("map-and_hist-plot.png")
+knitr::include_graphics(here::here("figures/map-and_hist-plot.png"))
 

### 2025-05-16 docs/undergraduate-research-symposium-images.qmd
simplify project structure

--- 2025-05-06 Extra/undergraduate-research-symposium-images.qmd
@@
      theme(
-       axis.text = element_text(size = rel(0.3)),
+       axis.text = element_text(size = rel(0.5)),
        axis.ticks = element_blank(),
@@
                           alpha=weight), 
-                      edge_colour = "black", 
+                      edge_colour = "grey50", 
                       show.legend= F,
@@
        scale_edge_width(range=c(0.5,1.5)) +
-       scale_edge_colour_gradient(low = "#CCCCCC",
-                                  high = "#000000") +
-       geom_node_point(alpha=1, size=5) +
+       scale_edge_colour_gradient(low = "grey80",
+                                  high = "grey50") +
+       geom_node_point(alpha=1, size=5, colour = "grey60") +
        geom_node_text(aes(label=row.names(artifact_data)), 
-                      size=3, 
+                      size=2, 
                       family= "Arial", 
@@
 ```{r}
-ser_net_trans <- plot_grid(seriation_trans, network_trans, align = "hv", labels = "AUTO", rel_widths = c(1.3, 1.4))
-ser_net_proto <- plot_grid(seriation_proto, network_proto, align = "hv", rel_widths = c(1.4, 1.4))
-ser_net_early <- plot_grid(seriation_early, network_early, align = "hv", rel_widths = c(1.65, 1.8))
-ser_net_evo <- plot_grid(seriation_evolved, network_evolved, align = "hv", rel_widths = c(1.4, 1.4))
+# BM: I think it might be easier to align if there are two separate columns of plots
+# top, right, bottom, left
 
-final <- plot_grid(ser_net_trans, ser_net_proto, ser_net_early, ser_net_evo, align = "hv", ncol = 1, labels = c("Transitional", "Proto", "Early", "Evolved"))
+ser_plot <- plot_grid(seriation_trans + theme(plot.margin =   unit(c( 0, 0, -10, 0), "cm")),
+                      seriation_proto + theme(plot.margin =   unit(c(-10, 0, -10, 0), "cm")), 
+                      seriation_early + theme(plot.margin =   unit(c(-10, 0, -10, 0), "cm")), 
+                      seriation_evolved + theme(plot.margin = unit(c(-10, 0,  0, 0), "cm")),
+                      align = "hv", 
+                      ncol = 1, 
+                      labels = c( 
+                                 "Transitional",   
+                                 "Proto",  
+                                 "Early",  
+                                 "Evolved"), 
+                      label_size = 9, 
+                      label_x = rep(0.35, 4), 
+                      hjust = 1, 
+                      label_y = rep(0.7, 4))
 
-ggsave(final, 
-      filename =  "poster.png",
-       bg = "white", h = 9, w = 4.5)
-```
+ggsave(ser_plot, 
+      filename =  here::here("Extra/poster4-seriation.png"), 
+      bg = "white", 
+      w = 4.5, 
+      h = 8)
 
-```{r}
-ser_net_trans <- plot_grid(seriation_trans, network_trans, align = "hv", labels = "AUTO", ncol = 1)
-ser_net_proto <- plot_grid(seriation_proto, network_proto, align = "hv", ncol = 1)
-ser_net_early <- plot_grid(seriation_early, network_early, align = "hv", ncol = 1)
-ser_net_evo <- plot_grid(seriation_evolved, network_evolved, align = "hv", ncol = 1)
+net_plot <- plot_grid(network_trans, NULL,
+                      network_proto, NULL,
+                      network_early, NULL,
+                      network_evolved, NULL,
+                      rel_heights = c(1, 0.1, 1, 0.1, 1, 0.1, 1, 0.1),
+                      align = "hv", 
+                      ncol = 1)
 
-final <- plot_grid(ser_net_trans, ser_net_proto, ser_net_early, ser_net_evo, align = "hv", ncol = 4, labels = c("Transitional", "Proto", "Early", "Evolved"))
+ggsave(net_plot, 
+      filename =  here::here("Extra/poster4-networks.png"), 
+      bg = "white", 
+      w = 3, 
+      h = 8)
 
-ggsave(final, 
-      filename =  "poster2.png",
-       bg = "white", w = 12)
-
-```
-
-```{r}
-ser_plot <- plot_grid(seriation_trans, seriation_proto, seriation_early, seriation_early, align = "hv", ncol = 1, labels = c("Transitional", "Proto", "Early", "Evolved"), label_size = 9)
-net_plot <- plot_grid(network_trans, network_proto, network_early, network_evolved, align = "hv", ncol = 1)
-
-  final <- plot_grid(ser_plot, net_plot, align = "hv", ncol = 2, rel_widths = c(1.3, 1), labels = "AUTO", label_x = c(0, -0.07))
-
-ggsave(final, 
-      filename =  "poster3.png", bg = "white", w = 5, h = 7)
 ```
@@
 sign_type_plot <- ggplot(sign_type_df) +
-  aes(reorder(sign_type, value), x = sign_type, y = value) +
+  aes(x = reorder(sign_type, -value), y = value) +
   geom_col() +
-  xlab("Sign type") +
+  xlab("") +
   ylab("Count") +
+  theme_minimal(base_size = 12) +
   theme(axis.title = element_text(size = 10)) +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, ))
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
 
 ggsave(sign_type_plot, 
-      filename =  "signtype.png",
-       bg = "white", w = 5, h = 3)
+      filename =  here::here("Extra/signtype.png"),
+       bg = "white", w = 5, h = 7)
 

### 2025-05-19 Extra/modularity.R
modularity

@@ -12,7 +12,7 @@ modularity_function<- function(modul_matrix, mem){
   nsteps <- 10000
   modul_table <- tibble("modularity" = c())
   for (i in 1:nsteps){
-    mem_random <- as_membership(c(sample(1:2, N, replace = T)))
+    mem_random <- as_membership(c(sample(1:N, N, replace = T)))
     modularity_permute <- modularity(modul_matrix, membership = mem_random)
     modul_row <- tibble("modularity" = modularity_permute)
     modul_table <- modul_table %>% rbind(modul_row)}
@@ -24,31 +24,6 @@ modularity_function<- function(modul_matrix, mem){
 
 
 modularity_function(modul_matrix, mem)
-
-
-
-
-modularity_function<- function(modul_matrix, mem){
-  gt <- modularity(modul_matrix, membership = mem)
-  N <- nrow(jac)
-  nsteps <- 10000
-  modul_table <- tibble("modularity" = c())
-  for (i in 1:nsteps){
-    graph_permute<- early_artifact_data[sample(1:N),]
-    jac2 <- vegdist(graph_permute %>% mutate_all(~ as.numeric(. > 0)), "jaccard", binary = TRUE)
-    dm2 <- as.matrix(jac2)
-    modul_matrix2 <-
-      graph_from_adjacency_matrix(1 - dm2, mode = "undirected")
-    modularity_permute <- modularity(modul_matrix2, membership = mem)
-    modul_row <- tibble("modularity" = modularity_permute)
-    modul_table <- modul_table %>% rbind(modul_row)}
-  num <- nrow(modul_table %>% 
-                filter(modularity > gt))
-  p_score <- num/10000
-  return(p_score)
-}
-
-
 
 
 

### 2025-05-19 paper/paper.qmd

@@
 The results of the network analysis supports the restricted-broad groupings identified in @fig-seriation, and provides additional insights into the evolution of connectivity between groups.
-In the Transitional phase, the network analysis reveals two distinct and disconnected groups.
-The restricted group consists of Geissenklösterle, Fumane, and Pod Hradem, and the broad group consisting of Hohle Fels, El Castillo, and Labeko Koba, with no connections between the groups.
+In the Transitional phase, the network analysis reveals one distinct and one disconnected group.
+The restricted group consists of Geissenklösterle, Fumane, and Pod Hradem, and the broad group consists of Hohle Fels, El Castillo, and Labeko Koba, with no connections between the groups.
 The Proto-Aurignacian phase also shows two clearly defined groups, the broad group consisting of Geissenklösterle, Spy, Vogelherd, and Grotte du Renne, and the restricted group consisting of Mladeč, Hohlenstein-Stadel, Gatzarria, and Abri Pataud.

### 2025-05-19 Extra/modularity.R
merge

@@ -1,29 +1,0 @@
-
-jac <- vegdist(trans_artifact_data %>% mutate_all(~ as.numeric(. > 0)), "jaccard", binary = TRUE)
-dm <- as.matrix(jac)
-modul_matrix <-
-  graph_from_adjacency_matrix(1 - dm, mode = "undirected")
-mem <- as_membership(parse_number(trans_unique_data$group))
-gt <- modularity(modul_matrix, membership = mem)
-
-modularity_function<- function(modul_matrix, mem){
-  gt <- modularity(modul_matrix, membership = mem)
-  N <- nrow(jac)
-  nsteps <- 10000
-  modul_table <- tibble("modularity" = c())
-  for (i in 1:nsteps){
-    mem_random <- as_membership(c(sample(1:N, N, replace = T)))
-    modularity_permute <- modularity(modul_matrix, membership = mem_random)
-    modul_row <- tibble("modularity" = modularity_permute)
-    modul_table <- modul_table %>% rbind(modul_row)}
-  num <- nrow(modul_table %>% 
-                filter(modularity > gt))
-  p_score <- num/10000
-  return(p_score)
-}
-
-
-modularity_function(modul_matrix, mem)
-
-
-

### 2025-05-27 endofquarterpaper.qmd
edit

--- 2025-05-14 endofquarterpaper.qmd
@@
 
-One potential identifier of aggregation sites comes from @fig-div-plot, which uses the tabula package [@Frerebeau_2025] to generate the Shannon-Weaver diversity index of sign types per site, and plot the expected diversity index for a given sample size against the actual diversity index per site [@Kintigh_1984].
+A potential identifier of aggregation sites comes from @fig-div-plot, which uses the tabula package [@Frerebeau_2025] to generate the Shannon-Weaver diversity index of sign types per site, and plot the expected diversity index for a given sample size against the actual diversity index per site [@Kintigh_1984].
 The plot identifies multiple sites that diverge from the diversity index predicted by sample size, including those with a higher index than expected, such as Hohle Fels in the Transitional phase, Spy, La Ferrassie, and Les Cottés in the Proto-Aurignacian, Solutré, Hohle Fels, and Grottes de Fonds-de-Forêt in the Early Aurignacian, and Vogelherd, Bockstein-Törle, and Vindija Cave in the Evolved Aurignacian.
 It is possible that those sites with a greater than expected diversity index acted as aggregation sites or served a similar purpose.
-@fig-neighbor-net also shows La Ferrassie in the Proto-Aurignacian as having multiple connections to sites in both groups, perhaps suggesting that it served as an inter-group aggregation site.
+La Ferrassie in the Proto-Aurignacian is also shown by @fig-neighbor-net to have multiple connections to sites in both groups, in addition to the diversity index shown in @fig-div-plot, perhaps suggesting that it served as a specifically inter-group aggregation site.
 
@@
 
-@fig-div-plot shows a great deal of general divergence from the expected diversity index for the sample size.
-Another potential explanation for the observed variation in diversity is that it may indicate variation in social structure.
-Despite the tendency to conceive of hunter-gatherers as homogenous, previous studies have established a potential for variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006].
+Another potential explanation for the observed general variation in diversity shown in 
+@fig-div-plot is that it may indicate variation in social structure.
+Despite the tendency to conceive of hunter-gatherers as homogeneous, previous studies have established a potential for variation in hunter-gatherer social structures [@Singh_Glowacki_2022; @Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006].
 Additional research has established how variation in ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982].

### 2025-05-27 paper/paper.qmd
stash

--- 2025-05-19 paper/paper.qmd
@@
 ggsave(maps_bar_plots, 
-      filename =  here::here("figures/map-and_hist-plot.png"),
+      filename =  "map-and_hist-plot.png",
        h = 10, w = 10,
@@
 
-knitr::include_graphics(here::here("figures/map-and_hist-plot.png"))
+knitr::include_graphics("map-and_hist-plot.png")
 

### 2025-05-27 endofquarterpaper_files/libs/quarto-html/quarto.js
tidy

@@ -1,939 +1,0 @@
-const sectionChanged = new CustomEvent("quarto-sectionChanged", {
-  detail: {},
-  bubbles: true,
-  cancelable: false,
-  composed: false,
-});
-
-const layoutMarginEls = () => {
-  // Find any conflicting margin elements and add margins to the
-  // top to prevent overlap
-  const marginChildren = window.document.querySelectorAll(
-    ".column-margin.column-container > *, .margin-caption, .aside"
-  );
-
-  let lastBottom = 0;
-  for (const marginChild of marginChildren) {
-    if (marginChild.offsetParent !== null) {
-      // clear the top margin so we recompute it
-      marginChild.style.marginTop = null;
-      const top = marginChild.getBoundingClientRect().top + window.scrollY;
-      if (top < lastBottom) {
-        const marginChildStyle = window.getComputedStyle(marginChild);
-        const marginBottom = parseFloat(marginChildStyle["marginBottom"]);
-        const margin = lastBottom - top + marginBottom;
-        marginChild.style.marginTop = `${margin}px`;
-      }
-      const styles = window.getComputedStyle(marginChild);
-      const marginTop = parseFloat(styles["marginTop"]);
-      lastBottom = top + marginChild.getBoundingClientRect().height + marginTop;
-    }
-  }
-};
-
-window.document.addEventListener("DOMContentLoaded", function (_event) {
-  // Recompute the position of margin elements anytime the body size changes
-  if (window.ResizeObserver) {
-    const resizeObserver = new window.ResizeObserver(
-      throttle(() => {
-        layoutMarginEls();
-        if (
-          window.document.body.getBoundingClientRect().width < 990 &&
-          isReaderMode()
-        ) {
-          quartoToggleReader();
-        }
-      }, 50)
-    );
-    resizeObserver.observe(window.document.body);
-  }
-
-  const tocEl = window.document.querySelector('nav.toc-active[role="doc-toc"]');
-  const sidebarEl = window.document.getElementById("quarto-sidebar");
-  const leftTocEl = window.document.getElementById("quarto-sidebar-toc-left");
-  const marginSidebarEl = window.document.getElementById(
-    "quarto-margin-sidebar"
-  );
-  // function to determine whether the element has a previous sibling that is active
-  const prevSiblingIsActiveLink = (el) => {
-    const sibling = el.previousElementSibling;
-    if (sibling && sibling.tagName === "A") {
-      return sibling.classList.contains("active");
-    } else {
-      return false;
-    }
-  };
-
[…]

### 2025-05-27 paper/paper.qmd
render to docx

@@
 ggsave(maps_bar_plots, 
-      filename =  "map-and_hist-plot.png",
+      filename =  here::here("figures/map-and-hist-plot.png"),
        h = 10, w = 10,
@@
 
-knitr::include_graphics("map-and_hist-plot.png")
+knitr::include_graphics(here::here("figures/map-and-hist-plot.png"))
 

### 2025-05-27 paper/paper.qmd
stash

--- 2025-05-27 paper/paper.qmd
@@
 
-aur_table <- read_excel(here::here("data/aurignacian-periods.xlsx"))
+aur_table <- read_excel("data/aurignacian-periods.xlsx")
 
@@
 library(rcarbon)
-signbase_full <- read_csv(here::here("data/signBase_Version1.0.csv"))
+signbase_full <- read_csv("data/signBase_Version1.0.csv")
 nsites <- signbase_full %>% 

### 2025-05-27 docs/modularity.R
investigating modularity score significance

@@ -1,15 +1,18 @@
 
-jac <- vegdist(trans_artifact_data %>% mutate_all(~ as.numeric(. > 0)), "jaccard", binary = TRUE)
+jac <- vegdist(trans_artifact_data %>% 
+                 mutate_all(~ as.numeric(. > 0)), "jaccard", binary = TRUE)
 dm <- as.matrix(jac)
 modul_matrix <-
   graph_from_adjacency_matrix(1 - dm, mode = "undirected")
 mem <- as_membership(parse_number(trans_unique_data$group))
 gt <- modularity(modul_matrix, membership = mem)
 
+
+n <- 10000
 modularity_function<- function(modul_matrix, mem){
   gt <- modularity(modul_matrix, membership = mem)
-  N <- nrow(jac)
-  nsteps <- 10000
+  N <-  nrow(jac)
+  nsteps <- n
   modul_table <- tibble("modularity" = c())
   for (i in 1:nsteps){
     mem_random <- as_membership(c(sample(1:N, N, replace = T)))
@@ -17,13 +20,14 @@ modularity_function<- function(modul_matrix, mem){
     modul_row <- tibble("modularity" = modularity_permute)
     modul_table <- modul_table %>% rbind(modul_row)}
   num <- nrow(modul_table %>% 
-                filter(modularity > gt))
-  p_score <- num/10000
+                filter(modularity >= gt))
+  p_score <- num/n
   return(p_score)
 }
 
+p_value <- modularity_function(modul_matrix, mem)
 
-modularity_function(modul_matrix, mem)
+paste0("p = ", p_value)
 
 
 

### 2025-05-27 paper/paper.qmd
fix paths

--- 2025-05-27 paper/paper.qmd
@@
 ---
-
 # Introduction

### 2025-05-27 docs/modularity.R
stash

@@ -6,19 +6,20 @@ modul_matrix <-
 mem <- as_membership(parse_number(trans_unique_data$group))
 gt <- modularity(modul_matrix, membership = mem)
 
+n <- 10000
 modularity_function<- function(modul_matrix, mem){
   gt <- modularity(modul_matrix, membership = mem)
   N <- nrow(jac)
-  nsteps <- 10000
+  nsteps <- n
   modul_table <- tibble("modularity" = c())
   for (i in 1:nsteps){
-    mem_random <- as_membership(c(sample(1:N, N, replace = T)))
+    mem_random <- as_membership(c(sample(1:2, N, replace = T)))
     modularity_permute <- modularity(modul_matrix, membership = mem_random)
     modul_row <- tibble("modularity" = modularity_permute)
     modul_table <- modul_table %>% rbind(modul_row)}
   num <- nrow(modul_table %>% 
-                filter(modularity > gt))
-  p_score <- num/10000
+                filter(modularity >= gt))
+  p_score <- num/n
   return(p_score)
 }
 

### 2025-05-27 paper/paper.qmd
stash

--- 2025-05-27 paper/paper.qmd
@@
 
-If the distribution of sign types can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases [.
+If the distribution of sign types can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases [@Hirsk  .
 This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019].
@@
 The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons.
-Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a diverse artifact assemblage [@conkley_agg].
+Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg].
 Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].

### 2025-06-01 paper/paper.qmd
edit

--- 2025-05-27 paper/paper.qmd
@@
 The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020], coupled with a biological transition from Neanderthals to early modern humans.
-Archaeologists have frequently relied upon variations in artwork, the large rise of which is a distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d].
+Archaeologists have frequently relied upon variations in artwork, the large rise of which is a distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d; @Fuentes_Lucas_Robert_2019].
 Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca.
@@
 We answer these questions by first identifying clusters within each phase using seriation and network analyses.
-We then test the strength of the clusters using modularity and perMANOVA tests.
+We then test the strength of the clusters using perMANOVA tests.
 All of our programming in this study was conducted in RStudio.
@@
   jac <- vegdist(artifact_data %>% mutate_all(~ as.numeric(. > 0)), "jaccard", binary = TRUE)
-  dm <- as.matrix(jac)
-  modul_matrix <-
-    graph_from_adjacency_matrix(1 - dm, mode = "undirected")
-  mem <- as_membership(parse_number(group_data$group))
-  as_group <- modularity(modul_matrix, membership = mem)
-  as_group <- round(as_group, 3)
-
   perman <- adonis2(jac~as.factor(group_data$group), 
@@
 
-  final_table <- data_frame("Modularity" = as_group, 
-                            "PerMANOVA R2" = perman_r,
+  final_table <- data_frame("PerMANOVA R2" = perman_r,
                             "PerMANOVA F" = perman_f,
@@
 
-If the distribution of sign types can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases [.
+If the distribution of sign types can be explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases [@Herskind2023; @Rigaud_dErrico_Vanhaeren_2015].
 This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019].
@@
 
-## Modularity and PerMANOVA for Cultural Connection Strength
-
-To evaluate the strength of the connections of the groups found in the seriation and network analyses, we used modularity and perMANOVA tests.
-The modularity test is a community structure network detection method, and quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006; @Newman_Girvan_2004].
-The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings.
-PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017].
+##  PerMANOVA for Cultural Connection Strength
+
+PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017; @Shennan_Crema_Kerig_2015].
 The results are expressed by an R^2^ statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, a pseudo-F statistic, and a p-value.
@@
 The p-value quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
-The modularity test and perMANOVA were performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.
+The perMANOVA was performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.
 
@@
 #| label: tbl-strength-table
-#| tbl-cap: "Table of the results per phase for the modularity test and perMANOVA"
+#| tbl-cap: "Table of the results per phase for the perMANOVA"
 #| tbl-cap-location: bottom
 
-## Modularity test
-
 strength_table <- full_table %>% 
-  dplyr::select(`Aurignacian phase`, Modularity, `PerMANOVA R2`, `PerMANOVA F`, `PerMANOVA p`)
+  dplyr::select(`Aurignacian phase`, `PerMANOVA R2`, `PerMANOVA F`, `PerMANOVA p`)
 
 kable(strength_table)
-
-proto_modularity <- proto_strength %>% 
-  dplyr::select(Modularity)
-evolved_modularity <- evolved_strength %>% 
-  dplyr::select(Modularity)
 
@@
 ```
-
-The modularity score quantifies the strength of the groupings, showing that the variations in grouping strength between phases are minimal.
-The Proto-Aurignacian exhibits the strongest groupings, with a score of `r proto_modularity`, and the Evolved Aurignacian has the weakest groupings, with a score of `r evolved_modularity`.
-There is a slight increase in modularity between the Transitional period and the Proto-Aurignacian, before an even slighter decrease into the Early Aurignacian, followed by a larger increase with the Evolved Aurignacian.
 
@@
 @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
-The modularity tests, as shown in @tbl-strength-table, show that the groupings are most distinct in the Transitional period and least distinct in the Early Aurignacian.
 The perMANOVA tests show that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.
-The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation.
+
+One interpretation of the inter-connectivity seen in @fig-neighbor-net is as a representation of non-utilitarian mobility.
+Non-utilitarian mobility serves to establish regional social networks which facilitate the spread of information and the survival of hunter-gatherers in uncertain environments [@Whallon_2006; @Fitzhugh_Phillips_Gjesfeld_2011; @Lovis_Donahue_2011; @Hitchcock_Ebert_2011; @Romano_Lozano_Fern2002].
+The exchange and production of decorative items is associated with non-utilitarian mobility and has been recognized as a sign of social contact in other regions and eras [@Whallon_2006; @Kelly_2024; @Kinahan_2013; @Newlander_2012; @Iizuka_Ferguson_Izuho_2022].
+Although it indicates a kind of cultural contact, whether this contact represents short-term information exchange or longer-term residential movement, such as a form of aggregation (see below) is still unclear, as existing data has shown a range of variation in mobility among hunter-gatherer groups [@Binford_1990; @Kelly_1983; @Johnson_2014; @Padilla-Iglesias_Bischoff_2024; @Singh_Glowacki_2022].
 
@@
 
+The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation.
 The geographic distribution of the groups change from phase to phase (@fig-group-map).
@@
 The plot identifies multiple sites that diverge from the diversity index predicted by sample size, including those with a higher index than expected, such as Hohle Fels in the Transitional phase, Spy, La Ferrassie, and Les Cottés in the Proto-Aurignacian, Solutré, Hohle Fels, and Grottes de Fonds-de-Forêt in the Early Aurignacian, and Vogelherd, Bockstein-Törle, and Vindija Cave in the Evolved Aurignacian.
-It is possible that those sites with a greater than expected diversity index acted as aggregation sites or served a similar purpose.
-La Ferrassie in the Proto-Aurignacian is also shown by @fig-neighbor-net to have multiple connections to sites in both groups, in addition to the diversity index shown in @fig-div-plot, perhaps suggesting that it served as a specifically inter-group aggregation site.
+It is possible that those sites with a greater than expected diversity index acted as aggregation sites for their groups or served a similar purpose.
+La Ferrassie in the Proto-Aurignacian is also shown by @fig-neighbor-net to have multiple connections to sites in both groups, in addition to the diversity index shown in @fig-div-plot, perhaps suggesting that it functioned as a specifically inter-group aggregation site.
 
@@
 
-@fig-div-plot shows a great deal of general divergence from the expected diversity index for the sample size.
+@fig-div-plot shows divergence from the expected diversity index for the sample size.
 Another potential explanation for the observed variation in diversity is that it may indicate variation in social structure.
-Despite the tendency to conceive of hunter-gatherers as homogenous, previous studies have established a potential for variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006].
-Additional research has established how variation in ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982].
-It is therefore possible that greater diversity in geometric signs indicates a wider variety of formal social roles [@Conkey_1985].
+Despite the tendency to conceive of hunter-gatherers as homogenous, previous studies have established variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006; @Johnson_2014; @Singh_Glowacki_2022; @Schwendler_2012].
+Additional research has established how variation in ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982; @Hodder1979; @Hodder1977; @Schwendler_2012].
+It is therefore possible that greater diversity in geometric signs indicates a wider variety of formal social roles [@Conkey_1985; @Schwendler_2012].
 This is especially relevant in regards to our larger grouping structure of broader versus restricted signs, as it may indicate two broad types of social structure, with the broader groups having a more formal social structure, and the restricted groups having a less formal social structure.
@@
 This study adds to our knowledge of social networks within the Aurignacian period, as our results document dynamic patterns of social exchange and connection.
-Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996].
+Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996; @Hill_Wood_Baggio_Hurtado_Boyd_2014; @Derex_Boyd_2016; @Romano_Lozano_Fern2022].
 @Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our results show evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted.

### 2025-06-02 paper/paper.qmd
edits

@@
 To investigate connectivity among clusters of sites in the seriation solution, we used network analysis to represent and quantify the relationship between sites [@Mills_2017].
-Each node is a site, and each edge is a measure of sign type distribution similarity, as computed by the Jaccard similarity algorothm.
+Each node is a site, and each edge is a measure of sign type distribution similarity, as computed by the Jaccard similarity algorithm.
 To avoid cluttering the graph, all edges below the value of 0.2 were removed.
@@
 @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
-The perMANOVA tests show that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.
+The perMANOVA tests show that the changes between the the Proto, Early, and Evolved Aurignacian phases are statistically significant, that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.
+The
 
 One interpretation of the inter-connectivity seen in @fig-neighbor-net is as a representation of non-utilitarian mobility.
-Non-utilitarian mobility serves to establish regional social networks which facilitate the spread of information and the survival of hunter-gatherers in uncertain environments [@Whallon_2006; @Fitzhugh_Phillips_Gjesfeld_2011; @Lovis_Donahue_2011; @Hitchcock_Ebert_2011; @Romano_Lozano_Fern2002].
+Non-utilitarian mobility establishes regional social networks which facilitate the spread of information and the survival of hunter-gatherers in uncertain environments [@Whallon_2006; @Fitzhugh_Phillips_Gjesfeld_2011; @Lovis_Donahue_2011; @Hitchcock_Ebert_2011; @Romano_Lozano_Fern2002].
 The exchange and production of decorative items is associated with non-utilitarian mobility and has been recognized as a sign of social contact in other regions and eras [@Whallon_2006; @Kelly_2024; @Kinahan_2013; @Newlander_2012; @Iizuka_Ferguson_Izuho_2022].
 Although it indicates a kind of cultural contact, whether this contact represents short-term information exchange or longer-term residential movement, such as a form of aggregation (see below) is still unclear, as existing data has shown a range of variation in mobility among hunter-gatherer groups [@Binford_1990; @Kelly_1983; @Johnson_2014; @Padilla-Iglesias_Bischoff_2024; @Singh_Glowacki_2022].
+
+The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons.
+Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg].
+Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
+The possibility that some sites functioned as aggregation sites may have influenced the geographic structure of sites (@fig-group-map), which change from phase to phase.
 
@@
 The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation.
-The geographic distribution of the groups change from phase to phase (@fig-group-map).
 In the Transitional phase, sites are widely spread out across Western Europe.
@@
 In the Evolved Aurignacian, we see a wider distribution and more distance between sites, which may be associated with the reduced connectivity seen in that phase.
-
-One factor that may influence the geographic structure of sites is the possibility that some sites functioned as aggregation sites.
-The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons.
-Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg].
-Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
 

### 2025-06-02 paper/paper.qmd

@@
 The exchange and production of decorative items is associated with non-utilitarian mobility and has been recognized as a sign of social contact in other regions and eras [@Whallon_2006; @Kelly_2024; @Kinahan_2013; @Newlander_2012; @Iizuka_Ferguson_Izuho_2022].
-Although it indicates a kind of cultural contact, whether this contact represents short-term information exchange or longer-term residential movement, such as a form of aggregation (see below) is still unclear, as existing data has shown a range of variation in mobility among hunter-gatherer groups [@Binford_1990; @Kelly_1983; @Johnson_2014; @Padilla-Iglesias_Bischoff_2024; @Singh_Glowacki_2022].
-
-The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons.
+Existing data has shown a range of variation in mobility among hunter-gatherer groups [@Binford_1990; @Kelly_1983; @Johnson_2014; @Padilla-Iglesias_Bischoff_2024; @Singh_Glowacki_2022]; some of the inter-connections in @fig-neighbor-net may represent shorter-term information exchange, while others may indicate longer-term residential movement.
+
+One form of longer-term residential movement is the aggregation/dispersion pattern, which is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons.
 Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg].
 Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
-The possibility that some sites functioned as aggregation sites may have influenced the geographic structure of sites (@fig-group-map), which change from phase to phase.
+The possibility that some sites functioned as aggregation sites may have influenced the geographic structure of sites (@fig-group-map), which changes from phase to phase.
 

### 2025-06-02 paper/paper.qmd
map

@@
 
-In identifying the groups, we referenced both the seriation and network analyses. The results of the above tests allowed us to find two main groups of sites within each phase (@fig-seriation).
-
-## Cultural Connection Strength
-
-```{r}
-#| label: tbl-strength-table
-#| tbl-cap: "Table of the results per phase for the perMANOVA"
-#| tbl-cap-location: bottom
-
-strength_table <- full_table %>% 
-  dplyr::select(`Aurignacian phase`, `PerMANOVA R2`, `PerMANOVA F`, `PerMANOVA p`)
-
-kable(strength_table)
-
-trans_perm_r <- trans_strength %>% 
-  dplyr::select(`PerMANOVA R2`)
-early_perm_r <- early_strength %>% 
-  dplyr::select(`PerMANOVA R2`)
-
-trans_perm_p <- trans_strength %>% 
-  dplyr::select(`PerMANOVA p`)
-
-trans_perm_f <- trans_strength %>% 
-  dplyr::select(`PerMANOVA F`)
-early_perm_f <- early_strength %>% 
-  dplyr::select(`PerMANOVA F`)
-```
-
-The perMANOVA R^2^ statistic shows the proportion of sign variation that can be explained by group membership.
-The Transitional period has the highest score of `r trans_perm_r` while the Early Aurignacian has the lowest score of `r early_perm_r`.
-We see a large decline in scores between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian periods.
-The pseudo-F statistic quantifies the difference between groups, with the Transitional phase having the highest value of `r trans_perm_f`, and the Early Aurignacian having the lowest value of `r early_perm_f`.
-Similar to the R^2^ statistic, we see a decline in values between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian period.
-The perMANOVA p-value indicates a statistically significant difference in the distribution of signs between the restricted and broad groups for the Proto , Early, and Evolved Aurignacian phases, but not the Transitional phase (p = `r trans_perm_p`).
-
-# Discussion
-
-## Grouping Dynamics Across the Four Phases of the Aurignacian
-
-Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and distribution.
-The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.
-@fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
-The perMANOVA tests show that the changes between the the Proto, Early, and Evolved Aurignacian phases are statistically significant, that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.
-The
-
-One interpretation of the inter-connectivity seen in @fig-neighbor-net is as a representation of non-utilitarian mobility.
-Non-utilitarian mobility establishes regional social networks which facilitate the spread of information and the survival of hunter-gatherers in uncertain environments [@Whallon_2006; @Fitzhugh_Phillips_Gjesfeld_2011; @Lovis_Donahue_2011; @Hitchcock_Ebert_2011; @Romano_Lozano_Fern2002].
-The exchange and production of decorative items is associated with non-utilitarian mobility and has been recognized as a sign of social contact in other regions and eras [@Whallon_2006; @Kelly_2024; @Kinahan_2013; @Newlander_2012; @Iizuka_Ferguson_Izuho_2022].
-Existing data has shown a range of variation in mobility among hunter-gatherer groups [@Binford_1990; @Kelly_1983; @Johnson_2014; @Padilla-Iglesias_Bischoff_2024; @Singh_Glowacki_2022]; some of the inter-connections in @fig-neighbor-net may represent shorter-term information exchange, while others may indicate longer-term residential movement.
-
-One form of longer-term residential movement is the aggregation/dispersion pattern, which is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons.
-Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg].
-Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
-The possibility that some sites functioned as aggregation sites may have influenced the geographic structure of sites (@fig-group-map), which changes from phase to phase.
+In identifying the groups, we referenced both the seriation and network analyses. The results of the above tests allowed us to find two main groups of sites within each phase (@fig-group-map).
 
@@
 
+## Cultural Connection Strength
+
+```{r}
+#| label: tbl-strength-table
+#| tbl-cap: "Table of the results per phase for the perMANOVA"
+#| tbl-cap-location: bottom
+
+strength_table <- full_table %>% 
+  dplyr::select(`Aurignacian phase`, `PerMANOVA R2`, `PerMANOVA F`, `PerMANOVA p`)
+
+kable(strength_table)
+
+trans_perm_r <- trans_strength %>% 
+  dplyr::select(`PerMANOVA R2`)
+early_perm_r <- early_strength %>% 
+  dplyr::select(`PerMANOVA R2`)
+
+trans_perm_p <- trans_strength %>% 
+  dplyr::select(`PerMANOVA p`)
+
+trans_perm_f <- trans_strength %>% 
+  dplyr::select(`PerMANOVA F`)
+early_perm_f <- early_strength %>% 
+  dplyr::select(`PerMANOVA F`)
+```
+
+The perMANOVA R^2^ statistic shows the proportion of sign variation that can be explained by group membership.
+The Transitional period has the highest score of `r trans_perm_r` while the Early Aurignacian has the lowest score of `r early_perm_r`.
+We see a large decline in scores between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian periods.
+The pseudo-F statistic quantifies the difference between groups, with the Transitional phase having the highest value of `r trans_perm_f`, and the Early Aurignacian having the lowest value of `r early_perm_f`.
+Similar to the R^2^ statistic, we see a decline in values between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian period.
+The perMANOVA p-value indicates a statistically significant difference in the distribution of signs between the restricted and broad groups for the Proto , Early, and Evolved Aurignacian phases, but not the Transitional phase (p = `r trans_perm_p`).
+
+# Discussion
+
+## Grouping Dynamics Across the Four Phases of the Aurignacian
+
+Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and distribution.
+The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.
+@fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
+The perMANOVA tests show that the changes between the the Proto, Early, and Evolved Aurignacian phases are statistically significant, that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.
+
+One interpretation of the connectivity seen in @fig-neighbor-net is as a representation of non-utilitarian mobility.
+Non-utilitarian mobility establishes regional social networks which facilitate the spread of information and the survival of hunter-gatherers in uncertain environments [@Whallon_2006; @Fitzhugh_Phillips_Gjesfeld_2011; @Lovis_Donahue_2011; @Hitchcock_Ebert_2011; @Romano_Lozano_Fern2002].
+The exchange and production of decorative items is associated with non-utilitarian mobility and has been recognized as a sign of social contact in other regions and eras [@Whallon_2006; @Kelly_2024; @Kinahan_2013; @Newlander_2012; @Iizuka_Ferguson_Izuho_2022].
+Previous studies have shown a range of variation in mobility among hunter-gatherer groups [@Binford_1990; @Kelly_1983; @Johnson_2014; @Padilla-Iglesias_Bischoff_2024; @Singh_Glowacki_2022]; some of the connections in @fig-neighbor-net may represent shorter-term information exchange, while others may indicate longer-term residential movement.
+
+One form of longer-term residential movement is the aggregation/dispersion pattern, which is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons.
+Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg].
+Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
+The possibility that some sites functioned as aggregation sites may have influenced the geographic structure of sites (@fig-group-map), which change from phase to phase.
+
+```{r fig.height= 7, fig.width=10}
+#| label: fig-group-map
+#| fig-cap: "Map of sites faceted by time period and colored by group number"
+
+ggplot(Europe) +
+  geom_sf() +
+  geom_sf(data = signbase_sf, aes(color = group)) +
+  coord_sf(xlim = c(-10,30), 
+           ylim = c(35,53), 
+           expand = FALSE) +
+  geom_text_repel(data = signbase_sf,
+                  aes(x = longitude ,
+                      y = latitude,
+                      label = site_name),
+                  max.overlaps = 30,
+                  size = 2,
+                  bg.color = "white",
+                  bg.r = 0.25) +
+  theme_minimal() +
+  facet_wrap(~time_period) +
+  theme(axis.text = element_blank(),
+        axis.title = element_blank(),
+        strip.text = element_text(size = 10))
+
+```
+
+The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation.
+In the Transitional phase, sites are widely spread out across Western Europe.
+This wider distribution may be a factor in weak connections seen in @fig-neighbor-net.
+In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the slightly stronger connectivity seen there.
+This phase is also associated with an increase in sites and sign types, which may be correlated with the increased proximity and population density.
+In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together, which may be associated with the increased connectivity in that phase.
+In the Evolved Aurignacian, we see a wider distribution and more distance between sites, which may be associated with the reduced connectivity seen in that phase.
+
 A potential identifier of aggregation sites comes from @fig-div-plot, which uses the tabula package [@Frerebeau_2025] to generate the Shannon-Weaver diversity index of sign types per site, and plot the expected diversity index for a given sample size against the actual diversity index per site [@Kintigh_1984].

### 2025-06-04 paper/paper.qmd
dupe
edit

@@
 
-In identifying the groups, we referenced both the seriation and network analyses. The results of the above tests allowed us to find two main groups of sites within each phase (@fig-group-map).
-
 ```{r fig.height= 7, fig.width=10}
@@
 
-The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation.
+In identifying the groups, we referenced both the seriation and network analyses.
+The results of the above tests allowed us to find two main groups of sites within each phase (@fig-group-map).
 In the Transitional phase, sites are widely spread out across Western Europe.
@@
 In the Evolved Aurignacian, we see a wider distribution and more distance between sites, which may be associated with the reduced connectivity seen in that phase.
+The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation.
 
@@
 ## Grouping Dynamics Across the Four Phases of the Aurignacian
-
 Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and distribution.
@@
 
+One consistent grouping dynamic throughout all four phases is the restricted/broader group divide, which may be influenced by variation in social structure.
+Despite the tendency to conceive of hunter-gatherers as homogeneous, previous studies have established variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006; @Johnson_2014; @Singh_Glowacki_2022; @Schwendler_2012], and additional studies have established how variation in ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982; @Hodder1979; @Hodder1977; @Schwendler_2012].
+It is therefore possible that, for some of the sites, the broader/restricted grouping indicates variation in social structure, with the broader groups having a greater variety of formalized social roles, and the restricted groups having fewer formalized social roles.
+
 One interpretation of the connectivity seen in @fig-neighbor-net is as a representation of non-utilitarian mobility.
@@
 One form of longer-term residential movement is the aggregation/dispersion pattern, which is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons.
-Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg].
+Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg], inclusing, in some cases, decorative objects [@Kelly_2024].
 Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
-The possibility that some sites functioned as aggregation sites may have influenced the geographic structure of sites (@fig-group-map), which change from phase to phase.
-
-```{r fig.height= 7, fig.width=10}
-#| label: fig-group-map
-#| fig-cap: "Map of sites faceted by time period and colored by group number"
-
-ggplot(Europe) +
-  geom_sf() +
-  geom_sf(data = signbase_sf, aes(color = group)) +
-  coord_sf(xlim = c(-10,30), 
-           ylim = c(35,53), 
-           expand = FALSE) +
-  geom_text_repel(data = signbase_sf,
-                  aes(x = longitude ,
-                      y = latitude,
-                      label = site_name),
-                  max.overlaps = 30,
-                  size = 2,
-                  bg.color = "white",
-                  bg.r = 0.25) +
-  theme_minimal() +
-  facet_wrap(~time_period) +
-  theme(axis.text = element_blank(),
-        axis.title = element_blank(),
-        strip.text = element_text(size = 10))
-
-```
-
-The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation.
-In the Transitional phase, sites are widely spread out across Western Europe.
-This wider distribution may be a factor in weak connections seen in @fig-neighbor-net.
-In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the slightly stronger connectivity seen there.
-This phase is also associated with an increase in sites and sign types, which may be correlated with the increased proximity and population density.
-In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together, which may be associated with the increased connectivity in that phase.
-In the Evolved Aurignacian, we see a wider distribution and more distance between sites, which may be associated with the reduced connectivity seen in that phase.
-
-A potential identifier of aggregation sites comes from @fig-div-plot, which uses the tabula package [@Frerebeau_2025] to generate the Shannon-Weaver diversity index of sign types per site, and plot the expected diversity index for a given sample size against the actual diversity index per site [@Kintigh_1984].
-The plot identifies multiple sites that diverge from the diversity index predicted by sample size, including those with a higher index than expected, such as Hohle Fels in the Transitional phase, Spy, La Ferrassie, and Les Cottés in the Proto-Aurignacian, Solutré, Hohle Fels, and Grottes de Fonds-de-Forêt in the Early Aurignacian, and Vogelherd, Bockstein-Törle, and Vindija Cave in the Evolved Aurignacian.
-It is possible that those sites with a greater than expected diversity index acted as aggregation sites for their groups or served a similar purpose.
-La Ferrassie in the Proto-Aurignacian is also shown by @fig-neighbor-net to have multiple connections to sites in both groups, in addition to the diversity index shown in @fig-div-plot, perhaps suggesting that it functioned as a specifically inter-group aggregation site.
+
+Aggregation activities may be identified by a higher than expected diversity index.
+@fig-div-plot plots the expected Shannon-Weaver diversity index [@Frerebeau_2025] for a given sample size against the actual diversity index per site [@Kintigh_1984].
+The plot identifies multiple sites with a higher index than expected, such as Hohle Fels in the Transitional phase, Spy, La Ferrassie, and Les Cottés in the Proto-Aurignacian, Solutré, Hohle Fels, and Grottes de Fonds-de-Forêt in the Early Aurignacian, and Vogelherd, Bockstein-Törle, and Vindija Cave in the Evolved Aurignacian.
+Notably, most of these sites are not shown as having connections between groups in @fig-neighbor-net, except for La Ferrassie in the Proto-Aurignacian.
+This suggests a distinction between shorter-term information exchange (e.g.
+the highly interconnected sites in @fig-neighbor-net) and longer-term residential movement (e.g.
+the sites with higher than expected diversity in @fig-div-plot).
 
@@
 
-@fig-div-plot shows divergence from the expected diversity index for the sample size.
-Another potential explanation for the observed variation in diversity is that it may indicate variation in social structure.
-Despite the tendency to conceive of hunter-gatherers as homogenous, previous studies have established variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006; @Johnson_2014; @Singh_Glowacki_2022; @Schwendler_2012].
-Additional research has established how variation in ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982; @Hodder1979; @Hodder1977; @Schwendler_2012].
-It is therefore possible that greater diversity in geometric signs indicates a wider variety of formal social roles [@Conkey_1985; @Schwendler_2012].
-This is especially relevant in regards to our larger grouping structure of broader versus restricted signs, as it may indicate two broad types of social structure, with the broader groups having a more formal social structure, and the restricted groups having a less formal social structure.
-
 ## Social Organization Dynamics Throughout the Aurignacian

### 2025-06-09 paper/paper.qmd
edits to text, some plots, formatting, front matter

@@
 ---
-title: "Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases"
-author: Sophie Cooper
-format: docx
+title: "Geometric signs reveal changes social structures and networks during the Western Eurasian Aurignacian"
+author:
+  - Sophie Cooper:
+      institute: uw
+      orcid: xxx
+  - Ben Marwick:
+      correspondence: "yes"
+      email: bmarwick@uw.edu
+      orcid: 0000-0001-7879-4531
+      institute:
+        - uw
+institute:
+  - uw: Department of Anthropology, University of Washington, Seattle, USA
+title-block-published: "Last updated"  
+date: now
+date-format: long
+format: 
+  docx:
+    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
+execute:
   echo: false
-execute: 
   warning: false
+  message: false
+  comment: "#>"
+  fig-path: "../figures/"
+  fig-dpi: 1000
+filters:
+  - ../templates/scholarly-metadata.lua
+  - ../templates/author-info-blocks.lua
+  - ../templates/pagebreak.lua
 bibliography: references.bib
-link-citations: true
+csl: "../templates/society-for-american-archaeology.csl" # Insert path for the bib-style
+abstract: |
+  ...
+keywords: |
+
 ---
+
 # Introduction
 
-The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020], coupled with a biological transition from Neanderthals to early modern humans.
-Archaeologists have frequently relied upon variations in artwork, the large rise of which is a distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d; @Fuentes_Lucas_Robert_2019].
+Archaeological evidence from the Upper Paleolithic period indicates increasingly complex social organization across Europe [@Boyd_Silk_2020], coinciding with the ongoing biological transition from Neanderthals to early modern humans.
+Archaeologists have frequently used ancient art, the abundance of which is distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d; @Fuentes_Lucas_Robert_2019].
 Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca.
-43 - 30 k BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020].
-In this paper, we investigate the social organization dynamics by analyzing the distribution and variation of geometric sign types found on mobile objects from the Upper Paleolithic in Europe across four phases of the Aurignacian period.
-
-Social organization dynamics of the Upper Paleolithic have previously been explored using statistical analyses of objects to understand patterns of cultural groups.
+43 - 30 k BP) is often linked to the dispersal of _Homo sapiens_ across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020].
+In this paper we investigate Aurignacian social organization dynamics implied by the distribution and variation of geometric sign types found on mobile objects in Europe across four phases of the Aurignacian period.
+
+Previous work has fruitfully explored social organization dynamics of the Upper Paleolithic using statistical analyses of objects to understand patterns of cultural groups.
 Multivariate analyses of beads from the Gravettian technocomplex suggests an east to west cline of nine cultural groups [@baker2024evidence].
-Seriation analysis, PCoA, and network analysis were used to identify the groups from 134 discrete bead types recovered from both burial and occupation sites.
+Seriation analysis, PCoA, and network analysis were used to identify groups from 134 discrete bead types recovered from both burial and occupation sites.
 To validate their claims of cultural groups, @baker2024evidence used a Mantel test to evaluate an isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance, finding that geographic distance alone did not solely account for the bead distribution.
@@
 Similar analyses of personal ornament types from the Aurignacian technocomplex identified fourteen geographically cohesive groups [@Vanhaeren_d].
-Drawing on ethnographic studies that show how body decoration and ornamentation indicates ethno-linguistic identity, @Vanhaeren_d used Aurignacian personal ornamentation as a proxy for ethno-linguistic diversity.
-The variation in personal ornament types is interpreted as evidence of long-lasting cultural differences and as robustly establishing the ethno-linguistic diversity of the Aurignacian.
-157 distinct ornament types from 98 Aurignacian sites in Europe and the Near East were analyzed, identifying geographically cohesive groups sharing similar ornament type associations by seriation, correspondence, and GIS analyses.
+Drawing on ethnographic studies that show how body decoration and ornamentation indicate ethno-linguistic identity, @Vanhaeren_d used Aurignacian personal ornamentation as a proxy for ethno-linguistic diversity.
+The variation in personal ornament types is interpreted as evidence of long-lasting cultural differences and as robustly establishing the ethno-linguistic diversity of the Aurignacian period.
+They used seriation, correspondence, and GIS analyses of 157 distinct ornament types from 98 Aurignacian sites in Europe and the Near East to identify geographically cohesive groups sharing similar ornament type associations.
 These groups sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.
 
-Cultural difference has also been identified from geometric signs in Upper Paleolithic parietal art from the Cantabrian region of Spain [@Sauvet_2018].
-Forty-two quadrilateral geometric signs from four cave sites, likely associated with the Magdalenian period, spanning a distance of 30 kilometers were recorded.
+Cultural groups have also been identified from forty-two quadrilateral geometric signs from four cave sites in the Cantabrian region of Spain, likely associated with the Magdalenian period, spanning a distance of 30 kilometers [@Sauvet_2018].
 Multivariate analysis of 45 morphological characteristics of the signs revealed two clusters.
-To interpret these clusters, @Sauvet_2018 draw on case studies of geometric signs as identity markers, including South African engraved ostrich shells from 60 k BP, maker's marks on Gallo-Roman pottery, and 19th century French craft guild members' personal marks.
-Using these case studies as analogies, @Sauvet_2018 interpret the Magdalenian geometric signs as identity markers, and the two clusters as representative of cultural groups.
-
-In this study, we extend the use of geometric signs as identity markers to investigate cultural groups in the Aurignacian period.
-The Aurignacian period spans roughly 43 - 30 k BP, and is largely characterized by material culture indicators, such as carinated scrapers, small-flake tools and split-based bone and ivory tools, as well as an ongoing evolutionary shift from Neanderthals to early modern humans [@Chu_Richter_2020].
-The Aurignacian can be divided into phases which are mainly distinguished by shifts in material culture; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015], as can be seen in @tbl-aur-table.
-@Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main phases, Aur-P1 (Proto/Early), which corresponds with the time period of 43–37 k BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods 37-32 k BP.
+To interpret these clusters, @Sauvet_2018 drew on case studies of geometric signs as identity markers, including South African engraved ostrich shells from 60 k BP, maker's marks on Gallo-Roman pottery, and 19th century French craft guild members' personal marks.
+Using these case studies as analogies, @Sauvet_2018 interpreted the Magdalenian geometric signs as identity markers, and the two clusters as representative of two cultural groups.
+
+In this study, we extend this previous work using geometric signs as identity markers to investigate cultural groups in the Aurignacian period.
+The Aurignacian period spans roughly 43-30 k BP, and is largely characterized by material culture indicators such as carinated scrapers, small-flake tools and split-based bone and ivory tools, as well as an ongoing evolutionary shift from Neanderthals to early modern humans [@Chu_Richter_2020].
+@Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main phases, Aur-P1 (Proto/Early, 43–37 k BP), and Aur-P2 (Evolved/Late, 37-32 k BP).
 @Banks_dErrico_Zilhao_2013 used Bayesian modeling methods to further divide Aur-P1 into the Proto and Early Aurignacian, placing the Proto-Aurignacian at 41.5 - 39.9 k BP, and the early Aurignacian at 39.8 - 37.8 k BP.
 They found that the Proto-Aurignacian occurs during the Greenland Interstadials (GI) 9 and 10, periods of climatic amelioration, while the Early Aurignacian began with the Henrich Stadial, which is characterized by dry and arid conditions, and ends with GI 8.
+Similar to @Banks_dErrico_Zilhao_2013, we divide the Aurignacian into four phases based on material culture changes; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015], as summarised in @tbl-aur-table.
 
@@
 
-We hypothesize that these phases, and the environmental changes across the Aurignacian, may correspond to changes in social dynamics of cultural groups, as represented by patterns in the distribution of geometric signs.
+We hypothesize that these phases, and the environmental changes across the them, may correspond to changes in dynamics of cultural groups, as represented by patterns in the distribution of geometric signs.
 We aim to address the following questions: How do grouping dynamics of geometric signs change across the four phases of the Aurignacian?
 How do the groupings vary in diversity, size, connectivity, and distribution?
-Lastly, what do the patterns tell us about dynamics of social organization throughout the Aurignacian?
-We answer these questions by first identifying clusters within each phase using seriation and network analyses.
-We then test the strength of the clusters using perMANOVA tests.
-All of our programming in this study was conducted in RStudio.
+What do these patterns tell us about dynamics of social organization throughout the Aurignacian?
+To answer these questions we use seriation analysis, network analysis, and PerMANOVA tests.
 
@@
 The data set consists of `r nartifacts` objects found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign types.
-We excluded signs if they fell under the "other" category, due to the wide range of potential possibilities.
-We also excluded seven sites from the sample (only one sign present: Willendorf; extreme geographic distance from the other sites: Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave; dated much later than the rest of the assemblages: Grotte De La Princesse Pauline and Šandalja II).
-These exclusions resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map).
+We excluded signs classified by @dutkiewicz2020 as "other".
+To improve the quality of our data we excluded seven sites from the sample: only Willendorf (one sign present); Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave (extreme geographic distance from the other sites); Grotte De La Princesse Pauline and Šandalja II (dated much later than the rest of the assemblages).
+This resulted in a sample of `r nartifacts_clean` objects found at `r nsites_clean` sites in `r ncountries_clean` countries, and `r nsigntypes_clean` sign types used in this study (@fig-site-map).
 
@@
 The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation.
-The p-value quantifies the probability of obtaining the observed correlation if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
+The p-value estimates the probability of obtaining the observed correlation if the null hypothesis of no correlation were true (we use an alpha value of 0.05 for the threshold of statistical significance).
 The Mantel tests were performed using the vegan package [@Vegan2001].
-The test indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`).
-
-Ages of the objects were determined by calibrating the uncalibrated ages provided by SignBase, using the rcarbon package [@Crema_Bevan_2021].
+The test indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`), validating our decision to use presence/absence data.
+
+Ages of the objects were determined by calibrating the uncalibrated ages provided by SignBase using the rcarbon package [@Crema_Bevan_2021].
 The calibrated radiocarbon dates of the sampled data range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years.
@@
 
-If the distribution of sign types can be explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases [@Herskind2023; @Rigaud_dErrico_Vanhaeren_2015].
+If the distribution of sign types can be mostly explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases [@Herskind2023; @Rigaud_dErrico_Vanhaeren_2015].
 This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019].
-To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geography for each phase.
+To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geographic distance for each phase.
 We used Mantel tests with a geographic distance matrix and a Jaccard similarity matrix, with the results shown in @tbl-mantel-table.
-None of the phases have a statistically significant correlation between sign type distribution and geographical distance, so the isolation-by-distance hypothesis is rejected.
+None of the phases have a statistically significant correlation between sign type distribution and geographical distance, so we reject the isolation-by-distance hypothesis as a primary explanation for variation in sign type distribution.
 
@@
 
-For each of the phases, seriation identifies at least two groups, one with a restricted range of sign types (1-3 types), which tend to be simpler signs, and one with a broader range (up to `r max_signs` types), which include more complex signs.
+For each of the phases, our seriation analysis identifies two groups, one with a restricted range of sign types (1-3 types), which tend to be visually simpler signs, and one with a broader range (up to `r max_signs` types), which include more complex signs.
 Simpler signs composed of fewer components, like line and notch, are the most prevalent across all time periods.
-Overall, across the four phases, the number of sites in the restricted group tends to be less than or equal to the number of sites in the broader group, with the exception of the final Evolved phase.
+Overall, across the four phases, the number of sites in the restricted range group tends to be less than or equal to the number of sites in the broader range group, with the exception of the final Evolved phase.
 
 In the Transitional phase, there is a relatively small range of sign types compared to other phases.
-The restricted group consists of Fumane, Pod Hradem, and Geissenklösterle, two of which only have a notch, and one of which has a notch and circumnotch.
-The broader group consists of Labeko Koba, Hohle Fels, and El Castillo, which vary in their sign composition and have only the line and oblique line (obline) signs in common.
+The restricted range group consists of Fumane, Pod Hradem, and Geissenklösterle, two of which only have a notch, and one of which has a notch and circumnotch.
+The broader range group consists of Labeko Koba, Hohle Fels, and El Castillo, which vary in their sign composition and have only the line and oblique line (obline) signs in common.
 
 In the Proto-Aurignacian we see an expansion in both the number of sites and range of signs, with no overlap in sites from the Transitional phase, except for Geissenklösterle, which is present in both phases.
-The restricted group consists of Abri Pataud, Hohlenstein-Stadel, and Mladeč, which only have notches and lines, and Gatzarria which also has a dashline.
-The broader group consists of Geissenklösterle, Spy, Vogelherd, Grotte du Renne, La Ferrassie, and Les Cottés, where we see a more consistent pattern of a more diverse range of signs, compared to the broader group of the Transitional phase.
-In addition to line and notch, the obline, radial notch (radnotch), cross, hatching, and circumferential notch (circumnotch) are also frequently found in the broader group.
+The restricted range group consists of Abri Pataud, Hohlenstein-Stadel, and Mladeč, which only have notches and lines, and Gatzarria which also has a dashline.
+The broader range group consists of Geissenklösterle, Spy, Vogelherd, Grotte du Renne, La Ferrassie, and Les Cottés, where we see a more consistent pattern of a more diverse range of signs, compared to the broader range group of the Transitional phase.
+In addition to line and notch, the obline, radial notch (radnotch), cross, hatching, and circumferential notch (circumnotch) are also frequently found in the broader range group.
 
 In the Early Aurignacian, we see appearance of completely new sites with no overlap with the Proto-Aurignacian sites, as well as increased diversity of patterns within the groups.
-The restricted group consists of Grottes de Fonds-de-Forêt (radnotch and notch), Riparo Bombrini (notch), and Grotte de la Verpillière I (notch and line).
-The broader group consists of Solutré, Hohle Fels, Castanet, Cellier, Blanchard, and Trou al'Wesse, which contain a wide range of signs in a variety of combinations.
-In addition to line and notch, the obline, dot and vulva are also frequently found in the broader group.
-
-Lastly, in the Evolved Aurignacian, we see slightly more consistent groups, again with new sites with no overlap with the Early Aurignacian sites, except for Hohle Fels, which also appears in the Transitional phase.
-Vogelherd also reappears, having appeared earlier in the Proto-Aurignacian.The restricted group consists of Les Rois, Gargas, La Viña, Sirgenstein Cave, and Vindija Cave, with the line, notch, and and obline signs.
-The broader group consists of Trou Magrite, Bockstein-Törle, Vogelherd, and Hohle Fels.
-Along with line, notch, and obline, the hatching, and oblique notch (obnotch) signs are also frequently found in the broader group.
+The restricted range group consists of Grottes de Fonds-de-Forêt (radnotch and notch), Riparo Bombrini (notch), and Grotte de la Verpillière I (notch and line).
+The broader range group consists of Solutré, Hohle Fels, Castanet, Cellier, Blanchard, and Trou al'Wesse, which contain a wide range of signs in a variety of combinations.
+In addition to line and notch, the obline, dot and vulva are also frequently found in the broader range group.
+
+Finally, in the Evolved Aurignacian, we see slightly more consistent groups, again with new sites with no overlap with the Early Aurignacian sites, except for Hohle Fels, which also appears in the Transitional phase.
+Vogelherd also reappears, having appeared earlier in the Proto-Aurignacian.
+The restricted range group consists of Les Rois, Gargas, La Viña, Sirgenstein Cave, and Vindija Cave, with the line, notch, and and obline signs.
+The broader range group consists of Trou Magrite, Bockstein-Törle, Vogelherd, and Hohle Fels.
+Along with line, notch, and obline, the hatching, and oblique notch (obnotch) signs are also frequently found in the broader range group.
 
@@
 
-The results of the network analysis supports the restricted-broad groupings identified in @fig-seriation, and provides additional insights into the evolution of connectivity between groups.
+The results of the network analysis supports the restricted-broad groupings identified in @fig-seriation, and provide additional insights into the evolution of connectivity between groups.
 In the Transitional phase, the network analysis reveals one distinct and one disconnected group.
-The restricted group consists of Geissenklösterle, Fumane, and Pod Hradem, and the broad group consists of Hohle Fels, El Castillo, and Labeko Koba, with no connections between the groups.
-The Proto-Aurignacian phase also shows two clearly defined groups, the broad group consisting of Geissenklösterle, Spy, Vogelherd, and Grotte du Renne, and the restricted group consisting of Mladeč, Hohlenstein-Stadel, Gatzarria, and Abri Pataud.
+The restricted range group consists of Geissenklösterle, Fumane, and Pod Hradem, and the broad range group consists of Hohle Fels, El Castillo, and Labeko Koba, with no connections between the groups.
+The Proto-Aurignacian phase also shows two clearly defined groups, the broad range group consisting of Geissenklösterle, Spy, Vogelherd, and Grotte du Renne, and the restricted range group consisting of Mladeč, Hohlenstein-Stadel, Gatzarria, and Abri Pataud.
 However, it also displays some connectivity between the two groups of sites, with La Ferrassie connected to both groups.
@@
 The Early Aurignacian shows more strongly inter-connected groups -- although the shape still suggests two main groups -- with Grotte de la Verpillière I in particular having multiple strong connections to both groups.
-Lastly, the Evolved Aurignacian again shows two main groups, although intra-group connectivity has decreased, with the broad group consisting of Hohle Fels, Trou Magrite, Vogelherd, and Bockstein-Törle, and the restricted group consisting of Gargas, La Viña, Sirgenstein Cave, Vindija Cave, and Les Rois.
+Lastly, the Evolved Aurignacian again shows two main groups, although intra-group connectivity has decreased, with the broad range group consisting of Hohle Fels, Trou Magrite, Vogelherd, and Bockstein-Törle, and the restricted range group consisting of Gargas, La Viña, Sirgenstein Cave, Vindija Cave, and Les Rois.
 As for the Early Aurignacian, inter-connectivity between groups continues to be evident in the Evolved Aurignacian, with connections between Trou Magrite, Gargas, and Vindija Cave, and Bockstein-Törle and Gargas, however the strength of inter-connectivity is weaker than in the Early Aurignacian.
@@
 #| label: fig-group-map
-#| fig-cap: "Map of sites faceted by time period and colored by group number"
+#| fig-cap: "Map of sites faceted by time period and colored by group type"
+
+signbase_sf_groups <- 
+  signbase_sf %>% 
+  mutate(group_map = case_when(
+    group == 1 ~ "Restricted range",
+    group == 2 ~ "Broader range",
+  ))
 
 ggplot(Europe) +
-  geom_sf() +
-  geom_sf(data = signbase_sf, aes(color = group)) +
+  geom_sf(fill = "white") +
+  geom_sf(data = signbase_sf_groups, 
+          aes(shape = group_map),
+          color = "black",
+          size = 3.5) +
+  geom_sf(data = signbase_sf_groups, 
+          aes(color = group_map,
+              shape = group_map),
+          size = 3) +
   coord_sf(xlim = c(-10,30), 
@@
            expand = FALSE) +
-  geom_text_repel(data = signbase_sf,
+  geom_text_repel(data = signbase_sf_groups,
                   aes(x = longitude ,
@@
                   max.overlaps = 30,
-                  size = 2,
+                  size = 3,
                   bg.color = "white",
-                  bg.r = 0.25) +
-  theme_minimal() +
+                  bg.r = 0.1) +
+  theme_bw() +
   facet_wrap(~time_period) +
@@
         axis.title = element_blank(),
-        strip.text = element_text(size = 10))
-
-```
-
-In identifying the groups, we referenced both the seriation and network analyses.
-The results of the above tests allowed us to find two main groups of sites within each phase (@fig-group-map).
+        strip.text = element_text(size = 10),
+        legend.position = "bottom")  +
+  scale_color_manual(name = "Group", 
+                     values = c("Restricted range" = "#FFFFB3", 
+                                "Broader range"    = "#8DD3C7")) +
+  scale_shape_manual(name = "Group",
+                     values = c("Restricted range" = 17, 
+                                "Broader range"    = 19))
+
+```
+
+While th results of the seriation and network analyses point to two main groups of sites within each phase, spatial clusters of sites are not highly evident in these groups, in part due to the small numbers of sites in each phase (@fig-group-map).
 In the Transitional phase, sites are widely spread out across Western Europe.
 This wider distribution may be a factor in weak connections seen in @fig-neighbor-net.
-In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the slightly stronger connectivity seen there.
+In the Proto-Aurignacian, we see a slightly narrower geographic distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the slightly stronger connectivity seen there.
 This phase is also associated with an increase in sites and sign types, which may be correlated with the increased proximity and population density.
@@
 Similar to the R^2^ statistic, we see a decline in values between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian period.
-The perMANOVA p-value indicates a statistically significant difference in the distribution of signs between the restricted and broad groups for the Proto , Early, and Evolved Aurignacian phases, but not the Transitional phase (p = `r trans_perm_p`).
+The perMANOVA p-value indicates a statistically significant difference in the distribution of signs between the restricted and broad range groups for the Proto , Early, and Evolved Aurignacian phases, but not the Transitional phase (p = `r trans_perm_p`).
 
@@
 ## Grouping Dynamics Across the Four Phases of the Aurignacian
+
 Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and distribution.
-The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.
-@fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
+The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type number and diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.
+@fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to very strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
 The perMANOVA tests show that the changes between the the Proto, Early, and Evolved Aurignacian phases are statistically significant, that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.
 
-One consistent grouping dynamic throughout all four phases is the restricted/broader group divide, which may be influenced by variation in social structure.
-Despite the tendency to conceive of hunter-gatherers as homogeneous, previous studies have established variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006; @Johnson_2014; @Singh_Glowacki_2022; @Schwendler_2012], and additional studies have established how variation in ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982; @Hodder1979; @Hodder1977; @Schwendler_2012].
-It is therefore possible that, for some of the sites, the broader/restricted grouping indicates variation in social structure, with the broader groups having a greater variety of formalized social roles, and the restricted groups having fewer formalized social roles.
-
-One interpretation of the connectivity seen in @fig-neighbor-net is as a representation of non-utilitarian mobility.
+One consistent grouping dynamic throughout all four phases is the restricted/broader range group divide, which may be influenced by variation in social structure.
+Variation in hunter-gatherer social structures is well-known from ethnographic and archaeological research  [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006; @Johnson_2014; @Singh_Glowacki_2022; @Schwendler_2012].
+Studies have established how variation in ornamentation and other forms of material culture can be used to communicate information about the numbers and types of social roles and statuses within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982; @Hodder1979; @Hodder1977; @Schwendler_2012].
+Drawing on this work, we interpret the broader/restricted range grouping in the geometric sign data as indicating variation in social structure.
+The broader range groups, with higher variety of signs and more visually complex signs, indicates communities with a greater variety of formalized social roles.
+Conversely, the sites with a restricted range of signs indicate communities with fewer formalized social roles.
+A single larger community could have both of these groups as sub-sets, rather than the restricted/broader range groups being distinct, non-overlapping cultural groups.
+
+The connectivity between sites represented in @fig-neighbor-net may reflect non-utilitarian mobility among these larger communities.
+Previous studies have shown a range of variation in magnitudes and types of mobility among hunter-gatherer groups [@Binford_1990; @Kelly_1983; @Johnson_2014; @Padilla-Iglesias_Bischoff_2024; @Singh_Glowacki_2022].
 Non-utilitarian mobility establishes regional social networks which facilitate the spread of information and the survival of hunter-gatherers in uncertain environments [@Whallon_2006; @Fitzhugh_Phillips_Gjesfeld_2011; @Lovis_Donahue_2011; @Hitchcock_Ebert_2011; @Romano_Lozano_Fern2002].
-The exchange and production of decorative items is associated with non-utilitarian mobility and has been recognized as a sign of social contact in other regions and eras [@Whallon_2006; @Kelly_2024; @Kinahan_2013; @Newlander_2012; @Iizuka_Ferguson_Izuho_2022].
-Previous studies have shown a range of variation in mobility among hunter-gatherer groups [@Binford_1990; @Kelly_1983; @Johnson_2014; @Padilla-Iglesias_Bischoff_2024; @Singh_Glowacki_2022]; some of the connections in @fig-neighbor-net may represent shorter-term information exchange, while others may indicate longer-term residential movement.
-
-One form of longer-term residential movement is the aggregation/dispersion pattern, which is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons.
-Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg], inclusing, in some cases, decorative objects [@Kelly_2024].
+The exchange and production of decorative items, such as objects with geometric signs, has been associated with non-utilitarian mobility and recognized as an indicator of social contact [@Whallon_2006; @Kelly_2024; @Kinahan_2013; @Newlander_2012; @Iizuka_Ferguson_Izuho_2022].
+We can divide the connections depicted in @fig-neighbor-net into shorter-term information exchange and longer-term residential movement.
+
+One form of longer-term residential movement that may be related to non-utilitarian mobility are aggregation/dispersion patterns, which are defined by @conkley_agg as a type of hunter-gather settlement pattern in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons.
+Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate.
+Some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg], including, in some cases, decorative objects [@Kelly_2024].
 Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
 
-Aggregation activities may be identified by a higher than expected diversity index.
+Following this previous work, we identify aggregation sites as those with a higher than expected sign diversity index.
 @fig-div-plot plots the expected Shannon-Weaver diversity index [@Frerebeau_2025] for a given sample size against the actual diversity index per site [@Kintigh_1984].
-The plot identifies multiple sites with a higher index than expected, such as Hohle Fels in the Transitional phase, Spy, La Ferrassie, and Les Cottés in the Proto-Aurignacian, Solutré, Hohle Fels, and Grottes de Fonds-de-Forêt in the Early Aurignacian, and Vogelherd, Bockstein-Törle, and Vindija Cave in the Evolved Aurignacian.
+The plot identifies multiple sites with a higher than expected diversity of signs, such as Hohle Fels in the Transitional phase, Spy, La Ferrassie, and Les Cottés in the Proto-Aurignacian, Solutré, Hohle Fels, and Grottes de Fonds-de-Forêt in the Early Aurignacian, and Vogelherd, Bockstein-Törle, and Vindija Cave in the Evolved Aurignacian.
 Notably, most of these sites are not shown as having connections between groups in @fig-neighbor-net, except for La Ferrassie in the Proto-Aurignacian.
 This suggests a distinction between shorter-term information exchange (e.g.
-the highly interconnected sites in @fig-neighbor-net) and longer-term residential movement (e.g.
+the highly interconnected sites in @fig-neighbor-net) and longer-term residential movement for aggregation (e.g.
 the sites with higher than expected diversity in @fig-div-plot).
 
-```{r}
+```{r fig.height= 7, fig.width=10}
 #| label: fig-div-plot
@@
   geom_text_repel(aes(label = diversity_sim@labels),
-                  size = 3, max.overlaps = 15) +
+                  size = 3, 
+                  max.overlaps = 15,
+                  bg.color = "white",
+                  bg.r = 0.1) +
   guides(colour = "none") +
@@
 plot_grid(plotlist =  signbase_full_clean_diversity_plot_by_period$plot,
-          labels = signbase_full_clean_diversity_plot_by_period$time_period,
-          label_size = 8)
+          labels = as.character(signbase_full_clean_diversity_plot_by_period$time_period),
+          label_size = 8,
+          hjust = -1,
+          label_y = 0.95)
 ```
@@
 
-The second aim was to determine what these patterns tell us about dynamics of social organization throughout the Aurignacian.
+The second aim was to determine what these patterns reveal about the dynamics of social organization throughout the Aurignacian.
 The results suggest that the Transitional phase was sparsely populated, with minimal interaction between loosely connected groups.
@@
 From 45 - 43.25 k BP, the population density in Europe was extremely low, due to the unfavorable environmental conditions of Greenland Stadial 12 (44 k BP) and the slow westward expansion of early modern humans [@Shao_Wegener_Klein_Schmidt_Weniger_2024].
-This was followed by rapid expansion from 43.25 - 41 k BP (Transitional/Proto), with the maximum extent of Proto/Early Aurignacian settlement being reached around 41 k BP.
+This was followed by rapid expansion from 43.25-41 k BP (Transitional/Proto), with the maximum extent of Proto/Early Aurignacian settlement being reached around 41 k BP.
 This rapid population expansion could be what stimulated the emergence of signs as a marker of group membership.
@@
 
-Similarly, @Banks_dErrico_Zilhao_2013 use eco-cultural niche modeling to show that the Heinrich Stadial 4 (HS4) at the very beginning of the Early Aurignacian resulted in a expansion of people into new environmental niches, which they hypothesize was associated with an expansion of social networks both within and between populations.
+Our results are consistent with @Banks_dErrico_Zilhao_2013 who used eco-cultural niche modeling to show that the Heinrich Stadial 4 (HS4) at the very beginning of the Early Aurignacian resulted in a expansion of people into new environmental niches, which they hypothesize was associated with an expansion of social networks both within and between populations.
 This correlates with our results which show a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased sign type correlation with geographic distance.
-Lastly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions.
+Similarly, @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 developed a human-existence probability model for the Aurignacian, finding an expanded, more dispersed settlement area for Evolved/Late Aurignacian theorizing that the environmental changes of HS4 lead to humans better adapted to survive in a broader range of climate conditions.
 @Shao_Wegener_Klein_Schmidt_Weniger_2024 also finds an increase in expansion marking the beginning of the Evolved Aurignacian.
@@
 
+<!-- connect back with the papers we discuss in the background section about Aurginacian cultural groups, did we find the same or different?
+Most of our conclusions are not similar in kind to theirs, that is fine, we just need to make a connection between their work and ours here.
+-->
+
 # Conclusion
@@
 This study adds to our knowledge of social networks within the Aurignacian period, as our results document dynamic patterns of social exchange and connection.
-Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996; @Hill_Wood_Baggio_Hurtado_Boyd_2014; @Derex_Boyd_2016; @Romano_Lozano_Fern2022].
+Previous work [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] has used material culture analysis to reconstruct Upper Paleolithic social networks, which have been argued to be a driver of cultural and evolutionary change [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996; @Hill_Wood_Baggio_Hurtado_Boyd_2014; @Derex_Boyd_2016; @Romano_Lozano_Fern2022].
 @Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our results show evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted.
@@
 This ambiguity may be addressed in the future through the usage of the data recently released in SignBase 2.0 covering sites in Germany, which contains more detailed information on the specific order and patterns in which signs occurred on individual objects.
+<!-- citation to signbase 2.0 ?-->
 

### 2025-06-09 paper/paper.qmd
added new paper citations

@@
 Archaeological evidence from the Upper Paleolithic period indicates increasingly complex social organization across Europe [@Boyd_Silk_2020], coinciding with the ongoing biological transition from Neanderthals to early modern humans.
-Archaeologists have frequently used ancient art, the abundance of which is distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d; @Fuentes_Lucas_Robert_2019].
+Archaeologists have frequently used ancient art, the abundance of which is distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@dErrico_Baker_Rigaud_2025; @baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d; @Fuentes_Lucas_Robert_2019].
 Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca.
@@
 Some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg], including, in some cases, decorative objects [@Kelly_2024].
-Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
+Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@dErrico_Baker_Rigaud_2025; @Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
 
@@
 This study adds to our knowledge of social networks within the Aurignacian period, as our results document dynamic patterns of social exchange and connection.
-Previous work [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] has used material culture analysis to reconstruct Upper Paleolithic social networks, which have been argued to be a driver of cultural and evolutionary change [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996; @Hill_Wood_Baggio_Hurtado_Boyd_2014; @Derex_Boyd_2016; @Romano_Lozano_Fern2022].
+Previous work [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] has used material culture analysis to reconstruct Upper Paleolithic social networks, which have been argued to be a driver of cultural and evolutionary change [@dErrico_Baker_Rigaud_2025; @Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996; @Hill_Wood_Baggio_Hurtado_Boyd_2014; @Derex_Boyd_2016; @Romano_Lozano_Fern2022].
 @Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our results show evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted.

### 2025-06-09 paper/paper.qmd
stash

--- 2025-06-04 paper/paper.qmd
@@
 ---
+
 # Introduction
@@
 The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020], coupled with a biological transition from Neanderthals to early modern humans.
-Archaeologists have frequently relied upon variations in artwork, the large rise of which is a distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d; @Fuentes_Lucas_Robert_2019].
+Archaeologists have frequently relied upon variations in artwork, the large rise of which is a distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d].
 Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca.
@@
 We answer these questions by first identifying clusters within each phase using seriation and network analyses.
-We then test the strength of the clusters using perMANOVA tests.
+We then test the strength of the clusters using modularity and perMANOVA tests.
 All of our programming in this study was conducted in RStudio.
@@
 ggsave(maps_bar_plots, 
-      filename =  here::here("figures/map-and-hist-plot.png"),
+      filename =  here::here("figures/map-and_hist-plot.png"),
        h = 10, w = 10,
@@
 
-knitr::include_graphics(here::here("figures/map-and-hist-plot.png"))
+knitr::include_graphics(here::here("figures/map-and_hist-plot.png"))
 
@@
   jac <- vegdist(artifact_data %>% mutate_all(~ as.numeric(. > 0)), "jaccard", binary = TRUE)
+  dm <- as.matrix(jac)
+  modul_matrix <-
+    graph_from_adjacency_matrix(1 - dm, mode = "undirected")
+  mem <- as_membership(parse_number(group_data$group))
+  as_group <- modularity(modul_matrix, membership = mem)
+  as_group <- round(as_group, 3)
+
   perman <- adonis2(jac~as.factor(group_data$group), 
@@
 
-  final_table <- data_frame("PerMANOVA R2" = perman_r,
+  final_table <- data_frame("Modularity" = as_group, 
+                            "PerMANOVA R2" = perman_r,
                             "PerMANOVA F" = perman_f,
@@
 
-If the distribution of sign types can be explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases [@Herskind2023; @Rigaud_dErrico_Vanhaeren_2015].
+If the distribution of sign types can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases [.
 This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019].
@@
 To investigate connectivity among clusters of sites in the seriation solution, we used network analysis to represent and quantify the relationship between sites [@Mills_2017].
-Each node is a site, and each edge is a measure of sign type distribution similarity, as computed by the Jaccard similarity algorithm.
+Each node is a site, and each edge is a measure of sign type distribution similarity, as computed by the Jaccard similarity algorothm.
 To avoid cluttering the graph, all edges below the value of 0.2 were removed.
@@
 
-##  PerMANOVA for Cultural Connection Strength
-
-PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017; @Shennan_Crema_Kerig_2015].
+## Modularity and PerMANOVA for Cultural Connection Strength
+
+To evaluate the strength of the connections of the groups found in the seriation and network analyses, we used modularity and perMANOVA tests.
+The modularity test is a community structure network detection method, and quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006; @Newman_Girvan_2004].
+The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings.
+PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017].
 The results are expressed by an R^2^ statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, a pseudo-F statistic, and a p-value.
@@
 The p-value quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
-The perMANOVA was performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.
+The modularity test and perMANOVA were performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.
 
@@
 As for the Early Aurignacian, inter-connectivity between groups continues to be evident in the Evolved Aurignacian, with connections between Trou Magrite, Gargas, and Vindija Cave, and Bockstein-Törle and Gargas, however the strength of inter-connectivity is weaker than in the Early Aurignacian.
+
+In identifying the groups, we referenced both the seriation and network analyses. The results of the above tests allowed us to find two main groups of sites within each phase (@fig-seriation).
+
+## Cultural Connection Strength
+
+```{r}
+#| label: tbl-strength-table
+#| tbl-cap: "Table of the results per phase for the modularity test and perMANOVA"
+#| tbl-cap-location: bottom
+
+## Modularity test
+
+strength_table <- full_table %>% 
+  dplyr::select(`Aurignacian phase`, Modularity, `PerMANOVA R2`, `PerMANOVA F`, `PerMANOVA p`)
+
+kable(strength_table)
+
+proto_modularity <- proto_strength %>% 
+  dplyr::select(Modularity)
+evolved_modularity <- evolved_strength %>% 
+  dplyr::select(Modularity)
+
+trans_perm_r <- trans_strength %>% 
+  dplyr::select(`PerMANOVA R2`)
+early_perm_r <- early_strength %>% 
+  dplyr::select(`PerMANOVA R2`)
+
+trans_perm_p <- trans_strength %>% 
+  dplyr::select(`PerMANOVA p`)
+
+trans_perm_f <- trans_strength %>% 
+  dplyr::select(`PerMANOVA F`)
+early_perm_f <- early_strength %>% 
+  dplyr::select(`PerMANOVA F`)
+```
+
+The modularity score quantifies the strength of the groupings, showing that the variations in grouping strength between phases are minimal.
+The Proto-Aurignacian exhibits the strongest groupings, with a score of `r proto_modularity`, and the Evolved Aurignacian has the weakest groupings, with a score of `r evolved_modularity`.
+There is a slight increase in modularity between the Transitional period and the Proto-Aurignacian, before an even slighter decrease into the Early Aurignacian, followed by a larger increase with the Evolved Aurignacian.
+
+The perMANOVA R^2^ statistic shows the proportion of sign variation that can be explained by group membership.
+The Transitional period has the highest score of `r trans_perm_r` while the Early Aurignacian has the lowest score of `r early_perm_r`.
+We see a large decline in scores between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian periods.
+The pseudo-F statistic quantifies the difference between groups, with the Transitional phase having the highest value of `r trans_perm_f`, and the Early Aurignacian having the lowest value of `r early_perm_f`.
+Similar to the R^2^ statistic, we see a decline in values between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian period.
+The perMANOVA p-value indicates a statistically significant difference in the distribution of signs between the restricted and broad groups for the Proto , Early, and Evolved Aurignacian phases, but not the Transitional phase (p = `r trans_perm_p`).
+
+# Discussion
+
+## Grouping Dynamics Across the Four Phases of the Aurignacian
+
+Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and distribution.
+The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.
+@fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
+The modularity tests, as shown in @tbl-strength-table, show that the groupings are most distinct in the Transitional period and least distinct in the Early Aurignacian.
+The perMANOVA tests show that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.
+The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation.
 
@@
 
-In identifying the groups, we referenced both the seriation and network analyses.
-The results of the above tests allowed us to find two main groups of sites within each phase (@fig-group-map).
+The geographic distribution of the groups change from phase to phase (@fig-group-map).
 In the Transitional phase, sites are widely spread out across Western Europe.
@@
 In the Evolved Aurignacian, we see a wider distribution and more distance between sites, which may be associated with the reduced connectivity seen in that phase.
-The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation.
-
-## Cultural Connection Strength
-
-```{r}
-#| label: tbl-strength-table
-#| tbl-cap: "Table of the results per phase for the perMANOVA"
-#| tbl-cap-location: bottom
-
-strength_table <- full_table %>% 
-  dplyr::select(`Aurignacian phase`, `PerMANOVA R2`, `PerMANOVA F`, `PerMANOVA p`)
-
-kable(strength_table)
-
-trans_perm_r <- trans_strength %>% 
-  dplyr::select(`PerMANOVA R2`)
-early_perm_r <- early_strength %>% 
-  dplyr::select(`PerMANOVA R2`)
-
-trans_perm_p <- trans_strength %>% 
-  dplyr::select(`PerMANOVA p`)
-
-trans_perm_f <- trans_strength %>% 
-  dplyr::select(`PerMANOVA F`)
-early_perm_f <- early_strength %>% 
-  dplyr::select(`PerMANOVA F`)
-```
-
-The perMANOVA R^2^ statistic shows the proportion of sign variation that can be explained by group membership.
-The Transitional period has the highest score of `r trans_perm_r` while the Early Aurignacian has the lowest score of `r early_perm_r`.
-We see a large decline in scores between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian periods.
-The pseudo-F statistic quantifies the difference between groups, with the Transitional phase having the highest value of `r trans_perm_f`, and the Early Aurignacian having the lowest value of `r early_perm_f`.
-Similar to the R^2^ statistic, we see a decline in values between the Transitional and the Proto-Aurignacian periods, before a slight increase in between the Early and Evolved Aurignacian period.
-The perMANOVA p-value indicates a statistically significant difference in the distribution of signs between the restricted and broad groups for the Proto , Early, and Evolved Aurignacian phases, but not the Transitional phase (p = `r trans_perm_p`).
-
-# Discussion
-
-## Grouping Dynamics Across the Four Phases of the Aurignacian
-Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and distribution.
-The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian.
-@fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
-The perMANOVA tests show that the changes between the the Proto, Early, and Evolved Aurignacian phases are statistically significant, that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.
-
-One consistent grouping dynamic throughout all four phases is the restricted/broader group divide, which may be influenced by variation in social structure.
-Despite the tendency to conceive of hunter-gatherers as homogeneous, previous studies have established variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006; @Johnson_2014; @Singh_Glowacki_2022; @Schwendler_2012], and additional studies have established how variation in ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982; @Hodder1979; @Hodder1977; @Schwendler_2012].
-It is therefore possible that, for some of the sites, the broader/restricted grouping indicates variation in social structure, with the broader groups having a greater variety of formalized social roles, and the restricted groups having fewer formalized social roles.
-
-One interpretation of the connectivity seen in @fig-neighbor-net is as a representation of non-utilitarian mobility.
-Non-utilitarian mobility establishes regional social networks which facilitate the spread of information and the survival of hunter-gatherers in uncertain environments [@Whallon_2006; @Fitzhugh_Phillips_Gjesfeld_2011; @Lovis_Donahue_2011; @Hitchcock_Ebert_2011; @Romano_Lozano_Fern2002].
-The exchange and production of decorative items is associated with non-utilitarian mobility and has been recognized as a sign of social contact in other regions and eras [@Whallon_2006; @Kelly_2024; @Kinahan_2013; @Newlander_2012; @Iizuka_Ferguson_Izuho_2022].
-Previous studies have shown a range of variation in mobility among hunter-gatherer groups [@Binford_1990; @Kelly_1983; @Johnson_2014; @Padilla-Iglesias_Bischoff_2024; @Singh_Glowacki_2022]; some of the connections in @fig-neighbor-net may represent shorter-term information exchange, while others may indicate longer-term residential movement.
-
-One form of longer-term residential movement is the aggregation/dispersion pattern, which is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons.
-Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg], inclusing, in some cases, decorative objects [@Kelly_2024].
+
+One factor that may influence the geographic structure of sites is the possibility that some sites functioned as aggregation sites.
+The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons.
+Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a diverse artifact assemblage [@conkley_agg].
 Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].
 
-Aggregation activities may be identified by a higher than expected diversity index.
-@fig-div-plot plots the expected Shannon-Weaver diversity index [@Frerebeau_2025] for a given sample size against the actual diversity index per site [@Kintigh_1984].
-The plot identifies multiple sites with a higher index than expected, such as Hohle Fels in the Transitional phase, Spy, La Ferrassie, and Les Cottés in the Proto-Aurignacian, Solutré, Hohle Fels, and Grottes de Fonds-de-Forêt in the Early Aurignacian, and Vogelherd, Bockstein-Törle, and Vindija Cave in the Evolved Aurignacian.
-Notably, most of these sites are not shown as having connections between groups in @fig-neighbor-net, except for La Ferrassie in the Proto-Aurignacian.
-This suggests a distinction between shorter-term information exchange (e.g.
-the highly interconnected sites in @fig-neighbor-net) and longer-term residential movement (e.g.
-the sites with higher than expected diversity in @fig-div-plot).
+A potential identifier of aggregation sites comes from @fig-div-plot, which uses the tabula package [@Frerebeau_2025] to generate the Shannon-Weaver diversity index of sign types per site, and plot the expected diversity index for a given sample size against the actual diversity index per site [@Kintigh_1984].
+The plot identifies multiple sites that diverge from the diversity index predicted by sample size, including those with a higher index than expected, such as Hohle Fels in the Transitional phase, Spy, La Ferrassie, and Les Cottés in the Proto-Aurignacian, Solutré, Hohle Fels, and Grottes de Fonds-de-Forêt in the Early Aurignacian, and Vogelherd, Bockstein-Törle, and Vindija Cave in the Evolved Aurignacian.
+It is possible that those sites with a greater than expected diversity index acted as aggregation sites or served a similar purpose.
+La Ferrassie in the Proto-Aurignacian is also shown by @fig-neighbor-net to have multiple connections to sites in both groups, in addition to the diversity index shown in @fig-div-plot, perhaps suggesting that it served as a specifically inter-group aggregation site.
 
@@
 This study adds to our knowledge of social networks within the Aurignacian period, as our results document dynamic patterns of social exchange and connection.
-Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996; @Hill_Wood_Baggio_Hurtado_Boyd_2014; @Derex_Boyd_2016; @Romano_Lozano_Fern2022].
+Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996].
 @Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our results show evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted.

### 2025-06-09 paper/paper.qmd
fix that citation

--- 2025-06-09 paper/paper.qmd
@@
 Previous studies have shown a range of variation in magnitudes and types of mobility among hunter-gatherer groups [@Binford_1990; @Kelly_1983; @Johnson_2014; @Padilla-Iglesias_Bischoff_2024; @Singh_Glowacki_2022].
-Non-utilitarian mobility establishes regional social networks which facilitate the spread of information and the survival of hunter-gatherers in uncertain environments [@Whallon_2006; @Fitzhugh_Phillips_Gjesfeld_2011; @Lovis_Donahue_2011; @Hitchcock_Ebert_2011; @Romano_Lozano_Fern2002].
+Non-utilitarian mobility establishes regional social networks which facilitate the spread of information and the survival of hunter-gatherers in uncertain environments [@Whallon_2006; @Fitzhugh_Phillips_Gjesfeld_2011; @Lovis_Donahue_2011; @Hitchcock_Ebert_2011; @Romano_Lozano_Fern2022].
 The exchange and production of decorative items, such as objects with geometric signs, has been associated with non-utilitarian mobility and recognized as an indicator of social contact [@Whallon_2006; @Kelly_2024; @Kinahan_2013; @Newlander_2012; @Iizuka_Ferguson_Izuho_2022].

### 2025-06-09 paper/paper.qmd
update citation style to NHB

@@
 ---
-title: "Geometric signs reveal changes social structures and networks during the Western Eurasian Aurignacian"
+title: "Geometric signs reveal changes in social structures and networks during the Western Eurasian Aurignacian"
 author:
@@
       orcid: xxx
+  - Aleah Rosner:
+      institute: uw
+      orcid: xxx
+  - Eva Oryn:
+      institute: uw
+      orcid: xxx
+  - Kenneth B. Vernon:
+      orcid: 0000-0003-0098-5092
+      institute:
+        - uu
   - Ben Marwick:
@@
   - uw: Department of Anthropology, University of Washington, Seattle, USA
+  - uu: Scientific Computing and Imaging Institute, University of Utah, Salt Lake City, USA
 title-block-published: "Last updated"  
@@
 bibliography: references.bib
-csl: "../templates/society-for-american-archaeology.csl" # Insert path for the bib-style
+csl: "../templates/nature.csl" # Insert path for the bib-style
 abstract: |

### 2025-07-07 paper/paper.qmd
editing and organising

@@
 
-# Introduction
-
-Archaeological evidence from the Upper Paleolithic period indicates increasingly complex social organization across Europe [@Boyd_Silk_2020], coinciding with the ongoing biological transition from Neanderthals to early modern humans.
+Archaeological evidence from the Upper Paleolithic period indicates increasingly complex social organization across Europe [@Boyd_Silk_2020].
 Archaeologists have frequently used ancient art, the abundance of which is distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@dErrico_Baker_Rigaud_2025; @baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d; @Fuentes_Lucas_Robert_2019].
@@
 43 - 30 k BP) is often linked to the dispersal of _Homo sapiens_ across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020].
-In this paper we investigate Aurignacian social organization dynamics implied by the distribution and variation of geometric sign types found on mobile objects in Europe across four phases of the Aurignacian period.
+We investigated human social organization dynamics implied by the distribution and variation of geometric sign types found on mobile objects in Europe across four phases of the Aurignacian period.
+
+## Cultural group identification in the Upper Paleolithic
 
@@
 Seriation analysis, PCoA, and network analysis were used to identify groups from 134 discrete bead types recovered from both burial and occupation sites.
-To validate their claims of cultural groups, @baker2024evidence used a Mantel test to evaluate an isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance, finding that geographic distance alone did not solely account for the bead distribution.
+To validate their claims of cultural groups, @baker2024evidence used a Mantel test to evaluate an isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance.
+They found that geographic distance alone did not solely account for the bead distribution.
 The nine geographically discrete groups encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and western Mediterranean regions.
 
-Similar analyses of personal ornament types from the Aurignacian technocomplex identified fourteen geographically cohesive groups [@Vanhaeren_d].
+Similar analyses of personal ornament types from the Aurignacian identified fourteen geographically cohesive groups [@Vanhaeren_d].
 Drawing on ethnographic studies that show how body decoration and ornamentation indicate ethno-linguistic identity, @Vanhaeren_d used Aurignacian personal ornamentation as a proxy for ethno-linguistic diversity.
 The variation in personal ornament types is interpreted as evidence of long-lasting cultural differences and as robustly establishing the ethno-linguistic diversity of the Aurignacian period.
-They used seriation, correspondence, and GIS analyses of 157 distinct ornament types from 98 Aurignacian sites in Europe and the Near East to identify geographically cohesive groups sharing similar ornament type associations.
+Seriation, correspondence, and GIS analyses of 157 distinct ornament types from 98 Aurignacian sites in Europe and the Near East identified geographically cohesive groups sharing similar ornament type associations.
 These groups sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.
@@
 
-In this study, we extend this previous work using geometric signs as identity markers to investigate cultural groups in the Aurignacian period.
-The Aurignacian period spans roughly 43-30 k BP, and is largely characterized by material culture indicators such as carinated scrapers, small-flake tools and split-based bone and ivory tools, as well as an ongoing evolutionary shift from Neanderthals to early modern humans [@Chu_Richter_2020].
+In this study, we extend this previous work using geometric signs as identity markers to investigate cultural group dynamics in the Aurignacian period, providing a novel approach to understanding how early groups of _Homo sapiens_ interacted with each other and their environments.
+Our archaeological and modelling results support the view that geometric signs are reliable proxies for reconstructing past cultural diversity and correlate with past population dynamics.
+
+## Periodization of the Aurignacian
+
+The Aurignacian period (43-30 k BP) is largely characterized by material culture indicators such as carinated scrapers, small-flake tools and split-based bone and ivory tools, as well as an ongoing evolutionary shift from Neanderthals to early modern humans [@Chu_Richter_2020].
 @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main phases, Aur-P1 (Proto/Early, 43–37 k BP), and Aur-P2 (Evolved/Late, 37-32 k BP).
@@
 To answer these questions we use seriation analysis, network analysis, and PerMANOVA tests.
-
-# Materials and Methods
 
@@
 
-## Data
+## Geometric signs from the Aurignacian
 
@@
 To minimize the influence of this skewed distribution, we followed @Lycett_2019 in using presence/absence data for each sign at each site, reducing the effect of variation in object count between sites.
-We used Mantel tests with a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types.
-The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse], (here 1000 permutations), producing a Mantel R statistic and a p-value.
-The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation.
-The p-value estimates the probability of obtaining the observed correlation if the null hypothesis of no correlation were true (we use an alpha value of 0.05 for the threshold of statistical significance).
-The Mantel tests were performed using the vegan package [@Vegan2001].
-The test indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`), validating our decision to use presence/absence data.
+To test for interactions between object abundance and sign type distribution, we used Mantel tests with a Euclidean object abundance distance matrix and a Jaccard distance matrix of sign types.
+The test indicated that there was no significant correlation between object abundance and sign type distribution (R = `r r_value_object_abundance_correlation`, p = `r p_value_object_abundance_correlation`), supporting our decision to use presence/absence data and remove frequency as a potentially confounding variable.
 
@@
 
-## Cultural Group Detection Using Seriation and Network Analysis
-
-Seriation analysis is the arrangement of data into a linear order to reveal patterns [@Hahsler_Hornik_Buchta].
-We used the Brower-Kile [-@Brower_Kile_1988] seriation algorithm from the seriation package [@seriation_pckge], in which a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal.
-To investigate connectivity among clusters of sites in the seriation solution, we used network analysis to represent and quantify the relationship between sites [@Mills_2017].
-Each node is a site, and each edge is a measure of sign type distribution similarity, as computed by the Jaccard similarity algorithm.
-To avoid cluttering the graph, all edges below the value of 0.2 were removed.
-Results were plotted with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths, and network analysis was performed using the vegan [@Vegan2001], statnet [@statnet_package] and igraph [@igraph_package] packages.
-
-##  PerMANOVA for Cultural Connection Strength
-
-PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017; @Shennan_Crema_Kerig_2015].
-The results are expressed by an R^2^ statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, a pseudo-F statistic, and a p-value.
-The pseudo-F statistic compares the total sum of squared dissimilarities between groups to total sum of squared dissimilarities within groups.
-Larger pseudo-F statistics indicate more separation between groups.
-The p-value quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
-The perMANOVA was performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.
-
 # Results
@@
 Les Cottés stands out with no connections to either group.
-The Early Aurignacian shows more strongly inter-connected groups -- although the shape still suggests two main groups -- with Grotte de la Verpillière I in particular having multiple strong connections to both groups.
+The Early Aurignacian shows more strongly inter-connected groups -- although two main groups remain evident -- with Grotte de la Verpillière I in particular having multiple strong connections to both groups.
 Lastly, the Evolved Aurignacian again shows two main groups, although intra-group connectivity has decreased, with the broad range group consisting of Hohle Fels, Trou Magrite, Vogelherd, and Bockstein-Törle, and the restricted range group consisting of Gargas, La Viña, Sirgenstein Cave, Vindija Cave, and Les Rois.
@@
 
-While th results of the seriation and network analyses point to two main groups of sites within each phase, spatial clusters of sites are not highly evident in these groups, in part due to the small numbers of sites in each phase (@fig-group-map).
+While the results of the seriation and network analyses point to two main groups of sites within each phase, spatial clusters of sites are not highly evident in these groups, in part due to the small numbers of sites in each phase (@fig-group-map).
 In the Transitional phase, sites are widely spread out across Western Europe.
 This wider distribution may be a factor in weak connections seen in @fig-neighbor-net.
-In the Proto-Aurignacian, we see a slightly narrower geographic distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the slightly stronger connectivity seen there.
+In the Proto-Aurignacian, we see a slightly narrower geographic distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the slightly stronger connectivity indicated in the network analysis results.
 This phase is also associated with an increase in sites and sign types, which may be correlated with the increased proximity and population density.
@@
 In the Evolved Aurignacian, we see a wider distribution and more distance between sites, which may be associated with the reduced connectivity seen in that phase.
-The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation.
 
@@
 ```{r}
-#| label: tbl-strength-table
-#| tbl-cap: "Table of the results per phase for the perMANOVA"
-#| tbl-cap-location: bottom
-
-strength_table <- full_table %>% 
-  dplyr::select(`Aurignacian phase`, `PerMANOVA R2`, `PerMANOVA F`, `PerMANOVA p`)
-
-kable(strength_table)
-
 trans_perm_r <- trans_strength %>% 
@@
 
-The perMANOVA R^2^ statistic shows the proportion of sign variation that can be explained by group membership.
+The perMANOVA R^2^ statistics presented in @tbl-strength-table show the proportion of sign variation that can be explained by group membership.
 The Transitional period has the highest score of `r trans_perm_r` while the Early Aurignacian has the lowest score of `r early_perm_r`.
@@
 The perMANOVA p-value indicates a statistically significant difference in the distribution of signs between the restricted and broad range groups for the Proto , Early, and Evolved Aurignacian phases, but not the Transitional phase (p = `r trans_perm_p`).
+
+```{r}
+#| label: tbl-strength-table
+#| tbl-cap: "Table of the results per phase for the perMANOVA"
+#| tbl-cap-location: bottom
+
+kable(strength_table)
+```
 
@@
 
+# Methods
+
+## Cultural Group Detection Using Seriation and Network Analysis
+
+Seriation analysis is the arrangement of data into a linear order to reveal patterns [@Hahsler_Hornik_Buchta].
+We used the Brower-Kile [-@Brower_Kile_1988] seriation algorithm from the seriation package [@seriation_pckge], in which a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal.
+To investigate connectivity among clusters of sites in the seriation solution, we used network analysis to represent and quantify the relationship between sites [@Mills_2017].
+Each node is a site, and each edge is a measure of sign type distribution similarity, as computed by the Jaccard similarity algorithm.
+To avoid cluttering the graph, all edges below the value of 0.2 were removed.
+Results were plotted with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths, and network analysis was performed using the vegan [@Vegan2001], statnet [@statnet_package] and igraph [@igraph_package] packages.
+
+##  PerMANOVA for Cultural Connection Strength
+
+PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017; @Shennan_Crema_Kerig_2015].
+The results are expressed by an R^2^ statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, a pseudo-F statistic, and a p-value.
+The pseudo-F statistic compares the total sum of squared dissimilarities between groups to total sum of squared dissimilarities within groups.
+Larger pseudo-F statistics indicate more separation between groups.
+The p-value quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance).
+The perMANOVA was performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.
+
+## Mantel tests
+
+The test seeks to determine the level of correspondence between the matrices through permutational evaluation of the null distribution [@Mantel_Smouse] (here 1000 permutations), producing a Mantel R statistic and a p-value.
+The R statistic falls between -1, meaning a strong negative correlation, and 1, meaning a strong positive correlation, while a value of 0 means no correlation.
+The p-value estimates the probability of obtaining the observed correlation if the null hypothesis of no correlation were true (we use an alpha value of 0.05 for the threshold of statistical significance).
+The Mantel tests were performed using the vegan package [@Vegan2001].
+
 # References

### 2025-07-07 paper/paper.qmd
minor edits

@@
 
+strength_table <- full_table %>% 
+  dplyr::select(`Aurignacian phase`, 
+                `PerMANOVA R2`, 
+                `PerMANOVA F`, 
+                `PerMANOVA p`)
+
 kable(strength_table)
@@
 The exchange and production of decorative items, such as objects with geometric signs, has been associated with non-utilitarian mobility and recognized as an indicator of social contact [@Whallon_2006; @Kelly_2024; @Kinahan_2013; @Newlander_2012; @Iizuka_Ferguson_Izuho_2022].
-We can divide the connections depicted in @fig-neighbor-net into shorter-term information exchange and longer-term residential movement.
-
-One form of longer-term residential movement that may be related to non-utilitarian mobility are aggregation/dispersion patterns, which are defined by @conkley_agg as a type of hunter-gather settlement pattern in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons.
-Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate.
+This non-utilitarian mobility can be divided into shorter-term information exchange and longer-term residential movement.
+
+One form of longer-term residential movement that may be related to non-utilitarian mobility are aggregation/dispersion patterns, which are a type of hunter-gather settlement pattern in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons [@conkley_agg].
+Aggregation sites, associated with high degrees of social connectivity, are the places where those group fragments or individuals congregate.
 Some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg], including, in some cases, decorative objects [@Kelly_2024].
@@
 
-Following this previous work, we identify aggregation sites as those with a higher than expected sign diversity index.
+We identify aggregation sites as those with a higher than expected sign diversity index.
 @fig-div-plot plots the expected Shannon-Weaver diversity index [@Frerebeau_2025] for a given sample size against the actual diversity index per site [@Kintigh_1984].
@@
 The blue line is the expected Heterogeneity value under the null model (simple increase in sample size means simple increase in diversity of sign types.
-The upper red line is the upper 95% confidence interval, if a site is near or above that line, it has a much higher diversity of signs than expected under the null model.The lower red line is the lower 95% confidence interval, if a site is near or below that line, it has a much lower diversity of signs than expected under the null model."
+The upper red line is the upper 95% confidence interval, if a site is near or above that line, it has a much higher diversity of signs than expected under the null model.
+The lower red line is the lower 95% confidence interval, if a site is near or below that line, it has a much lower diversity of signs than expected under the null model."
 

### 2025-08-22 paper/paper.qmd
editing
conclusion edits - add in Kovacevic paper

@@
 
+Our results do not suggest that geometric signs serve as markers of a structured ethnic identity, as found with Aurignacian personal ornaments in @Vanhaeren_d.
+However, the data used by @Vanhaeren_d was also analyzed by @Kovacevic_Shennan_Vanhaeren_dErrico_Thomas_2015 to find that cultural identity by-descent with modification and isolation-by-distance is a more likely explanation for the distribution of Aurignacian personal ornaments.
+Our results support the pattern of cultural identity by-descent with modification (although they do not support isolation-by-distance [@tbl-mantel-table]), as well as the patterns of signs and other artwork as a form of communication of cultural in-group and out-group boundaries proposed by @Sauvet_2018 and @baker2024evidence.
+
 <!-- connect back with the papers we discuss in the background section about Aurginacian cultural groups, did we find the same or different?
@@
 We attempted to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023] by dividing the data up into smaller time periods.
-Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019].
+Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019].Our results do not support the conclusions drawn in @Vanhaeren_d, where Aurignacian artwork is found to be a symbolic marker of structured ethnic identity, but are supported by @Kovacevic_Shennan_Vanhaeren_dErrico_Thomas_2015, which utilizes the same data as @Vanhaeren_d to conclude that cultural identity by-descent with modification is a more likely explanation for the distribution of Aurigancian ornaments.
 

### 2025-08-23 paper/paper.qmd
edit

@@
 However, the data used by @Vanhaeren_d was also analyzed by @Kovacevic_Shennan_Vanhaeren_dErrico_Thomas_2015 to find that cultural identity by-descent with modification and isolation-by-distance is a more likely explanation for the distribution of Aurignacian personal ornaments.
-Our results support the pattern of cultural identity by-descent with modification (although they do not support isolation-by-distance [@tbl-mantel-table]), as well as the patterns of signs and other artwork as a form of communication of cultural in-group and out-group boundaries proposed by @Sauvet_2018 and @baker2024evidence.
+Our results support the pattern of cultural identity by-descent with modification (although they do not support isolation-by-distance [@tbl-mantel-table]), as well as the patterns of signs and other artwork as a form of cultural in-group and out-group communication found by @Sauvet_2018 and
+@baker2024evidence.
 

### 2025-09-03 paper/paper.qmd
add some comments in the discussion text

--- 2025-07-07 paper/paper.qmd
@@
 
-<!-- connect back with the papers we discuss in the background section about Aurginacian cultural groups, did we find the same or different?
-Most of our conclusions are not similar in kind to theirs, that is fine, we just need to make a connection between their work and ours here.
+Our results do not suggest that geometric signs serve as markers of a structured ethnic identity <!-- BM: be more precise about what our findings do not support --> , as found with Aurignacian personal ornaments in @Vanhaeren_d.
+However, the data used by @Vanhaeren_d was also analyzed by @Kovacevic_Shennan_Vanhaeren_dErrico_Thomas_2015 <!-- BM: expand a bit more on what this study did and found -->to find that cultural identity by-descent with modification <!-- BM: expand this to define to show readers we know what this is --> and isolation-by-distance is a more likely explanation for the distribution of Aurignacian personal ornaments.
+Our results support the pattern of cultural identity by-descent with modification (although they do not support isolation-by-distance [@tbl-mantel-table]), as well as the patterns of signs and other artwork as a form of cultural in-group and out-group communication found by @Sauvet_2018 and @baker2024evidence.
+<!-- BM: want to engage specifically with Baker et al.
+maybe do this earlier in this para, start the para with reference specifically to Baker et al.
 -->
@@
 We attempted to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023] by dividing the data up into smaller time periods.
-Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019].
+Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019].Our results do not support the conclusions drawn in @Vanhaeren_d <!-- BM: mention Baker et al too -->, where Aurignacian artwork is found to be a symbolic marker of structured ethnic identity, but are supported by @Kovacevic_Shennan_Vanhaeren_dErrico_Thomas_2015, which utilizes the same data as @Vanhaeren_d to conclude that cultural identity by-descent with modification is a more likely explanation for the distribution of Aurigancian ornaments.
 

### 2025-10-28 paper/paper.qmd
fix typos spotted by Aleah

@@
 
-We hypothesize that these phases, and the environmental changes across the them, may correspond to changes in dynamics of cultural groups, as represented by patterns in the distribution of geometric signs.
+We hypothesize that these phases, and the environmental changes across them, may correspond to changes in dynamics of cultural groups, as represented by patterns in the distribution of geometric signs.
 We aim to address the following questions: How do grouping dynamics of geometric signs change across the four phases of the Aurignacian?
@@
 A Mantel test to evaluate the correlation between distance in time and distribution of sign types found a statistically significant correlation (R = `r chron_r`, p = `r chron_p`).
-This confirms that our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.
+This confirms our assumption that a meaningful change in sign types occurs over time during the Aurignacian period.
 
@@
 @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to very strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian.
-The perMANOVA tests show that the changes between the the Proto, Early, and Evolved Aurignacian phases are statistically significant, that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.
+The perMANOVA tests show that the changes between the Proto, Early, and Evolved Aurignacian phases are statistically significant, that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.
 
@@
 
-Our results are consistent with @Banks_dErrico_Zilhao_2013 who used eco-cultural niche modeling to show that the Heinrich Stadial 4 (HS4) at the very beginning of the Early Aurignacian resulted in a expansion of people into new environmental niches, which they hypothesize was associated with an expansion of social networks both within and between populations.
+Our results are consistent with @Banks_dErrico_Zilhao_2013 who used eco-cultural niche modeling to show that the Heinrich Stadial 4 (HS4) at the very beginning of the Early Aurignacian resulted in an expansion of people into new environmental niches, which they hypothesize was associated with an expansion of social networks both within and between populations.
 This correlates with our results which show a large increase in both inter-connectivity and intra-connectivity strength in the Early Aurignacian, as well as the increased sign type correlation with geographic distance.
@@
 
-Our results do not suggest that geometric signs serve as markers of a structured ethnic identity <!-- BM: be more precise about what our findings do not support --> , as found with Aurignacian personal ornaments in @Vanhaeren_d.
-However, the data used by @Vanhaeren_d was also analyzed by @Kovacevic_Shennan_Vanhaeren_dErrico_Thomas_2015 <!-- BM: expand a bit more on what this study did and found -->to find that cultural identity by-descent with modification <!-- BM: expand this to define to show readers we know what this is --> and isolation-by-distance is a more likely explanation for the distribution of Aurignacian personal ornaments.
-Our results support the pattern of cultural identity by-descent with modification (although they do not support isolation-by-distance [@tbl-mantel-table]), as well as the patterns of signs and other artwork as a form of cultural in-group and out-group communication found by @Sauvet_2018 and @baker2024evidence.
-<!-- BM: want to engage specifically with Baker et al.
-maybe do this earlier in this para, start the para with reference specifically to Baker et al.
--->
+Our results support a general theory of geometric signs as a form of cultural identification and communication during the Palaeolithic, as proposed by previous work [e.g.
+@Sauvet_2018 and @baker2024evidence].
+However, our results do not include geographically distinct cultural groups, and so our findings contrast with previous work such as @Vanhaeren_d and @baker2024evidence who found distinct cultural clusters in their analysis of personal ornaments.
+One possible explanation of this contrast is that a distinguishing factor between the Aurignacian and Gravettian is the appearance of geographically distinct cultural groups in the Gravettian, as reported by @baker2024evidence.
+We might hypothesize that the distribution found in the Evolved Aurignacian [@fig-group-map] evolved into the clusters proposed by @baker2024evidence for the Gravettian period.
+
+That said, our results also contrast with @Vanhaeren_d, who found nine ethnically structured groups using evidence from the Aurignacian period.
+One possible explanation for this is that the geometric signs studied here encode different kinds of group information compared to the personal ornament objects studied by @Vanhaeren_d.
+On one hand, we hypothesize that geometric signs are not a useful proxy for detecting geographically distinct cultural groups in the Palaeolithic, compared to personal ornaments.
+On the other hand, the data used by @Vanhaeren_d was also analyzed by @Kovacevic_Shennan_Vanhaeren_dErrico_Thomas_2015, who used Approximate Bayesian Computation to simulate and compare Aurignacian personal ornament distribution as a result of structured ethnic identity [e.g.
+] versus as a result of cultural identity by-descent with modification and isolation-by-distance.
+Their results show the latter was best supported by the data.
+Cultural identity by-descent with modification is defined as group difference generated by unconscious cultural mutation and drift, and largely influenced by spatial proximity to other groups.
+These results from  @Kovacevic_Shennan_Vanhaeren_dErrico_Thomas_2015 complicate claims by @Vanhaeren_d for geographically distinct cultural groups in the Aurignacian, and validate our findings.
+
+While our results do not entirely support the findings of @Kovacevic_Shennan_Vanhaeren_dErrico_Thomas_2015, given that our Mantel tests did not find a statistically significant correlation between geographic distance and sign type similarity (@tbl-mantel-table), cultural identity by-descent with modification seems a likely explanation for some of the patterns seen in our results.
 
@@
 The results are expressed by an R^2^ statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, a pseudo-F statistic, and a p-value.
-The pseudo-F statistic compares the total sum of squared dissimilarities between groups to total sum of squared dissimilarities within groups.
+The pseudo-F statistic compares the total sum of squared dissimilarities between groups to the total sum of squared dissimilarities within groups.
 Larger pseudo-F statistics indicate more separation between groups.
